---
title: "Excel as a Data-Analysis Workbench"
---

```{r setup-ch01}
#| include: false
library(tidyverse)
library(readxl)
library(here)
library(knitr)
library(scales)
```

::: {.callout-note appearance="simple"}
### Data for this chapter
This chapter uses the **BC silviculture** workbook
([`bc_disturbance_reforestation.xlsx`](../data/bc_disturbance_reforestation.xlsx){download="bc_disturbance_reforestation.xlsx"}).
Click the file name to download it directly, then save it in your `data/`
folder. For the source and details, see the
[Datasets Used in This Book](../datasets.qmd) page.
:::

## Chapter goals {.unnumbered}

By the end of this chapter you will be able to:

- Open a real BC Government Excel workbook and orient yourself to
  its contents on first read.
- Read the **About**, **ReadMe**, and **Data** sheets in the right
  order, and explain why that order matters.
- Identify each column's **type** (text, number, date) and
  **unit** (hectares, year).
- Use **freeze panes** and a few keyboard shortcuts to move through
  a multi-column file without losing context.
- Apply **basic number formats** that distinguish display from value.
- Write a few **simple formulas** (`SUM`, `AVERAGE`, `MIN`, `MAX`)
  and **check whether the results make sense**.
- Set up a clean **project folder structure** that separates raw
  data from cleaned files.

## Expected output for this chapter {.unnumbered}

At the end of this chapter, each learner (or group) will produce:

::: {.callout-tip appearance="default"}
### What you will hand in

1. The raw workbook **`bc_disturbance_reforestation.xlsx`** placed in
   your `data/` folder, **unchanged from the BC Data
   Catalogue download**.
2. A working copy you may modify, saved as
   **`bc_disturbance_reforestation_clean.xlsx`** in your
   `outputs/` folder, containing:
   - the original three sheets (**About**, **ReadMe**, **Data**) kept
     intact — you add to them, you do **not** create new sheets or
     delete the originals,
   - one short **data dictionary entry** in your own words for any
     column you choose, and
   - the five **simple formulas** from this chapter computed in any
     empty cell.
3. A short **group-lab worksheet** (or screenshot) submitted to the
   course site.

This list is the same as the rubric for the chapter assignment. Keep
it in view while you work.
:::

## Why we start with Excel

Excel has flaws. It hides decisions in invisible mouse clicks. It is
hard to reproduce or recreate. It is hard to version-control. We will spend most
of this book moving away from it.

We still start with Excel deliberately. The interface is forgiving.
Mistakes are visible. There is no console to fight. And the
fundamental discipline we want learners to leave the course with —
*inspect first, compute second, verify always* — is easier to teach
when the data is sitting on the screen in plain sight.

When we move to R in Chapter 4, we will use **the same dataset** we
analyse in Excel here. The continuity is the point: R is not a fresh
start; it is a more honest version of what you already do. (See *A
glimpse of Chapter 4* at the end of this chapter for a preview.)

### What "data analysis" actually means

People say *"data analysis"* to mean different things. In this
course it means a sequence of five repeatable steps:

1. **Inspect** the file. Know what is in it before computing
   anything.
2. **Document** the file. Write a data dictionary you can defend.
3. **Clean** the file. Remove what is wrong, document what was
   removed.
4. **Summarize** the file. Compute the numbers that answer your
   question.
5. **Communicate** the result. Tell the story with a chart, a
   table, or a sentence.

Chapter 1 covers steps 1 and 2 — *inspect* and *document*. The rest
of the book builds the other three steps on top of that foundation.

## The dataset for this chapter

This chapter uses one real, openly licensed Excel workbook from the
**BC Government / BC Data Catalogue**:

```{r dataset-summary}
#| echo: false
bc <- read_excel(here("data", "bc_disturbance_reforestation.xlsx"),
                 sheet = "Data")

tibble(
  Property = c("File name", "Source", "Catalogue page",
               "Licence", "Rows of data", "Columns",
               "Time span", "Spatial coverage", "Unit"),
  Value = c(
    "`bc_disturbance_reforestation.xlsx`",
    "Environmental Reporting BC — *Trends in Silviculture in B.C.*",
    "<https://catalogue.data.gov.bc.ca/dataset/indicator-summary-data-trends-in-silviculture-in-b-c->",
    "Open Government Licence — British Columbia",
    as.character(nrow(bc)),
    as.character(ncol(bc)),
    paste0(min(bc$Fiscal_Year), "–", max(bc$Fiscal_Year), " (fiscal years)"),
    "Province of British Columbia",
    "Hectares (ha)"
  )
) |> kable()
```

The file is **directly downloadable** from the BC Data Catalogue.
You do not need an account. There is no shapefile bundle to manage.
It is one Excel file with three sheets.

### Why this dataset matters for forestry

This is not a textbook dataset. Every number was published by the
BC Ministry of Forests under the **Trends in Silviculture in B.C.**
program. It is the same indicator that the province uses to track
whether reforestation is keeping pace with the area being harvested
and lost to natural disturbance.

A working forester or natural-resources analyst in BC will encounter
this dataset (or one with the same structure) within a few months of
starting any monitoring or reporting role. Learning to read it now
is real, applicable work.

The dataset captures the story of BC silviculture from 1987 onward:

- The shift in harvest practice from straight clearcutting toward
  **clearcutting-with-reserves** through the 1990s — a real policy
  change visible in the columns.
- BC's record-breaking **2017 wildfire season**, which dominates the
  natural-disturbance column.
- Steady growth in **reforestation** that, by the 2020s, exceeds
  total disturbance in most years.

::: {.callout-important}
### Where the data comes from — and what to do with it

This dataset is published by the BC Government and distributed
through the **BC Data Catalogue**. Two rules:

1. **Keep the raw downloaded file unchanged.** Save the file as
   `bc_disturbance_reforestation.xlsx` exactly as you downloaded
   it. Do not sort it, edit it, or save over it. The raw file is
   your point of reference if anything later goes wrong.
2. **Save any cleaned or modified version separately.** Use a
   different file name (typically
   `bc_disturbance_reforestation_clean.xlsx`) and a different
   folder. That way you can always go back to the original.

This is the single most important habit in this chapter.
:::

::: {.callout-tip}
### Why this dataset is right for this chapter

- **It is real.** Every number was published by the BC Ministry of
  Forests. You can verify any value against the official indicator
  page.
- **It is small.** 37 rows × 8 columns. You can scroll through
  every row and inspect every cell.
- **It has no missing values.** Cleaning comes later in the book.
- **The columns mean something.** Every column has a forestry
  meaning we can talk about.
- **It tells a story** — see *Why this dataset matters for forestry*
  above.
:::

### What the dataset measures

Each row is one **fiscal year** in BC. Each column is one
**area-based forestry indicator** — harvested area, naturally
disturbed area, reforested area. Units are always hectares (ha).

A fiscal year in BC runs from **April 1 to March 31** of the
following calendar year. So "Fiscal Year 2023" means April 2023
through March 2024.

## Project folder structure

Before you open the file, set up where it will live. The way you
organize files now saves a hundred *"where is that file?"* moments
later.

Create this folder on your computer (or on Posit Cloud):

```
frst232/
├── chapter01/
│   ├── data/                                Raw data files — never edit anything in here.
│   │   └── bc_disturbance_reforestation.xlsx
│   ├── outputs/                             Cleaned or modified files you produce.
│   │   └── bc_disturbance_reforestation_clean.xlsx   (you save this yourself)
│   ├── notes/                               Lecture notes, screenshots.
│   └── deliverables/                        What you submit to the course site.
└── shared/
    ├── data_dictionaries/                   ReadMe sheets you have written.
    └── references/                          Course PDFs, papers, links.
```

Three rules:

1. **Never edit raw data in place.** Always work on a copy that
   lives in `outputs/`.
2. **Use lowercase, snake_case names.** `bc_disturbance_reforestation.xlsx`
   — not `BC Disturbance Data (Final).xlsx`. Spaces and capitals
   cause problems when you move to R in Chapter 4.
3. **One folder per chapter.** Easier to find things in a dozen
   small folders than in one huge dump.

::: {.callout-tip}
### Where the file goes in this course

For every code chunk in this book to work, the data file should
live at:

```
your_project_root/data/bc_disturbance_reforestation.xlsx
```

When you see `here("data", "bc_disturbance_reforestation.xlsx")`
in a later chapter, that is the path it refers to.
:::

## The three-sheet pattern: About, ReadMe, Data

Download `bc_disturbance_reforestation.xlsx` from the BC Data
Catalogue and place it in your `data/` folder. Open the file. You
will see three tabs at the bottom:

- **About** — *what is this file?* Source, publisher, licence,
  citation, retrieval date.
- **ReadMe** — *what does each column mean?* One row per variable,
  with description, unit, and notes.
- **Data** — the actual 37 rows of data.

**Read these tabs in order. Always.** Reading the Data sheet
before reading About means computing something before you know
what it measures.

### About sheet

The About sheet answers *where did this file come from?*. Typical
fields:

| Field | Example value |
|---|---|
| Dataset | *Indicator Summary Data: Trends in Silviculture in B.C.* |
| Publisher | Resource Practices Branch, Ministry of Forests, B.C. |
| Source URL | catalogue.data.gov.bc.ca/... |
| Licence | Open Government Licence — British Columbia |
| Cite as | Environmental Reporting BC, *Trends in Silviculture in B.C.* |
| Retrieved | 2026-05-24 |

The point of the About sheet is **provenance**. A reader who finds
this file two years from now needs to know where it came from and
under what licence they can use it.

### ReadMe sheet

The ReadMe sheet answers *what does each column mean?*:

```{r readme-preview}
#| echo: false
read_excel(here("data", "bc_disturbance_reforestation.xlsx"),
           sheet = "ReadMe", skip = 3) |>
  setNames(c("Variable", "Description", "Unit", "Notes")) |>
  kable()
```

One row per column. Notice how the unit (hectares) appears in
every numeric row. Never assume the unit; check it.

::: {.callout-tip}
### A pattern to keep forever

Every dataset you create for this course should ship with About
and ReadMe sheets like these. They cost an hour to write and
save many hours of *"what does this column mean?"* arguments
later.
:::

### Data sheet

The Data sheet holds the 37 rows of forestry indicators. We will
inspect it next.

## Inspecting the Data sheet

::: {.callout-note collapse="true"}
### AI prompt to check your work
Use this **after** your own attempt. It should check your reasoning, not hand you the answer:

> *"I read the About, ReadMe, and Data sheets of `bc_disturbance_reforestation.xlsx` and concluded it has one row per fiscal year with area columns in hectares. Check whether my reading of the sheets and column meanings is complete and consistent, and tell me what a careful analyst might have missed on a first inspection. Do not summarise the file for me from scratch."*
:::

Open the Data sheet. Before computing anything, answer these five
questions:

1. **How many rows of data are there?** (Ignore the header.)
2. **How many columns?**
3. **What is the earliest and latest year?**
4. **Pick one column.** What does it measure, and in what units?
5. **Look at the first row.** Does anything look unexpected?

For this file:

```{r five-questions}
#| echo: false
tibble(
  Question = c("Rows of data",
               "Columns",
               "Earliest fiscal year",
               "Latest fiscal year",
               "Maximum Natural_Disturbance_ha (and its year)"),
  Answer = c(
    as.character(nrow(bc)),
    as.character(ncol(bc)),
    as.character(min(bc$Fiscal_Year)),
    as.character(max(bc$Fiscal_Year)),
    sprintf("%s ha in %d",
            format(max(bc$Natural_Disturbance_ha), big.mark = ","),
            bc$Fiscal_Year[which.max(bc$Natural_Disturbance_ha)])
  )
) |> kable()
```

The 2017 spike in `Natural_Disturbance_ha` is BC's record-breaking
wildfire season. The dataset would look like a flat trend without
it. The 2017 row alone is a lesson in why you *always look at the
extremes* of every column.

### The eight columns, one by one

| Column | Type | What it measures |
|---|---|---|
| `Fiscal_Year` | Integer year | The fiscal year the row describes (April–March, named by starting year). |
| `Clearcutting_ha` | Number (hectares) | Area harvested using the clearcutting silvicultural system. |
| `Clearcutting_with_reserves_ha` | Number (hectares) | Area harvested using clearcutting **with reserves** — some trees left standing. |
| `Partial_cutting_ha` | Number (hectares) | Area harvested using partial cutting (shelterwood, selection, etc.). |
| `Harvested_ha` | Number (hectares) | Sum of the three harvest types above. |
| `Natural_Disturbance_ha` | Number (hectares) | Area disturbed by natural causes (fire, pest, wind). |
| `Total_Disturbance_ha` | Number (hectares) | Total harvested + naturally disturbed area. |
| `Reforestation_ha` | Number (hectares) | Area planted or naturally regenerated that year. |

::: {.callout-tip}
### A small spot check

`Total_Disturbance_ha` is the sum of `Harvested_ha` and
`Natural_Disturbance_ha`. Pick any one row — say row 38 (year
2023) — and check that the formula `=E38 + F38 = G38` evaluates
to `TRUE`. If it does, the dataset is internally consistent.
:::

## Data types — and why they matter

Every cell in Excel has a **type**. The three you care about for
this chapter:

| Type | Example | Default alignment |
|---|---|---|
| Number | `216304.919` | right |
| Date | `2024-09-15` | right |
| Text | `Vancouver` | left |

Excel guesses the type when you open a file. Most of the time it
guesses right. But sometimes a number arrives stored as text
(maybe because of an extra apostrophe or a stray space) and is
then silently skipped when you `SUM` the column.

A quick test: select a numeric column and look at the status bar
at the bottom of the Excel window. It usually shows `Sum`,
`Average`, and `Count`. If `Count` matches your row count but
`Sum` is suspiciously small, some cells are stored as text.

## Navigation: freeze panes and a few shortcuts

A 37-row file does not need navigation tricks. A 37 000-row file
does. Learn these now while the file is small.

### Freeze panes

Click row 2 (the row below the header). Then choose
**View → Freeze Panes → Freeze Top Row**. The header row stays
visible while you scroll.

For wider files, select cell B2 and choose **View → Freeze Panes →
Freeze Panes**. Both the first row and the first column stay
visible.

### A few keyboard shortcuts

| Shortcut | What it does |
|---|---|
| `Ctrl + Arrow` | Jump to the edge of the data |
| `Ctrl + Home` | Jump to A1 |
| `Ctrl + End` | Jump to the last cell of the used range |
| `Ctrl + F` | Find |
| `Ctrl + ;` | Insert today's date as a value |

Learning a handful of shortcuts saves real time on bigger files.

## Number formats: display vs value

A single cell carries two pieces of information:

- The **value** (what the cell contains).
- The **display format** (how the value is shown).

These are different. Changing the format never changes the value.

### A worked example

Type `216304.919` into a cell. The cell displays `216304.919`.

Now right-click the cell → **Format Cells → Number**. Set decimal
places to 0. The cell now displays `216,305`. The underlying value
is still `216304.919`, just rounded for display.

Try **Format Cells → Currency** with two decimals. The cell shows
`$216,304.92`. Same underlying value.

Try **Format Cells → Custom** with the code `0 "ha"`. The cell
shows `216305 ha`. The string `"ha"` is part of the *format*, not
the value — formulas referencing this cell still treat it as a
plain number.

### Why this matters

Two cells can display the same thing but have different values.
If cell A holds `100000` and cell B holds `99999.999999`, both can
be formatted to display `100,000`. But `=A1=B1` returns `FALSE`.

This is the single most common silent bug in Excel-based analysis.
**Always be aware of the underlying value, not just the display.**

::: {.callout-tip}
### Two formats worth knowing

- `#,##0` — thousands separator, no decimals. Use for hectares.
- `yyyy-mm-dd` — ISO 8601 date format. Sorts correctly as text
  *and* as a date. Universally readable across regions.
:::

## Writing simple formulas

The five most-used aggregating formulas. Type each in any empty
cell on the Data sheet.

| Formula | What it does |
|---|---|
| `=SUM(H2:H38)` | Adds every value in column H (Reforestation_ha) |
| `=AVERAGE(G2:G38)` | Mean Total_Disturbance_ha across all rows |
| `=MIN(F2:F38)` | Smallest Natural_Disturbance_ha |
| `=MAX(F2:F38)` | Largest Natural_Disturbance_ha |
| `=COUNT(A2:A38)` | Number of rows with a numeric year |

On this file:

```{r expected-values}
#| echo: false
tibble(
  Formula = c("=SUM(H2:H38)", "=AVERAGE(G2:G38)",
              "=MIN(F2:F38)", "=MAX(F2:F38)", "=COUNT(A2:A38)"),
  Result = c(
    paste0(format(round(sum(bc$Reforestation_ha)), big.mark = ","), " ha"),
    paste0(format(round(mean(bc$Total_Disturbance_ha)), big.mark = ","), " ha"),
    paste0(format(round(min(bc$Natural_Disturbance_ha)), big.mark = ","), " ha"),
    paste0(format(round(max(bc$Natural_Disturbance_ha)), big.mark = ","), " ha"),
    "37"
  )
) |> kable()
```

## Checking whether values make sense

Before you trust any number you compute, run a spot check.

### Three quick checks

1. **Range check.** A mean should sit between the min and the max.
   If `MIN = 50,000`, `MAX = 250,000`, and your `AVERAGE` says
   `400,000`, something is wrong.
2. **Identity check.** If the dataset claims `Total_Disturbance_ha
   = Harvested_ha + Natural_Disturbance_ha`, pick one row and
   verify the identity.
3. **Order-of-magnitude check.** BC reforests roughly 200 000 ha
   per year. If your computed mean comes out at 20,000 or
   2,000,000, you have likely selected the wrong column or
   misformatted the result.

These three checks take seconds and catch most bugs.

::: {.callout-tip}
### Read the value, not the display

After you write a formula, click the result cell and look at the
formula bar. If the formula says `=SUM(B2:B38)` but you wanted
column H, you have selected the wrong column. Always read the
formula, not just the answer.
:::

## Building a data dictionary

::: {.callout-note collapse="true"}
### AI prompt to check your work
Use this **after** your own attempt. It should check your reasoning, not hand you the answer:

> *"Here is my data-dictionary entry for one column: `[paste your column name, type, unit, and plain-language description]`. Check whether my description matches the ReadMe, whether I named the correct type and unit, and what a reviewer might question. Do not rewrite the dictionary for me."*
:::

A data dictionary is a separate sheet that lists every column
with:

- **Variable** name (exactly as it appears in the data).
- **Description** in plain English — what does it measure?
- **Unit** of measurement, always.
- **Type** (number, text, date).
- **Notes** — anything a future reader should know.

The ReadMe sheet of `bc_disturbance_reforestation.xlsx` already
has one. Look at it.

### How to build your own

In any new workbook:

1. Add a new sheet, name it `ReadMe`.
2. In row 1, type the column headers: `Variable`, `Description`,
   `Unit`, `Type`, `Notes`.
3. In row 2 onward, add one row per column from your Data sheet.
4. Be **specific**. *"Reforestation, in hectares"* is okay. *"Area
   planted or naturally regenerated in that fiscal year, in
   hectares"* is better.

::: {.callout-warning}
### Test your dictionary

After you write a dictionary, hand the workbook to someone who
has never seen the data and ask them to compute one statistic
from it without asking you any questions. If they can, your
dictionary is good. If they ask what a column means, the
dictionary needs work.
:::

## The three habits: inspect first, compute second, verify always

If you take only one thing from this chapter, take this.

### Inspect first

Before you compute *any* number, look at the data. Open the file.
Read the About sheet. Read the ReadMe. Scroll through a few rows.
Find the extremes — the largest, the smallest, the oldest, the
newest.

This takes a few minutes and catches more bugs than any other
practice in data analysis.

### Compute second

Only after you have inspected do you start typing formulas. You
should be able to **predict** what your formula will return
before you press Enter. If the answer surprises you, that
surprise is information — either the data is interesting or the
formula is wrong.

### Verify always

Every time you compute a number, verify it against something:

- **Against a published total.** If the BC Ministry says the
  total reforestation in 2020 was X, your formula should equal X.
- **Against a spot check.** A mean cannot be larger than the
  max or smaller than the min.
- **Against a second method.** If you computed it with `SUM`,
  also compute it with a PivotTable (Chapter 3). If both agree,
  trust the number.

## Active learning

These short activities are designed to be done alongside the
chapter. Pick whichever match the pace of your session.

### Activity 1 — Read the About sheet aloud

In pairs, one learner reads the About sheet, the other listens
and asks one clarifying question. Switch.

The point is **slowing down**. Force yourself to read the
metadata before the data.

### Activity 2 — Five-question inspection

On your own, complete this inspection worksheet, then compare with a
partner. If your answers differ, go back to the About, ReadMe, and Data
sheets to resolve them.

| Check | Your answer |
|---|---|
| Number of rows | |
| Number of columns | |
| Earliest and latest fiscal year | |
| One column name, its unit, and meaning | |
| One value or pattern that seems important or surprising | |

### Activity 3 — Write five simple formulas

In your **working copy** (not the raw download), type each formula from
the *Writing simple formulas* section into an empty cell. Confirm each
result matches the expected value in the table.

### Activity 4 — Verify one identity

On any one row of the Data sheet, verify that
`Harvested_ha + Natural_Disturbance_ha = Total_Disturbance_ha`.
Use a formula like `=E2 + F2 - G2`. The result should be 0 (or
very close — Excel sometimes shows tiny non-zero values for
floating-point reasons).

### Activity 5 — Build a one-row data dictionary

Pick any column from the Data sheet. Write a one-sentence
description for it without looking at the ReadMe. Then compare
with the ReadMe. How close did you get?

## Practice Demo Lab

::: {.callout-important}
### Practice demo only
This is a **guided practice lab in the OER book**, designed to help you
learn the workflow before completing the official lab assignment on
**Canvas**. The Canvas lab is the graded assignment. Use this activity to
practice, compare your work with the reference solution, and learn how to
write an AI verification prompt.
:::

This demo lab has **six parts** — work them in order:

1. **Your attempt.** Work through the tasks below.
2. **Reference solution.** Compare against the **worked examples earlier in
   this chapter** (your public reference). The graded Canvas lab has its own
   private answer key — do not copy demo solutions into a Canvas submission.
3. **Compare your work with the reference.** Where did it match? Where did
   it differ? What caused the difference, and how did you fix it?
4. **Write your own AI verification prompt.** Ask an AI to *check* your
   reasoning, code, and outputs — not to produce the answer for you.
5. **Model AI verification prompt.**

   > *"I am doing a FRST 232 practice demo lab. My dataset is `[file name]`
   > (columns `[columns]`). I attempted `[paste your steps or code]` and got
   > `[paste the output]`. Explain what my work does line by line, tell me
   > whether it answers the task, and list the checks I can run to verify it
   > against the chapter's worked example. Do not give me the final answer."*

6. **Reflection: what changed after checking?** In two or three sentences —
   did your attempt hold up? what did you fix? what will you check first
   next time?


::: {.callout-note appearance="default"}
### Lab: Inspect a real BC dataset together

Work in groups of 3 or 4. Use one shared copy of
`bc_disturbance_reforestation.xlsx` (one machine per group, or
share the screen).

**Task 1 — Inspect the workbook.** Open the file. Take turns
reading the About sheet, the ReadMe sheet, and the Data sheet.
Each group member should be able to answer:

- Where did this file come from?
- What does each column measure?
- How many rows are there?

**Task 2 — Small data dictionary task.** As a group, pick **two
columns** of the Data sheet and write your own one-sentence
description of each in your own words. Compare your descriptions
to the ReadMe sheet's descriptions.

**Task 3 — Compute one or two summary values.** As a group,
choose one of these:

- Mean `Reforestation_ha` across all 37 fiscal years.
- Maximum `Natural_Disturbance_ha` and the year it happened.
- Total `Harvested_ha` across all 37 fiscal years.

Write the formula on the Data sheet. Then sanity-check the result
**with the check that fits your statistic**:

- **Mean** — must fall *between* the column's `MIN` and `MAX`.
- **Maximum** — must *equal* the single largest value in the column
  (and be ≥ every other row).
- **Total** — must be *at least as large as the* `MAX` (usually much
  larger), and never negative for an area column.

**Task 4 — Verify one formula.** Pick any row of the Data sheet.
Verify that
`Harvested_ha + Natural_Disturbance_ha = Total_Disturbance_ha`.
Use a formula like `=E10 + F10 - G10`. The result should be 0.

**Task 5 — Submit.** Submit either a short worksheet (a one-page
document with your group's answers to Tasks 1–4) **or** a
screenshot of the Data sheet showing all of your group's
formulas. Include all group members' names on the submission.

This lab is designed to fit comfortably inside a single
hands-on session.
:::

## Exercises

These are the take-home exercises that go with this chapter.

1. Download `bc_disturbance_reforestation.xlsx` from the BC Data
   Catalogue. Place it in the `data/` folder of your project.
   Open it and read the About sheet aloud.

2. In one sentence, write down what
   `Clearcutting_with_reserves_ha` measures **in your own words**,
   without looking at the ReadMe sheet. Then compare with the
   ReadMe.

3. Find the fiscal year with the smallest `Total_Disturbance_ha`.
   What year was it, and what was the value? *(Hint:
   `=MIN(...)` finds the value; sort the table or scan to find
   the year.)*

4. Add a new column `F_to_D` (reforestation-to-disturbance ratio)
   = `Reforestation_ha / Total_Disturbance_ha`. In what fraction
   of fiscal years was reforestation *greater than* total
   disturbance?

5. Set up the folder structure from the *Project folder
   structure* section above. Place a copy of the BC silviculture
   file in `data/`. Take a screenshot of your folder
   structure.

6. Save a cleaned copy of the workbook (rename it
   `bc_disturbance_reforestation_clean.xlsx`) into the
   `outputs/` folder. The raw file in `data/`
   should remain unchanged.

7. Apply ISO date format (`yyyy-mm-dd`) to any date in any file.
   Sort by that column. Confirm the sort works correctly.

## Optional, advanced

Items in this section are not required for the main path through
the chapter. Read them once you are comfortable with the basics
above.

### Named ranges

You can give a range of cells a name and refer to it by that
name in formulas:

- Select cells A2:A38 (the `Fiscal_Year` column).
- Type `years` into the Name Box (left of the formula bar).
- Press Enter.

Now anywhere in the workbook you can write `=MAX(years)` instead
of `=MAX(A2:A38)`. The formula is self-documenting. Manage named
ranges with **Formulas → Name Manager** (or `Ctrl + F3`).

Skip named ranges in Chapter 1 if you find them overwhelming —
the chapter does not depend on them. Chapter 2 will show
**structured tables**, which solve the same problem more cleanly.

### Custom number formats

Beyond the standard formats, Excel supports custom format codes.
For example, the custom format `0.000 "ha"` displays
`216304.919 ha`. The `"ha"` is part of the format, not the value.

This is useful for at-a-glance unit labels, but use sparingly —
too much custom formatting can make a workbook harder for a
collaborator to read.

## A glimpse ahead: From Excel to R

In Chapter 4 we will install R and re-open *this same file*
reproducibly. The Excel moves you learned in this chapter map
directly to R commands you will meet later. Here is a preview:

| What you did in Excel | What you will do in R (Ch 4–5) |
|---|---|
| Open the file by double-clicking | `read_excel(here("data", "bc_disturbance_reforestation.xlsx"), sheet = "Data")` |
| Look at the status bar `Count` | `dim(bc)` and `nrow(bc)` |
| Status-bar `Sum` of column H | `sum(bc$Reforestation_ha)` |
| Status-bar `Average` of column G | `mean(bc$Total_Disturbance_ha)` |
| Scan the Data sheet visually | `glimpse(bc)` and `summary(bc)` |
| Apply `MIN` and `MAX` | `range(bc$Natural_Disturbance_ha)` |

**The numbers will be exactly the same.** Same dataset, same
columns, same answers — but the R version is scriptable,
shareable, and reproducible.

You do not need to remember any of this now. The point is just to
see that R is not a fresh start. Everything you learn in Chapters
1–3 transfers directly. Chapter 4 will walk through each step in
detail.

## Self-assessment quiz {#sec-quiz-ch01}

::: {.content-visible when-format="pdf"}
*The self-assessment quiz is interactive: it appears in the online (HTML)
book, and you can also take it on Canvas. This PDF does not display the
interactive quiz questions.*
:::

::: {.content-visible when-format="html"}
Click the answer you think is correct. The right answer turns
green; wrong answers turn red and the correct one is revealed.
Explanations appear below each question.
:::

```{=html}
<style>
.quiz {
  background: #F4F7F4;
  border-left: 4px solid #2D6A4F;
  padding: 1em 1.5em;
  margin: 1.5em 0;
  border-radius: 6px;
}
.quiz .q {
  font-weight: 600;
  color: #1B4332;
  margin-bottom: 0.6em;
}
.quiz ul.qopts {
  list-style: none;
  padding: 0;
  margin: 0;
}
.quiz ul.qopts li {
  padding: 0.55em 0.9em;
  margin: 0.3em 0;
  background: white;
  border: 1px solid #d8e4d8;
  border-radius: 4px;
  cursor: pointer;
  transition: background 0.12s;
}
.quiz ul.qopts li:hover { background: #E8F1E9; }
.quiz ul.qopts li.correct {
  background: #C8E6C9;
  border-color: #2D6A4F;
  font-weight: 600;
}
.quiz ul.qopts li.wrong {
  background: #FFCDD2;
  border-color: #B05C50;
}
.quiz .explanation {
  display: none;
  margin-top: 0.7em;
  padding: 0.7em 1em;
  background: #FFF8E8;
  border-left: 3px solid #B7791F;
  border-radius: 4px;
  font-size: 0.95em;
}
.quiz .explanation.show { display: block; }
.quiz textarea {
  width: 100%;
  min-height: 4.5em;
  padding: 0.6em;
  border: 1px solid #d8e4d8;
  border-radius: 4px;
  font-family: inherit;
  font-size: 0.95em;
  resize: vertical;
  margin-top: 0.3em;
}
.quiz .reveal-btn {
  background: #2D6A4F;
  color: white;
  border: none;
  padding: 0.45em 0.9em;
  margin-top: 0.5em;
  border-radius: 4px;
  cursor: pointer;
  font-size: 0.9em;
}
.quiz .reveal-btn:hover { background: #1B4332; }
.quiz .sample {
  display: none;
  margin-top: 0.6em;
  padding: 0.6em 0.9em;
  background: #FFF8E8;
  border-left: 3px solid #B7791F;
  border-radius: 4px;
  font-size: 0.95em;
  font-style: italic;
}
.quiz .sample.show { display: block; }
</style>

<div class="quiz" id="q1" data-hints="What do you need to understand before the numbers can mean anything?||Which sheet records the source, licence, and citation? That context comes first.||The reading order runs from context, to the column dictionary, to the data itself.">
  <div class="q">1. Which sheet should you read first when you open <code>bc_disturbance_reforestation.xlsx</code>?</div>
  <ul class="qopts">
    <li data-correct="true">About</li>
    <li data-correct="false">Data</li>
    <li data-correct="false">ReadMe</li>
    <li data-correct="false">It does not matter</li>
  </ul>
  <div class="explanation">Read About first — it tells you the source, licence, and citation. Then ReadMe (variable dictionary). Then Data. Reading in the wrong order means you start computing before you know what the columns mean.</div>
</div>

<div class="quiz" id="q2" data-hints="Re-read what the About sheet says about who published it.||It is an open-government source, not a commercial or personal one.||Think BC Data Catalogue.">
  <div class="q">2. Where does the dataset come from?</div>
  <ul class="qopts">
    <li data-correct="false">A textbook publisher</li>
    <li data-correct="true">The BC Government, distributed through the BC Data Catalogue</li>
    <li data-correct="false">A personal blog</li>
    <li data-correct="false">A private consulting firm</li>
  </ul>
  <div class="explanation">Every dataset in this book is openly licensed and comes from a public source. This one is from Environmental Reporting BC and is distributed through the BC Data Catalogue under the Open Government Licence — British Columbia.</div>
</div>

<div class="quiz" id="q3" data-hints="What has one row per column of the Data sheet?||It tells you each column's meaning, units, and type.||That description is also called a data dictionary.">
  <div class="q">3. The ReadMe sheet is best understood as —</div>
  <ul class="qopts">
    <li data-correct="false">A backup copy of the data</li>
    <li data-correct="true">A data dictionary describing each column</li>
    <li data-correct="false">A blank sheet for personal notes</li>
    <li data-correct="false">The location of cleaned data</li>
  </ul>
  <div class="explanation">The ReadMe sheet (also called a data dictionary) has one row per column of the Data sheet — explaining what each column measures, its units, its type, and any notes.</div>
</div>

<div class="quiz" id="q4" data-hints="Ask: does formatting change what is stored, or only what is shown?||A formula on the cell still uses the full-precision number.||Display is not the same as value.">
  <div class="q">4. True or false: when you change the <em>display format</em> of a cell from "general" to "two decimals", the underlying value also changes.</div>
  <ul class="qopts">
    <li data-correct="false">True</li>
    <li data-correct="true">False</li>
  </ul>
  <div class="explanation">Display format is independent of value. The cell still holds <code>216304.919</code> even when it shows <code>216,304.92</code>. Formulas always operate on the underlying value, not the display.</div>
</div>

<div class="quiz" id="q5" data-hints="What two kinds of disturbance does the file track separately?||'Total' usually means the parts added together.||Check it on one row: harvested plus natural.">
  <div class="q">5. <code>Total_Disturbance_ha</code> should equal —</div>
  <ul class="qopts">
    <li data-correct="false">Just <code>Harvested_ha</code></li>
    <li data-correct="false">Just <code>Natural_Disturbance_ha</code></li>
    <li data-correct="true"><code>Harvested_ha + Natural_Disturbance_ha</code></li>
    <li data-correct="false"><code>Reforestation_ha</code></li>
  </ul>
  <div class="explanation">By definition, total disturbance is harvested area plus naturally disturbed area. Always verify this kind of identity on at least one row — it is a cheap way to confirm the data was assembled correctly.</div>
</div>

<div class="quiz" id="q6" data-hints="Which modifier key, held with an arrow, moves by blocks rather than one cell at a time?||It jumps to the next non-empty cell in the arrow's direction.||It is Ctrl held with an Arrow key.">
  <div class="q">6. What is the keyboard shortcut to jump to the edge of the data in Excel?</div>
  <ul class="qopts">
    <li data-correct="false">Ctrl + Tab</li>
    <li data-correct="true">Ctrl + Arrow</li>
    <li data-correct="false">Ctrl + Shift + V</li>
    <li data-correct="false">F1</li>
  </ul>
  <div class="explanation"><code>Ctrl + Arrow</code> jumps to the next non-empty cell in that direction — usually the edge of your data block.</div>
</div>

<div class="quiz" id="q7" data-hints="Which BC year had a record natural disturbance?||Think wildfire season.||2017.">
  <div class="q">7. The fiscal year with the largest <code>Natural_Disturbance_ha</code> in the BC dataset is —</div>
  <ul class="qopts">
    <li data-correct="false">1987</li>
    <li data-correct="false">2010</li>
    <li data-correct="true">2017</li>
    <li data-correct="false">2023</li>
  </ul>
  <div class="explanation">2017 was BC's record wildfire season — 218,275 ha of natural disturbance. It dominates the time series. Always look at the extremes of your data; the outliers tell you something about how the world worked that year.</div>
</div>

<div class="quiz" id="q8" data-hints="Raw, untouched downloads get their own folder.||Not outputs or deliverables.||data/.">
  <div class="q">8. Which folder is the right place to store the raw, untouched copy of a downloaded dataset?</div>
  <ul class="qopts">
    <li data-correct="true"><code>data/</code></li>
    <li data-correct="false"><code>outputs/</code></li>
    <li data-correct="false"><code>deliverables/</code></li>
    <li data-correct="false"><code>shared/</code></li>
  </ul>
  <div class="explanation">Raw data goes in <code>data/</code> and is never edited in place. Cleaning produces files that go in <code>outputs/</code>. Anything you submit goes in <code>deliverables/</code>.</div>
</div>

<div class="quiz" id="q9" data-hints="Freezing rows is a viewing option under one menu.||Look under View.||View, Freeze Panes, Freeze Top Row.">
  <div class="q">9. To freeze the top row so it stays visible while you scroll, which menu path do you use?</div>
  <ul class="qopts">
    <li data-correct="false">Home → Format → Freeze</li>
    <li data-correct="true">View → Freeze Panes → Freeze Top Row</li>
    <li data-correct="false">Data → Outline → Group</li>
    <li data-correct="false">Insert → Header & Footer</li>
  </ul>
  <div class="explanation"><strong>View → Freeze Panes → Freeze Top Row</strong> keeps row 1 in view while you scroll down.</div>
</div>

<div class="quiz" id="q10" data-hints="The comparison uses the stored values, not what is displayed.||The underlying numbers differ slightly.||FALSE.">
  <div class="q">10. Two cells both display <code>100,000</code>. Cell A's underlying value is <code>100000</code>; cell B's is <code>99999.999999</code>. The formula <code>=A1=B1</code> returns —</div>
  <ul class="qopts">
    <li data-correct="false">TRUE</li>
    <li data-correct="true">FALSE</li>
    <li data-correct="false">An error</li>
    <li data-correct="false">100,000</li>
  </ul>
  <div class="explanation">The comparison operates on the <em>values</em>, not the displays. The cells look identical but are not equal. Always remember: display ≠ value.</div>
</div>

<div class="quiz" id="q11" data-hints="Which format sorts correctly even when treated as plain text?||Year first, then month, then day, zero-padded.||ISO 8601 (yyyy-mm-dd).">
  <div class="q">11. Which date format sorts correctly both as text and as a date?</div>
  <ul class="qopts">
    <li data-correct="false"><code>15/09/2024</code></li>
    <li data-correct="false"><code>September 15, 2024</code></li>
    <li data-correct="true"><code>2024-09-15</code> (ISO 8601)</li>
    <li data-correct="false"><code>9/15/24</code></li>
  </ul>
  <div class="explanation">ISO 8601 (<code>yyyy-mm-dd</code>) is the only date format that sorts correctly when treated as text. It is also unambiguous across regions (no day/month confusion).</div>
</div>

<div class="quiz" id="q12" data-hints="Count only counts numeric cells.||Cells stored as text are skipped.||Some cells are stored as text and are not counted.">
  <div class="q">12. The status bar at the bottom of an Excel window shows <code>Sum</code>, <code>Average</code>, and <code>Count</code> when you select a numeric range. If <code>Count</code> is lower than the number of cells you selected, what is the most likely reason?</div>
  <ul class="qopts">
    <li data-correct="false">A sum that matches the visible total</li>
    <li data-correct="true">Some cells in the range are stored as text and being skipped</li>
    <li data-correct="false">An average is close to the median</li>
    <li data-correct="false">Excel is calculating slowly</li>
  </ul>
  <div class="explanation"><code>Count</code> only counts numeric cells. If you selected 37 cells but the status bar says <code>Count: 35</code>, two are stored as text and are being silently skipped by aggregation formulas.</div>
</div>

<div class="quiz" id="q13" data-hints="Never edit the original download in place.||Keep it untouched; save any changes separately.||Keep it unchanged in data/ and save edits elsewhere.">
  <div class="q">13. When you download <code>bc_disturbance_reforestation.xlsx</code>, what should you do with the raw file?</div>
  <ul class="qopts">
    <li data-correct="false">Open it and immediately start editing</li>
    <li data-correct="true">Keep it unchanged in a <code>data/</code> folder; save any cleaned version separately</li>
    <li data-correct="false">Delete the original after copying values</li>
    <li data-correct="false">Email it to yourself as a backup</li>
  </ul>
  <div class="explanation">Always keep the raw downloaded file unchanged. Save any cleaned or modified version with a different name (and ideally in a different folder, like <code>outputs/</code>) so you can always go back to the original.</div>
</div>

<div class="quiz" id="q14" data-hints="Documentation should travel with the data itself.||A sheet inside the workbook, one row per column.||A ReadMe (data dictionary) sheet.">
  <div class="q">14. Which approach is best for documenting the meaning of every column in a workbook?</div>
  <ul class="qopts">
    <li data-correct="false">Write a long email when you share the file</li>
    <li data-correct="false">Trust that the column headers are self-explanatory</li>
    <li data-correct="true">A dedicated ReadMe sheet in the workbook with one row per column</li>
    <li data-correct="false">A separate Word document</li>
  </ul>
  <div class="explanation">A ReadMe sheet travels with the data. Emails get lost. Headers are rarely self-explanatory.</div>
</div>

<div class="quiz" id="q15" data-hints="It is not the calendar year.||It starts in spring and ends the next spring.||April 1 to March 31.">
  <div class="q">15. The fiscal year in BC for forestry reporting runs —</div>
  <ul class="qopts">
    <li data-correct="false">January 1 to December 31</li>
    <li data-correct="true">April 1 to March 31 of the following calendar year</li>
    <li data-correct="false">July 1 to June 30</li>
    <li data-correct="false">September 1 to August 31</li>
  </ul>
  <div class="explanation">BC fiscal years run April–March. The fiscal year is named by its starting year — "Fiscal Year 2023" means April 2023 through March 2024.</div>
</div>

<div class="quiz" id="q16" data-hints="Total means add every value in the column.||Reforestation is column H, rows 2 to 38.||=SUM(H2:H38).">
  <div class="q">16. Which formula on the Data sheet computes the total reforestation across all 37 fiscal years?</div>
  <ul class="qopts">
    <li data-correct="false"><code>=AVERAGE(H2:H38)</code></li>
    <li data-correct="true"><code>=SUM(H2:H38)</code></li>
    <li data-correct="false"><code>=COUNT(H2:H38)</code></li>
    <li data-correct="false"><code>=MAX(H2:H38)</code></li>
  </ul>
  <div class="explanation"><code>SUM</code> adds every value in the range. <code>Reforestation_ha</code> is column H in this file. <code>AVERAGE</code> would give the mean per year, not the total.</div>
</div>

<div class="quiz" id="q17" data-hints="Re-running the same formula proves nothing.||Get to the answer a different way and compare.||Recompute it another way (or against a published total).">
  <div class="q">17. Which of these is the most reliable way to <em>verify</em> a number you computed?</div>
  <ul class="qopts">
    <li data-correct="false">Run the same formula twice</li>
    <li data-correct="true">Recompute the number a different way (a different formula, or against a published total) and see if you get the same answer</li>
    <li data-correct="false">Round it and check that it looks reasonable</li>
    <li data-correct="false">Ask AI whether it looks right</li>
  </ul>
  <div class="explanation">Verification means producing the same number a <em>second way</em>. Running the same formula twice just confirms that the formula has not been edited — not that it is correct.</div>
</div>

<div class="quiz" id="q18" data-hints="The course mantra — three verbs, in order.||Look before you compute; check after.||Inspect first, compute second, verify always.">
  <div class="q">18. The three habits of a good data analyst, in order, are —</div>
  <ul class="qopts">
    <li data-correct="false">Compute first, plot second, write last</li>
    <li data-correct="true">Inspect first, compute second, verify always</li>
    <li data-correct="false">Read the question, write the code, send the answer</li>
    <li data-correct="false">Open the file, sum the column, copy to email</li>
  </ul>
  <div class="explanation">Inspect (know your data before you compute). Compute (write the formulas). Verify (cross-check the answer). These three habits, applied every time, prevent more bugs than any single tool ever invented.</div>
</div>

<div class="quiz" id="q19" data-hints="DBH is a standard forestry tree measurement.||Forestry uses metric units for tree diameter.||Centimetres (cm).">
  <div class="q">19. You receive a file with a column called <code>DBH</code>. The ReadMe says "Diameter at breast height". What unit should you assume?</div>
  <ul class="qopts">
    <li data-correct="false">Centimetres — it is the most common</li>
    <li data-correct="false">Millimetres — for precision</li>
    <li data-correct="false">Inches — the column name sounds American</li>
    <li data-correct="true">Do not assume — check the ReadMe more carefully, or ask the data provider</li>
  </ul>
  <div class="explanation">Never assume units. Cm is most common for DBH in Canadian forestry, but it can be mm or inches depending on the source. Getting the unit wrong by a factor of 10 silently changes every later calculation.</div>
</div>

<div class="quiz" id="q20">
  <div class="q">20. <strong>Reflection.</strong> Describe one thing you noticed about <code>bc_disturbance_reforestation.xlsx</code> that surprised you — a column, a value, a pattern, anything.</div>
  <textarea placeholder="Type your answer here..."></textarea>
  <button class="reveal-btn" onclick="this.nextElementSibling.classList.toggle('show')">Show common answers</button>
  <div class="sample">Things learners typically notice: <em>the 2017 wildfire spike dwarfs every other year of natural disturbance</em> · <em>clearcutting-with-reserves was near-zero in the 1980s and grew through the 1990s — that is a real BC policy shift</em> · <em>reforestation is bigger than the total annual harvest for most years (because both planted and naturally regenerated area count)</em> · <em>the units are hectares — about the size of two football fields, which makes the numbers easier to visualize</em>.</div>
</div>

<script>
(function() {
  document.querySelectorAll('.quiz').forEach(function(q) {
    q.querySelectorAll('ul.qopts li').forEach(function(opt) {
      opt.addEventListener('click', function() {
        if (q.dataset.answered) return;
        q.dataset.answered = '1';
        var isCorrect = opt.getAttribute('data-correct') === 'true';
        if (isCorrect) {
          opt.classList.add('correct');
        } else {
          opt.classList.add('wrong');
          q.querySelectorAll('ul.qopts li').forEach(function(o) {
            if (o.getAttribute('data-correct') === 'true') {
              o.classList.add('correct');
            }
          });
        }
        var exp = q.querySelector('.explanation');
        if (exp) exp.classList.add('show');
      });
    });
  });
})();
</script>
```

## AI as a debugging companion

::: {.callout-tip}
### When to use AI in this chapter

- Ask AI to **define** a forestry term that appears in the ReadMe
  but is new to you. *"What does 'silvicultural system' mean?"*
- Ask AI to **explain** an Excel formula. Paste the formula
  verbatim and ask what it does.

### When *not* to use AI in this chapter

- AI is not the source of truth for what a BC dataset measures.
  The BC indicator page is. Use the ReadMe and the source URL
  first.
- Do not paste an exercise question into AI and copy the answer
  back. The point of an exercise is the thinking, not the result.
:::

## Reading

- *About* and *ReadMe* sheets of `bc_disturbance_reforestation.xlsx`.
- BC Government silviculture indicator background:
  <https://www.env.gov.bc.ca/soe/indicators/land/silviculture.html>
- BC Data Catalogue dataset page:
  <https://catalogue.data.gov.bc.ca/dataset/indicator-summary-data-trends-in-silviculture-in-b-c->
- Open Government Licence — British Columbia:
  <https://www2.gov.bc.ca/gov/content/data/open-data/open-government-licence-bc>

## Instructor notes {.unnumbered}

::: {.callout-note appearance="minimal" collapse="true"}
### Pacing and emphasis

- This chapter is deliberately beginner-friendly. The main path
  covers six skills: inspecting About / ReadMe / Data; identifying
  columns, units, and types; writing five simple formulas;
  checking whether values make sense; building a one-row data
  dictionary; setting up a folder structure.
- **Named ranges** and **custom number formats** are moved to the
  *Optional, advanced* section. Touch on them only if your group
  is moving quickly.
- The group lab is the central hands-on deliverable. Plan to give
  learners time to complete all five tasks in one session.

### Expected output checklist (matches the top-of-chapter callout)

1. Raw `bc_disturbance_reforestation.xlsx` in `data/`,
   unchanged.
2. Working `bc_disturbance_reforestation_clean.xlsx` in
   `outputs/` containing About / ReadMe / Data,
   a one-row data dictionary entry, and the five chapter formulas.
3. Group-lab worksheet or screenshot.

### Materials provided alongside this chapter

| File | Purpose |
|---|---|
| `bc_disturbance_reforestation.xlsx` | The raw open-data workbook (About + ReadMe + Data). Place in `data/`. |
| `bc_disturbance_reforestation_learner.xlsx` | Learner-facing exercises workbook with blanks and formula prompts. Distribute through the course site. |
| `bc_disturbance_reforestation_solutions.xlsx` | Instructor solutions with completed formulas and worked answers. **Do not distribute to learners.** |
| `quiz_ch01.html` | Standalone interactive quiz. Embedded in the rendered chapter; can also be hosted separately. |

### Suggested facilitation guidance for the Practice Demo Lab

*These are optional facilitation and feedback suggestions for instructors
running this book demo — **not** a grading key. The percentages below are
relative emphasis, not Canvas marks; the graded lab assignment is on Canvas.*

| Task | Weight | What to look for |
|---|---|---|
| Inspect workbook (Task 1) | 20 % | Group can articulate source, columns, row count. |
| Data dictionary (Task 2) | 20 % | Two columns described in own words; compared with ReadMe. |
| Summary value (Task 3) | 30 % | Correct formula AND a spot check. |
| Verify identity (Task 4) | 20 % | `Harvested + Natural = Total` confirmed on at least one row. |
| Submission (Task 5) | 10 % | Worksheet or screenshot is legible; group members named. |

### Common stumbling points

- Some learners try to compute the answer to Task 3 from the
  Summary sheet of the solutions file. Hide that file.
- Learners often write `=E2+F2=G2` and expect to see a number;
  they get `TRUE`/`FALSE` instead. That is correct — point out
  the difference between a comparison and an arithmetic
  expression.
- Some learners edit the raw
  `bc_disturbance_reforestation.xlsx` in place. Remind them that
  the raw file should stay unchanged.
- *Excel-to-R bridge:* If learners ask why we are starting in
  Excel "if we are going to move to R anyway", the answer is
  twofold: (a) the column meanings and forestry context are
  easier to learn with the data sitting visible on screen, and
  (b) Chapter 4 deliberately uses the *same file* to make the
  transition explicit. The "A glimpse ahead" table at the
  end of this chapter is the preview of that bridge.
:::
