---
title: "Clean and Summarize Data in Excel"
---

```{r setup-ch02}
#| include: false
library(tidyverse)
library(readxl)
library(here)
library(knitr)
library(janitor)
library(scales)
```

::: {.callout-note appearance="simple"}
## Data for this chapter
This chapter uses the **Ontario forest statistics** workbook
([`on_forest_statistics_1.xlsx`](../data/on_forest_statistics_1.xlsx){download="on_forest_statistics_1.xlsx"}).
Click the file name to download it directly, then save it in your `data/`
folder. For the source and details, see the
[Datasets Used in This Book](../datasets.qmd) page.
:::

# Chapter goals {.unnumbered}

By the end of this chapter you will be able to:

- **Sort** and **filter** large tables to surface unusual records.
- Detect **duplicate rows** on a single column and on a composite
  key.
- Use **structured tables** (`Ctrl+T`) to keep formulas connected
  to the data as it grows.
- Compute **single-cell summary statistics** — `SUM`, `AVERAGE`,
  `MEDIAN`, `MIN`, `MAX`, `COUNT`, `COUNTA`, `STDEV`.
- Use **relative, absolute, and mixed cell references** safely so
  a copied formula does what you expect.
- Apply **conditional aggregation** with `SUMIF`, `COUNTIF`, and
  `AVERAGEIF` (and their plural counterparts).
- Use **conditional logic** with `IF`, nested `IF`, and `IFS` to
  recode values into categories.
- **Look up** values from a reference table with `VLOOKUP` and
  `XLOOKUP`.
- Maintain a **cleaning log** that documents every change you made
  to the data.

# Expected output for this chapter {.unnumbered}

::: {.callout-tip appearance="default"}
## What you will hand in

1. The raw workbook **`on_forest_statistics_1.xlsx`** placed in
   your `chapter02/data/` folder, **unchanged from the Ontario Open
   Data download**.
2. A working copy you may modify, saved as
   **`on_forest_statistics_1_clean.xlsx`** in your
   `chapter02/outputs/` folder, containing:
   - the dataset converted to a **structured table** with a
     meaningful name,
   - a one-page **cleaning log sheet**,
   - a `size_class` column built with `IFS`, and
   - a small **forest-type lookup** sheet wired to the Data sheet
     with `XLOOKUP`.
3. A short **group-lab worksheet** (or screenshot) submitted to the
   course site.

This list is the same as the rubric for the chapter assignment.
Keep it in view while you work.
:::

# Why a cleaning log matters

Real data arrives messy. Duplicate stems. Inconsistent species
labels. Missing measurements. Unit ambiguity. The discipline
introduced in this chapter — *document every change, never edit
data silently* — is the foundation of every later chapter.

A **cleaning log** is a separate sheet (or document) that records
every change you made between the raw file and your analysis-ready
version:

| Date | Step | What I changed | Why |
|---|---|---|---|
| 2026-09-15 | 1 | Converted Data sheet into a structured table named `ontario_data` | Self-documenting formulas; auto-extends. |
| 2026-09-15 | 2 | Added a `size_class` column using `IFS` | Group small / medium / large areas for later summaries. |
| 2026-09-15 | 3 | Built a `codes` sheet mapping forest-type codes to plain-English names | Future joins / lookups need a clean reference. |
| 2026-09-15 | 4 | Verified the composite key has no duplicates | Confirms the file is a true cross-classification. |

Keep the cleaning log next to your data forever. When someone asks
*"why is your number different from the official report?"* — the
answer is in the cleaning log.

::: {.callout-warning}
## The two cardinal sins of data cleaning

1. **Editing raw data in place.** Once you change a cell, the
   original is gone. Always work on a *copy* of the raw file.
2. **Cleaning without a log.** Without documentation, your numbers
   become impossible to defend.

A cleaning log is your audit trail. It protects you when a manager,
reviewer, or collaborator questions your numbers months later.
:::

# The dataset for this chapter

This chapter uses one openly licensed Excel workbook from the
**Government of Ontario Open Data Catalogue**, alongside the BC
silviculture file you met in Chapter 1.

```{r ontario-summary}
#| echo: false
ontario <- read_excel(here("data", "on_forest_statistics_1.xlsx"),
                      sheet = "Data")

tibble(
  Property = c("File name", "Source", "Catalogue page",
               "Licence", "Rows in this teaching extract",
               "Columns", "Coverage", "Time period"),
  Value = c(
    "`on_forest_statistics_1.xlsx`",
    "Government of Ontario — *Forest Resources of Ontario 2021*",
    "<https://data.ontario.ca/dataset/forest-resources-of-ontario-2021>",
    "Open Government Licence — Ontario",
    as.character(nrow(ontario)),
    as.character(ncol(ontario)),
    "Region 3E only (teaching subset of the multi-region file)",
    "Snapshot for 2021"
  )
) |> kable()
```

::: {.callout-important}
## Where the data comes from — and what to do with it

This dataset is published by the **Government of Ontario** and
distributed through the **Ontario Open Data Catalogue**. Two rules
(same as Chapter 1):

1. **Keep the raw downloaded file unchanged.** Save the file as
   `on_forest_statistics_1.xlsx` exactly as you downloaded it. Do
   not sort it, edit it, or save over it.
2. **Save any cleaned or modified version separately** as
   `on_forest_statistics_1_clean.xlsx` in your `outputs/` folder.

The teaching extract in this book is **Region 3E only**. If you
need the full multi-region file (all landscape guide regions),
download it directly from the catalogue URL above.
:::

## Why this dataset matters for forestry

Where the BC silviculture file (Chapter 1) was a **time series** of
provincial totals, this Ontario file is a **cross-classification** —
every row is one combination of region × ownership × seral stage ×
forest type × age class, with the total area for that cell.

This shape is what a working forester sees most often:

- The Government of Ontario uses cross-classified tables like this
  to track stand structure across the landbase.
- BC's *Vegetation Resources Inventory* (VRI) and the National
  Forest Inventory both produce summary tables with the same
  shape.
- Any FRPA, FMP, or sustainable-forest-management plan eventually
  rolls up to a table like this.

Learning to navigate, summarize, and recode a cross-classified
file is one of the most transferable skills in forestry data work.

## The columns

| Column | Type | What it represents |
|---|---|---|
| `landscape_guide_region` | text | Ontario landscape guide region (here: 3E only) |
| `own_group` | text | Ownership group (here: CRN = Crown only) |
| `LGDS` | text | Seral stage code (P / S / I / M / L) |
| `SERAL_NAME` | text | Human-readable seral stage name |
| `PFT` | text | Provincial forest type code (3-letter) |
| `PFTName` | text | Human-readable forest type name |
| `AC_10` | number (year) | Age-class midpoint, in 10-year bins |
| `SumOfTotalHa` | number | Total area in hectares for that combination |

# Sorting

Click any cell inside the data, then choose **Data → Sort**.

A dialog opens. Choose:

- **Column**: `SumOfTotalHa`
- **Sort On**: Values
- **Order**: Largest to Smallest

Click OK. The largest cell rises to the top — a mature Conifer
Lowland area of over 300,000 ha. Look at the context: `MCL`,
`Mature`, an age class around 135. Does it make sense for a Region
3E mature lowland that has been growing for over a century? Yes —
old conifer lowland covers much of northern Ontario.

## Multi-key sort

Sort by `PFTName` (A→Z), then by `AC_10` (smallest to largest),
then by `SumOfTotalHa` (largest to smallest). Use **Data → Sort →
Add Level**.

Now the table reads like a story: for each forest type, you see
every age class in order, and within each age class the largest
area is on top. This is the human equivalent of `group_by() |>
arrange()` in R, which you will meet in Chapter 7.

::: {.callout-tip}
## Sorting changes the file

Unlike filtering, sorting **rearranges** the underlying rows. To
recover the original order:

- Add an `original_row_number` column before sorting (formula:
  `=ROW()` in the first row, dragged down), **or**
- Convert your data into a **structured table** first (next
  section) — sorts on a table do not destroy original order if
  you have an `original_row_number` column.
:::

# Filtering

Two ways: **AutoFilter** and **Slicers**.

## AutoFilter

Click any cell in your data, then **Data → Filter**. Each column
gets a dropdown arrow.

Click the arrow on `PFTName`, uncheck *Select All*, and check only
"Jack Pine". The view collapses to roughly two dozen rows. **The
data is unchanged** — you have only *hidden* rows, not deleted
them.

To re-show everything, click the arrow again and choose *Clear
Filter From*.

## Multiple filter criteria

Click the arrow on `SumOfTotalHa`. Choose **Number Filters →
Greater Than**. Type `50000`. The view filters to rows where the
area exceeds 50,000 hectares.

You can stack column filters: first filter `PFTName = "Jack Pine"`,
*then* filter `SumOfTotalHa > 5000`. Excel shows only the rows
matching **both** conditions — equivalent to
`filter(PFTName == "Jack Pine" & SumOfTotalHa > 5000)` in R.

## Slicers (with structured tables)

Slicers are visual filter panels. They only work on **structured
tables** (next section). Click any cell in a structured table,
then **Table Design → Insert Slicer**. Check which fields you
want.

Slicers are clickable buttons sitting beside the table. They are
visual, fun, and dangerous: a slicer can be on without the viewer
noticing. The number changes but the title does not. Always check
the slicer state before trusting a screenshot.

# Structured tables (`Ctrl+T`)

Excel's most undersold feature.

Click any cell in your data. Press **Ctrl+T** (Cmd+T on Mac).
Confirm *"My table has headers"*. Click OK.

What changed:

- The data became a *structured table* with a name (default
  `Table1`, which you should rename in the **Table Design** ribbon
  — try `ontario_data`).
- Headers got a coloured band.
- AutoFilter is automatically on.
- Formulas that reference table columns use *column names* instead
  of cell ranges:

```text
=AVERAGE(ontario_data[SumOfTotalHa])
```

Compare to the un-tabled version:

```text
=AVERAGE(H2:H180)
```

The structured version is **self-documenting** and
**self-resizing**. If you add a row at the bottom, the formula
automatically includes it.

::: {.callout-tip}
## Always start with Ctrl+T

When you open any new Excel file, the first thing to do is
convert each sheet's data into a structured table. The benefits
compound: slicers work, formulas are readable, growth is
automatic, and references are stable.
:::

## Renaming the table

Click anywhere in the table. Look at the **Table Design** ribbon.
In the leftmost field labelled *Table Name*, type a meaningful
name:

- ✅ `ontario_data`
- ✅ `bc_silviculture`
- ❌ `Table1`, `Sheet1!_FilterDatabase`

Use snake_case (lowercase with underscores). No spaces, no special
characters. This matches the R convention you will meet in
Chapter 4.

# Detecting duplicates

A real summary table should have **one row per combination** of
the grouping columns. If two rows have the same combination,
something went wrong upstream.

## On a single column

Select the column you want to check (e.g., `PFTName`). Choose
**Home → Conditional Formatting → Highlight Cells Rules →
Duplicate Values**. Duplicate values get highlighted.

For the Ontario file, every `PFTName` appears many times (because
the file has many seral × age combinations per forest type). That
is expected — not a bug. Duplicates on a *single* column only
signal a problem when the column should be a *unique key*.

## On a composite key

In a cross-classification like this file, the natural unique key
is `(landscape_guide_region, own_group, LGDS, PFTName, AC_10)`. A
composite key. To detect duplicates on it:

Add a helper column `composite_key` next to the data:

```text
=A2 & "|" & B2 & "|" & C2 & "|" & F2 & "|" & G2
```

The `|` separator prevents false collisions between, say, `"AB" +
"C"` and `"A" + "BC"`. If any concatenated key appears twice, the
cell highlights when you apply **Conditional Formatting →
Highlight Cells Rules → Duplicate Values** to the helper column.

```{r duplicate-check}
#| echo: false
dup_count <- ontario |>
  count(landscape_guide_region, own_group, LGDS, PFTName, AC_10) |>
  filter(n > 1) |>
  nrow()

cat(sprintf(
  "Verification in R: %d duplicate key combinations in the Ontario file.\n",
  dup_count))
```

For this file, the answer is zero. Good — the upstream pipeline is
clean.

## Removing duplicates

If you do find duplicates and want to remove them, **Data → Remove
Duplicates**. A dialog asks which columns to use as the key.
Always work on a copy. Document the removal in your cleaning log
with row counts before and after.

# Summary formulas — the essentials

The seven most-used aggregating functions:

| Formula | What it does | Example |
|---|---|---|
| `=SUM(range)` | Adds every numeric value | `=SUM(H2:H180)` |
| `=AVERAGE(range)` | Arithmetic mean (sum ÷ count) | `=AVERAGE(H2:H180)` |
| `=MEDIAN(range)` | Middle value when sorted; robust to outliers | `=MEDIAN(H2:H180)` |
| `=MIN(range)` | Smallest numeric value | `=MIN(H2:H180)` |
| `=MAX(range)` | Largest numeric value | `=MAX(H2:H180)` |
| `=COUNT(range)` | Number of *numeric* values | `=COUNT(H2:H180)` |
| `=COUNTA(range)` | Number of *non-empty* values (text + numeric) | `=COUNTA(A2:A180)` |
| `=STDEV(range)` | Standard deviation (sample) | `=STDEV(H2:H180)` |

The distinction between `COUNT` and `COUNTA` matters: `COUNT` skips
text cells. If your column should have 179 numeric entries and
`COUNT` says 177, two cells contain text or are blank.

## Worked examples on the Ontario file

```{r ontario-formulas-verify}
#| echo: false
tibble(
  Statistic = c("Total area covered (sum of SumOfTotalHa)",
                "Mean area per combination",
                "Median area per combination",
                "Maximum combination",
                "Number of rows"),
  Value = c(
    paste0(format(round(sum(ontario$SumOfTotalHa) / 1e6, 2), nsmall = 2), " million ha"),
    paste0(format(round(mean(ontario$SumOfTotalHa)), big.mark = ","), " ha"),
    paste0(format(round(median(ontario$SumOfTotalHa)), big.mark = ","), " ha"),
    paste0(format(round(max(ontario$SumOfTotalHa)), big.mark = ","), " ha"),
    as.character(nrow(ontario))
  )
) |> kable()
```

Look at the gap between mean and median. The mean is much larger —
a few very large rows pull the average up. The median is closer to
the *typical* combination. Both numbers are correct; they answer
different questions.

# Conditional aggregation: SUMIF, COUNTIF, AVERAGEIF

What if you want the sum, count, or average **only for rows that
match a condition**?

## `COUNTIF(range, criterion)`

How many rows describe Jack Pine?

```text
=COUNTIF(F2:F180, "Jack Pine")
```

The criterion can be a number, text, a comparison, or a wildcard:

| Criterion | Matches |
|---|---|
| `50000` | Exact 50000 |
| `">50000"` | Greater than 50000 |
| `"<>0"` | Anything not zero |
| `"Jack*"` | Text starting with "Jack" |
| `"*pine*"` | Text containing "pine" (case-insensitive) |

## `SUMIF(criterion_range, criterion, sum_range)`

What is the total area of "Mature" forest in Region 3E?

```text
=SUMIF(D2:D180, "Mature", H2:H180)
```

Read it: *"sum the `SumOfTotalHa` values for rows where
`SERAL_NAME` equals 'Mature'"*.

## `AVERAGEIF(criterion_range, criterion, average_range)`

What is the average area for "Jack Pine" combinations?

```text
=AVERAGEIF(F2:F180, "Jack Pine", H2:H180)
```

## `SUMIFS`, `COUNTIFS`, `AVERAGEIFS` — multiple criteria

The plural versions allow multiple criteria. The argument order
is different — the sum range comes **first**, then criterion
pairs:

```text
=SUMIFS(H2:H180, F2:F180, "Jack Pine", D2:D180, "Mature")
```

Read it: *"sum the `SumOfTotalHa` values for rows where `PFTName`
is 'Jack Pine' AND `SERAL_NAME` is 'Mature'"*.

::: {.callout-tip}
## When to use these vs a PivotTable

For a one-off number, `SUMIF` is fastest. For a whole table of
numbers (cross-classification across multiple categories), a
PivotTable is the right tool. We will meet PivotTables in
Chapter 3.
:::

# Safe cell references: the dollar sign

Type `=B2 + C2` in cell `D2`. Copy down. `D3` becomes `=B3 + C3`.
`D4` becomes `=B4 + C4`. The references **moved** with the
formula. That is a **relative** reference.

Type `=B2 + $C$2` in cell `D2`. Copy down. `D3` becomes `=B3 +
$C$2`. `D4` becomes `=B4 + $C$2`. The first part moves; the
second part **stays put**. The dollar signs locked it. That is an
**absolute** reference.

You can mix:

- `$A1` — column locked, row moves.
- `A$1` — row locked, column moves.
- `$A$1` — both locked.

## A worked bug

A common bug: you want to divide every value by the total. You
write `=H2 / H180` in cell `I2`, where `H180` is the grand total.
Copy down. `I3` becomes `=H3 / H181`. But `H181` is empty. Every
percentage below the first is wrong.

The fix is `=H2 / $H$180`. The denominator stays put.

::: {.callout-tip}
## A keyboard shortcut worth memorizing

While editing a cell reference, press **F4** to cycle through:

`$A$1` → `A$1` → `$A1` → `A1` → `$A$1` → ...

This is the fastest way to add or remove dollar signs.
:::

# Conditional logic: IF, nested IF, IFS

## `IF(condition, value_if_true, value_if_false)`

Basic form. Returns one of two values based on a logical test.

```text
=IF(H2 > 1000, "large", "small")
```

## Nested `IF` — multiple categories

Add a `size_class` column on the Ontario file based on
`SumOfTotalHa`:

```text
=IF(H2 < 1000, "small",
    IF(H2 < 50000, "medium", "large"))
```

This **nested IF** evaluates left to right. If the first condition
is true, return "small". Otherwise check the second. Otherwise
return "large".

## `IFS` — cleaner for many conditions

Available in Excel 2019+:

```text
=IFS(H2 < 1000, "small",
     H2 < 50000, "medium",
     TRUE, "large")
```

`IFS` evaluates each pair left to right. The first matching
condition returns its value. The trailing `TRUE` is a catch-all
for "everything else". Without it, unmatched rows return `#N/A`.

```{r case-when-verify}
#| echo: false
ontario |>
  mutate(size_class = case_when(
    SumOfTotalHa < 1000  ~ "small",
    SumOfTotalHa < 50000 ~ "medium",
    TRUE                 ~ "large"
  )) |>
  count(size_class) |>
  kable()
```

## `AND`, `OR`, `NOT` — combining conditions

```text
=IF(AND(H2 > 1000, F2 = "Jack Pine"), "flag", "")
```

Use `AND` when *all* conditions must be true. Use `OR` when *any*
condition can be true. Use `NOT` to reverse a condition.

# Lookups: VLOOKUP and XLOOKUP

Suppose you have a small reference sheet mapping forest-type
codes to plain-English names:

| Code | Name |
|---|---|
| PJK | Jack Pine |
| MCU | Conifer Upland |
| MCL | Conifer Lowland |
| MIX | Mixedwood |
| POP | Poplar |
| BWT | White Birch |
| PWR | Red and White Pine |
| TOL | Tolerant Hardwoods |

Place this on a new sheet called `codes`. Your data uses the code
in column E. In column I you want the plain-English name.

## `XLOOKUP` (Excel 2019+, recommended)

```text
=XLOOKUP(E2, codes!A:A, codes!B:B, "unknown")
```

Arguments: *value to find, where to look, what to return, default
if not found*. Clear and modern.

## `VLOOKUP` (everywhere)

```text
=VLOOKUP(E2, codes!A:B, 2, FALSE)
```

Arguments: *value to find, table range starting with the lookup
column, column number to return, exact match*. The `FALSE` at
the end forces exact match — **always include it**. The default
of `TRUE` (approximate match) silently returns the wrong value if
the lookup table is unsorted.

::: {.callout-important}
## When lookups go wrong

- **Missing key** — `#N/A`. Either the value is missing, or the
  key is misspelled. Wrap with `IFERROR` to flag rather than
  crash: `=IFERROR(VLOOKUP(...), "missing")`.
- **Duplicate key** — silently returns the *first* match. Always
  check that your lookup table has each key exactly once with
  `=COUNTIF(codes!A:A, "PJK")` (should return 1).
- **Wrong type** — looking up `"123"` (text) against `123`
  (number) fails. Inspect both sides.
- **Trailing spaces** — `"PJK "` does not match `"PJK"`. Use
  `=TRIM(...)` to clean strings first.
:::

# Putting it together — a worked cleaning sequence

You receive a colleague's spreadsheet. Here is a complete
cleaning sequence:

| Step | Action | Excel feature |
|---|---|---|
| 1 | Make a working copy of the raw file | File → Save As (with `_clean` suffix) |
| 2 | Open the working copy | — |
| 3 | Convert data to structured table | Ctrl+T, rename in Table Design |
| 4 | Apply `TRIM` to text columns to remove stray spaces | Helper column with `=TRIM(A2)` |
| 5 | Check for duplicates on the natural key | Conditional Formatting on composite key |
| 6 | Investigate missing values column by column | `=COUNTBLANK(...)` per column |
| 7 | Apply `IFS` to recode messy categories | New column with recoded values |
| 8 | Look up reference values where needed | `XLOOKUP` against reference sheet |
| 9 | Compute summary statistics | `SUMIF`, `COUNTIF`, `AVERAGEIF` |
| 10 | Write the cleaning log | New sheet, document every change |

Each step is reversible because we are working on a copy and
never modifying the raw values directly. Every step is documented
in the log. **This is what reproducible Excel looks like.**

# Active learning

These short activities are designed to be done alongside the
chapter.

## Activity 1 — Find duplicates in the Ontario file

Apply the duplicate-detection technique to the Ontario file. Use
the natural composite key. Report how many duplicates you found.

```{r act1-verify}
#| echo: false
dup_n <- ontario |>
  count(landscape_guide_region, own_group, LGDS, PFTName, AC_10) |>
  filter(n > 1) |>
  nrow()
cat(sprintf("Verification: %d duplicates on the natural key (should be zero).\n", dup_n))
```

## Activity 2 — Mature area by forest type

Using `SUMIF` (or `SUMIFS`), compute the total area of *mature*
forest in Region 3E for each forest type.

```{r act2-verify}
#| echo: false
ontario |>
  filter(SERAL_NAME == "Mature") |>
  group_by(PFTName) |>
  summarise(mature_ha = sum(SumOfTotalHa), .groups = "drop") |>
  arrange(desc(mature_ha)) |>
  mutate(mature_ha = comma(round(mature_ha))) |>
  kable()
```

## Activity 3 — Classify and count

On the Ontario file, add the `size_class` column from the
conditional-logic section. Then count rows in each class. Compare
with your neighbour.

## Activity 4 — Multi-criteria SUMIFS

Compute the total area where `PFTName = "Jack Pine"` **AND**
`SERAL_NAME = "Mature"` **AND** `AC_10 >= 95` (older mature jack
pine).

```{r act4-verify}
#| echo: false
ans <- ontario |>
  filter(PFTName == "Jack Pine", SERAL_NAME == "Mature", AC_10 >= 95) |>
  summarise(total_ha = sum(SumOfTotalHa)) |>
  pull()
cat(sprintf("Verification: %s ha of older mature jack pine.\n",
            format(round(ans), big.mark = ",")))
```

## Activity 5 — Build a lookup

Build a reference sheet `codes` with the 8 forest-type codes
(`PJK`, `MCU`, `MCL`, etc.) and their plain-English names.
Then use `XLOOKUP` (or `VLOOKUP`) to retrieve the name from the
`PFT` column of the Data sheet. Compare your retrieved column to
the existing `PFTName` column.

# Group lab activity

::: {.callout-note appearance="default"}
## Lab: Clean and summarize the Ontario forest file together

Work in groups of 3 or 4. Use one shared copy of
`on_forest_statistics_1.xlsx`.

**Task 1 — Save a working copy.** Save the workbook as
`on_forest_statistics_1_clean.xlsx` in your `outputs/` folder. The
raw file in `data/` should remain unchanged.

**Task 2 — Make it a structured table.** Convert the Data sheet
into a structured table. Rename it to `ontario_data`. Confirm
that autofilter dropdowns appear in the header row.

**Task 3 — Add a `size_class` column.** Use `IFS` (or nested
`IF`) to classify each row's `SumOfTotalHa` into
`small` / `medium` / `large` using the thresholds 1,000 and
50,000. Then use `COUNTIF` to count how many rows fall in each
class. Compare across groups.

**Task 4 — One conditional aggregation.** As a group, choose
**one** of these:

- Total area of *Jack Pine* in Region 3E (use `SUMIF`)
- Average area for *Mature* combinations (use `AVERAGEIF`)
- Count of rows where `SumOfTotalHa > 100,000` (use `COUNTIF`)

Write the formula. Confirm the result with a spot check.

**Task 5 — Submit.** Submit either a short worksheet (a one-page
document with your group's answers to Tasks 1–4) **or** a
screenshot of the cleaned workbook showing the structured table
and the new `size_class` column. Include all group members'
names on the submission.

This lab is designed to fit comfortably inside a single hands-on
session.
:::

# Exercises

These are the take-home exercises that go with this chapter.

1. Open `on_forest_statistics_1.xlsx`. Filter to `LGDS = "M"`
   (Mature). What is the total area, in hectares, of mature forest
   in Region 3E? Use `SUMIF`.

2. Open `bc_disturbance_reforestation.xlsx` (from Chapter 1). Add
   a `Decade` column using `=FLOOR(A2, 10)`. Then use `AVERAGEIF`
   to compute the mean `Harvested_ha` per decade.

3. Build a `codes` lookup sheet on the Ontario file with the eight
   forest-type codes and their plain-English names. Add a
   `PFTName_check` column that uses `XLOOKUP` to retrieve the name
   from the code. Use `IF` to flag any mismatches between the
   looked-up name and the existing `PFTName` column.

4. Compute the *coefficient of variation* (standard deviation
   divided by mean) of `SumOfTotalHa` on the Ontario file. Why
   might this be useful alongside the mean?

5. Apply conditional formatting to the `SumOfTotalHa` column with
   a three-colour scale: green for small, yellow for medium, red
   for large. Take a screenshot and submit.

6. Write a one-page cleaning log for the changes you made to the
   working copy of the Ontario file. Use the format shown at the
   top of this chapter. Submit the log next to the workbook.

7. **Optional, harder.** Build a small summary table on a new
   sheet of the Ontario file using only `SUMIFS` and `COUNTIFS`
   (no PivotTable yet — that's Chapter 3). Rows = forest type,
   columns = seral stage, values = sum of `SumOfTotalHa` and
   count of combinations.

# Optional, advanced

Items in this section are not required for the main path.

## Custom number formats

Beyond standard formats, Excel supports custom format codes. For
example, the custom format `#,##0 "ha"` displays `216,305 ha`.
The `"ha"` is part of the format, not the value — formulas still
treat the cell as a plain number.

## `INDEX` / `MATCH` — the classic lookup combo

`VLOOKUP` and `XLOOKUP` cover most cases. For old-Excel
compatibility or for *left-of-lookup-column* returns (which
`VLOOKUP` cannot do), use `INDEX` / `MATCH`:

```text
=INDEX(codes!B:B, MATCH(E2, codes!A:A, 0))
```

`MATCH(E2, codes!A:A, 0)` returns the row number of the match.
`INDEX(codes!B:B, n)` returns the value in column B at that row.
Combining them gives you a fully flexible lookup.

# A glimpse of Chapter 7: From Excel to R

In Chapter 7 (Summarize forestry data overall and by group) you
will do exactly what this chapter does — but in R. The Excel moves
you learned here map directly to R commands:

| What you did in Excel (Ch 2) | What you will do in R (Ch 6–7) |
|---|---|
| `=SUMIF(D2:D180, "Mature", H2:H180)` | `ontario \|> filter(SERAL_NAME == "Mature") \|> summarise(sum(SumOfTotalHa))` |
| `=COUNTIF(F2:F180, "Jack Pine")` | `ontario \|> filter(PFTName == "Jack Pine") \|> nrow()` |
| `=SUMIFS(H, F, "Jack Pine", D, "Mature")` | `ontario \|> filter(PFTName == "Jack Pine", SERAL_NAME == "Mature") \|> summarise(sum(SumOfTotalHa))` |
| `=IFS(H2<1000, "small", H2<50000, "medium", TRUE, "large")` | `mutate(size_class = case_when(SumOfTotalHa < 1000 ~ "small", SumOfTotalHa < 50000 ~ "medium", TRUE ~ "large"))` |
| `=VLOOKUP(E2, codes!A:B, 2, FALSE)` | `ontario \|> left_join(codes, by = "PFT")` |
| Structured table named `ontario_data` | `ontario` (the tibble) |
| Cleaning log on a separate sheet | A Markdown chunk in the same `.qmd` file |

**The intent is identical.** The R version is scriptable and
shareable; the Excel version is visible and forgiving. By the
time you reach Chapter 7, this whole chapter's logic will be
familiar — you will just be writing it in a different language.

# Self-assessment quiz {#sec-quiz-ch02}

Click the answer you think is correct. The right answer turns
green; wrong answers turn red and the correct one is revealed.
Explanations appear below each question.

```{=html}
<style>
.quiz {
  background: #F4F7F4;
  border-left: 4px solid #2D6A4F;
  padding: 1em 1.5em;
  margin: 1.5em 0;
  border-radius: 6px;
}
.quiz .q {
  font-weight: 600;
  color: #1B4332;
  margin-bottom: 0.6em;
}
.quiz ul.qopts {
  list-style: none;
  padding: 0;
  margin: 0;
}
.quiz ul.qopts li {
  padding: 0.55em 0.9em;
  margin: 0.3em 0;
  background: white;
  border: 1px solid #d8e4d8;
  border-radius: 4px;
  cursor: pointer;
  transition: background 0.12s;
}
.quiz ul.qopts li:hover { background: #E8F1E9; }
.quiz ul.qopts li.correct {
  background: #C8E6C9;
  border-color: #2D6A4F;
  font-weight: 600;
}
.quiz ul.qopts li.wrong {
  background: #FFCDD2;
  border-color: #B05C50;
}
.quiz .explanation {
  display: none;
  margin-top: 0.7em;
  padding: 0.7em 1em;
  background: #FFF8E8;
  border-left: 3px solid #B7791F;
  border-radius: 4px;
  font-size: 0.95em;
}
.quiz .explanation.show { display: block; }
.quiz textarea {
  width: 100%;
  min-height: 4.5em;
  padding: 0.6em;
  border: 1px solid #d8e4d8;
  border-radius: 4px;
  font-family: inherit;
  font-size: 0.95em;
  resize: vertical;
  margin-top: 0.3em;
}
.quiz .reveal-btn {
  background: #2D6A4F;
  color: white;
  border: none;
  padding: 0.45em 0.9em;
  margin-top: 0.5em;
  border-radius: 4px;
  cursor: pointer;
  font-size: 0.9em;
}
.quiz .reveal-btn:hover { background: #1B4332; }
.quiz .sample {
  display: none;
  margin-top: 0.6em;
  padding: 0.6em 0.9em;
  background: #FFF8E8;
  border-left: 3px solid #B7791F;
  border-radius: 4px;
  font-size: 0.95em;
  font-style: italic;
}
.quiz .sample.show { display: block; }
</style>

<div class="quiz" id="q1" data-hints="What happens to a reference with a dollar sign when the formula is copied?||The dollar signs pin the address so it does not shift.||$C$38 stays put; B2 moves relative to the new cell.">
  <div class="q">1. In the formula <code>=B2 / $C$38</code>, what does the dollar sign do?</div>
  <ul class="qopts">
    <li data-correct="false">Marks the cell as currency</li>
    <li data-correct="true">Locks the reference so it does not move when the formula is copied</li>
    <li data-correct="false">Indicates an error</li>
    <li data-correct="false">Nothing — Excel ignores it</li>
  </ul>
  <div class="explanation"><code>$C$38</code> stays as <code>$C$38</code> when copied. Without the dollar signs, the reference shifts relative to the formula's new position.</div>
</div>

<div class="quiz" id="q2" data-hints="Which counter includes text and dates, not only numbers?||COUNT is numbers-only; you need the all-non-empty version.||It is COUNTA.">
  <div class="q">2. Which function returns the <em>number of non-empty</em> cells in a range, including text cells?</div>
  <ul class="qopts">
    <li data-correct="false">COUNT</li>
    <li data-correct="true">COUNTA</li>
    <li data-correct="false">COUNTIF</li>
    <li data-correct="false">LEN</li>
  </ul>
  <div class="explanation"><code>COUNTA</code> counts every non-empty cell (text + numeric + dates). <code>COUNT</code> only counts numeric cells.</div>
</div>

<div class="quiz" id="q3" data-hints="You want to count how many times the key appears.||If it appears exactly once, the lookup is safe.||COUNTIF(codes!A:A, A2) should return 1.">
  <div class="q">3. To check that a lookup key appears <strong>exactly once</strong> in a reference table, which formula is best?</div>
  <ul class="qopts">
    <li data-correct="false">=VLOOKUP(A2, codes!A:B, 2, FALSE)</li>
    <li data-correct="true">=COUNTIF(codes!A:A, A2) — should return 1</li>
    <li data-correct="false">=IF(A2 = codes!A2, "OK", "not OK")</li>
    <li data-correct="false">=ISNUMBER(VLOOKUP(...))</li>
  </ul>
  <div class="explanation">If the result is 1, the lookup is safe. If it is 2 or more, your <code>VLOOKUP</code> would silently return the first match.</div>
</div>

<div class="quiz" id="q4" data-hints="What lets someone reproduce your clean file from the raw one?||Not just the final values — the changes and the reasons for them.||Record every change between raw and clean, with a reason.">
  <div class="q">4. A cleaning log should record:</div>
  <ul class="qopts">
    <li data-correct="false">Only the final cleaned data</li>
    <li data-correct="true">Every change you made between raw and clean, with a reason</li>
    <li data-correct="false">Only changes you cannot reverse</li>
    <li data-correct="false">The data dictionary</li>
  </ul>
  <div class="explanation">A cleaning log captures <strong>every</strong> change with a reason. This is what lets you defend your numbers months later.</div>
</div>

<div class="quiz" id="q5" data-hints="You are mapping exact codes to labels, not numeric ranges.||A long nested IF works but is hard to read and audit.||IFS (or SWITCH) is built for exact-match recoding.">
  <div class="q">5. Which function is best for recoding the value in a cell to one of several exact-match labels (e.g., "P" → "Pre-sapling", "S" → "Sapling")?</div>
  <ul class="qopts">
    <li data-correct="false">VLOOKUP</li>
    <li data-correct="false">A long chain of nested IF</li>
    <li data-correct="true">IFS (or SWITCH)</li>
    <li data-correct="false">SUMIF</li>
  </ul>
  <div class="explanation"><code>IFS</code> is designed for exact-match recoding. Cleaner and more readable than a long chain of nested <code>IF</code>.</div>
</div>

<div class="quiz" id="q6" data-hints="Does filtering remove rows, or just hide them?||Clearing the filter brings the rows back.||True — AutoFilter hides; only sorting rearranges permanently.">
  <div class="q">6. You apply AutoFilter to hide rows where <code>SumOfTotalHa &lt; 1000</code>. The other rows are not deleted — they are just hidden. True or false?</div>
  <ul class="qopts">
    <li data-correct="true">True — AutoFilter hides; it does not delete</li>
    <li data-correct="false">False — AutoFilter deletes the hidden rows</li>
  </ul>
  <div class="explanation">AutoFilter is reversible. Sorting, by contrast, <strong>rearranges</strong> the rows permanently.</div>
</div>

<div class="quiz" id="q7" data-hints="SUMIFS totals one range subject to criteria pairs.||Read each pair: a column and the value it must equal.||It sums H where F is Jack Pine AND D is Mature.">
  <div class="q">7. The formula <code>=SUMIFS(H2:H180, F2:F180, "Jack Pine", D2:D180, "Mature")</code> does what?</div>
  <ul class="qopts">
    <li data-correct="false">Sums every numeric value in column H</li>
    <li data-correct="true">Sums the H values for rows where column F is "Jack Pine" AND column D is "Mature"</li>
    <li data-correct="false">Returns "Jack Pine" if column D is "Mature"</li>
    <li data-correct="false">Counts the rows that match both conditions</li>
  </ul>
  <div class="explanation"><code>SUMIFS</code> sums the first range, subject to one or more criterion ranges.</div>
</div>

<div class="quiz" id="q8" data-hints="Think about what happens to formulas and ranges as the data grows.||Tables let you reference columns by name and auto-extend.||Formulas use column names and stay correct when rows are added.">
  <div class="q">8. What is the main advantage of converting your data into a structured table (Ctrl+T) before writing formulas?</div>
  <ul class="qopts">
    <li data-correct="false">It makes Excel faster</li>
    <li data-correct="true">Formulas use column names instead of cell ranges, and they auto-extend when rows are added</li>
    <li data-correct="false">It encrypts the data</li>
    <li data-correct="false">It is required for sorting</li>
  </ul>
  <div class="explanation">Structured tables make formulas self-documenting (<code>=AVERAGE(tbl[col])</code> instead of <code>=AVERAGE(H2:H180)</code>) and self-extending.</div>
</div>

<div class="quiz" id="q9" data-hints="Check the file's source/About note.||It is a provincial open-data source, not commercial.||Government of Ontario open data.">
  <div class="q">9. Where does the Ontario dataset come from?</div>
  <ul class="qopts">
    <li data-correct="false">A private consulting firm</li>
    <li data-correct="true">The Government of Ontario, distributed through the Ontario Open Data Catalogue</li>
    <li data-correct="false">A textbook publisher</li>
    <li data-correct="false">A research paper</li>
  </ul>
  <div class="explanation">The file is from the Ministry of Natural Resources and Forestry, published under the Open Government Licence — Ontario.</div>
</div>

<div class="quiz" id="q10" data-hints="Why are there so many rows for one region?||Each row is one combination of several categories.||One unique combination of categorical variables, with one numeric value.">
  <div class="q">10. The Ontario file in this chapter is a <strong>cross-classification</strong>. That means —</div>
  <ul class="qopts">
    <li data-correct="false">It has only one column of data</li>
    <li data-correct="true">Each row is one unique combination of categorical variables, with one numeric value</li>
    <li data-correct="false">The data is encrypted</li>
    <li data-correct="false">It is sorted by region</li>
  </ul>
  <div class="explanation">Every row is one unique combination of region × ownership × seral stage × forest type × age class, with the total area for that cell.</div>
</div>

<div class="quiz" id="q11" data-hints="No single column identifies a row uniquely here.||The key is the set of classifying columns taken together.||region + ownership + seral stage + forest type + age class.">
  <div class="q">11. What is the natural composite key for the Ontario file?</div>
  <ul class="qopts">
    <li data-correct="false">PFTName alone</li>
    <li data-correct="false">SumOfTotalHa alone</li>
    <li data-correct="true">landscape_guide_region, own_group, LGDS, PFTName, AC_10</li>
    <li data-correct="false">There is no key</li>
  </ul>
  <div class="explanation">The combination of region, ownership, seral stage, forest type, and age class uniquely identifies each row. If any combination repeats, something went wrong upstream.</div>
</div>

<div class="quiz" id="q12" data-hints="F4 changes how a reference is anchored.||It toggles the dollar signs on the reference.||It cycles absolute, mixed, and relative forms.">
  <div class="q">12. Press <strong>F4</strong> while editing a cell reference. What does it do?</div>
  <ul class="qopts">
    <li data-correct="false">Closes the file</li>
    <li data-correct="true">Cycles the reference through absolute / mixed / relative forms ($A$1 → A$1 → $A1 → A1)</li>
    <li data-correct="false">Inserts a function</li>
    <li data-correct="false">Triggers AutoFilter</li>
  </ul>
  <div class="explanation">F4 is the fastest way to add or remove dollar signs.</div>
</div>

<div class="quiz" id="q13" data-hints="Never overwrite the raw download.||Use a clear suffix and a separate folder.||A *_clean.xlsx in outputs/, with the raw left untouched in data/.">
  <div class="q">13. After cleaning the Ontario file, what should the file be named and where should it live?</div>
  <ul class="qopts">
    <li data-correct="false">Overwrite on_forest_statistics_1.xlsx in data/</li>
    <li data-correct="true">Save as on_forest_statistics_1_clean.xlsx in outputs/</li>
    <li data-correct="false">Email it to yourself</li>
    <li data-correct="false">Delete the original</li>
  </ul>
  <div class="explanation">The raw file should stay unchanged in <code>data/</code>. The cleaned version goes in <code>outputs/</code> with a clear suffix.</div>
</div>

<div class="quiz" id="q14" data-hints="AVERAGEIF averages values that meet one condition.||The condition is on the seral-stage column.||It returns the mean of H where the class is Mature.">
  <div class="q">14. <code>=AVERAGEIF(D2:D180, "Mature", H2:H180)</code> returns —</div>
  <ul class="qopts">
    <li data-correct="false">The total area of mature forest</li>
    <li data-correct="true">The mean SumOfTotalHa for rows where SERAL_NAME = "Mature"</li>
    <li data-correct="false">The number of mature rows</li>
    <li data-correct="false">The largest mature area</li>
  </ul>
  <div class="explanation"><code>AVERAGEIF</code> averages the third range, subject to the criterion. Use <code>SUMIF</code> for total, <code>COUNTIF</code> for count.</div>
</div>

<div class="quiz" id="q15" data-hints="What does the 4th argument TRUE mean in VLOOKUP?||Approximate match needs a sorted table and can silently return a wrong value.||TRUE = approximate; prefer FALSE for exact match.">
  <div class="q">15. Which is true about <code>=VLOOKUP(A2, codes!A:B, 2, TRUE)</code>?</div>
  <ul class="qopts">
    <li data-correct="false">It is the safest form of VLOOKUP</li>
    <li data-correct="true">The TRUE means approximate match — dangerous if the lookup table is unsorted; can silently return the wrong value</li>
    <li data-correct="false">It returns an error if the key is missing</li>
    <li data-correct="false">It looks up the value in column B</li>
  </ul>
  <div class="explanation">Always use <code>FALSE</code> (exact match) for VLOOKUP unless you really mean approximate. The default of <code>TRUE</code> is dangerous.</div>
</div>

<div class="quiz" id="q16" data-hints="What pulls a mean above the median?||A few very large rows stretch the mean upward.||The distribution is right-skewed.">
  <div class="q">16. The mean of <code>SumOfTotalHa</code> in the Ontario file is much larger than the median. What does this tell you?</div>
  <ul class="qopts">
    <li data-correct="false">The data is wrong</li>
    <li data-correct="true">The distribution is right-skewed — a few very large rows pull the mean up</li>
    <li data-correct="false">The data is sorted</li>
    <li data-correct="false">Half the rows are missing</li>
  </ul>
  <div class="explanation">When mean &gt; median by a lot, the distribution is right-skewed. The median is the more honest measure of a "typical" row in this case.</div>
</div>

<div class="quiz" id="q17" data-hints="Structured tables reference columns by name, not by cell range.||The syntax is TableName[ColumnName].||=AVERAGE(ontario_data[SumOfTotalHa]).">
  <div class="q">17. You convert your Ontario data to a structured table named <code>ontario_data</code>. Which formula references the SumOfTotalHa column correctly?</div>
  <ul class="qopts">
    <li data-correct="false">=AVERAGE(SumOfTotalHa)</li>
    <li data-correct="true">=AVERAGE(ontario_data[SumOfTotalHa])</li>
    <li data-correct="false">=AVERAGE("SumOfTotalHa")</li>
    <li data-correct="false">=AVERAGE(@SumOfTotalHa)</li>
  </ul>
  <div class="explanation">Structured-table syntax uses <code>table_name[column_name]</code>. Self-documenting and self-extending.</div>
</div>

<div class="quiz" id="q18" data-hints="Recall the usual VLOOKUP failure causes.||Trailing spaces, a text/number type mismatch, and a missing/TRUE match flag all break it.||The lookup table being on the same sheet is harmless — that is the answer.">
  <div class="q">18. Which of these does NOT cause a VLOOKUP to fail?</div>
  <ul class="qopts">
    <li data-correct="false">A trailing space in the lookup key</li>
    <li data-correct="false">The key is numeric but the lookup column is text</li>
    <li data-correct="true">The lookup table is on the same sheet as the data</li>
    <li data-correct="false">The exact-match flag is missing or TRUE</li>
  </ul>
  <div class="explanation">VLOOKUP works regardless of which sheet the lookup table is on. Trailing spaces, type mismatches, and the approximate-match default are the real culprits.</div>
</div>

<div class="quiz" id="q19" data-hints="What do structured-table references point to — cells or column names?||Column-name references survive sorts and filters.||Formulas stay connected to the right columns after rows are rearranged.">
  <div class="q">19. Why is sorting a structured table different from sorting a plain range?</div>
  <ul class="qopts">
    <li data-correct="false">Structured tables cannot be sorted</li>
    <li data-correct="true">Structured tables keep formulas connected to the right columns even after rows are rearranged</li>
    <li data-correct="false">Sorting a structured table makes Excel slower</li>
    <li data-correct="false">There is no difference</li>
  </ul>
  <div class="explanation">Structured-table column references stay valid through sorts and filters because they refer to column names, not cell ranges.</div>
</div>

<div class="quiz" id="q20">
  <div class="q">20. <strong>Reflection.</strong> Describe one formula you wrote this chapter that did not work the first time. What was wrong, and how did you find out?</div>
  <textarea placeholder="Type your answer here..."></textarea>
  <button class="reveal-btn" onclick="this.nextElementSibling.classList.toggle('show')">Show common answers</button>
  <div class="sample">Common answers: <em>VLOOKUP returned #N/A because of a trailing space in the lookup key</em> · <em>SUMIFS gave 0 because the criterion was a number formatted as text</em> · <em>a relative reference shifted incorrectly when copied down because I forgot the dollar signs on the denominator</em> · <em>IFS returned #N/A on some rows because the final TRUE catch-all was missing</em>.</div>
</div>

<script>
(function() {
  document.querySelectorAll('.quiz').forEach(function(q) {
    q.querySelectorAll('ul.qopts li').forEach(function(opt) {
      opt.addEventListener('click', function() {
        if (q.dataset.answered) return;
        q.dataset.answered = '1';
        var isCorrect = opt.getAttribute('data-correct') === 'true';
        if (isCorrect) {
          opt.classList.add('correct');
        } else {
          opt.classList.add('wrong');
          q.querySelectorAll('ul.qopts li').forEach(function(o) {
            if (o.getAttribute('data-correct') === 'true') {
              o.classList.add('correct');
            }
          });
        }
        var exp = q.querySelector('.explanation');
        if (exp) exp.classList.add('show');
      });
    });
  });
})();
</script>
```

# AI as a debugging companion

::: {.callout-tip}
## Useful prompts for this chapter

- *"In Excel, my formula `=VLOOKUP(A2, codes!A:B, 2, FALSE)`
  returns `#N/A` for some rows. The values in column A look the
  same as in the codes sheet. What are the three most common
  reasons VLOOKUP fails when the values look right?"*

- *"Explain step by step what this formula does:
  `=IFS(C2 < 1000, "small", C2 < 50000, "medium", TRUE, "large")`"*

- *"I want to write a `SUMIFS` that adds up the `SumOfTotalHa`
  column whenever `PFTName` is one of `{Jack Pine, Mixedwood,
  Poplar}`. Can I do this in a single formula?"*

## Verifying AI's answers

After AI tells you why your `VLOOKUP` fails, **check**:

1. Are there trailing spaces? `=LEN(A2)` and `=LEN(codes!A2)`
   should match.
2. Are both values the same type? `=ISTEXT(A2)` and
   `=ISTEXT(codes!A2)`.
3. Is the case the same? VLOOKUP is case-insensitive in Excel,
   but worth eliminating.

AI's answer is a *starting point*, not the final answer. Always
verify against the data.

## When not to use AI

- Do not use AI to *write the cleaning log for you*. The log is a
  record of *your* decisions.
- Do not use AI to *invent* lookup values that are not in the
  reference table. If a value is missing, document the gap.
:::

# Reading

- Microsoft Excel official documentation on [VLOOKUP](https://support.microsoft.com/en-us/office/vlookup-function-0bbc8083-26fe-4963-8ab8-93a18ad188a1).
- Microsoft Excel official documentation on [XLOOKUP](https://support.microsoft.com/en-us/office/xlookup-function-b7fd680e-6d10-43e6-84f9-88eae8bf5929).
- Ontario open-data dataset page: <https://data.ontario.ca/dataset/forest-resources-of-ontario-2021>
- Open Government Licence — Ontario: <https://www.ontario.ca/page/open-government-licence-ontario>

# Instructor notes {.unnumbered}

::: {.callout-note appearance="minimal" collapse="true"}
## Pacing and emphasis

- This chapter is heavier than Chapter 1. Plan to spread it
  across multiple sessions.
- The main path covers seven skills: sort, filter, structured
  tables, summary formulas, conditional aggregation, conditional
  logic, lookups. The cleaning log ties them together.
- **Custom number formats** and **INDEX/MATCH** are moved to the
  *Optional, advanced* section. Touch on them only if your group
  is moving quickly or already comfortable.
- The group lab is the central hands-on deliverable. Plan for
  one full session.

## Expected output checklist (matches the top-of-chapter callout)

1. Raw `on_forest_statistics_1.xlsx` in `chapter02/data/`,
   unchanged.
2. Working `on_forest_statistics_1_clean.xlsx` in
   `chapter02/outputs/` containing a structured table, a cleaning
   log sheet, a `size_class` column built with `IFS`, and a small
   `codes` lookup sheet wired with `XLOOKUP`.
3. Group-lab worksheet or screenshot.

## Materials provided alongside this chapter

| File | Purpose |
|---|---|
| `on_forest_statistics_1.xlsx` | The raw open-data workbook (About + ReadMe + Data). Place in `data/`. |
| `on_forest_statistics_1_learner.xlsx` | Learner-facing exercises workbook with blanks and formula prompts. Distribute through the course site. |
| `on_forest_statistics_1_solutions.xlsx` | Instructor solutions with completed formulas and worked answers. **Do not distribute to learners.** |
| `quiz_ch02.html` | Standalone interactive quiz. Embedded in the rendered chapter; can also be hosted separately. |

## Suggested grading guidance for the group lab

| Task | Weight | What to look for |
|---|---|---|
| Save working copy (Task 1) | 10 % | Cleaned file in `outputs/`; raw file untouched. |
| Structured table (Task 2) | 20 % | Table named meaningfully; autofilter present. |
| size_class column (Task 3) | 30 % | `IFS` formula correct; counts in each class confirmed. |
| Conditional aggregation (Task 4) | 30 % | Correct formula AND a spot check. |
| Submission (Task 5) | 10 % | Worksheet or screenshot is legible; group members named. |

## Common stumbling points

- Some learners forget the trailing `TRUE` in `IFS` and get
  `#N/A` on rows that don't match any explicit condition.
- Some try to write a structured-table formula by typing the
  column name without the table prefix. The right form is
  `ontario_data[SumOfTotalHa]`, not `SumOfTotalHa` alone.
- A few will edit the raw `on_forest_statistics_1.xlsx` in place.
  Remind them often that the raw file stays unchanged.
- *Excel-to-R bridge:* The "A glimpse of Chapter 7" section at
  the end of this chapter previews how the same logic appears in
  R. Refer to it when learners ask why we are still in Excel —
  the bridge is concrete and forward-pointing.
:::
