---
title: "Excel Visualization and PivotTables"
---

```{r setup-ch03}
#| include: false
library(tidyverse)
library(readxl)
library(here)
library(knitr)
library(scales)
```

# Chapter goals {.unnumbered}

By the end of this chapter you will be able to:

- Choose the **right chart type** for the question being asked
  (bar, column, line, scatter).
- Build clean Excel charts with **titles, axis labels, and
  captions** that another reader could understand on their own.
- Construct a **PivotTable** from a tidy dataset, choosing what
  goes in Rows, Columns, Values, and Filters.
- Switch the **aggregation function** (sum, count, average) inside
  a PivotTable and predict what each one returns.
- Connect a **PivotChart** to a PivotTable and keep them in sync.
- Add **Slicers** to a PivotTable to give a non-technical reader
  controls to explore the data.
- **Refresh** PivotTables when the underlying data changes, and
  recognize the silent-bug risk if you forget.
- Recreate a PivotTable manually with **`SUMIFS` and `COUNTIFS`** —
  the bridge to how the same logic works in R.
- Export a finished chart to a **PNG** with a caption ready for a
  report or a course-site submission.

# Expected output for this chapter {.unnumbered}

::: {.callout-tip appearance="default"}
## What you will hand in

1. **One PivotTable + PivotChart** built on
   `on_forest_statistics_1.xlsx`, answering one forestry question
   of your choice (e.g., "Total mature area by forest type" or
   "Count of combinations by seral stage and age class"). Exported
   as a **PNG with caption**.
2. **One time-series chart** built on
   `bc_disturbance_reforestation.xlsx`, showing reforestation,
   harvest, and natural disturbance from 1987 to 2023 on the same
   axes. Exported as a **PNG with caption**.
3. **One short worksheet** matching common forestry questions to
   the correct chart type (provided in the learner workbook).
4. A short **group-lab worksheet** (or screenshot) submitted to the
   course site.

This list is the same as the rubric for the chapter assignment.
Keep it in view while you work.
:::

# Why visualization matters

A number in a cell is a fact. A chart is an argument. The same
data, in the same workbook, can become a *story about reforestation
finally catching up to harvest* (a single time-series chart) or a
*story about a record-breaking wildfire season* (a bar chart with
2017 visibly above the rest). Choosing how to display data is part
of the analysis, not an after-the-fact decoration.

The transition matrix for this course explicitly treats Chapter 3
as the **bridge** from raw Excel work into tidy-data thinking. The
PivotTable you build in this chapter is conceptually the same as a
`group_by() |> summarise()` pipeline in R — which you will write in
Chapter 7. The chart you build is the conceptual ancestor of
`ggplot2`, which you will meet in Chapter 9.

For now, the goal is to be **fluent** in the Excel versions, so
that when we move to R, you already know what each command is
doing.

# The two datasets for this chapter

This chapter uses two openly licensed files you have already met:

```{r datasets-summary}
#| echo: false
bc <- read_excel(here("data", "bc_disturbance_reforestation.xlsx"),
                 sheet = "Data")
ontario <- read_excel(here("data", "on_forest_statistics_1.xlsx"),
                      sheet = "Data")

tibble(
  Property = c("File", "Source", "Used for", "Shape", "Rows × Cols"),
  `BC silviculture` = c(
    "`bc_disturbance_reforestation.xlsx`",
    "BC Government / BC Data Catalogue",
    "Time-series charts (line, area)",
    "Time series (one row per fiscal year)",
    paste(nrow(bc), "×", ncol(bc))
  ),
  `Ontario forest stats` = c(
    "`on_forest_statistics_1.xlsx`",
    "Government of Ontario / Ontario Open Data Catalogue",
    "PivotTables, cross-classification charts",
    "Cross-classification (region × ownership × seral × type × age)",
    paste(nrow(ontario), "×", ncol(ontario))
  )
) |> kable()
```

::: {.callout-important}
## File-handling rules carry over

Same two rules as Chapters 1 and 2:

1. **Keep the raw downloaded files unchanged** in your `data/`
   folder.
2. **Save any cleaned or chart-ready version** under
   `_clean.xlsx` in your `outputs/` folder.

You will work mostly on the *clean* copies for this chapter.
:::

# Choosing the right chart type

Before opening the chart dialog, answer one question: **what is the
chart for?**

| If your question is... | The natural chart is... |
|---|---|
| How has this number changed over time? | **Line chart** (or area chart) |
| How do these categories compare? | **Bar chart** (horizontal) or **column chart** (vertical) |
| How does Y relate to X across observations? | **Scatter plot** |
| What share does each part take of a whole? | **Stacked bar** (or pie, sparingly) |
| Where are the extremes / outliers? | **Bar chart** sorted by value |
| How does one group differ from another over time? | **Multi-series line chart** |

::: {.callout-warning}
## When **not** to use a pie chart

Pie charts are easy to draw and hard to read. The human eye is bad
at comparing angles. Use a sorted bar chart instead, except for:

- 2–3 categories where you specifically want to communicate
  *share of whole*,
- A clear majority slice (e.g., "97 % vs 3 %").

In every other case, a **sorted bar chart** is the better tool.
:::

## Worked example — choosing a chart for BC silviculture

Suppose you want to answer: *"How has BC reforestation kept pace
with disturbance since 1987?"*

The question is **about change over time**. So: line chart, with
two series:

- `Total_Disturbance_ha` over `Fiscal_Year`
- `Reforestation_ha` over `Fiscal_Year`

```{r bc-line-chart}
#| echo: false
#| fig.width: 7
#| fig.height: 3.5
#| fig-alt: "Line chart previewing an Excel time-series chart: BC total-disturbance and reforestation area in hectares across fiscal years, one coloured line per series."
bc |>
  pivot_longer(c(Total_Disturbance_ha, Reforestation_ha),
               names_to = "series", values_to = "ha") |>
  ggplot(aes(Fiscal_Year, ha, colour = series)) +
  geom_line(linewidth = 0.9) +
  scale_y_continuous(labels = comma) +
  scale_colour_manual(values = c("Reforestation_ha" = "#2D6A4F",
                                 "Total_Disturbance_ha" = "#B7791F"),
                      labels = c("Reforestation", "Total disturbance")) +
  labs(title = "BC reforestation has matched total disturbance since the late 1990s",
       subtitle = "Fiscal years 1987–2023, hectares",
       x = "Fiscal year", y = "Hectares", colour = NULL,
       caption = "Source: Environmental Reporting BC, Trends in Silviculture in B.C.") +
  theme_minimal(base_size = 11) +
  theme(legend.position = "top",
        plot.caption = element_text(hjust = 0, colour = "grey40", size = 8))
```

In Excel:

1. Select the three columns: `Fiscal_Year`, `Total_Disturbance_ha`,
   `Reforestation_ha`.
2. **Insert → Charts → Line chart**.
3. Add chart elements: title, axis titles, data labels (sparingly),
   caption / source line.

The 2017 spike — BC's record wildfire season — is the most visible
feature. The chart itself is the argument.

# Chart anatomy

A finished chart has six elements. Every one is editable; many are
hidden by default.

| Element | Why it matters | How to edit |
|---|---|---|
| **Title** | Tells the reader what the chart is *for*. Should be a one-line sentence. | Click chart → Chart Elements (+) → Chart Title |
| **Axis titles** | Tell the reader what each axis is measuring (and in what units). | Chart Elements → Axis Titles |
| **Legend** | Identifies each series. Skip if there is only one series. | Chart Elements → Legend |
| **Data labels** | Numbers on the data points. Use sparingly. | Chart Elements → Data Labels |
| **Source / caption** | One line at the bottom naming the data source and date. | Insert → Text Box, anchor below chart |
| **Theme** | Colour and font choices. Should match the rest of the deliverable. | Chart Design ribbon → Chart Styles |

## Title rules

A bad title is *"Chart 1"* or *"Reforestation_ha vs Fiscal_Year"*.

A good title is *"BC reforestation has matched total disturbance
since the late 1990s"* — a one-sentence claim the chart supports.

If you cannot write a one-sentence claim for the chart, the chart
is the wrong chart for your question.

## Axis labels and units

Always include units in axis labels: *"Hectares"*, not just
*"Area"*. *"Fiscal year"*, not just *"Year"*.

For very large numbers, use a thousands separator and consider
abbreviating the axis (200K instead of 200,000). Excel does this
through **Format Axis → Display units → Thousands**.

# PivotTables

A PivotTable is Excel's most underused feature for forestry data.
It takes a **tidy** table — one row per observation, one column per
variable — and rearranges it into a summary.

The Ontario file is tidy and cross-classified, which makes it
perfect for a PivotTable.

## What a PivotTable does

Given the 179-row Ontario file, a PivotTable can produce any of:

- A table of *total area* by forest type (one number per row).
- A table of *total area* by forest type × seral stage (rows × columns).
- A table of *average area* by forest type × seral stage.
- A *count* of how many age-class combinations exist for each forest
  type.

All without writing a single formula.

```{r pivot-preview}
#| echo: false
ontario |>
  group_by(PFTName, SERAL_NAME) |>
  summarise(total_ha = sum(SumOfTotalHa), .groups = "drop") |>
  pivot_wider(names_from = SERAL_NAME, values_from = total_ha) |>
  mutate(across(where(is.numeric), \(x) format(round(x), big.mark = ","))) |>
  kable(caption = "Total area (ha) by forest type × seral stage — what your PivotTable should look like")
```

## Building one — step by step

1. Click any cell inside the Ontario `Data` sheet.
2. **Insert → PivotTable**. A dialog opens.
3. Choose **New Worksheet** as the destination. Click OK.
4. A blank PivotTable appears on a new sheet, with a **PivotTable
   Fields** panel on the right.
5. Drag fields into the four zones:
   - **Rows**: `PFTName`
   - **Columns**: `SERAL_NAME`
   - **Values**: `SumOfTotalHa`
   - **Filters**: (leave empty for now)
6. Excel automatically picks `Sum` as the aggregation. Confirm.
7. Format the numbers: right-click any value → **Number Format → Number → 1000 separator → 0 decimals**.

You should now see the cross-tab shown above.

## Changing the aggregation

Right-click any value in the PivotTable → **Value Field Settings**.
A list of aggregations appears:

| Aggregation | What it returns | When to use |
|---|---|---|
| Sum | Total of the values | Default for most quantities. |
| Count | How many rows match | "How many combinations are there?" |
| Average | Mean of the values | "What is the typical row in this group?" |
| Max | Largest value | "Which group has the biggest single cell?" |
| StdDev | Variability around the mean | "How spread out are the values?" |

**The same PivotTable can answer multiple questions** by switching
the aggregation. Try it.

## Drag-and-drop pivot

The most useful PivotTable habit is **moving fields around**:

- Drag `PFTName` from Rows to Columns. The whole table flips.
- Drag `AC_10` into Rows below `PFTName`. You get nested groups.
- Drag `SERAL_NAME` into Filters. A dropdown appears above the
  table — choose one seral stage to filter the whole pivot.

This is the *fluency* PivotTables train. Drag, observe, drag back.

## Refreshing

A PivotTable is a **snapshot** of the data at the moment you built
it. If the underlying Data sheet changes, the PivotTable does
*not* automatically update.

To refresh: click anywhere in the PivotTable, then **PivotTable
Analyze → Refresh** (or right-click → **Refresh**).

::: {.callout-warning}
## A silent-bug warning

If you change the Data sheet and forget to refresh the PivotTable,
the chart you export will be based on **stale data**. There is no
warning. Always refresh before exporting.
:::

# PivotCharts

A PivotChart is a chart **wired** to a PivotTable. When you change
the PivotTable, the chart changes with it.

To create one:

1. Click anywhere inside an existing PivotTable.
2. **PivotTable Analyze → PivotChart**.
3. Choose a chart type. For a cross-classification, a **clustered
   column** or **stacked column** chart usually works best.

The chart appears with field buttons on it. Those buttons are
interactive — clicking them filters the chart (and the PivotTable
behind it).

```{r pivot-chart-preview}
#| echo: false
#| fig.width: 7
#| fig.height: 3.5
#| fig-alt: "Bar chart previewing an Excel PivotChart: total area in hectares by Ontario forest type (PFTName), sorted from largest to smallest."
ontario |>
  group_by(PFTName) |>
  summarise(total_ha = sum(SumOfTotalHa), .groups = "drop") |>
  arrange(desc(total_ha)) |>
  mutate(PFTName = fct_inorder(PFTName)) |>
  ggplot(aes(PFTName, total_ha)) +
  geom_col(fill = "#2D6A4F") +
  scale_y_continuous(labels = comma) +
  labs(title = "Total area by forest type, Region 3E",
       subtitle = "Ontario Forest Resources, 2021 snapshot",
       x = NULL, y = "Hectares",
       caption = "Source: Government of Ontario, Forest Resources of Ontario 2021") +
  theme_minimal(base_size = 11) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1),
        plot.caption = element_text(hjust = 0, colour = "grey40", size = 8))
```

## Slicers

A **Slicer** is a visual filter panel. Click any PivotTable →
**PivotTable Analyze → Insert Slicer**. Check `PFTName`. Click OK.

A panel of buttons appears, one per forest type. Click "Jack Pine"
— the PivotTable (and the PivotChart) filter to Jack Pine only.
`Ctrl+Click` to select multiple.

Slicers are excellent for **presentations** because they let a
non-technical reader explore the data without learning Excel.

# Recreating a PivotTable with formulas

PivotTables are fast and visual, but they are not the only way to
build a cross-tab. You can recreate one with `SUMIFS` and
`COUNTIFS` — the formula approach.

This matters because **the formula version is the closest thing to
what you will write in R**.

## Example: SUMIFS cross-tab on the Ontario file

Build a small table on a new sheet:

|   | Pre-Sapling | Sapling | Immature | Mature | Late Successional |
|---|---|---|---|---|---|
| Jack Pine          | `=SUMIFS(...)` | `=SUMIFS(...)` | ... | ... | ... |
| Conifer Upland     | ... | ... | ... | ... | ... |
| Conifer Lowland    | ... | ... | ... | ... | ... |
| Mixedwood          | ... | ... | ... | ... | ... |
| Poplar             | ... | ... | ... | ... | ... |
| White Birch        | ... | ... | ... | ... | ... |
| Red and White Pine | ... | ... | ... | ... | ... |
| Tolerant Hardwoods | ... | ... | ... | ... | ... |

Each cell is a `SUMIFS` formula like:

```text
=SUMIFS(Data!$H:$H, Data!$F:$F, $A2, Data!$D:$D, B$1)
```

- `Data!$H:$H` is `SumOfTotalHa`.
- `Data!$F:$F` is `PFTName`.
- `Data!$D:$D` is `SERAL_NAME`.
- `$A2` is the row label (a forest type).
- `B$1` is the column label (a seral stage).

Notice the **mixed references**: `$A2` (column locked, row moves)
and `B$1` (row locked, column moves). This is the cell-reference
pattern you learned in Chapter 2 — it lets one formula copied
across the whole grid produce the right answer in every cell.

The result is the same numbers as the PivotTable, but the
formulas are explicit, auditable, and version-controllable.

::: {.callout-tip}
## When to use which

- For **exploration** — drag-and-drop PivotTable. Fast.
- For **production reports** that need to be reproducible — formula
  cross-tab with `SUMIFS`. Auditable.
- For **scripted analysis** — neither, use R (Chapter 7).
:::

# Exporting a chart with a caption

A chart in Excel is not a deliverable. A chart **outside** Excel —
saved as a PNG with a caption — is a deliverable.

## To save as PNG

Right-click the chart border → **Save as Picture**. Choose PNG.
Save into your `outputs/` folder.

## Caption format

A caption is one short sentence that:

- Tells the reader what the chart shows.
- Names the data source.
- Notes the date or version.

Example:

> **Figure 1.** BC reforestation has tracked total disturbance
> since 1999, with the exception of the 2017 wildfire spike.
> Source: Environmental Reporting BC, *Trends in Silviculture in
> B.C.*, retrieved 2026-05-24.

Put the caption in the report document, not in the PNG itself.

# Connecting charts to the data dictionary

Every chart should be **traceable** back to the ReadMe sheet of the
file it came from. Three habits make this easy:

1. **Name the chart** after the variable it shows. *"Total
   reforestation by year"* is better than *"Chart 3"*.
2. **Save the PNG** with a snake_case name that includes the
   variable. `reforestation_by_year.png` is better than
   `chart3.png`.
3. **In the caption**, reference the variable by its exact ReadMe
   name. *"`Reforestation_ha`"* is better than *"reforestation"*.

This sounds fussy but pays off the first time someone asks *"which
column is this from?"*.

# Active learning

## Activity 1 — Choose the right chart

A list of forestry questions appears in the learner workbook.
Match each one to its best chart type from this menu: line, bar,
column, scatter, stacked column, pie. Defend your choices in pairs.

## Activity 2 — Build the BC time-series chart

Open `bc_disturbance_reforestation_clean.xlsx` (your working copy).
Select `Fiscal_Year`, `Total_Disturbance_ha`, and `Reforestation_ha`.
**Insert → Charts → Line**. Add a title, axis labels, and a
caption. Save as `outputs/bc_disturbance_reforestation_chart.png`.

## Activity 3 — Build a PivotTable

Open `on_forest_statistics_1_clean.xlsx`. Insert a PivotTable with
`PFTName` in Rows, `SERAL_NAME` in Columns, `SumOfTotalHa` in
Values. Verify the totals match the table in the chapter prose
above.

```{r act3-verify}
#| echo: false
ontario |>
  group_by(PFTName) |>
  summarise(total_ha = sum(SumOfTotalHa), .groups = "drop") |>
  arrange(desc(total_ha)) |>
  mutate(total_ha = comma(round(total_ha))) |>
  kable(caption = "Total area by forest type — should match the row totals of your PivotTable")
```

## Activity 4 — Switch the aggregation

Right-click any value in your PivotTable → **Value Field Settings →
Count**. The cell values change from total hectares to *number of
age-class combinations* for that group. How does the picture
change?

```{r act4-verify}
#| echo: false
ontario |>
  count(PFTName) |>
  arrange(desc(n)) |>
  kable(caption = "Number of age-class combinations per forest type")
```

## Activity 5 — Add a slicer

To your PivotTable, add a Slicer on `LGDS` (the seral-stage code).
Click each button in turn. Which forest types disappear at the
"Pre-Sapling" stage? Which dominate at "Mature"?

# Group lab activity

::: {.callout-note appearance="default"}
## Lab: Build a chart-and-PivotTable deliverable together

Work in groups of 3 or 4. One submission per group.

**Task 1 — Choose the question.** As a group, choose one forestry
question your PivotTable will answer. Options:

- *"Which forest type has the most mature area in Region 3E?"*
- *"Which age class has the most total area, across all types?"*
- *"How does conifer-vs-hardwood area compare across seral stages?"*

**Task 2 — Build the PivotTable.** Open
`on_forest_statistics_1_clean.xlsx`. Insert a PivotTable that
answers your question. Decide together what goes in Rows, Columns,
and Values.

**Task 3 — Build the PivotChart.** From your PivotTable, insert a
PivotChart of an appropriate type. Add a title, axis labels, and
ensure units (hectares) appear in the y-axis label.

**Task 4 — Add a slicer.** Insert at least one slicer (any
categorical field). Confirm that clicking a slicer button changes
both the PivotTable and the PivotChart.

**Task 5 — Export and caption.** Save the PivotChart as a PNG into
`outputs/`. Write a one-sentence caption that includes the data
source.

**Task 6 — Submit.** Submit a short worksheet showing your
question, your field choices, and the exported PNG + caption.
Include all group members' names.

This lab is designed to fit comfortably inside a single hands-on
session.
:::

# Exercises

These are the take-home exercises that go with this chapter.

1. Open `bc_disturbance_reforestation.xlsx`. Build a **column
   chart** of `Natural_Disturbance_ha` by `Fiscal_Year`. Add a
   title that names the 2017 wildfire spike. Export as PNG.

2. Open `on_forest_statistics_1.xlsx`. Build a PivotTable with
   `SERAL_NAME` in Rows and `SumOfTotalHa` in Values (using `Sum`).
   Sort the table by descending total area. Which seral stage
   covers the most area?

3. On the same PivotTable, change the aggregation from `Sum` to
   `Count`. Which seral stage has the most age-class
   combinations? Is it the same answer as Exercise 2?

4. Build a **manual SUMIFS cross-tab** with `PFTName` down the
   left and `SERAL_NAME` across the top. Verify your numbers
   match the PivotTable.

5. Add a **slicer** on `PFT` to your PivotTable. Take a
   screenshot showing the table when one forest-type slicer is
   active.

6. Write a one-page chart-caption sheet listing the captions you
   would use for the three exported PNG files (Exercises 1, 2,
   and the group lab output). Save as
   `outputs/chart_captions.docx` or `chart_captions.md`.

7. **Optional, harder.** Build a **dashboard sheet** containing
   your time-series chart from Exercise 1 and your PivotChart
   from Exercise 2 side by side, with a slicer that controls
   both. (Hint: a slicer can connect to multiple PivotTables
   built from the same data source.)

# Optional, advanced

## Sparklines

A sparkline is a tiny chart inside a single cell. Useful for
trend-at-a-glance summaries inside a table.

To insert: **Insert → Sparklines → Line** (or Column, or Win/Loss).
Select the data range, then choose the cell where the sparkline
goes.

## Conditional-formatted data bars

Inside a PivotTable, you can add **Data Bars** to a value column.
Select the values → **Home → Conditional Formatting → Data Bars**.
The cells now show a tiny in-cell horizontal bar proportional to
the value — a quick way to add visual rank to a table.

## Calculated fields

A **Calculated Field** is a column in a PivotTable derived from
other PivotTable fields. To add one: **PivotTable Analyze →
Fields, Items, & Sets → Calculated Field**.

Example: in the Ontario file you could add a `area_per_age_class`
calculated field equal to `SumOfTotalHa / AC_10`. Use sparingly —
calculated fields are powerful but easy to hide a typo inside.

# A glimpse of Chapter 9: From Excel to `ggplot2`

In Chapter 9 you will rebuild the BC and Ontario charts from this
chapter using `ggplot2` in R. The Excel moves you learned here map
to R commands you will meet later:

| What you did in Excel (Ch 3) | What you will do in R (Ch 9) |
|---|---|
| **Insert → Line chart** | `ggplot(bc, aes(Fiscal_Year, Reforestation_ha)) + geom_line()` |
| **Insert → Column chart** | `ggplot(...) + geom_col()` |
| **Chart title** in the chart area | `labs(title = "...")` |
| **Axis title** "Hectares" | `labs(y = "Hectares")` |
| **Add caption** in a text box | `labs(caption = "Source: ...")` |
| **PivotTable** with `PFTName` in Rows | `ontario \|> group_by(PFTName) \|> summarise(sum(SumOfTotalHa))` |
| **PivotTable** with `PFTName` × `SERAL_NAME` | `... \|> group_by(PFTName, SERAL_NAME) \|> summarise(...) \|> pivot_wider(...)` |
| **Slicer** on `PFTName` | `... \|> filter(PFTName == "Jack Pine")` |
| **Refresh PivotTable** (manual) | Automatic — re-running the code re-computes everything |
| **PNG export** via Save as Picture | `ggsave("outputs/bc_reforestation.png", width = 7, height = 4)` |

**The intent is identical.** R is scriptable; Excel is visual.
Chapter 9 will walk through each row of this table in detail. The
point right now is that the chart vocabulary you learn here
transfers directly.

# Self-assessment quiz {#sec-quiz-ch03}

Click the answer you think is correct. The right answer turns
green; wrong answers turn red and the correct one is revealed.
Explanations appear below each question.

```{=html}
<style>
.quiz {
  background: #F4F7F4;
  border-left: 4px solid #2D6A4F;
  padding: 1em 1.5em;
  margin: 1.5em 0;
  border-radius: 6px;
}
.quiz .q { font-weight: 600; color: #1B4332; margin-bottom: 0.6em; }
.quiz ul.qopts { list-style: none; padding: 0; margin: 0; }
.quiz ul.qopts li {
  padding: 0.55em 0.9em; margin: 0.3em 0; background: white;
  border: 1px solid #d8e4d8; border-radius: 4px; cursor: pointer;
  transition: background 0.12s;
}
.quiz ul.qopts li:hover { background: #E8F1E9; }
.quiz ul.qopts li.correct { background: #C8E6C9; border-color: #2D6A4F; font-weight: 600; }
.quiz ul.qopts li.wrong { background: #FFCDD2; border-color: #B05C50; }
.quiz .explanation {
  display: none; margin-top: 0.7em; padding: 0.7em 1em;
  background: #FFF8E8; border-left: 3px solid #B7791F;
  border-radius: 4px; font-size: 0.95em;
}
.quiz .explanation.show { display: block; }
.quiz textarea {
  width: 100%; min-height: 4.5em; padding: 0.6em;
  border: 1px solid #d8e4d8; border-radius: 4px;
  font-family: inherit; font-size: 0.95em; resize: vertical; margin-top: 0.3em;
}
.quiz .reveal-btn {
  background: #2D6A4F; color: white; border: none;
  padding: 0.45em 0.9em; margin-top: 0.5em; border-radius: 4px;
  cursor: pointer; font-size: 0.9em;
}
.quiz .reveal-btn:hover { background: #1B4332; }
.quiz .sample {
  display: none; margin-top: 0.6em; padding: 0.6em 0.9em;
  background: #FFF8E8; border-left: 3px solid #B7791F;
  border-radius: 4px; font-size: 0.95em; font-style: italic;
}
.quiz .sample.show { display: block; }
</style>

<div class="quiz" id="q1">
  <div class="q">1. You want to show how BC reforestation has changed from 1987 to 2023. Which chart type is best?</div>
  <ul class="qopts">
    <li data-correct="false">Pie chart</li>
    <li data-correct="true">Line chart</li>
    <li data-correct="false">Scatter plot</li>
    <li data-correct="false">Stacked bar</li>
  </ul>
  <div class="explanation">Time series with a single quantity over time → line chart. Pie shows share of a whole; scatter shows X vs Y relationships; stacked bar shows composition.</div>
</div>

<div class="quiz" id="q2">
  <div class="q">2. A PivotTable is built from a tidy table. What four field zones does it use?</div>
  <ul class="qopts">
    <li data-correct="false">Headers, Values, Footer, Title</li>
    <li data-correct="true">Rows, Columns, Values, Filters</li>
    <li data-correct="false">Sum, Count, Average, Max</li>
    <li data-correct="false">X-axis, Y-axis, Legend, Caption</li>
  </ul>
  <div class="explanation">Rows go down the left, Columns go across the top, Values are the numeric quantities being aggregated, Filters apply to the whole pivot.</div>
</div>

<div class="quiz" id="q3">
  <div class="q">3. You change the underlying Data sheet after building a PivotTable. What must you do for the PivotTable to show the new numbers?</div>
  <ul class="qopts">
    <li data-correct="false">Nothing — the PivotTable updates automatically</li>
    <li data-correct="true">Right-click the PivotTable → Refresh (or PivotTable Analyze → Refresh)</li>
    <li data-correct="false">Rebuild the PivotTable from scratch</li>
    <li data-correct="false">Delete the Data sheet</li>
  </ul>
  <div class="explanation">PivotTables are <em>snapshots</em>, not live views. Forgetting to refresh after a data change is the most common silent bug in PivotTable-based reporting.</div>
</div>

<div class="quiz" id="q4">
  <div class="q">4. A PivotChart is —</div>
  <ul class="qopts">
    <li data-correct="false">A type of chart that does not need data</li>
    <li data-correct="true">A chart wired to a PivotTable, which updates when the PivotTable changes</li>
    <li data-correct="false">A chart that automatically becomes a pie chart</li>
    <li data-correct="false">A chart with a slicer attached</li>
  </ul>
  <div class="explanation">PivotChart and its PivotTable are linked — change one and the other follows. This is why pivot-based charts are easier to keep current than hand-built charts.</div>
</div>

<div class="quiz" id="q5">
  <div class="q">5. Which chart title is better?</div>
  <ul class="qopts">
    <li data-correct="false">"Chart 1"</li>
    <li data-correct="false">"Reforestation_ha vs Fiscal_Year"</li>
    <li data-correct="true">"BC reforestation has tracked total disturbance since the late 1990s"</li>
    <li data-correct="false">"My chart"</li>
  </ul>
  <div class="explanation">A good title is a one-sentence claim the chart supports. If you cannot write such a claim, the chart is probably the wrong chart for your question.</div>
</div>

<div class="quiz" id="q6">
  <div class="q">6. In a PivotTable summarizing the Ontario file, you put <code>PFTName</code> in Rows, <code>SERAL_NAME</code> in Columns, and <code>SumOfTotalHa</code> in Values with aggregation <code>Sum</code>. One cell shows 178,000. That number is —</div>
  <ul class="qopts">
    <li data-correct="false">The total area of all forest types</li>
    <li data-correct="false">The number of rows with that forest type and seral stage</li>
    <li data-correct="true">The total hectares for one (forest type, seral stage) combination</li>
    <li data-correct="false">The average area for that forest type</li>
  </ul>
  <div class="explanation">Each interior cell of the PivotTable is the sum of <code>SumOfTotalHa</code> for the rows that match both the row label and the column label.</div>
</div>

<div class="quiz" id="q7">
  <div class="q">7. A Slicer is best understood as —</div>
  <ul class="qopts">
    <li data-correct="false">A new chart type</li>
    <li data-correct="true">A clickable visual filter panel that controls one or more PivotTables / PivotCharts</li>
    <li data-correct="false">A way to delete rows</li>
    <li data-correct="false">A function in the formula bar</li>
  </ul>
  <div class="explanation">Slicers are visual filters. They are excellent for presentations because non-technical readers can explore the data by clicking buttons.</div>
</div>

<div class="quiz" id="q8">
  <div class="q">8. Which Excel formula recreates a PivotTable cell that sums <code>SumOfTotalHa</code> for "Jack Pine" × "Mature" combinations?</div>
  <ul class="qopts">
    <li data-correct="false">=SUM(SumOfTotalHa)</li>
    <li data-correct="false">=SUMIF(PFTName, "Jack Pine", SumOfTotalHa)</li>
    <li data-correct="true">=SUMIFS(SumOfTotalHa, PFTName, "Jack Pine", SERAL_NAME, "Mature")</li>
    <li data-correct="false">=COUNTIFS(PFTName, "Jack Pine", SERAL_NAME, "Mature")</li>
  </ul>
  <div class="explanation"><code>SUMIFS</code> sums one range subject to multiple criteria — exactly what a PivotTable cell does. <code>COUNTIFS</code> would return a row count, not a sum.</div>
</div>

<div class="quiz" id="q9">
  <div class="q">9. Why is a pie chart usually a bad choice for comparing forest type areas across 8 forest types?</div>
  <ul class="qopts">
    <li data-correct="false">Excel cannot draw pie charts</li>
    <li data-correct="true">Humans are bad at comparing angles; a sorted bar chart is easier to read</li>
    <li data-correct="false">Pie charts are not allowed in forestry</li>
    <li data-correct="false">Pie charts always use the wrong colors</li>
  </ul>
  <div class="explanation">The eye compares lengths much better than angles. For 8 categories, a sorted bar chart lets the reader rank values at a glance.</div>
</div>

<div class="quiz" id="q10">
  <div class="q">10. Where does the Ontario forest statistics file come from?</div>
  <ul class="qopts">
    <li data-correct="false">A textbook</li>
    <li data-correct="true">The Government of Ontario, distributed through the Ontario Open Data Catalogue</li>
    <li data-correct="false">A research blog</li>
    <li data-correct="false">A private consulting firm</li>
  </ul>
  <div class="explanation">Both datasets in this chapter are open government data. The Ontario file is from the Ministry of Natural Resources and Forestry, published under the Open Government Licence — Ontario.</div>
</div>

<div class="quiz" id="q11">
  <div class="q">11. You build a column chart of <code>Natural_Disturbance_ha</code> by <code>Fiscal_Year</code>. One year's column is much taller than every other. Which year?</div>
  <ul class="qopts">
    <li data-correct="false">1987</li>
    <li data-correct="false">2010</li>
    <li data-correct="true">2017</li>
    <li data-correct="false">2023</li>
  </ul>
  <div class="explanation">2017 was BC's record wildfire season — 218,275 ha. Always look at the chart extremes; the outlier tells you something about the world.</div>
</div>

<div class="quiz" id="q12">
  <div class="q">12. When exporting a PivotChart for a report, which file format is best?</div>
  <ul class="qopts">
    <li data-correct="false">XLSX</li>
    <li data-correct="true">PNG (right-click chart → Save as Picture)</li>
    <li data-correct="false">CSV</li>
    <li data-correct="false">TXT</li>
  </ul>
  <div class="explanation">PNG (or SVG) is a stand-alone image that any reader can open without Excel. CSV and TXT are data formats, not image formats.</div>
</div>

<div class="quiz" id="q13">
  <div class="q">13. A chart caption should always include —</div>
  <ul class="qopts">
    <li data-correct="false">The Excel version you used</li>
    <li data-correct="true">A one-sentence description and the data source</li>
    <li data-correct="false">Your full name</li>
    <li data-correct="false">A list of every formula in the workbook</li>
  </ul>
  <div class="explanation">A good caption tells the reader what the chart shows and where the data came from. Date or version is also useful.</div>
</div>

<div class="quiz" id="q14">
  <div class="q">14. What does the <em>mixed reference</em> <code>$A2</code> mean in a formula copied across a grid?</div>
  <ul class="qopts">
    <li data-correct="false">Both the row and column are locked</li>
    <li data-correct="true">The column is locked; the row moves as the formula is copied</li>
    <li data-correct="false">The row is locked; the column moves</li>
    <li data-correct="false">Nothing is locked</li>
  </ul>
  <div class="explanation">The <code>$</code> before <code>A</code> locks the column. When you copy the formula down, <code>$A2</code> becomes <code>$A3</code>, <code>$A4</code>, etc. When you copy it right, it stays <code>$A2</code>.</div>
</div>

<div class="quiz" id="q15">
  <div class="q">15. In a PivotTable, switching the Value aggregation from <code>Sum</code> to <code>Count</code> changes —</div>
  <ul class="qopts">
    <li data-correct="false">The colors of the cells</li>
    <li data-correct="false">The row and column labels</li>
    <li data-correct="true">The interior values, from "total hectares" to "number of rows in that group"</li>
    <li data-correct="false">The Data sheet</li>
  </ul>
  <div class="explanation">The aggregation function controls what the interior cells display. <code>Sum</code> adds the values; <code>Count</code> counts how many rows fall in the group.</div>
</div>

<div class="quiz" id="q16">
  <div class="q">16. You add a Slicer for <code>PFTName</code> to your PivotTable. You click "Jack Pine". What happens to the PivotChart wired to the same PivotTable?</div>
  <ul class="qopts">
    <li data-correct="false">Nothing — slicers do not affect charts</li>
    <li data-correct="true">The chart filters to show only Jack Pine data, in sync with the PivotTable</li>
    <li data-correct="false">The chart shows an error</li>
    <li data-correct="false">The slicer creates a new chart</li>
  </ul>
  <div class="explanation">Slicers, PivotTables, and PivotCharts are linked. A slicer filters the PivotTable, and the chart follows automatically.</div>
</div>

<div class="quiz" id="q17">
  <div class="q">17. What is the closest R equivalent of an Excel PivotTable cell built with <code>Sum</code>?</div>
  <ul class="qopts">
    <li data-correct="false">read_excel(...)</li>
    <li data-correct="true">group_by(...) |> summarise(sum(...))</li>
    <li data-correct="false">ggplot(...) + geom_line()</li>
    <li data-correct="false">filter(...)</li>
  </ul>
  <div class="explanation">A PivotTable aggregating by a category is conceptually <code>group_by()</code> followed by <code>summarise()</code>. This is the bridge to Chapter 7.</div>
</div>

<div class="quiz" id="q18">
  <div class="q">18. The PNG file you submit for the chapter assignment should live in which folder of your project?</div>
  <ul class="qopts">
    <li data-correct="false">data/</li>
    <li data-correct="true">outputs/ (or deliverables/, if the file is the final submission)</li>
    <li data-correct="false">shared/</li>
    <li data-correct="false">It does not matter</li>
  </ul>
  <div class="explanation"><code>data/</code> is for raw input files only. Generated files like PNG charts go in <code>outputs/</code>, and the version you actually submit goes in <code>deliverables/</code>.</div>
</div>

<div class="quiz" id="q19">
  <div class="q">19. A reader asks "which column did your chart come from?" You should answer by —</div>
  <ul class="qopts">
    <li data-correct="false">Guessing from the chart title</li>
    <li data-correct="true">Pointing to the ReadMe sheet entry for the exact variable named in the caption</li>
    <li data-correct="false">Recreating the chart from memory</li>
    <li data-correct="false">Saying it does not matter</li>
  </ul>
  <div class="explanation">This is why captions reference variables by their exact ReadMe names. Traceability from chart → caption → ReadMe → Data column is what makes the work auditable.</div>
</div>

<div class="quiz" id="q20">
  <div class="q">20. <strong>Reflection.</strong> Describe one chart or PivotTable you built for this chapter that surprised you — anything you noticed in the data that you did not expect.</div>
  <textarea placeholder="Type your answer here..."></textarea>
  <button class="reveal-btn" onclick="this.nextElementSibling.classList.toggle('show')">Show common answers</button>
  <div class="sample">Common observations: <em>the 2017 wildfire spike is hard to even see on a linear axis — it is so much bigger than anything else that other years look flat</em> · <em>mature Conifer Lowland dominates the Ontario file in terms of area</em> · <em>most forest types have very few rows in the Pre-Sapling stage, which makes sense for old commercial landbases</em> · <em>switching from Sum to Count completely changes which forest type "leads" — total area and number of combinations tell different stories</em>.</div>
</div>

<script>
(function() {
  document.querySelectorAll('.quiz').forEach(function(q) {
    q.querySelectorAll('ul.qopts li').forEach(function(opt) {
      opt.addEventListener('click', function() {
        if (q.dataset.answered) return;
        q.dataset.answered = '1';
        var isCorrect = opt.getAttribute('data-correct') === 'true';
        if (isCorrect) {
          opt.classList.add('correct');
        } else {
          opt.classList.add('wrong');
          q.querySelectorAll('ul.qopts li').forEach(function(o) {
            if (o.getAttribute('data-correct') === 'true') {
              o.classList.add('correct');
            }
          });
        }
        var exp = q.querySelector('.explanation');
        if (exp) exp.classList.add('show');
      });
    });
  });
})();
</script>
```

# AI as a debugging companion

::: {.callout-tip}
## Useful prompts for this chapter

- *"In Excel, my PivotTable shows only zeros for one row. The
  underlying Data sheet has values. What are the three most common
  reasons a PivotTable shows zeros?"*

- *"Explain step by step how to add a slicer to an existing
  PivotTable and link it to a second PivotTable on the same sheet."*

- *"I want to recreate this PivotTable cell with a formula — sum
  of `SumOfTotalHa` for `PFTName = 'Jack Pine'` and `SERAL_NAME =
  'Mature'`. What is the SUMIFS formula?"*

## Verifying AI's answers

After AI suggests a chart type or PivotTable layout, **check**:

1. Does the chart actually answer the question you asked?
2. Does the title state a one-sentence claim?
3. Do the units appear in the axis label?
4. Did the AI refresh the PivotTable in its instructions? (It
   often forgets.)

AI's chart suggestions are starting points. The interpretation is
yours.

## When *not* to use AI

- Do not let AI choose the chart type without you stating the
  question. The chart serves the question; the question serves
  you.
- Do not let AI invent a caption that names a source you have not
  verified.
:::

# Reading

- Microsoft Excel official documentation on [PivotTables](https://support.microsoft.com/en-us/office/create-a-pivottable-to-analyze-worksheet-data-a9a84538-bfe9-40a9-a8e9-f99134456576).
- Microsoft Excel official documentation on [PivotCharts](https://support.microsoft.com/en-us/office/create-a-pivotchart-c1b1e057-6990-4c38-b52b-8255538e7b1c).
- Cole Nussbaumer Knaflic, *Storytelling with Data* (chapter on
  chart choice — short, opinionated, practical).
- BC Government silviculture indicator background:
  <https://www.env.gov.bc.ca/soe/indicators/land/silviculture.html>
- Ontario open-data dataset page:
  <https://data.ontario.ca/dataset/forest-resources-of-ontario-2021>

# Instructor notes {.unnumbered}

::: {.callout-note appearance="minimal" collapse="true"}
## Pacing and emphasis

- This chapter is the **bridge chapter** from raw Excel work into
  tidy-data thinking. Spend time on the PivotTable → R parallel
  in the "A glimpse of Chapter 9" section — it is the most
  important conceptual move of the first half of the book.
- The main path covers six skills: chart choice, chart anatomy,
  PivotTable construction, aggregation switching, PivotChart, and
  Slicers. The SUMIFS cross-tab section is essential for the
  Excel-to-R bridge.
- **Sparklines**, **data bars in pivots**, and **calculated
  fields** are moved to the *Optional, advanced* section. Mention
  them only if your group is ahead of pace.
- The group lab is the central hands-on deliverable. Plan for one
  full session.

## Expected output checklist (matches the top-of-chapter callout)

1. One PivotTable + PivotChart on `on_forest_statistics_1.xlsx`,
   exported as PNG with caption.
2. One time-series chart on
   `bc_disturbance_reforestation.xlsx`, exported as PNG with
   caption.
3. A short chart-choice worksheet (from the learner workbook).
4. Group-lab worksheet or screenshot.

## Materials provided alongside this chapter

| File | Purpose |
|---|---|
| `bc_disturbance_reforestation.xlsx` | Raw BC silviculture workbook (carried forward from Chapter 1). |
| `on_forest_statistics_1.xlsx` | Raw Ontario forest stats workbook (carried forward from Chapter 2). |
| `frst232_ch03_pivot_learner.xlsx` | Learner-facing workbook with both datasets and chapter exercises. |
| `frst232_ch03_pivot_solutions.xlsx` | Instructor solutions. **Do not distribute to learners.** |
| `quiz_ch03.html` | Standalone interactive quiz. Embedded in the rendered chapter; can also be hosted separately. |

## Suggested grading guidance for the group lab

| Task | Weight | What to look for |
|---|---|---|
| Question + PivotTable (Tasks 1–2) | 25 % | Question is specific and answerable; PivotTable layout matches the question. |
| PivotChart + axis labels (Task 3) | 25 % | Chart type matches the question; axis labels include units. |
| Slicer (Task 4) | 15 % | Slicer present and functional. |
| PNG export + caption (Task 5) | 25 % | PNG file submitted; one-sentence caption includes source. |
| Submission (Task 6) | 10 % | Worksheet is legible; group members named. |

## Common stumbling points

- Some learners build a PivotTable, change the Data sheet, and
  then export from the stale PivotTable. Emphasize the **Refresh
  before export** habit.
- Some forget that a chart title and an axis label are different
  things. A good chart has both.
- A few will choose a pie chart for 8-category comparisons.
  Redirect to a sorted bar chart.
- Some will compute "average" on a column already summed in a
  PivotTable, double-aggregating. Walk through what `Sum` vs
  `Average` actually does at the cell level.
- *Excel-to-R bridge:* The "A glimpse of Chapter 9" section is the
  single most important paragraph in Chapter 3 for the long arc
  of the course. Refer to it explicitly when learners ask why we
  are still in Excel.
:::
