```{r setup-excelcs}
#| include: false
library(tidyverse)
library(readxl)
library(here)
library(knitr)
```

# Guided Case Study: From Excel to R {.unnumbered}

::: {.callout-important}
## Practice demo only
This activity is a **guided practice lab in the OER book**. It is designed
to help you learn the workflow before completing the official lab
assignment on Canvas. **The Canvas lab is the graded assignment.** Use this
book activity to practice, compare your work with the reference solution,
and learn how to write an AI verification prompt.
:::

::: {.callout-note appearance="simple"}
## Data for this chapter
This chapter uses the **BC silviculture** workbook
([`bc_disturbance_reforestation.xlsx`](../data/bc_disturbance_reforestation.xlsx){download="bc_disturbance_reforestation.xlsx"}) —
the same file you used in the Excel chapters. Keep it in your `data/`
folder. For the source and details, see the
[Datasets Used in This Book](../datasets.qmd) page.
:::

# What this chapter is {.unnumbered}

This is your **first, guided case study** — an early bridge from Excel
(Chapters 1–3) to R (Chapters 4–5). You will answer one small forestry
question **twice**: once with the Excel skills you already have, and once
with the first R tools. Seeing the *same answer* on both sides is the point.

It deliberately uses **only the skills taught so far**: Excel inspection,
cleaning, formulas, PivotTables, and charting, plus early R setup, import,
inspection, basic functions, the native pipe `|>`, and Quarto. It does
**not** need joins, reshaping, grouped `summarise()`, or `ggplot2` — those
come later, and the full **independent** case study at the end of the
course (Chapter [Integrated Case Study](10-case-study.qmd)) uses them all.

# Chapter goals {.unnumbered}

By the end you will be able to:

- restate a small forestry question and decide what you need to answer it;
- answer it in **Excel** (inspect → formula / PivotTable → chart);
- answer the same question in **early R** (import → inspect → a basic
  calculation with the native pipe) and **confirm the two agree**;
- verify a result instead of trusting it, and write an AI prompt that
  *checks* your work rather than doing it for you.

# The question

> *"What was the average annual reforested area in British Columbia over
> the whole record, and which single year was the highest?"*

Nothing here needs a join or a grouped summary — just an average, a
maximum, and careful reading of the file.

# Part A — answer it in Excel

You already know these moves from Chapters 1–3:

1. **Inspect.** Open the **Data** sheet. Confirm the columns (`Fiscal_Year`,
   `Reforestation_ha`, …), the units (hectares), and that there are no
   missing values.
2. **Average.** In an empty cell, `=AVERAGE(H2:H38)` on the
   `Reforestation_ha` column.
3. **Maximum + its year.** `=MAX(H2:H38)` for the largest value, and
   `=INDEX(A2:A38, MATCH(MAX(H2:H38), H2:H38, 0))` for the fiscal year it
   happened in.
4. **Chart (optional).** A quick line chart of `Reforestation_ha` by
   `Fiscal_Year` shows the shape behind the numbers.

Write your three answers down — you will check them against R next.

# Part B — answer it in early R

The same three answers, now reproducibly. This uses only Chapter 4–5
skills: `read_excel()`, inspection, base functions, and the pipe.

```{r excelcs-import}
bc <- read_excel(here("data", "bc_disturbance_reforestation.xlsx"),
                 sheet = "Data")
dim(bc)        # rows and columns
names(bc)      # column names — confirm Reforestation_ha is here
```

```{r excelcs-answer}
# Average reforested area (same as Excel =AVERAGE)
bc$Reforestation_ha |> mean()

# The maximum, and the fiscal year it happened in
bc$Reforestation_ha |> max()
bc$Fiscal_Year[which.max(bc$Reforestation_ha)]
```

The pipe `|>` reads *"take the `Reforestation_ha` column, and then take its
mean."* No new packages, no `group_by()` — just the moves you already have.

**The bridge:** the R average should match your Excel `=AVERAGE` to the
cent, and the R maximum and its year should match your Excel `MAX` /
`INDEX`/`MATCH`. If they *don't*, something is off in one of the two — that
mismatch is a verification win, not a failure.

# Practice Demo Lab

::: {.callout-important}
## Practice demo only (not the graded Canvas lab)
Work through the six steps below. The graded version is on Canvas.
:::

**1 — Your attempt.** For a *different* column — **`Harvested_ha`** —
compute in R: (a) the average harvested area, (b) the maximum, and (c) the
fiscal year of the maximum. Then confirm each against an Excel formula.

**2 — Reference solution.**

```{r excelcs-ref}
bc$Harvested_ha |> mean()                              # (a) average
bc$Harvested_ha |> max()                               # (b) maximum
bc$Fiscal_Year[which.max(bc$Harvested_ha)]             # (c) year of max
```

*(In Excel: `=AVERAGE(E2:E38)`, `=MAX(E2:E38)`, and
`=INDEX(A2:A38, MATCH(MAX(E2:E38), E2:E38, 0))`.)*

**3 — Compare your work with the reference solution.** Where did your
numbers match? Where did they differ? What caused any difference (wrong
column? a stray blank cell in the Excel range? the wrong sheet)? How did
you fix it?

**4 — Write your own AI verification prompt.** Draft a prompt that asks an
AI to *check* your reasoning and outputs — not to hand you the answer.
Include the file name, the column, what you computed, and the checks you
want it to confirm.

**5 — Model AI verification prompt.**

> *"I am using R in FRST 232. I have a tibble `bc` from
> `bc_disturbance_reforestation.xlsx` (Data sheet), columns including
> `Fiscal_Year` and `Harvested_ha`. I computed the mean, the max, and the
> year of the max of `Harvested_ha`. Here is my code and output: `[paste]`.
> Please explain what each line does, tell me whether it correctly answers
> the question, and list checks I can run to confirm it (for example,
> comparing to an Excel `=AVERAGE`/`=MAX`, and checking for blank cells).
> Do not give me a final assignment answer — help me verify my own."*

**6 — Reflection: what changed after checking?** In two or three sentences:
did Excel and R agree the first time? If not, what was wrong? What will you
check *first* next time?

# Done

You have answered one forestry question in two tools and **verified they
agree**. That habit — compute, then confirm — is the foundation of every
later chapter, and of the full independent case study at the end of the
course.
