```{r setup-excelcs}
#| include: false
library(tidyverse)
library(readxl)
library(here)
library(knitr)
```

# Guided Case Study: From Excel to R {.unnumbered}

::: {.callout-important}
## Practice demo only
This activity is a **guided practice lab in the OER book**. It is designed
to help you learn the workflow before completing the official lab
assignment on Canvas. **The Canvas lab is the graded assignment.** Use this
book activity to practice, compare your work with the reference solution,
and learn how to write an AI verification prompt.
:::

::: {.callout-note appearance="simple"}
## Data for this chapter
This chapter uses the **BC silviculture** workbook
([`bc_disturbance_reforestation.xlsx`](../data/bc_disturbance_reforestation.xlsx){download="bc_disturbance_reforestation.xlsx"}) тАФ
the same file you used in the Excel chapters. Keep it in your `data/`
folder. For the source and details, see the
[Datasets Used in This Book](../datasets.qmd) page.
:::

# What this chapter is {.unnumbered}

This is your **first, guided case study** тАФ an early bridge from Excel
(Chapters 1тАУ3) to R (Chapters 4тАУ5). You will answer one small forestry
question **twice**: once with the Excel skills you already have, and once
with the first R tools. Seeing the *same answer* on both sides is the point.

It deliberately uses **only the skills taught so far**: Excel inspection,
cleaning, formulas, PivotTables, and charting, plus early R setup, import,
inspection, basic functions, the native pipe `|>`, and Quarto. It does
**not** need joins, reshaping, grouped `summarise()`, or `ggplot2` тАФ those
come later, and the full **independent** case study at the end of the
course (Chapter [Integrated Case Study](10-case-study.qmd)) uses them all.

# Chapter goals {.unnumbered}

By the end you will be able to:

- restate a small forestry question and decide what you need to answer it;
- answer it in **Excel** (inspect тЖТ formula / PivotTable тЖТ chart);
- answer the same question in **early R** (import тЖТ inspect тЖТ a basic
  calculation with the native pipe) and **confirm the two agree**;
- verify a result instead of trusting it, and write an AI prompt that
  *checks* your work rather than doing it for you.

# The question

> *"What was the average annual reforested area in British Columbia over
> the whole record, and which single year was the highest?"*

Nothing here needs a join or a grouped summary тАФ just an average, a
maximum, and careful reading of the file.

# Part A тАФ answer it in Excel

You already know these moves from Chapters 1тАУ3:

1. **Inspect.** Open the **Data** sheet. Confirm the columns (`Fiscal_Year`,
   `Reforestation_ha`, тАж), the units (hectares), and that there are no
   missing values.
2. **Average.** In an empty cell, `=AVERAGE(H2:H38)` on the
   `Reforestation_ha` column.
3. **Maximum + its year.** `=MAX(H2:H38)` for the largest value, and
   `=INDEX(A2:A38, MATCH(MAX(H2:H38), H2:H38, 0))` for the fiscal year it
   happened in.
4. **Chart (optional).** A quick line chart of `Reforestation_ha` by
   `Fiscal_Year` shows the shape behind the numbers.

Write your three answers down тАФ you will check them against R next.

# Part B тАФ answer it in early R

::: {.callout-note collapse="true"}
## AI prompt to check your work
Use this **after** your own attempt. It should check your reasoning, not hand you the answer:

> *"My question is `[question]`, and I plan to answer it with these steps: `[list your Excel and early-R steps]`. Check whether my plan uses the simplest tools that answer the question, whether the order makes sense, and which checks would confirm the result. Do not solve the case study for me."*
:::

The same three answers, now reproducibly. This uses only Chapter 4тАУ5
skills: `read_excel()`, inspection, base functions, and the pipe.

```{r excelcs-import}
bc <- read_excel(here("data", "bc_disturbance_reforestation.xlsx"),
                 sheet = "Data")
dim(bc)        # rows and columns
names(bc)      # column names тАФ confirm Reforestation_ha is here
```

```{r excelcs-answer}
# Average reforested area (same as Excel =AVERAGE)
bc$Reforestation_ha |> mean()

# The maximum, and the fiscal year it happened in
bc$Reforestation_ha |> max()
bc$Fiscal_Year[which.max(bc$Reforestation_ha)]
```

The pipe `|>` reads *"take the `Reforestation_ha` column, and then take its
mean."* No new packages, no `group_by()` тАФ just the moves you already have.

**The bridge:** the R average should match your Excel `=AVERAGE` to the
cent, and the R maximum and its year should match your Excel `MAX` /
`INDEX`/`MATCH`. If they *don't*, something is off in one of the two тАФ that
mismatch is a verification win, not a failure.

# Practice Demo Lab

::: {.callout-important}
## Practice demo only (not the graded Canvas lab)
Work through the six steps below. The graded version is on Canvas.
:::

**1 тАФ Your attempt.** For a *different* column тАФ **`Harvested_ha`** тАФ
compute in R: (a) the average harvested area, (b) the maximum, and (c) the
fiscal year of the maximum. Then confirm each against an Excel formula.

**2 тАФ Reference solution.**

```{r excelcs-ref}
bc$Harvested_ha |> mean()                              # (a) average
bc$Harvested_ha |> max()                               # (b) maximum
bc$Fiscal_Year[which.max(bc$Harvested_ha)]             # (c) year of max
```

*(In Excel: `=AVERAGE(E2:E38)`, `=MAX(E2:E38)`, and
`=INDEX(A2:A38, MATCH(MAX(E2:E38), E2:E38, 0))`.)*

**3 тАФ Compare your work with the reference solution.** Where did your
numbers match? Where did they differ? What caused any difference (wrong
column? a stray blank cell in the Excel range? the wrong sheet)? How did
you fix it?

**4 тАФ Write your own AI verification prompt.** Draft a prompt that asks an
AI to *check* your reasoning and outputs тАФ not to hand you the answer.
Include the file name, the column, what you computed, and the checks you
want it to confirm.

**5 тАФ Model AI verification prompt.**

> *"I am using R in FRST 232. I have a tibble `bc` from
> `bc_disturbance_reforestation.xlsx` (Data sheet), columns including
> `Fiscal_Year` and `Harvested_ha`. I computed the mean, the max, and the
> year of the max of `Harvested_ha`. Here is my code and output: `[paste]`.
> Please explain what each line does, tell me whether it correctly answers
> the question, and list checks I can run to confirm it (for example,
> comparing to an Excel `=AVERAGE`/`=MAX`, and checking for blank cells).
> Do not give me a final assignment answer тАФ help me verify my own."*

**6 тАФ Reflection: what changed after checking?** In two or three sentences:
did Excel and R agree the first time? If not, what was wrong? What will you
check *first* next time?

# Done

You have answered one forestry question in two tools and **verified they
agree**. That habit тАФ compute, then confirm тАФ is the foundation of every
later chapter, and of the full independent case study at the end of the
course.
