---
title: "Start of Computing in R: Install, Project, First Import"
---

```{r setup-ch04}
#| include: false
library(tidyverse)
library(readxl)
library(here)
library(knitr)
library(scales)
```

::: {.callout-note appearance="simple"}
## Data for this chapter
This chapter uses the **BC silviculture**
([`bc_disturbance_reforestation.xlsx`](../data/bc_disturbance_reforestation.xlsx){download="bc_disturbance_reforestation.xlsx"})
and **Ontario forest statistics**
([`on_forest_statistics_1.xlsx`](../data/on_forest_statistics_1.xlsx){download="on_forest_statistics_1.xlsx"})
workbooks. Click each file name to download it directly, then save it in
your `data/` folder. For sources and details, see the
[Datasets Used in This Book](../datasets.qmd) page.
:::

# Chapter goals {.unnumbered}

By the end of this chapter you will be able to:

- Explain the difference between **R**, **RStudio**, **Quarto**,
  and **Posit Cloud**, and choose which you will use.
- Install R and RStudio on your computer, **or** create a free
  Posit Cloud account (recommended for getting started).
- Create an **RStudio Project** for FRST 232 with a clean folder
  structure (`data/`, `outputs/`, `R/`, `notes/`).
- Install and load the **`tidyverse`**, **`readxl`**, and
  **`here`** packages.
- Use the **`here()`** function for portable, project-relative
  file paths.
- **Import** the BC silviculture file from Chapter 1 into R using
  `readxl::read_excel()`.
- Inspect the imported tibble with `dim()`, `nrow()`, `ncol()`,
  `names()`, `glimpse()`, `summary()`, and `head()`.
- **Confirm** the numbers R computes match the numbers you
  computed in Excel in Chapters 1–3.
- Use the **native pipe `|>`** to chain operations.
- Recognize and read common error messages.

# Expected output for this chapter {.unnumbered}

::: {.callout-tip appearance="default"}
## What you will hand in

1. A **working RStudio Project** named `frst232/` (or
   `frst232-yourname/`), either on your local machine or on Posit
   Cloud, with this folder structure:

   ```
   frst232/
   ├── frst232.Rproj
   ├── data/bc_disturbance_reforestation.xlsx
   ├── outputs/
   ├── R/
   └── notes/
   ```

2. A rendered Quarto HTML file (`frst232_ch04_learner.html`)
   showing the result of importing and inspecting the BC
   silviculture file. The file should include:
   - the working directory printed by `here::here()`,
   - `dim(bc)`, `names(bc)`, and `glimpse(bc)` output,
   - the five basic statistics (`sum`, `mean`, `min`, `max`,
     `median`) of `bc$Reforestation_ha`,
   - a one-line confirmation that **the R numbers match the Excel
     numbers** from Chapter 1.

3. A short **group-lab worksheet** (or screenshot) submitted to
   the course site.

This list is the same as the rubric for the chapter assignment.
Keep it in view while you work.
:::

# Why R, why now

For three chapters, every analysis lived inside Excel. You wrote
formulas, dragged them across grids, built PivotTables, made
charts. Real work.

But you also saw the limits:

- **PivotTable refresh.** Change the Data sheet → forget to
  refresh → silent stale data.
- **Cell-reference shifts.** Copy a formula one cell over → forget
  the `$` → silently wrong percentages.
- **Display vs value.** Two cells show `100,000` → one is
  `99999.999`. `=A1=B1` is `FALSE` and nobody notices.
- **No history.** Six months later, *"how did I get this number?"*
  is unanswerable.

R fixes these — not by being smarter, but by making every step
**a written instruction**. The instruction is the audit trail. Run
it again, get the same number. Send it to a colleague, get the
same number. Edit one line, re-run, see what changed.

Chapter 4 is where the language changes. The data does not. The
file you import today is the same
`bc_disturbance_reforestation.xlsx` from Chapter 1. The numbers
you compute will be identical. **What changes is that the steps
are now scriptable.**

# R, RStudio, Quarto, Posit Cloud — what each thing is

These four words show up everywhere. They are not the same thing.

| Name | What it is | When you need it |
|---|---|---|
| **R** | The computing language and engine. The thing that does the math. | Always. R is the underlying tool. |
| **RStudio** | A friendly editor for R, with a console, file browser, and project manager. | Almost always. You *could* use R without it, but you do not want to. |
| **Quarto** | A document format that lets you mix R code, R output, and prose into one rendered HTML / PDF / Word file. | When you want to write a report (Chapter 5 onward). |
| **Posit Cloud** | A free, browser-based version of RStudio that runs in the cloud — no install needed. | When you do not want to (or cannot) install R locally. |

::: {.callout-tip}
## The recommended path for this course

If you have never used R before, **start with Posit Cloud**. No
installation. No version conflicts. Works on any laptop, any
operating system, any course-lab computer.

Switch to a local install only if (a) you need to work offline or
(b) you are comfortable troubleshooting R installation issues
yourself.
:::

# Path A — Posit Cloud (recommended)

1. Go to <https://posit.cloud>.
2. Click **Sign Up** (free plan is fine for this course).
3. Once signed in, click **New Project → New RStudio Project**.
4. Rename the project `frst232` (top-left of the window).
5. You now have a full RStudio environment in your browser.

That is it. Skip Path B and go to *Your first R commands*.

# Path B — Install R and RStudio locally

If you prefer (or need) a local install:

## 1. Install R

- Windows: <https://cran.r-project.org/bin/windows/base/>
- macOS: <https://cran.r-project.org/bin/macosx/>
- Pick the latest release. Accept all defaults.

## 2. Install RStudio

- All platforms: <https://posit.co/download/rstudio-desktop/>
- Install **after** R is installed. RStudio finds R automatically.

## 3. Install Quarto

- <https://quarto.org/docs/get-started/>
- Quarto often comes bundled with recent RStudio; if so, you can
  skip this step.

## 4. Verify

Open RStudio. In the **Console** pane (bottom-left), type:

```r
R.version.string
```

You should see something like `"R version 4.5.0 (2025-04-11)"`.
If you do, you are ready.

# Creating an RStudio Project

Whether on Posit Cloud or local, **always work inside an RStudio
Project**. Never run R code without a project.

A project is just a folder with a `.Rproj` file inside it. The
`.Rproj` file tells RStudio:

- where your working directory is,
- which files belong to this project,
- how to keep your R session separate from other projects.

## To create the project

In RStudio: **File → New Project → New Directory → New Project**.

- **Directory name**: `frst232`
- **Create project as subdirectory of**: pick a sensible parent
  folder (e.g., your `Documents/` folder, or — on Posit Cloud —
  the default).
- Click **Create Project**.

RStudio opens a fresh session inside the new folder. The top-right
corner of the window now reads `frst232`.

## Why this matters

Without a project, your R session has **no anchor**. You end up
typing absolute paths like
`/Users/yourname/Desktop/FRST232/Lab1/data.csv` — which break the
moment you move the folder, switch computers, or share with a
classmate.

With a project, you type **relative paths** like
`data/bc_disturbance_reforestation.xlsx`. R knows the project root
and resolves the rest. Portable forever.

# Folder structure (same as Chapters 1–3, now used by R)

Inside your `frst232` project folder, create these subfolders.
In RStudio, use the **Files** pane (bottom-right) → **New Folder**.

```
frst232/
├── frst232.Rproj                                   ← created automatically
├── data/                                           ← raw input files
│   └── bc_disturbance_reforestation.xlsx
├── outputs/                                        ← cleaned files, figures
├── R/                                              ← reusable R scripts (.R files)
├── notes/                                          ← scribbles, screenshots
└── deliverables/                                   ← what you submit
```

Same three rules from Chapter 1:

1. **Never edit raw data in place.** `data/` is read-only.
2. **Use lowercase, snake_case names.** No spaces, no capitals.
3. **One folder per chapter** (or per topic). Easier to find
   things later.

## Place the BC file

If you do not have it already, download
`bc_disturbance_reforestation.xlsx` from the BC Data Catalogue
(<https://catalogue.data.gov.bc.ca/dataset/indicator-summary-data-trends-in-silviculture-in-b-c->)
and drop it into `data/`.

On Posit Cloud, you upload files through the **Files pane →
Upload** button.

# Installing packages

R out of the box can do a lot, but most real work uses **packages**
— pre-written collections of functions. You install a package once,
then load it at the start of every R session that needs it.

The three packages we use this chapter:

| Package | What it does |
|---|---|
| `tidyverse` | A bundle of packages for data manipulation, plotting, and reshaping. The backbone of modern R. |
| `readxl` | Reads `.xlsx` and `.xls` files. |
| `here` | Builds portable, project-relative file paths. |

## To install

In the RStudio Console, type:

```r
install.packages(c("tidyverse", "readxl", "here"))
```

Press Enter. This will take a minute or two the first time — R
downloads each package and its dependencies.

::: {.callout-warning}
## install vs library

- **`install.packages(...)`** — runs **once per machine**.
  Downloads and installs.
- **`library(...)`** — runs **once per R session**. Loads the
  installed package so its functions are available.

Confusing these two is the most common first-day R bug. *"I
installed it but R says it can't find the function"* — almost
always means you forgot the `library()` call.
:::

## To load

At the top of every R script or Quarto document for this course,
include:

```r
library(tidyverse)
library(readxl)
library(here)
```

# `here()` — portable paths

The `here()` function builds a path relative to your project root.
Type this in the Console:

```r
here("data", "bc_disturbance_reforestation.xlsx")
```

You should see a string like:

```text
/cloud/project/data/bc_disturbance_reforestation.xlsx
```

(or, on a local install, something like
`/Users/yourname/Documents/frst232/data/bc_disturbance_reforestation.xlsx`)

The point: **the same code works for everyone**. You do not need
to know the full path; `here()` figures it out.

::: {.callout-tip}
## Why this matters for collaboration

When you share your project folder with a classmate (or with a TA
for grading), they may unzip it to a completely different
location. **Because every file path uses `here()`, the code still
works** — no edits required.

This is the reproducibility win that pure-Excel work cannot give
you.
:::

# Your first R commands

Open a new R script: **File → New File → R Script**. Save it
into `R/` as `01_first_import.R`.

Type these lines (or copy-paste). Run them one at a time by
clicking on the line and pressing **Ctrl+Enter** (Cmd+Enter on
Mac):

```r
# Load packages
library(tidyverse)
library(readxl)
library(here)

# Import the BC silviculture file (same one you used in Chapter 1)
bc <- read_excel(here("data", "bc_disturbance_reforestation.xlsx"),
                 sheet = "Data")

# Inspect
dim(bc)
names(bc)
glimpse(bc)
```

What you should see:

```{r first-import-demo}
#| echo: false
bc <- read_excel(here("data", "bc_disturbance_reforestation.xlsx"),
                 sheet = "Data")
cat("dim(bc):", dim(bc), "\n\n")
cat("names(bc):\n")
print(names(bc))
cat("\nglimpse(bc):\n")
glimpse(bc)
```

37 rows × 8 columns — exactly the BC file you knew from Chapter 1,
now living inside R.

## The assignment operator `<-`

You probably noticed the `<-` symbol. It is the **assignment
operator** — *"store this on the right into this name on the
left"*. In R, `<-` is preferred over `=` for assignment because
`=` has other uses (function arguments).

A keyboard shortcut: **Alt+-** (Windows / Linux) or **Option+-**
(Mac) inserts `<- ` with a space.

# The Excel-to-R bridge — side by side

The following table is what Chapters 1, 2, and 3 have been
previewing. Now you can actually run the right-hand column.

| What you did in Excel | What you do in R | Result on the BC file |
|---|---|---|
| Open the file by double-clicking | `bc <- read_excel(here("data", "bc_disturbance_reforestation.xlsx"), sheet = "Data")` | A tibble in memory called `bc` |
| Read the status-bar `Count` | `nrow(bc)` | `r nrow(bc)` |
| Read the status-bar `Sum` for column H | `sum(bc$Reforestation_ha)` | `r format(round(sum(bc$Reforestation_ha)), big.mark = ",")` |
| Read the status-bar `Average` for column G | `mean(bc$Total_Disturbance_ha)` | `r format(round(mean(bc$Total_Disturbance_ha)), big.mark = ",")` |
| Scan the Data sheet visually | `glimpse(bc)` and `summary(bc)` | text summaries |
| `=MIN(F2:F38)` | `min(bc$Natural_Disturbance_ha)` | `r format(round(min(bc$Natural_Disturbance_ha)), big.mark = ",")` |
| `=MAX(F2:F38)` | `max(bc$Natural_Disturbance_ha)` | `r format(round(max(bc$Natural_Disturbance_ha)), big.mark = ",")` |

**Every number on the right matches the value you saw in Chapter 1.**
The data is the same. The columns are the same. The only thing
that changed is *how you ask for the answer*.

# Accessing columns with `$`

`bc$Reforestation_ha` means *"the column called Reforestation_ha
inside the tibble bc"*. The `$` is the column selector.

Try these in the Console:

```r
bc$Fiscal_Year                  # the year column (37 numbers)
bc$Reforestation_ha             # the reforestation column
sum(bc$Reforestation_ha)        # total reforestation
mean(bc$Reforestation_ha)       # mean annual reforestation
length(bc$Reforestation_ha)     # how many values (= 37)
```

## Tab completion saves typos

In RStudio, after you type `bc$`, press **Tab**. A dropdown shows
every column name. Click one (or arrow-down + Tab) to insert it.
This prevents the *"I spelled the column wrong"* class of bugs.

# Computing the chapter's five statistics

```r
sum(bc$Reforestation_ha)
mean(bc$Reforestation_ha)
min(bc$Reforestation_ha)
max(bc$Reforestation_ha)
median(bc$Reforestation_ha)
```

```{r five-stats}
#| echo: false
tibble(
  Statistic = c("Total reforestation",
                "Mean annual reforestation",
                "Min annual reforestation",
                "Max annual reforestation",
                "Median annual reforestation"),
  Result = c(
    paste0(format(round(sum(bc$Reforestation_ha)), big.mark = ","), " ha"),
    paste0(format(round(mean(bc$Reforestation_ha)), big.mark = ","), " ha"),
    paste0(format(round(min(bc$Reforestation_ha)), big.mark = ","), " ha"),
    paste0(format(round(max(bc$Reforestation_ha)), big.mark = ","), " ha"),
    paste0(format(round(median(bc$Reforestation_ha)), big.mark = ","), " ha")
  )
) |> kable()
```

**Match against Chapter 1.** Open
`bc_silviculture_ch01_workbook.xlsx` from your Chapter 1
deliverables and compare. Total reforestation should match to the
last decimal. If anything does not match, something is wrong —
either with your import or with the file you handed in for
Chapter 1.

# The native pipe `|>`

The pipe operator `|>` takes the thing on its left and sends it
into the first argument of the function on its right. It reads
left-to-right, like a sentence.

Without the pipe:

```r
round(mean(bc$Reforestation_ha), 0)
```

With the pipe:

```r
bc$Reforestation_ha |> mean() |> round(0)
```

Same answer. The piped version reads top to bottom: *"take the
reforestation column → take its mean → round to zero decimals"*.

This is the same logical pattern as **chaining slicers in
PivotTables**: a sequence of operations applied one after the
other. The pipe is how that sequence looks in R.

::: {.callout-tip}
## When pipes help, when they hurt

- Pipes **help** when you have 3 or more steps that read naturally
  as a sequence. `bc |> filter() |> group_by() |> summarise()`.
- Pipes **hurt** when there is only one step. Just write
  `mean(bc$Reforestation_ha)` directly.

Use the pipe when it makes the code easier to read. Skip it when
it does not.
:::

# Reading error messages

Errors in R look intimidating. They are not.

A typical first-day error:

```r
> sum(bc$Refrestation_ha)
Error in `$.tbl_df`(bc, Refrestation_ha) :
  Column `Refrestation_ha` doesn't exist.
Did you mean `Reforestation_ha`?
```

Two facts from that message:

1. The error happens in `$` — that is the column-selector.
2. R suggests a fix — *"Did you mean `Reforestation_ha`?"*. You
   misspelled it.

R errors usually tell you **where** and **why**. Read every word.
The fix is often in the message itself.

## A second common error

```r
> library(tidyverse)
Error in library(tidyverse) : there is no package called 'tidyverse'
```

R is saying *"you tried to load a package you haven't installed"*.
The fix is `install.packages("tidyverse")`, then try again.

::: {.callout-tip}
## How to read any R error in 30 seconds

1. **Read the function name** in the error (e.g., `$`, `library`,
   `read_excel`). That tells you where the problem is.
2. **Read the suggested fix** if there is one (R is good at this).
3. **Copy-paste the error** into your search engine if you cannot
   solve it. Someone else has had the same problem.
4. **Ask AI** with the full error text *and* the code that caused
   it. Vague *"my code doesn't work"* prompts give vague answers.
:::

# Active learning

## Activity 1 — Spot-check the install

In the Console, type:

```r
R.version.string
sessionInfo()
```

The first line should report your R version (4.5 or newer). The
second should list `tidyverse`, `readxl`, and `here` among the
loaded packages (after you have run `library(...)` on each).

## Activity 2 — Verify your folder structure

In the Console, type:

```r
list.files(here("data"))
```

You should see `"bc_disturbance_reforestation.xlsx"`. If you see
an empty result, the file is not where you think it is.

## Activity 3 — Import and inspect

Run the import code from the *Your first R commands* section.
Then answer in your own words:

- What is the *type* of `bc`? (Hint: `class(bc)`)
- How many rows does it have?
- What are the column names?

## Activity 4 — Compute and compare

Compute the five statistics from this chapter. Compare each value
to the same statistic in your Chapter 1 deliverable workbook. Do
they match?

## Activity 5 — Write your first pipe

Take this sequence:

```r
round(mean(bc$Total_Disturbance_ha), 0)
```

Rewrite it as a pipe with three steps. Confirm the answer is the
same.

# Group lab activity

::: {.callout-note appearance="default"}
## Lab: Set up a working R project together

Work in groups of 3 or 4. Each member follows the same setup
steps on their own computer (or in their own Posit Cloud project).
Use the lab session to debug together.

**Task 1 — Posit Cloud accounts.** Every member creates a free
Posit Cloud account (if they have not already).

**Task 2 — Create the project.** Each member creates a project
named `frst232` and sets up the folder structure from the
*Folder structure* section.

**Task 3 — Install packages.** Each member runs:

```r
install.packages(c("tidyverse", "readxl", "here"))
```

While packages install, discuss: *what is a package, and why are
there so many?*

**Task 4 — Upload the data file.** Each member uploads
`bc_disturbance_reforestation.xlsx` into `data/`. Confirm with
`list.files(here("data"))`.

**Task 5 — First import.** Each member runs the import code from
the chapter. Confirm everyone sees the same `dim(bc)` output
(`37 8`).

**Task 6 — Compute and compare.** Each member computes
`sum(bc$Reforestation_ha)`. Confirm everyone gets the same number
(`r format(round(sum(bc$Reforestation_ha)), big.mark = ",")`).
Now compare this to the Chapter 1 workbook value. **They should
match.**

**Task 7 — Submit.** Submit a single screenshot showing one
group member's RStudio with the BC file loaded, the column names
visible, and the `sum()` result. Include all group members' names
on the screenshot or in the submission caption.

This lab is designed to fit comfortably inside a single hands-on
session.
:::

# Exercises

These are the take-home exercises that go with this chapter.

1. Compute `mean(bc$Harvested_ha)` and confirm it equals the same
   number you computed in Excel for Chapter 1 (Activity 3, the
   `=AVERAGE(...)` formula).

2. Use the pipe to compute the median of
   `Natural_Disturbance_ha`, rounded to the nearest hectare.

3. Use `which.max()` to find the **index** of the largest value
   in `bc$Natural_Disturbance_ha`. Then use it to find the year
   of the maximum:

   ```r
   bc$Fiscal_Year[which.max(bc$Natural_Disturbance_ha)]
   ```

   What year does R report? Does it match Chapter 1?

4. Use `summary(bc)` and explain in one sentence what each row of
   the summary shows (Min, 1st Qu, Median, Mean, 3rd Qu, Max).

5. Save your `R/01_first_import.R` script. Close the project.
   Re-open it. Run the whole script again from top to bottom
   (**Code → Run Region → Run All**). Confirm everything works
   without errors.

6. Open the Ontario file from Chapter 2 the same way:

   ```r
   ontario <- read_excel(here("data", "on_forest_statistics_1.xlsx"),
                         sheet = "Data")
   dim(ontario)
   ```

   How many rows does R report? Does it match Chapter 2?

7. **Optional, harder.** Write a tiny R function that takes a
   numeric vector and returns a named list with `mean`, `median`,
   `min`, and `max`. Apply it to `bc$Reforestation_ha`.

# Optional, advanced

## The old pipe `%>%`

You may see `%>%` (the magrittr pipe) in older R code. It does
almost the same thing as the native `|>` but is not built into
base R. This course uses the native `|>` because it is now part
of R itself (since R 4.1) and does not require a package.

If you see `%>%` in a tutorial, treat it as equivalent to `|>` for
most purposes. The differences are minor and only matter in
advanced cases.

## RStudio Project options

In the project menu (top-right of RStudio), choose **Project
Options → General**. Recommended settings for this course:

- **Restore .RData into workspace on startup**: **No**
- **Save workspace to .RData on exit**: **Never**

These two changes guarantee that your R session starts fresh
every time. This sounds inconvenient but is the right discipline:
if your code only works because of *leftover variables from
yesterday*, your code is broken.

## Reading multiple sheets at once

`readxl::excel_sheets()` returns the sheet names of a file:

```r
excel_sheets(here("data", "bc_disturbance_reforestation.xlsx"))
```

You can then `purrr::map()` over them to read all sheets. We will
get to `purrr` later in the course; just know that batch-import is
possible.

# A glimpse of Chapter 5: From Excel to Quarto

In Chapter 5 you will write your first **Quarto** document. Quarto
mixes:

- prose (written in Markdown),
- R code (in chunks like ` ```{r} ... ``` `), and
- the output of that R code (numbers, tables, charts)

…into one rendered HTML, PDF, or Word file. The chapter you are
reading right now is a Quarto document.

| In Excel (Ch 1–3) | In Quarto (Ch 5+) |
|---|---|
| Submit a `.xlsx` workbook | Submit a rendered `.html` (or `.pdf`) file |
| Charts as PNG inserts | Charts as code chunks that re-render |
| Cleaning log on a separate sheet | Cleaning log as Markdown next to the code |
| Hand-edited values inside cells | Code-computed values that update on re-render |
| File is the deliverable | The `.qmd` source AND the rendered file together |

The principle is the same as Excel's *About / ReadMe / Data*
pattern: documentation lives with the data. The difference is that
the documentation, the code, and the output are now one file.

# Self-assessment quiz {#sec-quiz-ch04}

Click the answer you think is correct. The right answer turns
green; wrong answers turn red and the correct one is revealed.
Explanations appear below each question.

```{=html}
<style>
.quiz {
  background: #F4F7F4;
  border-left: 4px solid #2D6A4F;
  padding: 1em 1.5em;
  margin: 1.5em 0;
  border-radius: 6px;
}
.quiz .q { font-weight: 600; color: #1B4332; margin-bottom: 0.6em; }
.quiz ul.qopts { list-style: none; padding: 0; margin: 0; }
.quiz ul.qopts li {
  padding: 0.55em 0.9em; margin: 0.3em 0; background: white;
  border: 1px solid #d8e4d8; border-radius: 4px; cursor: pointer;
  transition: background 0.12s;
}
.quiz ul.qopts li:hover { background: #E8F1E9; }
.quiz ul.qopts li.correct { background: #C8E6C9; border-color: #2D6A4F; font-weight: 600; }
.quiz ul.qopts li.wrong { background: #FFCDD2; border-color: #B05C50; }
.quiz .explanation {
  display: none; margin-top: 0.7em; padding: 0.7em 1em;
  background: #FFF8E8; border-left: 3px solid #B7791F;
  border-radius: 4px; font-size: 0.95em;
}
.quiz .explanation.show { display: block; }
.quiz textarea {
  width: 100%; min-height: 4.5em; padding: 0.6em;
  border: 1px solid #d8e4d8; border-radius: 4px;
  font-family: inherit; font-size: 0.95em; resize: vertical; margin-top: 0.3em;
}
.quiz .reveal-btn {
  background: #2D6A4F; color: white; border: none;
  padding: 0.45em 0.9em; margin-top: 0.5em; border-radius: 4px;
  cursor: pointer; font-size: 0.9em;
}
.quiz .reveal-btn:hover { background: #1B4332; }
.quiz .sample {
  display: none; margin-top: 0.6em; padding: 0.6em 0.9em;
  background: #FFF8E8; border-left: 3px solid #B7791F;
  border-radius: 4px; font-size: 0.95em; font-style: italic;
}
.quiz .sample.show { display: block; }
</style>

<div class="quiz" id="q1" data-hints="One is the engine; the other is the dashboard you drive it from.||R runs the code; RStudio is the editor/IDE around it.||R = language/engine, RStudio = interface.">
  <div class="q">1. What is the difference between R and RStudio?</div>
  <ul class="qopts">
    <li data-correct="false">They are the same thing</li>
    <li data-correct="true">R is the language and engine that runs the math; RStudio is a friendly editor that uses R</li>
    <li data-correct="false">R is the editor; RStudio is the language</li>
    <li data-correct="false">RStudio is an older version of R</li>
  </ul>
  <div class="explanation">R does the computation. RStudio is the interface that lets you write, run, and organize R code more easily.</div>
</div>

<div class="quiz" id="q2" data-hints="What makes relative paths work for anyone who opens the folder?||A Project sets the working directory automatically.||So portable paths like data/file.xlsx work on any machine.">
  <div class="q">2. Why is it recommended to always work inside an RStudio Project?</div>
  <ul class="qopts">
    <li data-correct="false">RStudio refuses to run code without one</li>
    <li data-correct="true">A project sets the working directory automatically, so relative paths (like data/file.xlsx) work for anyone who opens the project</li>
    <li data-correct="false">Projects make R faster</li>
    <li data-correct="false">Projects encrypt your code</li>
  </ul>
  <div class="explanation">The Project anchor lets you use portable file paths. Without it, you end up with absolute paths that break when the folder moves.</div>
</div>

<div class="quiz" id="q3" data-hints="Is installing a per-session or a per-machine action?||Downloading/installing happens once; loading is per session.||Once per machine.">
  <div class="q">3. <code>install.packages("tidyverse")</code> needs to be run —</div>
  <ul class="qopts">
    <li data-correct="false">At the top of every R script, every time</li>
    <li data-correct="true">Once per machine (the package downloads and installs)</li>
    <li data-correct="false">Never — tidyverse is built into R</li>
    <li data-correct="false">Only on Mondays</li>
  </ul>
  <div class="explanation">Install is a one-time download. Loading the package (<code>library()</code>) is what you do every session.</div>
</div>

<div class="quiz" id="q4" data-hints="Loading is different from installing.||R forgets which packages are loaded when it restarts.||Once per R session, at the top of each script.">
  <div class="q">4. <code>library(tidyverse)</code> needs to be run —</div>
  <ul class="qopts">
    <li data-correct="false">Once per machine</li>
    <li data-correct="true">Once per R session (at the top of every script that uses tidyverse functions)</li>
    <li data-correct="false">Never if you have already installed it</li>
    <li data-correct="false">Only inside an RStudio Project</li>
  </ul>
  <div class="explanation">Loading a package is per-session. R forgets which packages are loaded when you restart it.</div>
</div>

<div class="quiz" id="q5" data-hints="here() builds a path from the project root.||The result is a path string, not the file contents.||An absolute path to the file.">
  <div class="q">5. What does <code>here("data", "bc_disturbance_reforestation.xlsx")</code> return?</div>
  <ul class="qopts">
    <li data-correct="false">The contents of the file</li>
    <li data-correct="true">The absolute path to the file, computed relative to the RStudio Project root</li>
    <li data-correct="false">An error message</li>
    <li data-correct="false">The file's modification date</li>
  </ul>
  <div class="explanation"><code>here()</code> builds a portable path. The result is an absolute path string that points to the file regardless of where the project lives on disk.</div>
</div>

<div class="quiz" id="q6" data-hints="Where can the whole group run the same project with no install?||It is RStudio running in the browser.||A free cloud RStudio.">
  <div class="q">6. Posit Cloud is —</div>
  <ul class="qopts">
    <li data-correct="false">A paid alternative to R</li>
    <li data-correct="true">A free, browser-based RStudio environment that runs in the cloud — no local install needed</li>
    <li data-correct="false">A statistics textbook</li>
    <li data-correct="false">An Excel competitor</li>
  </ul>
  <div class="explanation">Posit Cloud is RStudio running in your browser. Recommended for getting started with R because it avoids installation headaches.</div>
</div>

<div class="quiz" id="q7" data-hints="It is not loaded by library(tidyverse).||It is the tidyverse-adjacent reader for .xlsx files.||readxl.">
  <div class="q">7. The function <code>read_excel()</code> comes from which package?</div>
  <ul class="qopts">
    <li data-correct="false">tidyverse</li>
    <li data-correct="true">readxl</li>
    <li data-correct="false">here</li>
    <li data-correct="false">base R</li>
  </ul>
  <div class="explanation"><code>readxl</code> is the tidyverse-adjacent package for reading <code>.xlsx</code> and <code>.xls</code> files. It is not loaded by <code>library(tidyverse)</code> — you need <code>library(readxl)</code> separately.</div>
</div>

<div class="quiz" id="q8" data-hints="How do you select one column by name from a data frame?||Use the dollar sign.||bc$Reforestation_ha.">
  <div class="q">8. After loading the BC file with <code>bc &lt;- read_excel(...)</code>, how do you access just the <code>Reforestation_ha</code> column?</div>
  <ul class="qopts">
    <li data-correct="false">bc["Reforestation_ha"]</li>
    <li data-correct="true">bc$Reforestation_ha</li>
    <li data-correct="false">Reforestation_ha</li>
    <li data-correct="false">bc.Reforestation_ha</li>
  </ul>
  <div class="explanation">The <code>$</code> is the column selector. Tab completion after <code>bc$</code> shows you every column name.</div>
</div>

<div class="quiz" id="q9" data-hints="dim reports the shape — which number comes first?||Rows first, then columns.||37 rows and 8 columns.">
  <div class="q">9. <code>dim(bc)</code> returns <code>c(37, 8)</code>. What does that mean?</div>
  <ul class="qopts">
    <li data-correct="false">37 columns and 8 rows</li>
    <li data-correct="true">37 rows and 8 columns</li>
    <li data-correct="false">37 sheets and 8 files</li>
    <li data-correct="false">37 megabytes and 8 packages</li>
  </ul>
  <div class="explanation"><code>dim()</code> returns rows first, then columns. Use <code>nrow()</code> or <code>ncol()</code> if you only want one of them.</div>
</div>

<div class="quiz" id="q10" data-hints="Same file, same operation — what should the totals be?||Nothing changed but the tool.||Identical.">
  <div class="q">10. You ran <code>sum(bc$Reforestation_ha)</code> in R and got a value. In Excel last chapter you also computed total reforestation. The two numbers should be —</div>
  <ul class="qopts">
    <li data-correct="false">Different — R rounds differently</li>
    <li data-correct="true">Identical — same data, same computation, same answer</li>
    <li data-correct="false">Off by 10%</li>
    <li data-correct="false">It depends on the day</li>
  </ul>
  <div class="explanation">Whole point of the Excel-to-R bridge: the data did not change, the operation did not change, so the answer cannot change. If they differ, something is wrong.</div>
</div>

<div class="quiz" id="q11" data-hints="Which pipe is built into R 4.1+ and needs no package?||Not the older magrittr %>%.||The native pipe |> .">
  <div class="q">11. The native pipe in R is —</div>
  <ul class="qopts">
    <li data-correct="false">%&gt;%</li>
    <li data-correct="true">|&gt;</li>
    <li data-correct="false">-&gt;&gt;</li>
    <li data-correct="false">==&gt;</li>
  </ul>
  <div class="explanation"><code>|&gt;</code> is the native pipe, built into R 4.1+. <code>%&gt;%</code> is the older magrittr pipe, which still works but requires a package.</div>
</div>

<div class="quiz" id="q12" data-hints="The pipe inserts the left value as the first argument of the right function.||So x flows into mean() as its argument.||mean(x).">
  <div class="q">12. <code>x |> mean()</code> is equivalent to —</div>
  <ul class="qopts">
    <li data-correct="false">x + mean()</li>
    <li data-correct="true">mean(x)</li>
    <li data-correct="false">mean(mean(x))</li>
    <li data-correct="false">x * mean</li>
  </ul>
  <div class="explanation">The pipe takes the thing on its left and inserts it as the first argument of the function on its right. Useful when chaining several operations.</div>
</div>

<div class="quiz" id="q13" data-hints="Read the error: the package was never installed (different from not loaded).||Install it first, then load it.||Run install.packages('tidyverse'), then library(tidyverse).">
  <div class="q">13. You see this error: <code>Error in library(tidyverse) : there is no package called 'tidyverse'</code>. What do you do?</div>
  <ul class="qopts">
    <li data-correct="false">Reinstall R</li>
    <li data-correct="true">Run install.packages("tidyverse") first, then library(tidyverse)</li>
    <li data-correct="false">Switch operating systems</li>
    <li data-correct="false">Ignore it</li>
  </ul>
  <div class="explanation">R is telling you the package was never installed on this machine. Install once, then load. The error message itself names the package — read every word.</div>
</div>

<div class="quiz" id="q14" data-hints="What anchors the working directory in a Project?||It is the folder containing the .Rproj file.||The project root folder.">
  <div class="q">14. Inside an RStudio Project, the working directory is —</div>
  <ul class="qopts">
    <li data-correct="false">Your Desktop</li>
    <li data-correct="true">The project root folder (where the .Rproj file lives)</li>
    <li data-correct="false">/tmp/</li>
    <li data-correct="false">Wherever you ran the install</li>
  </ul>
  <div class="explanation">The Project anchors the working directory. <code>here()</code> uses this anchor to resolve relative paths.</div>
</div>

<div class="quiz" id="q15" data-hints="Raw inputs have a dedicated folder.||Keep the download untouched there.||In data/.">
  <div class="q">15. Where should the raw <code>bc_disturbance_reforestation.xlsx</code> file live inside your RStudio Project?</div>
  <ul class="qopts">
    <li data-correct="false">In the project root</li>
    <li data-correct="true">In data/</li>
    <li data-correct="false">In outputs/</li>
    <li data-correct="false">On your Desktop</li>
  </ul>
  <div class="explanation">Same folder structure as Chapter 1: raw files in <code>data/</code>, generated files in <code>outputs/</code>, scripts in <code>R/</code>, deliverables in <code>deliverables/</code>.</div>
</div>

<div class="quiz" id="q16" data-hints="The course prefers the arrow, not the equals sign.||It keeps assignment visually distinct from function arguments.||The &lt;- operator.">
  <div class="q">16. Which is the assignment operator preferred in this course?</div>
  <ul class="qopts">
    <li data-correct="false">=</li>
    <li data-correct="true">&lt;-</li>
    <li data-correct="false">::</li>
    <li data-correct="false">==</li>
  </ul>
  <div class="explanation"><code>&lt;-</code> is the conventional R assignment operator. <code>=</code> works in most cases but is also used for function arguments, so <code>&lt;-</code> keeps the two uses visually distinct. The keyboard shortcut is <strong>Alt+-</strong> (Windows) / <strong>Option+-</strong> (Mac).</div>
</div>

<div class="quiz" id="q17" data-hints="glimpse is a compact, transposed overview of the data.||It lists each column with its type and first values.||A compact summary of every column's type and sample values.">
  <div class="q">17. <code>glimpse(bc)</code> shows you —</div>
  <ul class="qopts">
    <li data-correct="false">A pie chart of the data</li>
    <li data-correct="true">A compact summary of every column with its type and the first few values</li>
    <li data-correct="false">Only the row count</li>
    <li data-correct="false">Only the column names</li>
  </ul>
  <div class="explanation"><code>glimpse()</code> is one of the most useful first-look functions. It shows every column, its type, and a preview of the values — like a richer version of Excel's status bar.</div>
</div>

<div class="quiz" id="q18" data-hints="Think total, average, the two extremes, and the middle value.||These mirror Excel's SUM, AVERAGE, MIN, MAX, MEDIAN.||sum, mean, min, max, median.">
  <div class="q">18. The five basic statistics from this chapter are —</div>
  <ul class="qopts">
    <li data-correct="false">add, subtract, multiply, divide, exponent</li>
    <li data-correct="true">sum, mean, min, max, median</li>
    <li data-correct="false">x, y, z, w, v</li>
    <li data-correct="false">SUMIF, COUNTIF, AVERAGEIF, MAXIF, MINIF</li>
  </ul>
  <div class="explanation">Five base R functions — <code>sum()</code>, <code>mean()</code>, <code>min()</code>, <code>max()</code>, <code>median()</code> — directly equivalent to Excel's <code>SUM</code>, <code>AVERAGE</code>, <code>MIN</code>, <code>MAX</code>, <code>MEDIAN</code>.</div>
</div>

<div class="quiz" id="q19" data-hints="Are R error messages noise, or clues?||They name the function at fault and often suggest a fix.||They are informative — read every word.">
  <div class="q">19. Which is true about R errors?</div>
  <ul class="qopts">
    <li data-correct="false">They are random and impossible to read</li>
    <li data-correct="true">They usually tell you where the problem is (which function) and often suggest a fix</li>
    <li data-correct="false">You should always reinstall R when you see one</li>
    <li data-correct="false">They mean your computer is broken</li>
  </ul>
  <div class="explanation">R errors are written by people for people. Read every word. Often the fix is in the message ("Did you mean ..."), and copy-pasting the error into a search engine usually finds someone with the same problem.</div>
</div>

<div class="quiz" id="q20">
  <div class="q">20. <strong>Reflection.</strong> Describe one moment from this chapter where R felt easier than Excel — or harder. What made the difference?</div>
  <textarea placeholder="Type your answer here..."></textarea>
  <button class="reveal-btn" onclick="this.nextElementSibling.classList.toggle('show')">Show common answers</button>
  <div class="sample">Common reflections: <em>easier — being able to re-run the whole import in one click instead of clicking through Excel menus</em> · <em>harder — remembering install vs library the first time</em> · <em>easier — typing sum(bc$Reforestation_ha) and getting the same number as my Excel formula, instantly</em> · <em>harder — getting used to writing code instead of clicking</em> · <em>aha moment — realizing the data is the same, just the language is different</em>.</div>
</div>

<script>
(function() {
  document.querySelectorAll('.quiz').forEach(function(q) {
    q.querySelectorAll('ul.qopts li').forEach(function(opt) {
      opt.addEventListener('click', function() {
        if (q.dataset.answered) return;
        q.dataset.answered = '1';
        var isCorrect = opt.getAttribute('data-correct') === 'true';
        if (isCorrect) opt.classList.add('correct');
        else {
          opt.classList.add('wrong');
          q.querySelectorAll('ul.qopts li').forEach(function(o) {
            if (o.getAttribute('data-correct') === 'true') o.classList.add('correct');
          });
        }
        var exp = q.querySelector('.explanation');
        if (exp) exp.classList.add('show');
      });
    });
  });
})();
</script>
```

# AI as a debugging companion

::: {.callout-tip}
## Useful prompts for this chapter

- *"I am brand new to R. I just installed it and I'm seeing this
  error message when I run `library(tidyverse)`. What does it
  mean?"* (paste the full error)

- *"Explain step by step what this R code does:
  `bc |> filter(Fiscal_Year >= 2010) |> summarise(total = sum(Reforestation_ha))`"*

- *"I'm getting `Error in `$.tbl_df`(bc, Refrestation_ha)`. The
  column name in my Excel file is `Reforestation_ha`. What's wrong
  and how do I fix it?"*

## Verifying AI's answers

After AI suggests a fix, **check**:

1. Does the fix run without error in your console?
2. Does the resulting number match what you expect from Chapter 1?
3. Did AI invent a function that doesn't exist? (Run
   `?function_name` — if the help page doesn't open, the function
   is fictional.)

AI's R suggestions are usually correct but occasionally include
non-existent functions (called "hallucinations"). The two-second
check is to look up the function in R's own help.

## When *not* to use AI

- Do not paste an exercise question into AI and copy the answer
  back. The point of the exercise is the thinking.
- Do not let AI make code-style decisions for you (pipe vs
  nested, snake_case vs camelCase) without understanding why.
- Do not let AI invent a *replacement* for `here()` that uses
  absolute paths. Portable paths are non-negotiable for this
  course.
:::

# Reading

- The *R for Data Science* book by Wickham, Çetinkaya-Rundel,
  & Grolemund — free at <https://r4ds.hadley.nz/>. Chapter 2
  ("Workflow: basics") and Chapter 8 ("Data import") are
  directly relevant.
- The `here` package vignette:
  <https://here.r-lib.org/articles/here.html>
- Jenny Bryan's classic *"Project-Oriented Workflow"*:
  <https://www.tidyverse.org/blog/2017/12/workflow-vs-script/>
- RStudio Education resources:
  <https://education.rstudio.com/learn/beginner/>

# Instructor notes {.unnumbered}

::: {.callout-note appearance="minimal" collapse="true"}
## Pacing and emphasis

- Chapter 4 is structurally different from Chapters 1–3. The
  first half is **installation and setup**, which can swallow an
  entire lab session if your group is mostly new to R.
- The **Excel-to-R bridge table** in the *Side by side* section
  is the single most important content in this chapter. If
  learners only remember one thing, it should be that the
  numbers they computed in Excel are the same numbers R computes,
  using different syntax for the same data.
- **The old pipe `%>%`** is briefly mentioned in *Optional,
  advanced*. Keep the main path on the native `|>` only — the
  difference rarely matters for beginners, and consistency makes
  the course easier to teach.

## Posit Cloud vs local install

- **Strongly recommend Posit Cloud** for the first lab. It
  eliminates 90% of "my install isn't working" tickets.
- Learners can switch to a local install later in the term once
  they are comfortable with the workflow.
- For the lab session, have one TA ready to help with local
  installs and another with Posit Cloud account creation.

## Expected output checklist (matches the top-of-chapter callout)

1. RStudio Project named `frst232/` with the correct folder
   structure (verify with `list.files(here("data"))`).
2. Rendered Quarto HTML showing `dim(bc)`, `names(bc)`,
   `glimpse(bc)`, and the five basic statistics.
3. Group-lab worksheet or screenshot.

## Materials provided alongside this chapter

| File | Purpose |
|---|---|
| `bc_disturbance_reforestation.xlsx` | Raw BC silviculture workbook (carried forward from Chapter 1). |
| `frst232_ch04_learner.qmd` | Quarto document with TODO chunks for learners to complete. |
| `frst232_ch04_solutions.qmd` | Completed Quarto document with all code chunks filled in. **Do not distribute to learners.** |
| `frst232_ch04_starter.zip` | A pre-built starter project containing the .Rproj file, folder structure, learner .qmd, and a placeholder data folder. |
| `quiz_ch04.html` | Standalone interactive quiz. Embedded in the rendered chapter; can also be hosted separately. |

## Suggested grading guidance for the group lab

| Task | Weight | What to look for |
|---|---|---|
| Posit Cloud accounts (Task 1) | 10 % | Everyone has an account. |
| Project + folders (Task 2) | 20 % | All four folders present; project named correctly. |
| Packages installed (Task 3) | 15 % | No load errors for tidyverse, readxl, here. |
| Data file uploaded (Task 4) | 15 % | `list.files(here("data"))` returns the .xlsx name. |
| Successful import (Task 5) | 15 % | `dim(bc)` returns `37 8`. |
| Sum matches Excel (Task 6) | 15 % | `sum(bc$Reforestation_ha)` matches the Chapter 1 deliverable to the cent. |
| Screenshot submission (Task 7) | 10 % | Screenshot legible; all group members named. |

## Common stumbling points

- **install vs library** confusion. Walk through the distinction
  explicitly at the start of the lab.
- **Forgot to load `readxl`.** `read_excel()` errors with
  *"could not find function"* if `readxl` is not loaded.
- **Wrong working directory.** Learners who skip the RStudio
  Project step end up with absolute paths that work on their
  machine but break in submission. Insist on `here()` from day
  one.
- **Misspelled column names.** R is case-sensitive. `bc$Refor`
  is not `bc$Reforestation_ha`. Tab completion fixes most of
  these.
- **The "saved workspace" trap.** RStudio's default is to save
  your workspace on exit and restore it on startup. This means
  yesterday's variables are still around — until they aren't,
  and your code mysteriously breaks. Recommend the **No / Never**
  setting from *Optional, advanced*.
- *Excel-to-R bridge:* Refer back to the side-by-side table
  often. When a learner asks *"is this the same as the Excel
  formula?"*, the answer is always *"yes — here's the row in the
  bridge table."*
:::
