---
title: "Import and Inspect Forestry Data with R and Quarto"
---

```{r setup-ch05}
#| include: false
library(tidyverse)
library(readxl)
library(here)
library(knitr)
library(scales)
```

::: {.callout-note appearance="simple"}
## Data for this chapter
This chapter uses the **BC silviculture**
([`bc_disturbance_reforestation.xlsx`](../data/bc_disturbance_reforestation.xlsx){download="bc_disturbance_reforestation.xlsx"}),
**Ontario forest statistics**
([`on_forest_statistics_1.xlsx`](../data/on_forest_statistics_1.xlsx){download="on_forest_statistics_1.xlsx"}),
and **BC air-quality**
([`airdata3.csv`](../data/airdata3.csv){download="airdata3.csv"}) files.
Click each file name to download it directly, then save it in your `data/`
folder. For sources and details, see the
[Datasets Used in This Book](../datasets.qmd) page.
:::

# Chapter goals {.unnumbered}

By the end of this chapter you will be able to:

- Explain what a **vector** is in R and recognize the four
  atomic types: logical, integer, double, character.
- Use `c()` to combine values into a vector, and apply
  vectorized functions (`sum`, `mean`, `length`) directly to a
  vector.
- Distinguish a **[tibble](../glossary.qmd#tibble)** from a
  **[data.frame](../glossary.qmd#data-frame)** and explain why
  the tidyverse uses tibbles.
- Identify column types from `glimpse()` output (`<int>`,
  `<dbl>`, `<chr>`, `<lgl>`, `<date>`).
- Import an Excel file with `readxl::read_excel()` and a CSV file
  with `readr::read_csv()`.
- Inspect a tibble with `dim()`, `names()`, `glimpse()`,
  `summary()`, `head()`, and `tail()`.
- Count missing values per column with `colSums(is.na(...))` and
  distinguish `NA` from `0`.
- Write a small **data dictionary** in Quarto next to the code
  that produced it.
- Render a complete **Quarto HTML report** that another reader
  could open without R.

# Expected output for this chapter {.unnumbered}

::: {.callout-tip appearance="default"}
## What you will hand in

1. A **rendered Quarto HTML report** named
   `frst232_ch05_learner.html`, containing:
   - The result of importing both
     `bc_disturbance_reforestation.xlsx` and
     `on_forest_statistics_1.xlsx`.
   - The inspection output — `dim()`, `names()`, `glimpse()`,
     `summary()`, `head()`, and `tail()` — for each file (the
     inspection moves taught in this chapter).
   - A **missing-value summary** for each file (one cell per
     column showing how many `NA` values it has).
   - A **data dictionary table** for the BC file, written by you
     in Markdown.
   - A short prose paragraph (two or three sentences) describing
     what is in each file.
2. The corresponding **`.qmd` source file**.
3. Notification that you have completed the **Excel midterm** in
   the lab session for this chapter (see *Lab session — Excel
   midterm* below).
4. A short **group peer-review worksheet** of another learner's
   rendered HTML.

This list is the same as the rubric for the chapter assignment.
Keep it in view while you work.
:::

# Where we are in the course

In Chapter 4 you:

- Installed R, RStudio, and Quarto (or set up Posit Cloud).
- Created the `frst232/` project with `data/`, `outputs/`, `R/`,
  `notes/`.
- Imported `bc_disturbance_reforestation.xlsx` with
  `readxl::read_excel()`.
- Computed five statistics (`sum`, `mean`, `min`, `max`,
  `median`) on `bc$Reforestation_ha`.
- **Confirmed those numbers match the values you saw in Excel
  for Chapter 1.**

Chapter 5 deepens the R fundamentals: vectors, tibbles, column
types, missing values, and your first full **Quarto report**.

# Excel-to-R bridge — the inspection moves

From Chapter 4 you already know:

| Excel inspection | R command | Returns |
|---|---|---|
| Status-bar `Count` | `dim(bc)` / `nrow(bc)` | Row and column count |
| Status-bar `Sum` | `sum(bc$ColumnName)` | Sum of one column |
| Status-bar `Average` | `mean(bc$ColumnName)` | Mean of one column |

This chapter adds three more — the *missing-value* and *type*
moves that Excel cannot do cleanly:

| Excel pattern | R command | Returns |
|---|---|---|
| "Cells: blank or text? Hard to tell." | `glimpse(bc)` | Each column with type **and** first values |
| "Count = 35 but 37 rows? Two cells are text." | `colSums(is.na(bc))` | Number of missing values per column |
| "Format Cells → Number / Text" | `class(bc$ColumnName)` | The type ("numeric", "character", etc.) |

By the end of this chapter, every move you used to do by *looking*
at Excel is now a scripted R call.

# Vectors — the building block

Almost everything in R is a **vector**: an ordered sequence of
values, all of the same type.

```{r vectors}
# A vector of numbers
years <- c(1987, 1988, 1989, 1990, 1991)
years

# A vector of text
species <- c("Pl", "Fd", "Sx")
species

# A vector of TRUE/FALSE values
is_old <- c(TRUE, FALSE, TRUE, TRUE, FALSE)
is_old
```

The `c()` function (**c** for *combine*) is how you build a
vector. Type the values, comma-separated, inside `c(...)`.

## The four atomic types

Every value in R has a type. The four most common:

| Type | Examples | R shorthand |
|---|---|---|
| **logical** | `TRUE`, `FALSE`, `NA` | `<lgl>` |
| **integer** | `1L`, `2L`, `2026L` | `<int>` |
| **double** | `1`, `3.14`, `-2.5` | `<dbl>` |
| **character** | `"Vancouver"`, `"Pl"` | `<chr>` |

Numbers without an `L` suffix are `<dbl>` (double-precision) by
default. Most numeric R work uses doubles.

## Vectorized operations

Most R functions work on a *whole vector at once*, not one
element at a time:

```{r vectorized}
ha <- c(220000, 180000, 250000, 190000)

# Each of these acts on the whole vector
sum(ha)
mean(ha)
min(ha)
max(ha)
length(ha)

# Arithmetic is vectorized too — element-by-element
ha / 1000      # convert each value to thousands of hectares
ha + 1000      # add 1000 to every value
ha > 200000    # comparison returns a logical vector
```

That last result — `ha > 200000` returns a vector of
`TRUE`/`FALSE` — is the bedrock of every filter and condition you
will write in R.

## Subsetting with `[`

You can pull out specific positions from a vector with square
brackets:

```{r subsetting}
ha[1]                 # first value
ha[c(1, 3)]           # first and third values
ha[ha > 200000]       # only values greater than 200,000
```

That third example reads: *"return the elements of `ha` where
`ha > 200000` is TRUE"*. This is the **logical-vector
subsetting** pattern — the conceptual ancestor of `dplyr::filter()`
which you will meet in Chapter 6.

# Tibbles — vectors arranged in columns

A **tibble** is a tidyverse-flavored data.frame: a rectangular
table where every column is a vector and all columns have the
same length.

When you ran:

```r
bc <- read_excel(here("data", "bc_disturbance_reforestation.xlsx"),
                 sheet = "Data")
```

…in Chapter 4, R built a tibble called `bc`. Each column of `bc`
is a vector. The whole tibble is a list of vectors that R prints
as a rectangle.

```{r tibble-demo}
#| echo: false
bc <- read_excel(here("data", "bc_disturbance_reforestation.xlsx"),
                 sheet = "Data")
print(bc, n = 6)
```

Look at the top of the printout. Just below the column names, you
see things like `<dbl>`, `<dbl>`, ... — the column types. **A
tibble always shows its column types when printed.** That alone
is reason enough to use tibbles over base R data.frames.

## Tibble vs data.frame

For most purposes, a tibble *is* a data.frame. The differences
that matter for this course:

| Feature | data.frame | tibble |
|---|---|---|
| Prints | All rows by default (slow on big data) | First 10 rows only |
| Column types | Hidden until you ask | Always shown |
| Subset with `[, "col"]` | Returns a vector | Returns a tibble |
| Created by | `read.csv()`, `data.frame()` | `read_excel()`, `read_csv()`, `tibble()` |

You can always convert **in either direction**: `as_tibble(df)` turns a
data.frame *into* a tibble, and `as.data.frame(tib)` turns a tibble *back
into* a data.frame. For tidyverse work, stay with tibbles.

# Importing files

::: {.callout-note collapse="true"}
## AI prompt to check your work
Use this **after** your own attempt. It should check your reasoning, not hand you the answer:

> *"I imported `[file]` and got a tibble of `[rows] x [cols]`. Check whether the column types look right, whether any column that should be numeric came in as text (or a date as character), and which inspection commands I should run next. Do not hand me a finished import script."*
:::

Two import functions cover almost every real-world file:

## `readxl::read_excel()`

```{r read-excel}
bc <- read_excel(here("data", "bc_disturbance_reforestation.xlsx"),
                 sheet = "Data")
```

Arguments to remember:

- The first argument is the **file path** — always build it with
  `here()`.
- `sheet = ` names the sheet to read. If you skip it, `read_excel`
  reads the first sheet.
- `skip = 3` skips the first 3 rows (useful when the file has a
  banner or title above the headers).
- `na = c("", "NA", "n/a")` tells `read_excel` which strings to
  treat as missing.

## `readr::read_csv()`

```r
# Example (you do not need to run this; we use it in Chapter 6).
airdata <- read_csv(here("data", "airdata3.csv"))
```

Arguments to remember:

- The first argument is the file path.
- `col_types = ` can lock in column types if `read_csv` guesses
  wrong (advanced — leave it off for now).
- `na = c("", "NA", "-9999")` is useful for fields that use
  sentinel values for missing.

## Other readr siblings (briefly)

- `read_tsv()` — tab-separated.
- `read_delim()` — generic, any delimiter.

All readr functions return a tibble. All accept a file path as
the first argument.

# Inspecting a tibble — the eight moves

Whenever you load a new file, run these eight commands. They take
seconds and catch most problems.

```{r six-moves}
dim(bc)         # rows and columns
nrow(bc)        # rows only
ncol(bc)        # columns only
names(bc)       # column names
head(bc, 3)     # first three rows
tail(bc, 3)     # last three rows
```

```{r glimpse-summary}
glimpse(bc)     # every column, type, and first values
summary(bc)     # min / quartiles / mean / max for each numeric column
```

Each one answers a different question:

| What you want to know | Use |
|---|---|
| How many rows × columns? | `dim()` |
| Just rows | `nrow()` |
| Just columns | `ncol()` |
| The column names | `names()` |
| What the first / last rows look like | `head()` / `tail()` |
| Column types + a preview of values | `glimpse()` |
| Distributions of numeric columns | `summary()` |

`glimpse()` is the single most valuable function in this list.
Make it your default first move on any tibble.

# Column types — and why they matter

In the `glimpse(bc)` output above, you see types like `<dbl>`
next to each column name. The type tells you *what R thinks
this column contains*.

| Type | Means | Example |
|---|---|---|
| `<int>` | integer (whole numbers, no decimals) | `1987L`, `2L` |
| `<dbl>` | double (any real number) | `216304.919` |
| `<chr>` | character (text) | `"Vancouver"` |
| `<lgl>` | logical | `TRUE` / `FALSE` |
| `<fct>` | factor (categorical with a fixed set of levels) | rare on imported files |
| `<date>` | date | `2024-09-15` |

If `read_excel` reads `Fiscal_Year` as `<dbl>` but you want it as
`<int>`, you can coerce after import:

```{r coerce}
bc$Fiscal_Year <- as.integer(bc$Fiscal_Year)
class(bc$Fiscal_Year)
```

Most of the time the inferred types are fine. The only case
where it matters is when something *looks* numeric but
arrives as text (often because of an extra space or a stray
character).

## How to spot a type bug

If you compute `sum(bc$Reforestation_ha)` and get the wrong
answer — or an error message like *"non-numeric argument to
binary operator"* — the column is probably character. Check:

```{r class-check}
class(bc$Reforestation_ha)
```

If it returns `"character"`, fix it:

```r
bc$Reforestation_ha <- as.numeric(bc$Reforestation_ha)
```

# Missing values — `NA` is its own thing

::: {.callout-note collapse="true"}
## AI prompt to check your work
Use this **after** your own attempt. It should check your reasoning, not hand you the answer:

> *"`colSums(is.na(df))` shows `[which columns]` have missing values. Check whether my plan to handle them (`[drop / keep / na.rm = TRUE]`) is appropriate for this dataset, and explain what could go wrong if I treated `NA` as zero. Do not decide the whole cleaning strategy for me."*
:::

In R, missing values are represented by `NA`. `NA` is **not**
zero. It is **not** an empty string. It is its own thing: a
flag meaning *"we do not know this value"*. If that distinction
is new to you, skim **`NA` vs `0`** (below) first — treating an
unknown value as `0` silently corrupts every sum and mean.

## Counting NAs in one column

```{r is-na-vector}
is.na(bc$Reforestation_ha)            # logical vector — TRUE for each NA
sum(is.na(bc$Reforestation_ha))       # how many NAs in this column
```

`is.na()` returns a logical vector of the same length as its
input — `TRUE` for each `NA`, `FALSE` for each real value. Sum
that vector to count the missing values (`TRUE` is treated as
`1`, `FALSE` as `0`).

## Counting NAs in every column at once

```{r colsums-isna}
colSums(is.na(bc))
```

This is the **fastest missing-value audit in R**. One line.
Every column. Always run it after import.

```{r colsums-isna-out}
#| echo: false
colSums(is.na(bc)) |>
  tibble::enframe(name = "column", value = "n_missing") |>
  kable()
```

For the BC silviculture file, every column shows `0` — there are
no missing values, which matches what you saw in Excel for
Chapter 1. (The Ontario file is the same — we will check in the
exercises.)

## `NA` vs `0` — a habit worth forming

| Situation | What to record |
|---|---|
| The variable was measured and the value really is zero | `0` |
| The variable was not measured | `NA` |
| The plot did not exist that year | `NA` |
| The measurement was below the detection limit | depends — see your data dictionary |

Mixing these up is a silent bug class. *"Mean reforestation"*
including 12 unmeasured years coded as `0` will be 12 zeros
pulled into the average. Coded as `NA`, they would be skipped
(if you use `na.rm = TRUE`).

# Skipping `NA` in calculations

```{r na-rm}
# If a column had NAs, the default behaviour is:
mean(c(1, 2, NA, 4))    # returns NA

# Add na.rm = TRUE to skip missing values:
mean(c(1, 2, NA, 4), na.rm = TRUE)   # returns 7/3 = 2.33
```

`na.rm = TRUE` is available on every aggregation function (`sum`,
`mean`, `min`, `max`, `median`, `sd`). Add it whenever your data
might have missing values. Forgetting it is the most common
silent bug in first-year R.

# A worked inspection — the BC file end-to-end

Here is the canonical inspection sequence on a freshly imported
file:

```{r worked-inspection}
# 1. Import
bc <- read_excel(here("data", "bc_disturbance_reforestation.xlsx"),
                 sheet = "Data")

# 2. Dimensions and column names
dim(bc)
names(bc)

# 3. Type + preview
glimpse(bc)

# 4. Numeric summaries
summary(bc)

# 5. Missing-value audit
colSums(is.na(bc))

# 6. Spot-check the verification identity from Chapter 1
identical_check <- all.equal(
  bc$Harvested_ha + bc$Natural_Disturbance_ha,
  bc$Total_Disturbance_ha
)
identical_check
```

Six lines. Six confirmations. You now know:

- The file has 37 rows and 8 columns (matches Chapter 1).
- Column names match the ReadMe (you can compare them by eye).
- Every column is `<dbl>` — no surprise text-as-number issues.
- The `summary()` table tells you each column's range.
- No missing values anywhere.
- The arithmetic identity from Chapter 1 still holds: `Harvested +
  Natural = Total` for every row.

That sequence is **what professional R analysts do every time
they open a new file**. Do it on every dataset you touch.

# First Quarto report

You have been *reading* Quarto documents through this whole book.
Now you write one.

## The four parts of a `.qmd` file

A Quarto file has four ingredients:

| Part | Looks like | What it does |
|---|---|---|
| **YAML header** | `---` block at the top | Tells Quarto how to render |
| **Markdown text** | Plain prose, headings, bullets | The reading content |
| **Code chunks** | ```` ```{r} ... ``` ```` | R code that gets executed |
| **Inline code** | <code>&#96;r expression&#96;</code> | One-line R inside prose |

A minimal `.qmd` file:

```` markdown
---
title: "My first Quarto report"
format: html
---

# Introduction

This report inspects the BC silviculture file.

```{r}
library(readxl)
library(here)
bc <- read_excel(here("data", "bc_disturbance_reforestation.xlsx"),
                 sheet = "Data")
dim(bc)
```

The file has `r nrow(bc)` rows.
````

Save as `my_first_report.qmd`. Click **Render** (or
**Ctrl+Shift+K** / **Cmd+Shift+K**). You get a single self-contained
HTML file.

## Common chunk options

Chunk options go right after the ```` ```{r} ```` line, prefixed
with `#|`:

```` markdown
```{r}
#| label: import-bc
#| echo: false
#| message: false
#| warning: false

bc <- read_excel(here("data", "bc_disturbance_reforestation.xlsx"),
                 sheet = "Data")
```
````

| Option | What it does |
|---|---|
| `label:` | Names the chunk (helpful when one errors) |
| `echo: false` | Hide the code in the rendered output (just show the result) |
| `include: false` | Run the code but hide both code and output (useful for setup) |
| `message: false` | Hide messages from R (e.g., the tidyverse load message) |
| `warning: false` | Hide warnings |
| `eval: false` | Show the code but do not run it |
| `fig.width: 6` | Plot width in inches (Chapter 9) |
| `fig.cap: "..."` | Caption for figures |

## What rendering does

When you click **Render**, Quarto:

1. Reads your `.qmd` file.
2. Runs every R chunk in order, from top to bottom, in a fresh R
   session.
3. Captures the output (text, tables, charts).
4. Stitches code + text + output into a single HTML file.
5. Saves the HTML alongside the `.qmd`.

The **fresh session** part is important: rendering does not use
your interactive console. If a chunk depends on a variable that
was only created in your console (and not in an earlier chunk),
rendering will fail.

::: {.callout-tip}
## A reproducibility test

If your `.qmd` renders cleanly from a fresh R session, your work
is reproducible. If it does not, your work depends on state that
no one else can see — and that is the bug.

**Restart R often.** In RStudio: **Session → Restart R**
(Ctrl+Shift+F10 / Cmd+Shift+F10). Then render. If it still works,
you are good.
:::

# Writing a data dictionary in Quarto

A data dictionary lives next to your code. In Quarto, it is just
a Markdown table:

```` markdown
| Column | Type | Unit | Description |
|---|---|---|---|
| `Fiscal_Year` | integer | year | Fiscal year (April–March) |
| `Harvested_ha` | double | hectares | Total area harvested |
| ... | | | |
````

That renders as:

| Column | Type | Unit | Description |
|---|---|---|---|
| `Fiscal_Year` | integer | year | Fiscal year (April–March) |
| `Harvested_ha` | double | hectares | Total area harvested |
| ... | | | |

The point: **the dictionary travels with the analysis**. When you
hand your `.qmd` to a colleague, they see the table and the
code that built the numbers, in the same file.

::: {.callout-tip}
## Starter for the BC file
The BC silviculture file has these eight columns. Copy this table into
your report and fill in the **Type**, **Unit**, and **Description**
yourself (use `glimpse(bc)` for the types):

| Column | Type | Unit | Description |
|---|---|---|---|
| `Fiscal_Year` | | | |
| `Clearcutting_ha` | | | |
| `Clearcutting_with_reserves_ha` | | | |
| `Partial_cutting_ha` | | | |
| `Harvested_ha` | | | |
| `Natural_Disturbance_ha` | | | |
| `Total_Disturbance_ha` | | | |
| `Reforestation_ha` | | | |

**File-summary template.** After the dictionary, add two or three
sentences that turn your inspection output into prose, e.g.:

> *"The BC silviculture file has 37 rows and 8 columns. The variables are
> numeric areas in hectares plus a fiscal-year integer, and the
> missing-value audit shows no missing values. The main variables describe
> harvested, natural-disturbance, total-disturbance, and reforested area by
> fiscal year."*
:::

# Active learning

## Activity 1 — Vectors

```r
ha <- c(220000, 180000, 250000, 190000, 210000)
```

In the console, compute:

- the sum,
- the mean,
- the count of values greater than 200,000 (hint: `sum(ha > 200000)`),
- the values greater than 200,000 themselves.

## Activity 2 — Inspect two files

Run the six-move inspection on the BC silviculture file, then
on the Ontario forest stats file. Note the differences:

```r
ontario <- read_excel(here("data", "on_forest_statistics_1.xlsx"),
                      sheet = "Data")
dim(ontario); glimpse(ontario)
```

How are the column types different from the BC file? Which file
has any character columns?

## Activity 3 — Missing-value audit

Run `colSums(is.na(bc))` and `colSums(is.na(ontario))`. Are there
any missing values? Confirm against what you saw in Excel for
Chapters 1 and 2.

## Activity 4 — Render your first Quarto report

Open the chapter learner file `frst232_ch05_learner.qmd`.
Render it. Confirm the HTML opens and shows all expected output.

## Activity 5 — Break and fix

In `frst232_ch05_learner.qmd`, deliberately misspell
`Reforestation_ha` somewhere. Render. Read the error message
carefully. Fix it. Re-render. The error message taught you
something useful — what?

# Lab session — Excel midterm

::: {.callout-important appearance="default"}
## The lab session for Chapter 5 is the Excel midterm

There is no separate practice lab this chapter. The lab session
is given over to the **Excel midterm**, covering material from
Chapters 1–3:

- Importing and inspecting an Excel workbook (Chapter 1).
- Cleaning, summarizing, and writing formulas (Chapter 2).
- Visualization and PivotTables (Chapter 3).

The midterm is closed-AI but open-notes. Bring a laptop with
Excel installed (or use the lab computers). The teaching team will
provide a midterm-specific workbook at the start of the session.

**What to do before the lab:**

- Re-read the Chapter 1, 2, and 3 *Expected output* callouts.
- Make sure your laptop has Excel working.
- Bring your `bc_disturbance_reforestation.xlsx` and
  `on_forest_statistics_1.xlsx` files in case the midterm
  references them.

**What is NOT on the midterm:**

- R code.
- Quarto rendering.
- Anything from Chapter 4 or Chapter 5.

The R material is assessed through homework exercises, the
chapter quizzes, and the final exam at the end of the term.
:::

# Group activity — Peer review of Quarto reports

Outside the midterm session, run this peer-review activity at any
time during the chapter (e.g., during a lecture session).

::: {.callout-note appearance="default"}
## Group: Peer review a rendered HTML report

Work in pairs (or threes). Each member should have a rendered
`frst232_ch05_learner.html` to share.

**Task 1 — Swap.** Each person sends their rendered HTML to the
person on their left. Open it in a browser.

**Task 2 — Score each other's report** using this rubric:

| Criterion | 0 | 1 | 2 |
|---|---|---|---|
| Renders without error | Did not render | Renders with warnings | Renders cleanly |
| `dim()`, `names()`, `glimpse()` visible | Missing | Present but unclear | Clearly visible |
| Missing-value summary present | Missing | Present but unexplained | Present with note |
| Data dictionary table | Missing | Incomplete | Complete and readable |
| Two-sentence prose summary | Missing | Vague | Specific to the file |

**Task 3 — Give one piece of constructive feedback** in writing.
What is the strongest part of the report? What would you change?

**Task 4 — Submit.** Submit your peer-review rubric (with your
own name and the author's name) to the course site.

This activity is designed to fit inside a single lecture session.
:::

# Exercises

These are the take-home exercises that go with this chapter.

1. Run the **six-move inspection** sequence on the BC silviculture
   file. Copy the output of each function into your `.qmd` file
   inside R chunks.

2. Run the same six-move inspection on the **Ontario file**.
   Compare the two outputs. Which file has character columns?
   How many?

3. Compute the **missing-value summary** for both files using
   `colSums(is.na(...))`. Are there any missing values?

4. Build a **data dictionary table** for the BC silviculture file
   in your `.qmd`, in the format shown in this chapter. Each row
   should have *Column*, *Type*, *Unit*, *Description*.

5. Use **vectorized arithmetic** to compute the ratio
   `Reforestation_ha / Total_Disturbance_ha` for every year. Save
   the result as a new vector. How many years had a ratio greater
   than 1?

6. Use **`summary(bc)`** to find the **median** of
   `Total_Disturbance_ha`. Confirm it matches the value you would
   compute with `median(bc$Total_Disturbance_ha)`.

7. **Restart R** (Session → Restart R). Re-render your
   `frst232_ch05_learner.qmd` from scratch. Confirm it renders
   without errors. (If it does not, fix what depends on hidden
   console state.)

8. **Optional, harder.** Write a tiny R function called
   `inspect_file()` that takes a tibble and prints `dim`,
   `names`, `glimpse`, and `colSums(is.na(...))` all at once.
   Apply it to `bc` and `ontario`.

# Optional, advanced

## Lists — the catch-all data structure

A **list** in R is like a vector except its elements can be of
different types. A tibble is technically a list of equal-length
vectors.

```r
my_list <- list(name = "BC silviculture",
                rows = 37,
                cols = 8,
                tibble = bc)
my_list$name
my_list$tibble
```

You will use lists in Chapter 11 when working with spatial data
(where each "row" is a complex geometry).

## Factors — categorical variables with order

A **factor** is a vector with a fixed set of allowed values
(levels). Useful when you want ordered categories (`small` <
`medium` < `large`) or when a categorical variable has a known
set of valid values.

```r
size <- factor(c("small", "large", "medium", "large"),
               levels = c("small", "medium", "large"))
size
```

We will use factors more in Chapter 9 (ggplot2 axis ordering).

## Locking column types at import time

If `read_excel` or `read_csv` guesses a column type wrong, you
can override it:

```r
read_csv("file.csv",
         col_types = cols(
           plot_id = col_character(),
           dbh_cm = col_double(),
           date_measured = col_date(format = "%Y-%m-%d")
         ))
```

This is overkill for clean BC and Ontario files. It becomes
useful when you import messy real-world CSVs in Chapter 6.

# A glimpse ahead: From inspection to cleaning

In Chapter 6 you move from *inspecting* to *cleaning*. The core
moves of `dplyr`:

| What you want to do | Excel (Ch 2) | dplyr verb (Ch 6) |
|---|---|---|
| Keep some rows | AutoFilter | `filter()` |
| Add a new column | `=A2+B2` in a new column | `mutate()` |
| Sort | Data → Sort | `arrange()` |
| Keep only some columns | Delete columns | `select()` |
| Group and summarise | PivotTable | `group_by() |> summarise()` |
| Look up from a reference | `VLOOKUP` | `left_join()` |
| Recode values | `IFS` | `case_when()` |

**Every Excel cleaning move you learned in Chapter 2 has a one-line
dplyr equivalent.** The names are short, the pattern is
consistent, and once you know the six verbs above, you can clean
almost any real-world dataset.

Chapter 6 walks through each verb on the BC air-quality dataset
(`airdata3.csv`) — a messier file than what you have seen so far,
with actual missing values, mixed date formats, and the kind of
problems that real forestry data presents.

# Self-assessment quiz {#sec-quiz-ch05}

::: {.content-visible when-format="pdf"}
*The self-assessment quiz is interactive: it appears in the online (HTML)
book, and you can also take it on Canvas. This PDF does not display the
interactive quiz questions.*
:::

::: {.content-visible when-format="html"}
Click the answer you think is correct. The right answer turns
green; wrong answers turn red and the correct one is revealed.
Explanations appear below each question.
:::

```{=html}
<style>
.quiz {
  background: #F4F7F4;
  border-left: 4px solid #2D6A4F;
  padding: 1em 1.5em;
  margin: 1.5em 0;
  border-radius: 6px;
}
.quiz .q { font-weight: 600; color: #1B4332; margin-bottom: 0.6em; }
.quiz ul.qopts { list-style: none; padding: 0; margin: 0; }
.quiz ul.qopts li {
  padding: 0.55em 0.9em; margin: 0.3em 0; background: white;
  border: 1px solid #d8e4d8; border-radius: 4px; cursor: pointer;
  transition: background 0.12s;
}
.quiz ul.qopts li:hover { background: #E8F1E9; }
.quiz ul.qopts li.correct { background: #C8E6C9; border-color: #2D6A4F; font-weight: 600; }
.quiz ul.qopts li.wrong { background: #FFCDD2; border-color: #B05C50; }
.quiz .explanation {
  display: none; margin-top: 0.7em; padding: 0.7em 1em;
  background: #FFF8E8; border-left: 3px solid #B7791F;
  border-radius: 4px; font-size: 0.95em;
}
.quiz .explanation.show { display: block; }
.quiz textarea {
  width: 100%; min-height: 4.5em; padding: 0.6em;
  border: 1px solid #d8e4d8; border-radius: 4px;
  font-family: inherit; font-size: 0.95em; resize: vertical; margin-top: 0.3em;
}
.quiz .reveal-btn {
  background: #2D6A4F; color: white; border: none;
  padding: 0.45em 0.9em; margin-top: 0.5em; border-radius: 4px;
  cursor: pointer; font-size: 0.9em;
}
.quiz .reveal-btn:hover { background: #1B4332; }
.quiz .sample {
  display: none; margin-top: 0.6em; padding: 0.6em 0.9em;
  background: #FFF8E8; border-left: 3px solid #B7791F;
  border-radius: 4px; font-size: 0.95em; font-style: italic;
}
.quiz .sample.show { display: block; }
</style>

<div class="quiz" id="q1" data-hints="Think of one column of values, all the same kind.||They are stored in order.||An ordered collection of values of the same type.">
  <div class="q">1. What is a vector in R?</div>
  <ul class="qopts">
    <li data-correct="false">A graphical chart</li>
    <li data-correct="true">An ordered sequence of values, all of the same type</li>
    <li data-correct="false">A file format</li>
    <li data-correct="false">The same as a list</li>
  </ul>
  <div class="explanation">Vectors are the building block of R. A vector has a single type — logical, integer, double, or character — and a length.</div>
</div>

<div class="quiz" id="q2" data-hints="These are plain numbers that could have decimals.||R calls that type double.||Numeric (double).">
  <div class="q">2. <code>c(1987, 1988, 1989)</code> is a vector of which type?</div>
  <ul class="qopts">
    <li data-correct="false">integer (<int>)</li>
    <li data-correct="true">double (<dbl>)</li>
    <li data-correct="false">character (<chr>)</li>
    <li data-correct="false">logical (<lgl>)</li>
  </ul>
  <div class="explanation">Numbers without an <code>L</code> suffix are doubles by default. <code>c(1987L, 1988L, 1989L)</code> would be integer.</div>
</div>

<div class="quiz" id="q3" data-hints="It is the tidyverse's version of a table.||Columns are variables; rows are observations.||A modern data frame (a rectangular table).">
  <div class="q">3. A tibble is —</div>
  <ul class="qopts">
    <li data-correct="false">A new programming language</li>
    <li data-correct="true">A tidyverse-flavored data.frame — a rectangular table where each column is a vector</li>
    <li data-correct="false">A type of chart</li>
    <li data-correct="false">A statistical model</li>
  </ul>
  <div class="explanation">Tibbles are the tidyverse's default rectangular structure. They print better than data.frames and show column types.</div>
</div>

<div class="quiz" id="q4" data-hints="The angle-bracket tag under a column is its type.||dbl is short for double.||A numeric (double) column.">
  <div class="q">4. In a tibble printout, you see <code>&lt;dbl&gt;</code> under a column name. What does that mean?</div>
  <ul class="qopts">
    <li data-correct="false">The column has no values</li>
    <li data-correct="true">The column is a double (any real number)</li>
    <li data-correct="false">The column is text</li>
    <li data-correct="false">The column has been deleted</li>
  </ul>
  <div class="explanation"><code>&lt;dbl&gt;</code> = double-precision number. <code>&lt;int&gt;</code> = integer. <code>&lt;chr&gt;</code> = character (text). <code>&lt;lgl&gt;</code> = logical.</div>
</div>

<div class="quiz" id="q5" data-hints="Which tidyverse reader handles comma-separated files?||Not read_excel — that is for .xlsx.||read_csv().">
  <div class="q">5. To import a CSV file in R, you would use —</div>
  <ul class="qopts">
    <li data-correct="false">read_excel()</li>
    <li data-correct="true">read_csv()</li>
    <li data-correct="false">read_pdf()</li>
    <li data-correct="false">open()</li>
  </ul>
  <div class="explanation"><code>read_csv()</code> from the <code>readr</code> package (part of tidyverse) reads comma-separated files into a tibble. <code>read_excel()</code> is for <code>.xlsx</code> files only.</div>
</div>

<div class="quiz" id="q6" data-hints="is.na() flags each cell TRUE/FALSE; colSums adds those up per column.||TRUE counts as 1.||The number of missing values in each column.">
  <div class="q">6. <code>colSums(is.na(bc))</code> returns —</div>
  <ul class="qopts">
    <li data-correct="false">The sum of every column</li>
    <li data-correct="true">A named vector giving the number of missing (NA) values in each column</li>
    <li data-correct="false">Whether the file is complete</li>
    <li data-correct="false">An error</li>
  </ul>
  <div class="explanation"><code>is.na(bc)</code> returns a logical matrix (TRUE for each NA). <code>colSums()</code> adds the TRUEs per column. One line, complete missing-value audit.</div>
</div>

<div class="quiz" id="q7" data-hints="NA is not zero and not empty text.||It marks a value that is unknown.||A missing value.">
  <div class="q">7. <code>NA</code> in R means —</div>
  <ul class="qopts">
    <li data-correct="false">Zero</li>
    <li data-correct="false">An empty string</li>
    <li data-correct="true">Missing — the value is unknown</li>
    <li data-correct="false">North America</li>
  </ul>
  <div class="explanation"><code>NA</code> is its own value type. Mixing it up with 0 is a silent bug: an unmeasured year recorded as 0 will be pulled into averages; recorded as NA, it will be skipped if you use <code>na.rm = TRUE</code>.</div>
</div>

<div class="quiz" id="q8" data-hints="By default, any NA in the input makes the result NA.||You would need na.rm = TRUE to ignore it.||NA.">
  <div class="q">8. <code>mean(c(1, 2, NA, 4))</code> returns —</div>
  <ul class="qopts">
    <li data-correct="false">2.33</li>
    <li data-correct="true">NA</li>
    <li data-correct="false">An error</li>
    <li data-correct="false">7</li>
  </ul>
  <div class="explanation">By default, R refuses to silently ignore missing values. Add <code>na.rm = TRUE</code> to skip them: <code>mean(c(1,2,NA,4), na.rm = TRUE)</code> returns 2.33.</div>
</div>

<div class="quiz" id="q9" data-hints="Which function lists each column with its type and first values?||Not summary (that gives per-column stats).||glimpse().">
  <div class="q">9. Which function gives the most useful one-line preview of every column in a tibble (name, type, and first values)?</div>
  <ul class="qopts">
    <li data-correct="false">dim()</li>
    <li data-correct="false">summary()</li>
    <li data-correct="true">glimpse()</li>
    <li data-correct="false">head()</li>
  </ul>
  <div class="explanation"><code>glimpse()</code> shows every column, its type, and a preview of the first values — like a richer version of Excel's status bar.</div>
</div>

<div class="quiz" id="q10" data-hints="summary gives distribution statistics for each numeric column.||Min, quartiles, median, mean, max.||A five-number summary plus the mean per column.">
  <div class="q">10. <code>summary(bc)</code> on a tibble of numeric columns returns —</div>
  <ul class="qopts">
    <li data-correct="false">The R version</li>
    <li data-correct="true">For each column: Min, 1st Quartile, Median, Mean, 3rd Quartile, Max (and NA count if any)</li>
    <li data-correct="false">Only the column names</li>
    <li data-correct="false">A chart</li>
  </ul>
  <div class="explanation"><code>summary()</code> produces the five-number summary plus the mean for each numeric column. For character columns it just notes the type.</div>
</div>

<div class="quiz" id="q11" data-hints="Recall the four parts: YAML header, prose, code chunks, inline code.||A stylesheet can be added but is not required.||A standalone CSS file is NOT one of the four.">
  <div class="q">11. A Quarto <code>.qmd</code> file has four main parts. Which is NOT one of them?</div>
  <ul class="qopts">
    <li data-correct="false">YAML header</li>
    <li data-correct="false">Markdown text</li>
    <li data-correct="false">Code chunks</li>
    <li data-correct="true">A standalone CSS file</li>
  </ul>
  <div class="explanation">The four parts: YAML header, Markdown text, code chunks (```` ```{r} ... ``` ````), and inline R code (<code>&#96;r ...&#96;</code>). CSS can be customized but is not a required part.</div>
</div>

<div class="quiz" id="q12" data-hints="echo controls whether the code is shown, not whether it runs.||The output still appears.||It hides the code but keeps the output.">
  <div class="q">12. Inside a Quarto code chunk, what does <code>#| echo: false</code> do?</div>
  <ul class="qopts">
    <li data-correct="false">Skips the chunk entirely</li>
    <li data-correct="true">Hides the code in the rendered HTML; still shows the output</li>
    <li data-correct="false">Hides the output; still shows the code</li>
    <li data-correct="false">Stops the chunk after one second</li>
  </ul>
  <div class="explanation"><code>echo: false</code> shows the output but hides the source code. Use it for the setup chunk and for any code your reader does not need to see.</div>
</div>

<div class="quiz" id="q13" data-hints="Render runs the code and weaves prose with output.||It produces a finished, shareable document.||It executes the chunks and produces an HTML (or other) report.">
  <div class="q">13. When you click <strong>Render</strong> in RStudio, Quarto —</div>
  <ul class="qopts">
    <li data-correct="false">Just exports the .qmd file as text</li>
    <li data-correct="true">Runs every code chunk in order in a fresh R session and produces a self-contained HTML (or PDF) file</li>
    <li data-correct="false">Asks an AI to render the document</li>
    <li data-correct="false">Saves the file to the cloud</li>
  </ul>
  <div class="explanation">Render = run every chunk top-to-bottom in a fresh session + assemble code, text, and output into one HTML. The fresh session is what makes rendering a reproducibility test.</div>
</div>

<div class="quiz" id="q14" data-hints="The dollar sign pulls one column out of the tibble.||A single column is itself a vector.||A vector — one column of the tibble.">
  <div class="q">14. <code>bc$Fiscal_Year</code> is —</div>
  <ul class="qopts">
    <li data-correct="false">A tibble</li>
    <li data-correct="true">A vector — the Fiscal_Year column of the bc tibble</li>
    <li data-correct="false">An error</li>
    <li data-correct="false">A function</li>
  </ul>
  <div class="explanation">Using <code>$</code> on a tibble returns a single column as a vector. <code>bc[, "Fiscal_Year"]</code> on a tibble returns a one-column tibble, not a vector.</div>
</div>

<div class="quiz" id="q15" data-hints="A comparison is evaluated for each element of the vector.||Each value becomes TRUE or FALSE.||A logical vector of TRUE/FALSE.">
  <div class="q">15. <code>ha &gt; 200000</code> returns —</div>
  <ul class="qopts">
    <li data-correct="false">A single number</li>
    <li data-correct="true">A logical vector — TRUE or FALSE for each element of ha</li>
    <li data-correct="false">An error</li>
    <li data-correct="false">The number of values greater than 200,000</li>
  </ul>
  <div class="explanation">Comparison in R is vectorized: each element of <code>ha</code> is compared to 200000, returning a parallel logical vector. To count how many are TRUE, wrap with <code>sum()</code>.</div>
</div>

<div class="quiz" id="q16" data-hints="Render starts a FRESH R session — your console objects are gone.||A chunk can only see what it (or an earlier chunk) creates.||It errors: the object is not found.">
  <div class="q">16. Inside a Quarto chunk, you write code that uses a variable created only in your interactive console. When you Render, what happens?</div>
  <ul class="qopts">
    <li data-correct="false">It works fine — Quarto sees the console</li>
    <li data-correct="true">It fails — Render runs in a fresh R session that does not see console state</li>
    <li data-correct="false">It works only on Posit Cloud</li>
    <li data-correct="false">It runs the chunk twice</li>
  </ul>
  <div class="explanation">Render uses a fresh session. Everything the document depends on must be created inside a chunk. This is the whole point of Quarto being reproducible.</div>
</div>

<div class="quiz" id="q17" data-hints="Check the course schedule — it is in a lab session.||It is the Chapter 5 lab.||In the Chapter 5 lab session.">
  <div class="q">17. The Excel midterm is held —</div>
  <ul class="qopts">
    <li data-correct="false">Online, asynchronously</li>
    <li data-correct="true">During the Chapter 5 lab session</li>
    <li data-correct="false">At the end of the term</li>
    <li data-correct="false">There is no midterm</li>
  </ul>
  <div class="explanation">The Chapter 5 lab session is given over to the midterm. The midterm covers Chapter 1, 2, and 3 (Excel) material.</div>
</div>

<div class="quiz" id="q18" data-hints="That Excel warning is about values silently skipped because they are missing or non-numeric.||In R you surface those by counting the NAs.||colSums(is.na(...)) reveals the skipped/missing cells.">
  <div class="q">18. Which is the closest equivalent of Excel's status-bar "Count = 35 but I selected 37 cells" warning?</div>
  <ul class="qopts">
    <li data-correct="false">summary(bc)</li>
    <li data-correct="false">dim(bc)</li>
    <li data-correct="true">colSums(is.na(bc)) — a non-zero entry means that column has missing or non-numeric cells</li>
    <li data-correct="false">head(bc)</li>
  </ul>
  <div class="explanation">In Excel, status-bar Count &lt; selection size meant some cells were text-not-number. In R, <code>colSums(is.na(...))</code> per column tells you the same thing more precisely.</div>
</div>

<div class="quiz" id="q19" data-hints="Both are tables; the difference is printing and strictness.||A tibble prints types and only the first rows, and never silently changes types.||A tibble prints nicer and is stricter, but is otherwise a data frame.">
  <div class="q">19. What is the main difference between a tibble and a data.frame in this course?</div>
  <ul class="qopts">
    <li data-correct="false">Tibbles cannot hold text</li>
    <li data-correct="true">Tibbles print better (first 10 rows, column types shown) and are the default in tidyverse imports</li>
    <li data-correct="false">data.frames are faster</li>
    <li data-correct="false">There is no difference</li>
  </ul>
  <div class="explanation">For most operations, a tibble IS a data.frame (it inherits from it). The differences matter for printing, subsetting (<code>tibble[, "col"]</code> returns a tibble, not a vector), and consistency with tidyverse functions.</div>
</div>

<div class="quiz" id="q20">
  <div class="q">20. <strong>Reflection.</strong> Now that you have rendered a Quarto report, how does the experience compare to submitting an Excel workbook? What is easier? What is harder?</div>
  <textarea placeholder="Type your answer here..."></textarea>
  <button class="reveal-btn" onclick="this.nextElementSibling.classList.toggle('show')">Show common answers</button>
  <div class="sample">Common reflections: <em>easier — every number in the report is traceable to the code that produced it</em> · <em>harder — getting the YAML header right the first time</em> · <em>easier — re-running with a new dataset doesn't require redoing every chart and table</em> · <em>harder — error messages on render are scary at first</em> · <em>aha moment — the Quarto report carries its own data dictionary AND its own analysis, in one file.</em></div>
</div>

<script>
(function() {
  document.querySelectorAll('.quiz').forEach(function(q) {
    q.querySelectorAll('ul.qopts li').forEach(function(opt) {
      opt.addEventListener('click', function() {
        if (q.dataset.answered) return;
        q.dataset.answered = '1';
        var isCorrect = opt.getAttribute('data-correct') === 'true';
        if (isCorrect) opt.classList.add('correct');
        else {
          opt.classList.add('wrong');
          q.querySelectorAll('ul.qopts li').forEach(function(o) {
            if (o.getAttribute('data-correct') === 'true') o.classList.add('correct');
          });
        }
        var exp = q.querySelector('.explanation');
        if (exp) exp.classList.add('show');
      });
    });
  });
})();
</script>
```

# AI as a debugging companion

::: {.callout-tip}
## Useful prompts for this chapter

- *"My Quarto document fails to render with the error
  `object 'bc' not found`. The variable works fine in my console.
  What might be wrong?"* (Hint: fresh-session rendering.)

- *"Explain this output of `glimpse(bc)` line by line:"* (paste
  the glimpse output)

- *"I imported a CSV with `read_csv` and one of my numeric columns
  was read as character. What's the most common cause and how do
  I fix it?"*

## A prompt template for the rest of the course

When you ask AI for help with R code, include four things:

1. Your R version (`R.version.string`).
2. The packages you have loaded (the top of your `.qmd`).
3. The exact code you ran.
4. The exact error message (copy-paste, do not paraphrase).

Without these four, the answer will be too generic to use.

## Verifying AI's answers

After AI suggests a fix, **check**:

1. Run the suggested code in your console. Does it work?
2. Does the result match what you expect from
   `glimpse()`, `summary()`, or a manual count?
3. If AI rewrites a chunk for you, render the whole document
   afterwards to confirm nothing else broke.

## When *not* to use AI

- Do not let AI write your data dictionary. The dictionary
  describes *your* analysis decisions.
- Do not let AI hide an unexplained behaviour with a `try()` or
  `suppressWarnings()`. Warnings exist for a reason — read them.
:::

# Reading

- *R for Data Science* — chapter 2 *Workflow: basics* and chapter
  8 *Data import*. Free at <https://r4ds.hadley.nz/>.
- *Quarto official guide* — <https://quarto.org/docs/guide/>.
  Start with *Authoring* → *Markdown Basics* and *Computations*
  → *R*.
- Jenny Bryan, *"Naming things"* — short lecture on file naming
  conventions: <https://speakerdeck.com/jennybc/how-to-name-files>.
- The `readxl` documentation: <https://readxl.tidyverse.org/>.

# Instructor notes {.unnumbered}

::: {.callout-note appearance="minimal" collapse="true"}
## Pacing and emphasis

- The chapter has two big new ideas: **vectors / tibbles** (the
  R type system) and **Quarto** (the new deliverable format).
  Spend time on both; do not rush past the YAML header.
- The **missing-value audit** (`colSums(is.na(bc))`) is the
  single most useful one-line R command in this chapter. Use
  it in your live demo.
- **Quarto rendering must succeed before submission.** Many
  learners will hand in a `.qmd` that does not render. Build
  the "render from a fresh session" check into the grading
  rubric.

## Excel midterm logistics

- The lab session for this chapter is the midterm.
- Material is Chapter 1, 2, 3 Excel only.
- Recommended midterm structure: one 90-minute Excel deliverable
  (single workbook) with sections on importing, cleaning,
  summarizing, and a PivotTable. Provide a starter file at the
  start of the session.
- Closed-AI, open-notes. Learners may use the chapter materials
  but not generative-AI tools.
- TAs should be prepared for *"my Excel won't open"* support
  questions at the start of the session — have a backup
  computer ready.

## Expected output checklist (matches the top-of-chapter callout)

1. `frst232_ch05_learner.html` (rendered) with imports for both
   BC and Ontario files.
2. `frst232_ch05_learner.qmd` (source).
3. Excel midterm completed in the lab session.
4. Peer-review worksheet (during the lecture peer-review
   activity).

## Materials provided alongside this chapter

| File | Purpose |
|---|---|
| `bc_disturbance_reforestation.xlsx` | Raw BC silviculture (from Ch 1, unchanged). |
| `on_forest_statistics_1.xlsx` | Raw Ontario forest stats (from Ch 2, unchanged). |
| `frst232_ch05_learner.qmd` | Quarto starter with TODO chunks for each exercise. |
| `frst232_ch05_solutions.qmd` | Completed Quarto document. **Do not distribute to learners.** |
| `quiz_ch05.html` | Standalone interactive quiz. |

## Common stumbling points

- **"Object not found on render."** Almost always: the variable
  was created in the console, not in a chunk. Solution: Restart
  R, then re-render.
- **YAML errors.** Quarto is fussy about the YAML header.
  Indentation must use spaces (no tabs); the `---` fences must
  be on their own lines.
- **`library(tidyverse)` not run in the setup chunk.** Then the
  first chunk that uses a tidyverse function errors. Make sure
  the setup chunk has `include: false` (runs but is hidden)
  and loads every package the document needs.
- **NA confusion.** Some learners conflate NA with 0 when they
  see a "missing value count of 0" — they think *"so there are
  zero values?"*. Walk through the distinction explicitly.
- *Excel-to-R bridge:* The most important sentence in this
  chapter is *"every move you used to do by looking at Excel is
  now a scripted R call"*. Refer back to it whenever a learner
  asks why a particular function exists.
:::


---

# Working with AI on this chapter {.unnumbered}

Use AI as a **verification and debugging partner**, not an answer machine.
The goal is not to make AI do the work — it is to use AI to help you
**understand, check, and improve your own work.**

::: {.callout-note collapse="true"}
## AI prompt to check your work
Fill in the brackets and paste this into your AI assistant. It should
*explain and check*, not hand you a final assignment answer:

> *"I am working on FRST 232, this chapter. My dataset is `[file name]`
> with columns `[columns]`. My task is `[what you are trying to do]`. I
> tried `[paste your code or steps]` and got `[paste the exact output or
> error]`. Explain what this does line by line, tell me whether it answers
> the question, and list the checks I should run to verify it. Do not give
> me the final answer."*
:::

**Verify it yourself** — before trusting any result (yours *or* an AI's):

- the rows, columns, and units are what you expect;
- the row count changes sensibly after a `filter()` or a join;
- group sizes and summary values are plausible;
- the figure answers the question and axis labels include units.

**Common AI mistakes to watch for:** inventing a column, function, or value
that is not in your data; choosing a tool before reading the question; and
(in Excel) suggesting the wrong cell range or forgetting to refresh a
PivotTable.

::: {.callout-tip collapse="true"}
## Compare your solution with the reference
When a reference solution is available, do **not** just copy it — use it to
check your own work:

- **Expected result** — does your output match the target shape and values?
- **Required components** — did you include every step the task asked for?
- **If your result differs** — say *where* it matched, *where* it differed,
  *what* caused the difference, *how* you fixed it, and *what* you learned.
:::

::: {.callout-tip collapse="true"}
## AI replication challenge
Write a prompt that would guide an AI *toward* the reference solution
without handing over the final answer. Include the dataset, the columns,
the task goal, the output format you need, and the verification checks.
Then paste (or summarise) the AI's response and answer: did its approach
match the book's? did it miss a verification step? did it invent a column,
function, or value? what did your group change after checking it?
:::
