---
title: "Clean and Prepare Forestry Data in R"
---

```{r setup-ch06}
#| include: false
library(tidyverse)
library(readr)
library(lubridate)
library(here)
library(knitr)
library(scales)
```

::: {.callout-note appearance="simple"}
## Data for this chapter
This chapter uses the **BC air-quality** file
([`airdata3.csv`](../data/airdata3.csv){download="airdata3.csv"}). Click
the file name to download it directly, then save it in your `data/`
folder. For the source and details, see the
[Datasets Used in This Book](../datasets.qmd) page.
:::

# Chapter goals {.unnumbered}

By the end of this chapter you will be able to:

- Recognize the **six core dplyr verbs**: `filter()`, `select()`,
  `mutate()`, `arrange()`, `group_by()`, `summarise()`.
- Drop unwanted columns with **`select()`** and drop unwanted
  rows with **`filter()`**.
- Add new columns or transform existing ones with **`mutate()`**.
- Sort tibbles with **`arrange()`** (ascending) and `desc()`
  (descending).
- Parse messy text dates into proper R dates with
  **`lubridate::mdy()`** and friends.
- Recode values into categories with **`case_when()`** (the R
  equivalent of `IFS` from Chapter 2).
- Look up reference values with **`left_join()`** (the R
  equivalent of `VLOOKUP`).
- Chain operations into a single readable pipeline with the
  native [pipe](../glossary.qmd#pipe) **`|>`**.
- Maintain a **cleaning log** inside a Quarto document — text and
  code side by side, every change documented.

# Expected output for this chapter {.unnumbered}

::: {.callout-tip appearance="default"}
## What you will hand in

1. A **cleaned tibble** of `airdata3.csv`, saved as
   `outputs/airdata3_clean.csv`, containing only the useful
   columns, with the date parsed properly, and missing-value
   rows removed where appropriate.
2. A **rendered Quarto HTML report** (`frst232_ch06_learner.html`)
   showing the full cleaning sequence and a cleaning log.
3. A **station-level summary table** of mean O₃ by
   monitoring station, sorted in descending order.
4. A short **group-lab worksheet** (or screenshot) submitted to
   the course site.

This list is the same as the rubric for the chapter assignment.
Keep it in view while you work.
:::

# Why dplyr

In Chapter 2 you cleaned data in Excel using AutoFilter, sorts,
helper columns with `IF`/`IFS`, `VLOOKUP`, and `SUMIFS`. Every
move was visible — but every move was also a click that left no
record. *"How did you get this number?"* months later was hard to
answer.

**dplyr** is the tidyverse package for data cleaning in R. Its
verbs do exactly what your Excel moves did, with three changes:

1. Every move is a **written instruction** that can be re-run.
2. Every move is **named after the operation** (`filter` not
   "AutoFilter"; `mutate` not "add helper column").
3. Moves **chain together** with the pipe `|>`, so a multi-step
   cleaning is one block of code that reads top to bottom.

# Excel-to-dplyr bridge

This is the cleaning bridge that Chapter 2 previewed and Chapter
5 introduced. Now you write the right-hand column for real.

| What you did in Excel (Ch 2) | What you do in R (Ch 6) |
|---|---|
| AutoFilter to hide rows | `filter(data, condition)` |
| Sort A→Z | `arrange(data, column)` |
| Sort Z→A | `arrange(data, desc(column))` |
| Delete unwanted columns | `select(data, -unwanted_col)` |
| Helper column with `=A2+B2` | `mutate(data, new_col = A + B)` |
| `IFS(x<1000, "small", ...)` | `case_when(x < 1000 ~ "small", ...)` |
| `VLOOKUP(...)` | `left_join(data, reference, by = "key")` |
| PivotTable (sum by group) | `group_by(col) |> summarise(sum = sum(x))` |
| Cleaning log on a separate sheet | Markdown prose between code chunks |

**Every cell of the right-hand column is one line of code.**
After this chapter, you will be able to clean a dataset in R as
fluently as you cleaned one in Excel — with the bonus that the
work is now reproducible.

# The dataset for this chapter

This chapter uses one openly licensed file from the **BC
Government — Environmental Reporting BC** program:

```{r dataset-summary}
#| echo: false
tibble(
  Property = c("File", "Source", "Catalogue", "Licence", "Rows",
               "Columns", "Coverage", "Time period"),
  Value = c(
    "`airdata3.csv`",
    "BC Government — Environmental Reporting BC, BC Air Data",
    "<https://envistaweb.env.gov.bc.ca/>",
    "Open Government Licence — British Columbia",
    "11,904",
    "27 (most of them empty)",
    "16 air-quality monitoring stations across Metro Vancouver",
    "January 2000 (one month of hourly readings)"
  )
) |> kable()
```

## Why this dataset matters for forestry

Forest health and air quality are tightly linked. **Ground-level
ozone (O₃)** is one of the most damaging air pollutants for
vegetation: it enters leaves and needles through the stomata,
injures foliage, and reduces photosynthesis and growth over a
season. Unlike a pollutant that is emitted directly, ozone *forms*
when sunlight reacts with pollutants from vehicles, industry, and
— during fire season — wildfire emissions. The 16 stations in this
file are part of the network BC uses to monitor air quality across
the Lower Mainland.

A working analyst at a BC ministry, regional district, or
consultancy will use exactly this format — hourly readings from
many stations — to assess pollutant exposure. In this chapter we
focus on **O₃**, which is recorded at eight of these stations, so
that later chapters can compare readings across stations and
regions.

::: {.callout-important}
## The file is intentionally messy

Unlike the BC silviculture and Ontario forest stats files (which
were pre-cleaned by their publishers), `airdata3.csv` arrives in
a state much closer to **real-world raw data**:

- Several columns are **entirely missing** (every row is `NA`).
  We need to detect and drop these.
- Two **redundant index columns** (`"Unnamed: 0"` and similar)
  were added by an upstream tool — we drop these too.
- **Dates are stored as text** in `M/D/YYYY` format
  (`"1/1/2000"`). R does not recognise these as dates until we
  parse them.
- **Many rows have no O₃ measurement** because not every
  station measures every pollutant.

Real forestry data looks like this. The cleaning work in this
chapter is the cleaning work you will do all the time as a
working analyst.
:::

# Importing the file

In your `frst232/` project (from Chapter 4 or 5), the file should
live at `data/airdata3.csv`. If you do not have it yet, download
it from the BC Government link above and place it there.

```{r import}
airdata_raw <- read_csv(here("data", "airdata3.csv"),
                        show_col_types = FALSE)
dim(airdata_raw)
```

Notice the variable name: **`airdata_raw`**. We never modify this
in place. Every cleaning step creates a new variable. The raw
tibble stays untouched in memory in case we need to start over.

# Inspect first — the six-move check from Chapter 5

```{r inspect-airdata}
dim(airdata_raw)
glimpse(airdata_raw)
colSums(is.na(airdata_raw)) |> sort(decreasing = TRUE) |> head(15)
```

What we see:

- 11,904 rows × 27 columns.
- Many `<dbl>` columns. Some `<chr>` columns (`location`, `Date`,
  `Time`).
- **Several columns are 100% missing** — the count of NAs equals
  the total row count.

The 100%-missing columns are an artefact of how this dataset is
distributed: the schema reserves room for variables that *some*
stations might measure, but the stations in this extract do not
all measure them. Our first cleaning move is to drop them.

# Verb 1 — `select()` (keep or drop columns)

`select()` chooses which columns to keep. The simplest form is to
list the columns you want:

```{r select-keep}
airdata_raw |>
  select(location, Date, Time, O3, PM10, NO2, SO2, CO) |>
  head(3)
```

The pipe `|>` reads: *"take `airdata_raw`, then pass it into
`select()`, then take just the first 3 rows"*.

## Dropping columns with `-`

To drop named columns, prefix them with `-`:

```{r select-drop}
# Drop the redundant index columns added by an upstream tool.
airdata_step1 <- airdata_raw |>
  select(-`...1`, -starts_with("Unnamed"))
ncol(airdata_raw)        # columns before
ncol(airdata_step1)      # columns after — a few fewer
names(airdata_step1)     # and here is exactly what remains
```

`starts_with("Unnamed")` is a **selection helper** — a pattern
match that grabs every column whose name starts with `"Unnamed"`.
Other helpers: `ends_with()`, `contains()`, `matches()` (regex),
and `where()` (predicate).

## Dropping all-NA columns programmatically

We saw above that several columns are entirely missing. We can
drop them in one move with `where()`:

```{r drop-all-na}
airdata_step2 <- airdata_step1 |>
  select(where(\(col) !all(is.na(col))))
ncol(airdata_step2)
names(airdata_step2)
```

Read it: *"keep columns where it is **not** the case that all
values are NA"*. We just removed twelve dead-weight columns with
one line.

# Verb 2 — `filter()` (keep or drop rows)

`filter()` keeps rows where a condition is `TRUE`:

```{r filter-o3}
# Keep only rows where O₃ was actually measured.
airdata_step3 <- airdata_step2 |>
  filter(!is.na(O3))
nrow(airdata_step3)
```

`!is.na(O3)` reads *"O3 is NOT missing"*. The `!` is the
logical NOT operator.

## Combining conditions

You can combine conditions with `&` (AND), `|` (OR), or by adding
multiple conditions separated by commas (which dplyr treats as
AND):

```{r filter-multi}
# Keep only Burnaby South rows with measured O₃
burnaby_o3 <- airdata_step3 |>
  filter(location == "Burnaby South",
         !is.na(O3))
nrow(burnaby_o3)
```

Note the **double-equals `==`** for "is equal to" — single `=`
would be an assignment.

## Common filter patterns

| You want | Filter expression |
|---|---|
| One specific value | `filter(col == "Burnaby South")` |
| Not equal | `filter(col != "Burnaby South")` |
| One of several | `filter(col %in% c("A", "B", "C"))` |
| Numeric threshold | `filter(O3 > 10)` |
| Between two values | `filter(O3 >= 10, O3 < 20)` |
| Non-missing | `filter(!is.na(O3))` |
| Date range | `filter(date >= "2000-01-15")` |

The right-hand column reads naturally in English. *"Filter the
data where location is in 'A', 'B', or 'C'."* That's the dplyr
design.

# Verb 3 — `mutate()` (add or change columns)

`mutate()` adds new columns or replaces existing ones. The most
common use in this chapter is **parsing the text date column into
a real R date**.

```{r mutate-date}
airdata_step4 <- airdata_step3 |>
  mutate(date_parsed = mdy(Date))
glimpse(airdata_step4 |> select(Date, date_parsed))
```

`mdy()` from the **lubridate** package reads a text date in
*month/day/year* order and returns a proper `<date>` column.
Other parsers:

| Source format | lubridate function |
|---|---|
| `"1/15/2000"` (US) | `mdy()` |
| `"15/1/2000"` (UK / Canadian common) | `dmy()` |
| `"2000-01-15"` (ISO 8601) | `ymd()` |
| `"15-Jan-2000"` | `dmy()` (handles abbreviations) |

::: {.callout-warning}
## Date ambiguity

A string like `"3/4/2000"` could be March 4 or April 3 depending
on convention. **lubridate trusts the function name** — call
`mdy()` and it interprets as March 4; call `dmy()` and it
interprets as April 3.

Always check: pick one row, look at the original `Date` string,
and confirm the parsed date is what you expected. A wrong date
parser is a silent bug that ruins every later analysis.
:::

## Multiple new columns at once

```{r mutate-many}
airdata_step5 <- airdata_step4 |>
  mutate(date_parsed = mdy(Date),
         year = year(date_parsed),
         month = month(date_parsed),
         day = day(date_parsed),
         o3_kg_per_m3 = O3 / 1000)
glimpse(airdata_step5 |> select(Date, date_parsed, year, month, day, o3_kg_per_m3))
```

`year()`, `month()`, `day()` are lubridate accessors. Each
returns a vector of the same length.

# Verb 4 — `arrange()` (sort rows)

`arrange()` sorts the tibble:

```{r arrange-asc}
airdata_step5 |>
  arrange(O3) |>
  select(location, date_parsed, O3) |>
  head(5)
```

Ascending by default. For descending, wrap in `desc()`:

```{r arrange-desc}
airdata_step5 |>
  arrange(desc(O3)) |>
  select(location, date_parsed, O3) |>
  head(5)
```

The top of the list is the highest-O₃ hourly readings —
useful for spotting the most polluted moments.

## Multi-key sort

```{r arrange-multi}
airdata_step5 |>
  arrange(location, desc(O3)) |>
  select(location, date_parsed, O3) |>
  head(8)
```

Sorts by `location` (alphabetical), then within each location by
`O3` descending. The equivalent in Excel was **Data → Sort →
Add Level**.

# Verb 5 — `case_when()` (the R `IFS`)

`case_when()` recodes values into categories based on rules. It
is the direct R equivalent of Excel's `IFS`.

```{r case-when}
airdata_step6 <- airdata_step5 |>
  mutate(o3_band = case_when(
    is.na(O3)    ~ "missing",
    O3 < 10      ~ "low",
    O3 < 20      ~ "moderate",
    O3 < 30      ~ "high",
    O3 >= 30     ~ "very high",
    TRUE           ~ "other"
  ))
airdata_step6 |> count(o3_band)
```

Read it line by line: *"if O3 is missing, return 'missing';
otherwise if it is less than 10, return 'low'; otherwise less
than 20, 'moderate'; ..."*. The first match wins. The trailing
`TRUE ~ "other"` is the catch-all, like the `TRUE` in Excel's
`IFS`.

::: {.callout-tip}
## Why `case_when` beats nested `if_else`

You could write the same logic with nested `if_else()`:

```r
if_else(O3 < 10, "low",
  if_else(O3 < 20, "moderate",
    if_else(O3 < 30, "high", "very high")))
```

It works, but it nests deeply and is hard to maintain. `case_when`
is flat — one row per rule — and easy to extend.
:::

# Verb 6 — `summarise()` overall, `group_by()` for groups

`summarise()` collapses a tibble to **one row** of summary
statistics:

```{r summarise-overall}
airdata_step6 |>
  summarise(
    n_obs = n(),
    mean_o3 = mean(O3, na.rm = TRUE),
    median_o3 = median(O3, na.rm = TRUE),
    max_o3 = max(O3, na.rm = TRUE)
  )
```

The function `n()` (with no arguments, only used inside
`summarise()` or `mutate()`) returns the row count.

Add `group_by()` *before* `summarise()` to get one row per
group:

```{r group-by}
airdata_step6 |>
  group_by(location) |>
  summarise(
    n_obs = n(),
    n_with_o3 = sum(!is.na(O3)),
    mean_o3 = mean(O3, na.rm = TRUE)
  ) |>
  arrange(desc(mean_o3))
```

This is the **PivotTable equivalent** from Chapter 3. Same
operation. Same answer. Now scriptable.

::: {.callout-tip}
## Always remember `na.rm = TRUE`

When summarising a column that *might* have missing values, add
`na.rm = TRUE` to every aggregation function: `mean()`, `sum()`,
`min()`, `max()`, `median()`, `sd()`. Forgetting it returns `NA`
for any group with even one missing value, which is rarely what
you want.
:::

## `count()` — the most common summarise

`count(col)` is a shortcut for `group_by(col) |> summarise(n =
n())`:

```{r count}
airdata_step6 |>
  count(location, sort = TRUE)
```

`sort = TRUE` orders the result descending by count. Useful for
spotting which categories dominate.

# Verb 7 — `left_join()` (the R `VLOOKUP`)

A real-world cleaning task often needs a **lookup**: take a code
column, attach a description from a reference table. `left_join`
does this without the silent failures of `VLOOKUP`.

Suppose we have a small reference tibble mapping stations to
regions:

```{r station-regions}
station_regions <- tribble(
  ~location,                          ~region,
  "Burnaby South",                    "Burrard Peninsula",
  "Coquitlam Douglas College",        "North-East Sector",
  "Langley Central",                  "Fraser Valley West",
  "Maple Ridge Golden Ears School",   "North-East Sector",
  "New Westminster Sapperton Park",   "Burrard Peninsula",
  "North Delta",                      "South of Fraser",
  "North Vancouver Second Narrows",   "North Shore",
  "Pitt Meadows Airport",             "North-East Sector",
  "Port Coquitlam North",             "North-East Sector",
  "Port Moody Rocky Point Park",      "North-East Sector"
)
```

Attach the `region` column to the data:

```{r left-join}
airdata_step7 <- airdata_step6 |>
  left_join(station_regions, by = "location")
airdata_step7 |>
  select(location, region, date_parsed, O3) |>
  head(5)
```

Every row gets its region. The `by = "location"` argument names
the column to match on. Stations not in the reference table get
`region = NA` (and the join does *not* drop them — that is what
`left` means).

::: {.callout-tip}
## `left_join` vs `inner_join` vs `right_join`

| Join | Keeps |
|---|---|
| `left_join(x, y)` | **All rows of `x`**, plus matched rows from `y` (NA otherwise) |
| `inner_join(x, y)` | Only rows that match in both |
| `right_join(x, y)` | All rows of `y`, plus matched rows from `x` |
| `full_join(x, y)` | All rows from both, NA where no match |

For lookups, `left_join` is the default — it keeps every row in
your main data, even if the lookup misses.
:::

# Putting it all together — the full cleaning pipeline

Now the entire chapter in one pipeline:

```{r full-pipeline}
airdata_clean <- read_csv(here("data", "airdata3.csv"),
                          show_col_types = FALSE) |>
  # Drop redundant index columns
  select(-starts_with("Unnamed"), -any_of("...1")) |>
  # Drop columns that are 100% missing
  select(where(\(col) !all(is.na(col)))) |>
  # Parse the date and add date parts
  mutate(date_parsed = mdy(Date),
         year = year(date_parsed),
         month = month(date_parsed),
         day = day(date_parsed)) |>
  # Keep only rows with a measured O₃
  filter(!is.na(O3)) |>
  # Classify O₃ into bands
  mutate(o3_band = case_when(
    O3 < 10  ~ "low",
    O3 < 20  ~ "moderate",
    O3 < 30  ~ "high",
    TRUE       ~ "very high"
  )) |>
  # Attach region info
  left_join(station_regions, by = "location") |>
  # Final sort
  arrange(date_parsed, location)

dim(airdata_clean)
glimpse(airdata_clean)
```

**One pipeline.** Eight steps. Reads top-to-bottom in English.
Reproducible. Documented. This is what cleaning data in R looks
like once you are fluent.

## Save the cleaned tibble

```{r write-csv}
#| eval: false
write_csv(airdata_clean,
          here("outputs", "airdata3_clean.csv"))
```

The cleaned file is your **Expected output** for the chapter
assignment. The raw `data/airdata3.csv` stays untouched.

# The cleaning log

In Excel you kept a cleaning log on a separate sheet. In Quarto
you keep it as **Markdown text between the code chunks**.

For this chapter, the log might look like:

````markdown
## Cleaning log

| Date | Step | What I did | Why |
|---|---|---|---|
| 2026-09-22 | 1 | Dropped 2 redundant index columns (`Unnamed: 0`, etc.) | Added by an upstream pandas export; no meaning. |
| 2026-09-22 | 2 | Dropped 12 columns that were 100% missing | No measurements at any station for this month. |
| 2026-09-22 | 3 | Parsed `Date` from text to `<date>` using `lubridate::mdy()` | Original was `"M/D/YYYY"` text. |
| 2026-09-22 | 4 | Added `year`, `month`, `day` columns | Convenience for later grouping. |
| 2026-09-22 | 5 | Filtered to rows where O3 was measured | Reduced rows from 11,904 to 742. |
| 2026-09-22 | 6 | Recoded O3 into 4 bands using `case_when` | For the air-quality summary table. |
| 2026-09-22 | 7 | Left-joined a station→region reference table | To enable regional summaries. |
| 2026-09-22 | 8 | Arranged by date then location | Final sort. |
````

That table goes in your `.qmd` directly above the cleaning
pipeline. Anyone reading the rendered report sees what you did,
why, and the code that did it.

# Active learning

## Activity 1 — Find the dirty columns

Run `colSums(is.na(airdata_raw)) |> sort(decreasing = TRUE)`.
How many columns are 100% missing? How many are partially
missing?

## Activity 2 — Filter and count

How many rows did the file have before filtering for non-missing
O3? How many after? What fraction of rows did you keep?

## Activity 3 — Try a different case_when

Replace the four-band O3 classification with a three-band
version (`good` / `moderate` / `unhealthy`). Re-run the pipeline.
How does the count of each category change?

## Activity 4 — Add a new station to the lookup

Add one row to `station_regions` for a station you noticed in the
data but is missing from the reference. Re-run the `left_join`.
Confirm the new station now has a region.

## Activity 5 — Save and re-import

Save your cleaned tibble to `outputs/airdata3_clean.csv`. Then in
a fresh R session, import it with `read_csv()`. Confirm the date
column is still `<date>` (not `<chr>`). If it is, your earlier
parsing held.

# Practice Demo Lab

::: {.callout-important}
## Practice demo only
This is a **guided practice lab in the OER book**, designed to help you
learn the workflow before completing the official lab assignment on
**Canvas**. The Canvas lab is the graded assignment. Use this activity to
practice, compare your work with the reference solution, and learn how to
write an AI verification prompt.
:::

This demo lab has **six parts** — work them in order:

1. **Your attempt.** Work through the tasks below.
2. **Reference solution.** Compare against the **worked examples earlier in
   this chapter** (your public reference). The graded Canvas lab has its own
   private answer key — do not copy demo solutions into a Canvas submission.
3. **Compare your work with the reference.** Where did it match? Where did
   it differ? What caused the difference, and how did you fix it?
4. **Write your own AI verification prompt.** Ask an AI to *check* your
   reasoning, code, and outputs — not to produce the answer for you.
5. **Model AI verification prompt.**

   > *"I am doing a FRST 232 practice demo lab. My dataset is `[file name]`
   > (columns `[columns]`). I attempted `[paste your steps or code]` and got
   > `[paste the output]`. Explain what my work does line by line, tell me
   > whether it answers the task, and list the checks I can run to verify it
   > against the chapter's worked example. Do not give me the final answer."*

6. **Reflection: what changed after checking?** In two or three sentences —
   did your attempt hold up? what did you fix? what will you check first
   next time?


::: {.callout-note appearance="default"}
## Lab: Clean the air-quality file together

Work in groups of 3 or 4. Each member should be inside their own
`frst232/` RStudio Project. The file `airdata3.csv` must be in
each member's `data/` folder.

**Task 1 — Inspect together.** Each member runs the six-move
inspection (`dim`, `names`, `glimpse`, `summary`,
`colSums(is.na(...))`). Compare what you see. How many columns
have any data? How many are 100% missing?

**Task 2 — Drop redundant columns.** As a group, decide which
columns to drop. Write the `select()` call together.

**Task 3 — Parse the date.** Try `mdy(Date)`. Did it work for
every row, or did some fail? (Check with `summary(parsed_date)`
— any `NA` count from parsing.)

**Task 4 — Filter and group.** Write a `filter()` + `group_by()`
+ `summarise()` pipeline that returns mean O₃ per station.
**Sort descending** by mean O₃.

**Task 5 — Discuss the result.** Which station has the highest
mean O₃ in this month? Does that make sense given what the
group members know about Metro Vancouver geography?

**Task 6 — Submit.** Each group submits:
- the final pipeline code (paste into a Quarto chunk in your
  shared learner file),
- the station-ranked summary table,
- a one-sentence interpretation of the result.

This lab fits comfortably inside a single hands-on session.
:::

# Exercises

These are the take-home exercises for this chapter.

1. **Read and inspect.** Import `airdata3.csv`. Report the row
   count, column count, and the number of all-NA columns.

2. **Clean step by step.** Build the pipeline yourself, one step
   at a time, in separate chunks. After each step, print
   `nrow()` and `ncol()`. The final cleaned tibble should have
   742 rows (rows with O₃ measured).

3. **Per-station summary.** Group by `location` and compute
   `n_obs`, `mean_o3`, `median_o3`, `max_o3`. Sort
   descending by `mean_o3`.

4. **Per-day summary.** Group by your parsed `date_parsed` column
   and compute mean O₃ across all stations per day. Which day
   in January 2000 had the highest provincial mean O₃?

5. **Add a category.** Use `case_when()` to add a column
   `aq_alert` that is `"alert"` if O₃ ≥ 25 and `"normal"`
   otherwise. How many `"alert"` hours are in the file?

6. **Lookup challenge.** Build a small `station_lat_lon` tibble
   with at least 5 stations and made-up latitude/longitude pairs.
   `left_join` it onto your cleaned tibble. Confirm the join did
   not drop any rows.

7. **Cleaning log.** Write a cleaning log Markdown table in your
   `.qmd`, with one row per step you did in Exercise 2.

8. **Optional, harder.** Use `pivot_wider()` to reshape your
   per-station summary so each station is a column. (We will
   meet `pivot_wider` formally in Chapter 8 — try it now if you
   want to look ahead.)

# Optional, advanced

## Renaming columns

`rename()` lets you give a column a new name:

```r
airdata_step1 |>
  rename(station = location,
         o3_ugm3 = O3)
```

## `if_else()` for binary recodes

When you only need two outcomes, `if_else()` is briefer than
`case_when()`:

```r
mutate(aq_alert = if_else(O3 >= 25, "alert", "normal"))
```

## `across()` for applying a function to many columns

If you want the mean of *every* numeric column grouped by
station, you can use `across()`:

```r
airdata_clean |>
  group_by(location) |>
  summarise(across(where(is.numeric), \(x) mean(x, na.rm = TRUE)))
```

This is one of the most powerful patterns in dplyr — but also
one of the most confusing. We will use it sparingly in this
course.

# A glimpse ahead: From cleaning to summarizing

In Chapter 7 we go deeper into `summarise()`. The verbs are
already familiar; what we add is:

- Multiple summary statistics in one call.
- `group_by()` with two or more grouping variables.
- `count()` and `tally()` shortcuts.
- Reshaping summary output with `pivot_wider()` for cross-tabs.
- The connection to the Excel **PivotTable** from Chapter 3 —
  same answer, now scriptable.

```{r ch7-preview}
#| eval: false
airdata_clean |>
  group_by(region, o3_band) |>
  summarise(n = n(), .groups = "drop") |>
  pivot_wider(names_from = o3_band, values_from = n, values_fill = 0)
```

That's a cross-tabulation: rows = region, columns = O₃ band,
values = count. Equivalent to a PivotTable with `region` in
Rows, `o3_band` in Columns, and `Count of n_obs` in Values —
but as one line of dplyr.

Chapter 7 walks through this in detail.

# Self-assessment quiz {#sec-quiz-ch06}

::: {.content-visible when-format="pdf"}
*The self-assessment quiz is interactive: it appears in the online (HTML)
book, and you can also take it on Canvas. This PDF does not display the
interactive quiz questions.*
:::

::: {.content-visible when-format="html"}
Click the answer you think is correct. The right answer turns
green; wrong answers turn red and the correct one is revealed.
Explanations appear below each question.
:::

```{=html}
<style>
.quiz { background: #F4F7F4; border-left: 4px solid #2D6A4F; padding: 1em 1.5em; margin: 1.5em 0; border-radius: 6px; }
.quiz .q { font-weight: 600; color: #1B4332; margin-bottom: 0.6em; }
.quiz ul.qopts { list-style: none; padding: 0; margin: 0; }
.quiz ul.qopts li { padding: 0.55em 0.9em; margin: 0.3em 0; background: white; border: 1px solid #d8e4d8; border-radius: 4px; cursor: pointer; transition: background 0.12s; }
.quiz ul.qopts li:hover { background: #E8F1E9; }
.quiz ul.qopts li.correct { background: #C8E6C9; border-color: #2D6A4F; font-weight: 600; }
.quiz ul.qopts li.wrong { background: #FFCDD2; border-color: #B05C50; }
.quiz .explanation { display: none; margin-top: 0.7em; padding: 0.7em 1em; background: #FFF8E8; border-left: 3px solid #B7791F; border-radius: 4px; font-size: 0.95em; }
.quiz .explanation.show { display: block; }
.quiz textarea { width: 100%; min-height: 4.5em; padding: 0.6em; border: 1px solid #d8e4d8; border-radius: 4px; font-family: inherit; font-size: 0.95em; resize: vertical; margin-top: 0.3em; }
.quiz .reveal-btn { background: #2D6A4F; color: white; border: none; padding: 0.45em 0.9em; margin-top: 0.5em; border-radius: 4px; cursor: pointer; font-size: 0.9em; }
.quiz .reveal-btn:hover { background: #1B4332; }
.quiz .sample { display: none; margin-top: 0.6em; padding: 0.6em 0.9em; background: #FFF8E8; border-left: 3px solid #B7791F; border-radius: 4px; font-size: 0.95em; font-style: italic; }
.quiz .sample.show { display: block; }
</style>

<div class="quiz" id="q1" data-hints="Which verb subsets ROWS by a condition?||Not select — that is for columns.||filter().">
  <div class="q">1. Which dplyr verb keeps rows where a condition is TRUE?</div>
  <ul class="qopts">
    <li data-correct="false">select()</li>
    <li data-correct="true">filter()</li>
    <li data-correct="false">mutate()</li>
    <li data-correct="false">arrange()</li>
  </ul>
  <div class="explanation"><code>filter()</code> works on rows. <code>select()</code> works on columns. Easy to confuse — name them out loud when you write them.</div>
</div>

<div class="quiz" id="q2" data-hints="Which verb subsets COLUMNS by name?||Not filter — that is for rows.||select().">
  <div class="q">2. Which dplyr verb keeps or drops columns by name?</div>
  <ul class="qopts">
    <li data-correct="true">select()</li>
    <li data-correct="false">filter()</li>
    <li data-correct="false">mutate()</li>
    <li data-correct="false">group_by()</li>
  </ul>
  <div class="explanation"><code>select()</code> takes column names (or selection helpers like <code>starts_with()</code>) and returns a tibble with only those columns.</div>
</div>

<div class="quiz" id="q3" data-hints="The minus sign in select() means remove.||It keeps everything except that column.||It drops the Date column.">
  <div class="q">3. <code>airdata_raw |> select(-Date)</code> does what?</div>
  <ul class="qopts">
    <li data-correct="false">Selects only the Date column</li>
    <li data-correct="true">Drops the Date column, keeps everything else</li>
    <li data-correct="false">Sorts by Date</li>
    <li data-correct="false">Filters where Date is missing</li>
  </ul>
  <div class="explanation">The minus sign means "drop this column". <code>select()</code> can drop one column or many.</div>
</div>

<div class="quiz" id="q4" data-hints="desc() reverses the default ascending order.||Largest values come to the top.||Sorts by O3 from highest to lowest.">
  <div class="q">4. <code>arrange(desc(O3))</code> sorts the tibble —</div>
  <ul class="qopts">
    <li data-correct="false">Ascending by O3</li>
    <li data-correct="true">Descending by O3 (largest first)</li>
    <li data-correct="false">Alphabetically</li>
    <li data-correct="false">By row number</li>
  </ul>
  <div class="explanation"><code>arrange()</code> is ascending by default. Wrap in <code>desc()</code> for descending.</div>
</div>

<div class="quiz" id="q5" data-hints="Which verb creates or transforms a column?||Not summarise — that collapses rows.||mutate().">
  <div class="q">5. Which dplyr verb adds a new column or transforms an existing one?</div>
  <ul class="qopts">
    <li data-correct="false">filter()</li>
    <li data-correct="false">select()</li>
    <li data-correct="true">mutate()</li>
    <li data-correct="false">summarise()</li>
  </ul>
  <div class="explanation"><code>mutate()</code> is the column-creation/transformation verb. <code>mutate(date_parsed = mdy(Date))</code> adds a parsed date column.</div>
</div>

<div class="quiz" id="q6" data-hints="The text is month, then day, then year.||lubridate names its parsers by that order.||mdy().">
  <div class="q">6. The lubridate function that parses <code>"1/15/2000"</code> (month/day/year text) into a real R date is —</div>
  <ul class="qopts">
    <li data-correct="false">ymd()</li>
    <li data-correct="true">mdy()</li>
    <li data-correct="false">dmy()</li>
    <li data-correct="false">parse_date()</li>
  </ul>
  <div class="explanation"><code>mdy()</code> reads month-first. <code>dmy()</code> reads day-first. <code>ymd()</code> reads ISO 8601 (year-month-day).</div>
</div>

<div class="quiz" id="q7" data-hints="case_when recodes through a series of conditions.||Excel's multi-condition recoder.||IFS (or a nested IF).">
  <div class="q">7. <code>case_when()</code> in dplyr is the R equivalent of which Excel function?</div>
  <ul class="qopts">
    <li data-correct="false">VLOOKUP</li>
    <li data-correct="false">SUMIF</li>
    <li data-correct="true">IFS (or nested IF)</li>
    <li data-correct="false">COUNTA</li>
  </ul>
  <div class="explanation"><code>case_when()</code> walks down a list of conditions, returning the value for the first matching one. The trailing <code>TRUE ~ ...</code> is the catch-all, matching Excel's <code>TRUE</code> in <code>IFS</code>.</div>
</div>

<div class="quiz" id="q8" data-hints="Pulling matching values from a reference table by a key — which Excel function?||It looks a value up in another table.||VLOOKUP (or XLOOKUP).">
  <div class="q">8. <code>left_join()</code> in dplyr is the R equivalent of which Excel function?</div>
  <ul class="qopts">
    <li data-correct="false">SUMIFS</li>
    <li data-correct="true">VLOOKUP (or XLOOKUP)</li>
    <li data-correct="false">IFS</li>
    <li data-correct="false">PIVOT</li>
  </ul>
  <div class="explanation">A left join attaches a column from a reference table to your main data based on a matching key — exactly what VLOOKUP does, with fewer silent failures.</div>
</div>

<div class="quiz" id="q9" data-hints="Aggregating a value within each category — which Excel tool does that?||It is the drag-and-drop summary tool.||A PivotTable.">
  <div class="q">9. <code>group_by(location) |> summarise(mean(O3, na.rm = TRUE))</code> is equivalent to which Excel feature?</div>
  <ul class="qopts">
    <li data-correct="false">AutoFilter</li>
    <li data-correct="false">VLOOKUP</li>
    <li data-correct="true">A PivotTable with location in Rows and Average of O3 in Values</li>
    <li data-correct="false">Conditional Formatting</li>
  </ul>
  <div class="explanation">Group-then-summarise is the conceptual ancestor of every PivotTable. Same answer, scriptable.</div>
</div>

<div class="quiz" id="q10" data-hints="What does mean() return if any value is NA?||na.rm drops the missing values before averaging.||Without it the result would be NA.">
  <div class="q">10. Why include <code>na.rm = TRUE</code> in <code>mean(O3, na.rm = TRUE)</code>?</div>
  <ul class="qopts">
    <li data-correct="false">It makes R faster</li>
    <li data-correct="true">Without it, the mean returns NA whenever any value in the column is NA</li>
    <li data-correct="false">It removes the O3 column</li>
    <li data-correct="false">It is required syntax</li>
  </ul>
  <div class="explanation">By default, R refuses to silently skip missing values. <code>na.rm = TRUE</code> tells the aggregation function to skip them. Forgetting it is the most common silent bug in summarise pipelines.</div>
</div>

<div class="quiz" id="q11" data-hints="The pipe feeds the left value into the next function.||It goes in as that function's first argument.||Inserts it as the first argument of the function on its right.">
  <div class="q">11. The native pipe <code>|></code> takes the thing on its left and —</div>
  <ul class="qopts">
    <li data-correct="false">deletes it</li>
    <li data-correct="true">passes it as the first argument of the function on its right</li>
    <li data-correct="false">divides it by the right side</li>
    <li data-correct="false">prints it</li>
  </ul>
  <div class="explanation"><code>x |> f()</code> is equivalent to <code>f(x)</code>. Chained pipes let a multi-step pipeline read top-to-bottom in English.</div>
</div>

<div class="quiz" id="q12" data-hints="Testing membership in a set has its own operator — cleaner than chained OR.||Read it as 'is in this set'.||location %in% c('A','B','C').">
  <div class="q">12. To filter rows where <code>location</code> is one of <code>"A"</code>, <code>"B"</code>, or <code>"C"</code>, the cleanest expression is —</div>
  <ul class="qopts">
    <li data-correct="false">filter(location == "A" | location == "B" | location == "C")</li>
    <li data-correct="true">filter(location %in% c("A", "B", "C"))</li>
    <li data-correct="false">filter(location = c("A", "B", "C"))</li>
    <li data-correct="false">filter(location == c("A", "B", "C"))</li>
  </ul>
  <div class="explanation"><code>%in%</code> is the membership operator. The first option works but is wordy. The last is a common bug — <code>==</code> does element-wise comparison and will not do what you expect.</div>
</div>

<div class="quiz" id="q13" data-hints="The exclamation mark means NOT; is.na flags missing values.||So you keep the values that are present.||Rows where O3 is not missing.">
  <div class="q">13. <code>airdata_raw |> filter(!is.na(O3))</code> keeps —</div>
  <ul class="qopts">
    <li data-correct="false">Only rows where O3 is missing</li>
    <li data-correct="true">Only rows where O3 is NOT missing</li>
    <li data-correct="false">All rows, unchanged</li>
    <li data-correct="false">An error</li>
  </ul>
  <div class="explanation"><code>is.na()</code> returns TRUE for missing values. The <code>!</code> negates it. So this keeps rows where O3 is non-missing.</div>
</div>

<div class="quiz" id="q14" data-hints="count tallies rows per value; sort orders the result.||It is group_by then summarise(n()) then arrange(desc()).||A frequency table of locations, most common first.">
  <div class="q">14. <code>count(location, sort = TRUE)</code> is equivalent to —</div>
  <ul class="qopts">
    <li data-correct="false">filter(location == ...)</li>
    <li data-correct="true">group_by(location) |> summarise(n = n()) |> arrange(desc(n))</li>
    <li data-correct="false">select(location)</li>
    <li data-correct="false">mutate(location_count = ...)</li>
  </ul>
  <div class="explanation"><code>count()</code> is the dplyr shortcut for group-then-tally. <code>sort = TRUE</code> orders the result descending by count.</div>
</div>

<div class="quiz" id="q15" data-hints="You want to keep a column only if it is not entirely missing.||select() can take a where() test applied to each column.||select(where()) keeping columns that are not all NA.">
  <div class="q">15. Which is the cleanest way to drop every column that is 100% missing?</div>
  <ul class="qopts">
    <li data-correct="false">Manually list every empty column in select()</li>
    <li data-correct="true">select(where(\(col) !all(is.na(col))))</li>
    <li data-correct="false">filter(!is.na(.))</li>
    <li data-correct="false">drop_na()</li>
  </ul>
  <div class="explanation"><code>where()</code> applies a predicate to each column. The lambda <code>\(col) !all(is.na(col))</code> reads "keep the column if it is not the case that all of its values are NA". <code>drop_na()</code> drops rows, not columns.</div>
</div>

<div class="quiz" id="q16" data-hints="Check the file's source note.||BC government open data, an air-quality monitoring network.||BC Ministry of Environment open data (BC Data Catalogue).">
  <div class="q">16. Where does the BC air-quality file come from?</div>
  <ul class="qopts">
    <li data-correct="false">A textbook</li>
    <li data-correct="true">The BC Government, Environmental Reporting BC — the same publisher as the silviculture file</li>
    <li data-correct="false">A research paper</li>
    <li data-correct="false">A private consulting firm</li>
  </ul>
  <div class="explanation">Both the BC silviculture (Ch 1) and the BC air-quality (Ch 6) files come from Environmental Reporting BC, published under the Open Government Licence — British Columbia.</div>
</div>

<div class="quiz" id="q17" data-hints="A clean dataset should be saved separately from the raw file.||Write it out so later chapters can reuse it.||write_csv() it to outputs/.">
  <div class="q">17. After you build a cleaning pipeline, what should you do with the cleaned tibble?</div>
  <ul class="qopts">
    <li data-correct="false">Overwrite the raw CSV in data/</li>
    <li data-correct="true">Save it to outputs/ with a clear name (e.g., airdata3_clean.csv) using write_csv()</li>
    <li data-correct="false">Email it to yourself</li>
    <li data-correct="false">Delete the raw file</li>
  </ul>
  <div class="explanation">Same convention as Excel: raw stays in <code>data/</code>, cleaned versions go in <code>outputs/</code>.</div>
</div>

<div class="quiz" id="q18" data-hints="A pipe reads left to right, each step feeding the next.||The first verb runs first.||In order: filter, then mutate, then arrange.">
  <div class="q">18. In a pipe like <code>x |> filter(...) |> mutate(...) |> arrange(...)</code>, in what order do the operations run?</div>
  <ul class="qopts">
    <li data-correct="false">arrange first, then mutate, then filter</li>
    <li data-correct="true">filter first, then mutate, then arrange — top to bottom, left to right</li>
    <li data-correct="false">All three run simultaneously</li>
    <li data-correct="false">In alphabetical order</li>
  </ul>
  <div class="explanation">Pipes are evaluated left to right. The output of each step becomes the input of the next. Reading a pipe is like reading a sentence.</div>
</div>

<div class="quiz" id="q19" data-hints="SUMIFS totals values meeting conditions; here, one station and O3 above 25.||In dplyr: filter to those rows, then summarise a sum.||filter(location=='Burnaby South', O3>25) then summarise(sum(O3)).">
  <div class="q">19. The Excel function <code>=SUMIFS(O3_col, location_col, "Burnaby South", O3_col, "&gt;25")</code> has which dplyr equivalent?</div>
  <ul class="qopts">
    <li data-correct="false">select(location, O3)</li>
    <li data-correct="true">filter(location == "Burnaby South", O3 > 25) |> summarise(total = sum(O3))</li>
    <li data-correct="false">mutate(SUMIFS = ...)</li>
    <li data-correct="false">arrange(O3)</li>
  </ul>
  <div class="explanation">SUMIFS = sum the values subject to multiple criteria. In dplyr: filter to the criteria, then summarise the sum. Same answer.</div>
</div>

<div class="quiz" id="q20">
  <div class="q">20. <strong>Reflection.</strong> Of the six core dplyr verbs you met, which felt most natural coming from Excel? Which felt most foreign?</div>
  <textarea placeholder="Type your answer here..."></textarea>
  <button class="reveal-btn" onclick="this.nextElementSibling.classList.toggle('show')">Show common answers</button>
  <div class="sample">Common reflections: <em>filter felt natural — it is just AutoFilter typed out</em> · <em>select was easy once I learned the minus-sign for dropping</em> · <em>case_when felt foreign at first but turned out cleaner than nested IF</em> · <em>left_join is way clearer than VLOOKUP — no silent first-match bug</em> · <em>group_by + summarise was the most powerful new tool — one line replaces a whole PivotTable</em>.</div>
</div>

<script>
(function() {
  document.querySelectorAll('.quiz').forEach(function(q) {
    q.querySelectorAll('ul.qopts li').forEach(function(opt) {
      opt.addEventListener('click', function() {
        if (q.dataset.answered) return;
        q.dataset.answered = '1';
        var isCorrect = opt.getAttribute('data-correct') === 'true';
        if (isCorrect) opt.classList.add('correct');
        else {
          opt.classList.add('wrong');
          q.querySelectorAll('ul.qopts li').forEach(function(o) {
            if (o.getAttribute('data-correct') === 'true') o.classList.add('correct');
          });
        }
        var exp = q.querySelector('.explanation');
        if (exp) exp.classList.add('show');
      });
    });
  });
})();
</script>
```

# AI as a debugging companion

::: {.callout-tip}
## Useful prompts for this chapter

- *"In R, my `case_when()` returns NA for some rows even though I
  have a TRUE catch-all at the end. What might be wrong?"*

- *"Explain step by step what this pipeline does:
  `data |> filter(!is.na(O3)) |> group_by(location) |>
  summarise(mean_o3 = mean(O3))`"*

- *"I parsed a date with `mdy()` but the result is `NA` for some
  rows. The original strings look like '1/15/2000'. What are the
  most common reasons mdy() returns NA?"*

## A cleaning prompt template

When you ask AI for cleaning help, include four things:

1. The output of `glimpse(your_tibble)`.
2. The cleaning step you are trying to do (in plain English).
3. The dplyr code you tried.
4. The output (or error) you got.

Without these four, AI's suggestion will be too generic to use
directly.

## When *not* to use AI

- Do not let AI write your cleaning log. The log records *your*
  decisions about *your* data.
- Do not let AI invent column names that are not in your data
  (a common hallucination). Cross-check every suggested column
  name against `names(your_tibble)`.
:::

# Reading

- *R for Data Science* — chapter 4 *Data transformation*, the
  canonical dplyr reference. Free at <https://r4ds.hadley.nz/>.
- The dplyr cheatsheet (PDF): <https://rstudio.github.io/cheatsheets/data-transformation.pdf>
- The lubridate cheatsheet (PDF): <https://rstudio.github.io/cheatsheets/lubridate.pdf>
- BC Air Data background:
  <https://www2.gov.bc.ca/gov/content/environment/air-land-water/air/air-quality/air-data>

# Instructor notes {.unnumbered}

::: {.callout-note appearance="minimal" collapse="true"}
## Pacing and emphasis

- The chapter has **seven new R concepts** (the six core verbs
  plus `case_when` / `left_join`). Plan to teach the verbs in
  two clusters: row-and-column (filter, select, arrange) in one
  session, transform-and-summarize (mutate, case_when,
  group_by + summarise, left_join) in the next.
- The **Excel-to-dplyr bridge** table at the top of the chapter
  is the most important content. Refer to it whenever a learner
  asks why R has a particular verb.
- The **cleaning log** habit is the chapter's most transferable
  skill. Insist learners write one for every cleaning task,
  starting with this chapter.

## Expected output checklist (matches the top-of-chapter callout)

1. `outputs/airdata3_clean.csv` (saved tibble).
2. `frst232_ch06_learner.html` (rendered Quarto report).
3. Station-ranked summary table.
4. Group-lab worksheet or screenshot.

## Materials provided alongside this chapter

| File | Purpose |
|---|---|
| `airdata3.csv` | The raw BC air-quality file. Place in `data/`. |
| `frst232_ch06_learner.qmd` | Quarto starter with TODO chunks for each verb. |
| `frst232_ch06_solutions.qmd` | Completed solutions. **Do not distribute to learners.** |
| `quiz_ch06.html` | Standalone interactive quiz. |

## Common stumbling points

- **Confusing `filter` with `select`.** Walk through "rows vs
  columns" explicitly. *"`filter` chooses some of the rows.
  `select` chooses some of the columns."*
- **`==` vs `=`.** Learners write `filter(location = "Burnaby")`
  and get an error. Insist on `==` for comparison.
- **`%in%` confusion.** Some learners try
  `filter(location == c("A", "B"))` and get unexpected results
  (it does element-wise comparison). Teach `%in%` explicitly.
- **Date parsing failures.** Calling `mdy()` on a column that's
  in `dmy()` format silently returns wrong dates. Walk through
  the format-checking habit.
- **Forgetting `na.rm = TRUE`.** Resulting in summaries full of
  `NA`. Emphasise that for any column that might be missing,
  add `na.rm = TRUE` to every aggregation.
- *Excel-to-R bridge:* When a learner asks how to do an Excel
  cleaning move, point them to the bridge table at the top of
  this chapter. Almost every Excel cleaning question has a
  one-line dplyr answer.
:::
