---
title: "Summarize Forestry Data Overall and by Group"
---

```{r setup-ch07}
#| include: false
library(tidyverse)
library(lubridate)
library(here)
library(knitr)
library(scales)

# Reuse the cleaning pipeline from Chapter 6
airdata_raw <- read_csv(here("data", "airdata3.csv"),
                        show_col_types = FALSE)

station_regions <- tribble(
  ~location,                          ~region,
  "Burnaby South",                    "Burrard Peninsula",
  "Coquitlam Douglas College",        "North-East Sector",
  "Langley Central",                  "Fraser Valley West",
  "Maple Ridge Golden Ears School",   "North-East Sector",
  "New Westminster Sapperton Park",   "Burrard Peninsula",
  "North Delta",                      "South of Fraser",
  "North Vancouver Second Narrows",   "North Shore",
  "Pitt Meadows Airport",             "North-East Sector",
  "Port Coquitlam North",             "North-East Sector",
  "Port Moody Rocky Point Park",      "North-East Sector"
)

airdata_clean <- airdata_raw |>
  select(-starts_with("Unnamed"), -any_of("...1")) |>
  select(where(\(col) !all(is.na(col)))) |>
  mutate(date_parsed = mdy(Date),
         year = year(date_parsed),
         month = month(date_parsed),
         day = day(date_parsed)) |>
  filter(!is.na(PM25)) |>
  mutate(pm25_band = case_when(
    PM25 < 10  ~ "low",
    PM25 < 25  ~ "moderate",
    PM25 < 50  ~ "high",
    TRUE       ~ "very high"
  )) |>
  left_join(station_regions, by = "location")
```

# Chapter goals {.unnumbered}

By the end of this chapter you will be able to:

- Write an **overall summary** of one or more columns with
  `summarise()` — counts, means, medians, standard deviations,
  ranges.
- Write a **by-group summary** by adding `group_by()` before
  `summarise()`.
- Group by **multiple columns** at once (e.g., region × day) to
  build cross-tabulations.
- Use the **shortcut functions** `count()`, `tally()`, and
  `n_distinct()` for common patterns.
- Reshape a summary with **`pivot_wider()`** so each group level
  becomes a column — the closest R has to a PivotTable.
- Recognize that a `group_by() |> summarise()` pipeline is the
  conceptual equivalent of an Excel **PivotTable** from Chapter 3
  — and produce identical numbers.
- Use **`across()`** to apply the same summary to many columns
  at once.
- Write summary tables that read cleanly in a rendered Quarto
  report.

# Expected output for this chapter {.unnumbered}

::: {.callout-tip appearance="default"}
## What you will hand in

1. A **rendered Quarto HTML report**
   (`frst232_ch07_learner.html`) containing:
   - an **overall summary** of PM2.5 across the whole cleaned
     air-quality file,
   - a **by-station** summary table (one row per station),
   - a **by-region × by-day** summary table reshaped with
     `pivot_wider()` so each region is a column,
   - one short prose paragraph interpreting which station and
     which day had the highest mean PM2.5.
2. The corresponding **`.qmd` source file**.
3. A short **group-lab worksheet** (or screenshot) submitted to
   the course site.

This list is the same as the rubric for the chapter assignment.
Keep it in view while you work.
:::

# Where we are

In Chapter 6 you cleaned `airdata3.csv` with the six dplyr verbs.
The cleaned tibble (`airdata_clean`) is the starting point for
this chapter. Chapter 7 picks up where Chapter 6 leaves off:
**now that the data is clean, how do you turn it into numbers
your reader can use?**

Chapter 6 introduced `group_by() |> summarise()` briefly. This
chapter is that pattern at full depth.

# Excel-to-R bridge — summarising

| What you did in Excel | What you do in R |
|---|---|
| `=AVERAGE(C2:C743)` over a column | `summarise(mean = mean(PM25, na.rm = TRUE))` |
| `=MEDIAN(C2:C743)` | `summarise(median = median(PM25, na.rm = TRUE))` |
| `=COUNT(C2:C743)` | `summarise(n_obs = n())` |
| `=SUMIF(loc, "Burnaby", pm25)` | `filter(location == "Burnaby") |> summarise(sum = sum(PM25))` |
| `=AVERAGEIF(loc, "Burnaby", pm25)` | `filter(...) |> summarise(mean = mean(PM25))` |
| `=SUMIFS(...)` multiple criteria | `filter(...) |> summarise(...)` with multiple conditions |
| PivotTable (rows = station) | `group_by(location) |> summarise(...)` |
| PivotTable (rows × columns) | `group_by(region, day) |> summarise(...) |> pivot_wider(...)` |
| PivotTable count by category | `count(category)` |
| PivotTable refresh after data change | Re-render the document |

The right-hand column is the rest of this chapter. **Every Excel
summary move from Chapters 2 and 3 has a one-line dplyr
equivalent.**

# The starting point — the cleaned tibble

For the rest of this chapter, assume `airdata_clean` is already
in memory (built by the Chapter 6 pipeline).

```{r preview-clean}
glimpse(airdata_clean)
```

`r nrow(airdata_clean)` rows × `r ncol(airdata_clean)` columns,
all with a measured PM2.5 value, parsed dates, a PM2.5 band, and
a region.

# Overall summary — one row

The simplest case: collapse the whole tibble to one row of
summary statistics:

```{r overall-summary}
airdata_clean |>
  summarise(
    n_obs       = n(),
    mean_pm25   = mean(PM25, na.rm = TRUE),
    median_pm25 = median(PM25, na.rm = TRUE),
    sd_pm25     = sd(PM25, na.rm = TRUE),
    min_pm25    = min(PM25, na.rm = TRUE),
    max_pm25    = max(PM25, na.rm = TRUE)
  )
```

Every column in the result is a single number. The arguments to
`summarise()` are named: `<name> = <expression>`.

## The functions you will use most

| Function | Returns | Notes |
|---|---|---|
| `n()` | row count | No arguments. Only inside `summarise()` / `mutate()`. |
| `sum()` | total | Add `na.rm = TRUE` if missing values are possible. |
| `mean()` | arithmetic mean | Same. |
| `median()` | middle value | Robust to outliers. |
| `min()` / `max()` | extremes | Sensitive to outliers. |
| `sd()` | standard deviation | Spread around the mean. |
| `n_distinct()` | unique values | Useful as `n_distinct(location)`. |
| `quantile()` | any percentile | `quantile(PM25, 0.9)` for the 90th. |
| `IQR()` | interquartile range | Robust spread. |

# By-group summary — one row per group

Add `group_by()` *before* `summarise()`:

```{r by-station}
station_summary <- airdata_clean |>
  group_by(location) |>
  summarise(
    n_obs       = n(),
    mean_pm25   = mean(PM25, na.rm = TRUE),
    median_pm25 = median(PM25, na.rm = TRUE),
    max_pm25    = max(PM25, na.rm = TRUE),
    .groups = "drop"
  ) |>
  arrange(desc(mean_pm25))

station_summary
```

This is the dplyr equivalent of an Excel PivotTable with
`location` in Rows and `Average of PM25` in Values. Same
operation. Same answer.

::: {.callout-tip}
## Always add `.groups = "drop"` (or your code will warn)

dplyr keeps the grouping after `summarise()` by default for
backward compatibility, which causes confusing behaviour in
downstream pipelines. Explicitly say `.groups = "drop"` to
clear the grouping after summarise. Treat it as a habit, not a
choice.
:::

## Multi-key grouping

Group by two or more variables at once. Each unique combination
becomes one row:

```{r region-day}
region_day_summary <- airdata_clean |>
  group_by(region, day) |>
  summarise(
    mean_pm25 = mean(PM25, na.rm = TRUE),
    .groups = "drop"
  )

region_day_summary |> head(10)
```

The result is in **long format**: one row per (region, day)
combination. Useful for plotting (Chapter 9) but hard to scan.

# Reshaping with `pivot_wider()` — the PivotTable equivalent

For human-readable cross-tabulations, reshape the long-format
summary into a wide table:

```{r pivot-wider}
region_day_summary |>
  pivot_wider(
    names_from  = region,
    values_from = mean_pm25
  ) |>
  arrange(day)
```

What we just built is a true cross-tab: one row per day, one
column per region, values = mean PM2.5. **This is what an Excel
PivotTable produces** when you put `day` in Rows, `region` in
Columns, and `Average of PM25` in Values.

## `pivot_wider()` arguments

| Argument | What it does |
|---|---|
| `names_from` | the column whose values become new column names |
| `values_from` | the column whose values fill the cells |
| `values_fill` | what to put in cells where no data exists (default: NA; often `0` for counts) |

::: {.callout-tip}
## When to use long vs wide

- **Long format** (one row per combination) is what
  `summarise()` and `ggplot2` prefer. Use it for analysis.
- **Wide format** (one column per group) is what humans prefer
  to read. Use it for tables in a report.

Convert between them with `pivot_longer()` and `pivot_wider()`.
The two are inverses.
:::

# Shortcut: `count()` and `tally()`

The two most common summarise patterns:

```{r count-shortcut}
airdata_clean |> count(region, sort = TRUE)
airdata_clean |> count(region, pm25_band)
```

`count(x)` is exactly `group_by(x) |> summarise(n = n())`. Use
it for tallies.

`count(x, y)` cross-tabulates. To pivot it wide:

```{r count-pivot}
airdata_clean |>
  count(region, pm25_band) |>
  pivot_wider(names_from = pm25_band, values_from = n, values_fill = 0)
```

That's a complete cross-tab in three lines. The Excel equivalent
was a PivotTable with `region` in Rows, `pm25_band` in Columns,
`Count of n` in Values, plus the "Show items with no data" toggle
to fill empty cells with 0. Three clicks in Excel; three lines in
R; same answer.

# `n_distinct()` — unique values

```{r n-distinct}
airdata_clean |>
  summarise(
    n_stations = n_distinct(location),
    n_regions  = n_distinct(region),
    n_days     = n_distinct(day)
  )
```

Useful when you want *"how many unique stations report PM2.5?"*
in one line.

# Multiple statistics with `across()`

When you want the same summary (e.g., mean) on **several
columns**, `across()` is the cleaner pattern:

```{r across-mean}
airdata_clean |>
  group_by(region) |>
  summarise(
    across(c(PM25, PM10, O3, NO2),
           \(x) mean(x, na.rm = TRUE)),
    .groups = "drop"
  )
```

This summarises four pollutant columns at once, one mean per
column per region. Far less repetitive than writing four
`mean(...)` calls manually.

## Multiple statistics on the same column

You can also use `across()` to apply multiple statistics to one
column:

```{r across-multi-stat}
airdata_clean |>
  group_by(region) |>
  summarise(
    across(PM25,
           list(mean = \(x) mean(x, na.rm = TRUE),
                sd   = \(x) sd(x, na.rm = TRUE),
                max  = \(x) max(x, na.rm = TRUE))),
    .groups = "drop"
  )
```

The result has columns `PM25_mean`, `PM25_sd`, `PM25_max`.

::: {.callout-tip}
## `across()` is advanced — use it when it helps

For two or three columns, plain `summarise(mean1 = mean(x),
mean2 = mean(y), ...)` is more readable. For five or more,
`across()` shines. Use whichever is clearer for your reader.
:::

# Putting it together — a worked summary report

Here is what a complete summary section of a forestry data
report might look like.

## Overall

```{r report-overall}
overall <- airdata_clean |>
  summarise(
    n_obs       = n(),
    n_stations  = n_distinct(location),
    n_regions   = n_distinct(region),
    mean_pm25   = round(mean(PM25, na.rm = TRUE), 1),
    median_pm25 = round(median(PM25, na.rm = TRUE), 1),
    max_pm25    = round(max(PM25, na.rm = TRUE), 1)
  )
overall |> kable()
```

## By station

```{r report-by-station}
airdata_clean |>
  group_by(location) |>
  summarise(
    n_obs     = n(),
    mean_pm25 = round(mean(PM25, na.rm = TRUE), 1),
    max_pm25  = max(PM25, na.rm = TRUE),
    .groups = "drop"
  ) |>
  arrange(desc(mean_pm25)) |>
  kable()
```

## By region × day (wide)

```{r report-region-day}
airdata_clean |>
  group_by(region, day) |>
  summarise(mean_pm25 = round(mean(PM25, na.rm = TRUE), 1),
            .groups = "drop") |>
  pivot_wider(names_from = region, values_from = mean_pm25,
              values_fill = NA) |>
  arrange(day) |>
  head(10) |>
  kable()
```

## By PM2.5 band — count cross-tab

```{r report-band}
airdata_clean |>
  count(region, pm25_band) |>
  pivot_wider(names_from = pm25_band, values_from = n, values_fill = 0) |>
  kable()
```

That entire summary section — four tables — is **fifteen lines
of dplyr code** and renders to a polished HTML report. The same
output in Excel would be three or four PivotTables, each requiring
a manual refresh after every data update. **The dplyr version
updates automatically when you re-render.**

# Active learning

## Activity 1 — Overall summary

Write a single `summarise()` call that returns the count, mean,
median, sd, min, and max of PM2.5 across the whole cleaned
tibble. Compare with the table in the chapter.

## Activity 2 — By-station ranking

Group by `location`, summarise `mean_pm25`, and arrange
descending. Which station has the highest mean PM2.5? Which has
the lowest?

## Activity 3 — `count()` shortcut

Use `count(region)` and `count(region, sort = TRUE)`. What is
the difference?

## Activity 4 — Cross-tabulation

Build a cross-tab where rows are `region`, columns are
`pm25_band`, and cells are counts. Use `pivot_wider()` with
`values_fill = 0` so empty cells show `0` instead of `NA`.

## Activity 5 — `across()` for many means

Use `across(c(PM25, PM10, O3, NO2), \(x) mean(x, na.rm = TRUE))`
inside a `group_by(region) |> summarise(...)` pipeline. How does
the output differ from writing four separate mean calls?

# Group lab activity

::: {.callout-note appearance="default"}
## Lab: Produce a summary report together

Work in groups of 3 or 4. Each member should already have
`airdata_clean` in memory (from Chapter 6, or by re-running the
Chapter 6 pipeline at the top of their `frst232_ch07_learner.qmd`
setup chunk).

**Task 1 — Agree on the question.** As a group, pick **one**
forestry question your summary report will answer:

- Which BC region had the highest mean PM2.5 in January 2000?
- Which day in January 2000 was the worst air-quality day, on
  average, across all stations?
- Which monitoring station had the most "very high" PM2.5
  hours?

**Task 2 — Build the summary.** Write a single dplyr pipeline
that answers your question. Use `group_by()`, `summarise()`,
and `arrange()` as needed.

**Task 3 — Reshape if useful.** If your summary has two grouping
variables, use `pivot_wider()` to produce a cross-tab readable
in a single screen. If it has one grouping variable, sort it.

**Task 4 — Caption it.** Write a one-sentence caption above
your table. The caption should state the **headline finding**
(*"Region X had the highest mean PM2.5..."*).

**Task 5 — Submit.** Submit:
- the dplyr pipeline (in a Quarto chunk),
- the rendered table,
- the one-sentence caption,
- all group members' names.

This lab fits comfortably inside a single hands-on session.
:::

# Exercises

These are the take-home exercises for this chapter.

1. **Overall summary.** Write a single `summarise()` call that
   returns `n_obs`, `n_stations`, `mean_pm25`, `median_pm25`,
   `sd_pm25`, and `max_pm25` for the cleaned air-quality data.

2. **By region.** Group by `region`, summarise mean PM2.5,
   arrange descending. Which region has the highest mean? Does
   the result make sense given Metro Vancouver geography?

3. **By day.** Group by `day` (the day-of-month integer), then
   summarise `n_obs` and `mean_pm25` per day. Which day in
   January 2000 had the highest provincial mean PM2.5?

4. **Cross-tabulation.** Build the region × pm25_band cross-tab
   from Activity 4. Use `pivot_wider()` with `values_fill = 0`.
   Submit the rendered table.

5. **Top station per region.** Group by `region`, then within
   each region find the station with the highest mean PM2.5.
   Hint: `group_by(region) |> slice_max(mean_pm25, n = 1)` after
   you have already computed per-station means.

6. **`across()` for multiple pollutants.** Group by `region`
   and compute the mean of every pollutant column (`PM25`,
   `PM10`, `O3`, `NO2`, `SO2`, `CO`). Use `across()`.

7. **PivotTable parity.** Save your cleaned tibble to
   `outputs/airdata3_clean.csv`. Open it in Excel. Build a
   PivotTable with `region` in Rows, `pm25_band` in Columns,
   `Count of PM25` in Values. **Confirm the numbers match your
   dplyr cross-tab to the last unit.**

8. **Optional, harder.** Compute the 90th percentile of PM2.5
   per station: `summarise(p90 = quantile(PM25, 0.9, na.rm =
   TRUE))`. Which station has the highest 90th-percentile PM2.5?

# Optional, advanced

## `slice_max()` and `slice_min()` — top-N per group

```r
airdata_clean |>
  group_by(location) |>
  slice_max(PM25, n = 3, with_ties = FALSE)
```

Returns the three highest-PM2.5 hours at each station.

## `summarise()` with custom functions

You can define a helper function and call it inside
`summarise()`:

```r
robust_mean <- function(x) mean(x, na.rm = TRUE, trim = 0.05)

airdata_clean |>
  group_by(region) |>
  summarise(rmean_pm25 = robust_mean(PM25), .groups = "drop")
```

The 5% trimmed mean drops the top 5% and bottom 5% before
averaging — more robust to outliers than a plain mean.

## `summary()` of a tibble vs `summarise()` of a tibble

These two are easily confused:

| Function | Returns |
|---|---|
| `summary(tibble)` | A base-R text dump: five-number summary per column |
| `summarise(tibble, ...)` | A new tibble with the columns you asked for |

Use `summary()` for quick interactive inspection. Use
`summarise()` for output that goes into a report.

# A glimpse of Chapter 8: From summarising to reshaping and combining

Chapter 7 used `pivot_wider()` for the first time. In Chapter 8
you will meet:

- `pivot_longer()` — the inverse of `pivot_wider()`. Useful when
  data arrives in wide format but you need long format for
  analysis.
- `inner_join()`, `right_join()`, `full_join()` — the other
  three join types, alongside `left_join()` from Chapter 6.
- `bind_rows()` and `bind_cols()` for stacking tibbles.
- The principles of **tidy data** — why one row per observation
  and one column per variable matter so much.

The verbs in Chapter 8 are the *last* dplyr verbs you need for
this course. After Chapter 8, every cleaning, summarising,
reshaping, and joining task you face has a one-line dplyr
answer.

# Self-assessment quiz {#sec-quiz-ch07}

Click the answer you think is correct. The right answer turns
green; wrong answers turn red and the correct one is revealed.
Explanations appear below each question.

```{=html}
<style>
.quiz { background: #F4F7F4; border-left: 4px solid #2D6A4F; padding: 1em 1.5em; margin: 1.5em 0; border-radius: 6px; }
.quiz .q { font-weight: 600; color: #1B4332; margin-bottom: 0.6em; }
.quiz ul.qopts { list-style: none; padding: 0; margin: 0; }
.quiz ul.qopts li { padding: 0.55em 0.9em; margin: 0.3em 0; background: white; border: 1px solid #d8e4d8; border-radius: 4px; cursor: pointer; transition: background 0.12s; }
.quiz ul.qopts li:hover { background: #E8F1E9; }
.quiz ul.qopts li.correct { background: #C8E6C9; border-color: #2D6A4F; font-weight: 600; }
.quiz ul.qopts li.wrong { background: #FFCDD2; border-color: #B05C50; }
.quiz .explanation { display: none; margin-top: 0.7em; padding: 0.7em 1em; background: #FFF8E8; border-left: 3px solid #B7791F; border-radius: 4px; font-size: 0.95em; }
.quiz .explanation.show { display: block; }
.quiz textarea { width: 100%; min-height: 4.5em; padding: 0.6em; border: 1px solid #d8e4d8; border-radius: 4px; font-family: inherit; font-size: 0.95em; resize: vertical; margin-top: 0.3em; }
.quiz .reveal-btn { background: #2D6A4F; color: white; border: none; padding: 0.45em 0.9em; margin-top: 0.5em; border-radius: 4px; cursor: pointer; font-size: 0.9em; }
.quiz .reveal-btn:hover { background: #1B4332; }
.quiz .sample { display: none; margin-top: 0.6em; padding: 0.6em 0.9em; background: #FFF8E8; border-left: 3px solid #B7791F; border-radius: 4px; font-size: 0.95em; font-style: italic; }
.quiz .sample.show { display: block; }
</style>

<div class="quiz" id="q1">
  <div class="q">1. <code>summarise()</code> with no <code>group_by()</code> before it returns —</div>
  <ul class="qopts">
    <li data-correct="false">One row per row of the original tibble</li>
    <li data-correct="true">One row of overall summary statistics</li>
    <li data-correct="false">An error</li>
    <li data-correct="false">A chart</li>
  </ul>
  <div class="explanation"><code>summarise()</code> collapses the tibble to one row. With <code>group_by()</code> first, it collapses to one row per group.</div>
</div>

<div class="quiz" id="q2">
  <div class="q">2. <code>group_by(location) |> summarise(mean = mean(PM25, na.rm = TRUE))</code> returns —</div>
  <ul class="qopts">
    <li data-correct="false">One row total</li>
    <li data-correct="true">One row per unique location</li>
    <li data-correct="false">A chart</li>
    <li data-correct="false">An error</li>
  </ul>
  <div class="explanation">Each unique value of the grouping column becomes one row of the summary.</div>
</div>

<div class="quiz" id="q3">
  <div class="q">3. Inside <code>summarise()</code>, what does <code>n()</code> return?</div>
  <ul class="qopts">
    <li data-correct="false">The number of columns</li>
    <li data-correct="true">The row count of the group (or the whole tibble if ungrouped)</li>
    <li data-correct="false">The number of NAs</li>
    <li data-correct="false">The sum of the values</li>
  </ul>
  <div class="explanation"><code>n()</code> is a special dplyr function — no arguments, only used inside <code>summarise()</code> or <code>mutate()</code>. Returns the row count of the current group.</div>
</div>

<div class="quiz" id="q4">
  <div class="q">4. <code>count(region, pm25_band)</code> is equivalent to —</div>
  <ul class="qopts">
    <li data-correct="false">filter(region == ..., pm25_band == ...)</li>
    <li data-correct="true">group_by(region, pm25_band) |> summarise(n = n())</li>
    <li data-correct="false">select(region, pm25_band)</li>
    <li data-correct="false">arrange(region, pm25_band)</li>
  </ul>
  <div class="explanation"><code>count(...)</code> is the dplyr shortcut for group-then-tally. Multiple grouping variables produce a row per combination.</div>
</div>

<div class="quiz" id="q5">
  <div class="q">5. <code>pivot_wider(names_from = region, values_from = mean_pm25)</code> does what?</div>
  <ul class="qopts">
    <li data-correct="false">Drops the region column</li>
    <li data-correct="true">Turns each unique value of region into a new column, with mean_pm25 values inside</li>
    <li data-correct="false">Sorts by region</li>
    <li data-correct="false">Counts the regions</li>
  </ul>
  <div class="explanation">This produces a cross-tab. The original long-format summary becomes wide-format with one column per region.</div>
</div>

<div class="quiz" id="q6">
  <div class="q">6. The closest Excel equivalent of <code>group_by(region, day) |> summarise(mean(PM25)) |> pivot_wider(...)</code> is —</div>
  <ul class="qopts">
    <li data-correct="false">A bar chart</li>
    <li data-correct="false">An AutoFilter</li>
    <li data-correct="true">A PivotTable with day in Rows, region in Columns, and Average of PM25 in Values</li>
    <li data-correct="false">A VLOOKUP</li>
  </ul>
  <div class="explanation">The dplyr summarise + pivot_wider pipeline produces exactly what a PivotTable produces — same data, same answer, scriptable instead of clickable.</div>
</div>

<div class="quiz" id="q7">
  <div class="q">7. <code>summarise(p90 = quantile(PM25, 0.9, na.rm = TRUE))</code> returns —</div>
  <ul class="qopts">
    <li data-correct="false">The 9th percentile</li>
    <li data-correct="true">The 90th percentile of PM25 (the value below which 90% of observations fall)</li>
    <li data-correct="false">A 90-row summary</li>
    <li data-correct="false">An error</li>
  </ul>
  <div class="explanation"><code>quantile(x, p)</code> returns the value at the p-th percentile of x. 0.9 is the 90th percentile.</div>
</div>

<div class="quiz" id="q8">
  <div class="q">8. <code>n_distinct(location)</code> returns —</div>
  <ul class="qopts">
    <li data-correct="false">The total row count</li>
    <li data-correct="true">The number of unique values in the location column</li>
    <li data-correct="false">The first location</li>
    <li data-correct="false">An error</li>
  </ul>
  <div class="explanation"><code>n_distinct(x)</code> counts unique values. Useful for "how many unique stations are in this dataset?".</div>
</div>

<div class="quiz" id="q9">
  <div class="q">9. <code>.groups = "drop"</code> in <code>summarise()</code> —</div>
  <ul class="qopts">
    <li data-correct="false">Drops the grouped columns</li>
    <li data-correct="false">Drops the rows</li>
    <li data-correct="true">Clears the grouping after summarise, so downstream operations are not grouped</li>
    <li data-correct="false">Is mandatory or the code errors</li>
  </ul>
  <div class="explanation">dplyr keeps the grouping by default after summarise (for backward compatibility). <code>.groups = "drop"</code> clears it, which is usually what you want. Treat it as a habit.</div>
</div>

<div class="quiz" id="q10">
  <div class="q">10. <code>pivot_wider(values_fill = 0)</code> means —</div>
  <ul class="qopts">
    <li data-correct="false">Fill the entire table with zeros</li>
    <li data-correct="true">Where no observation existed for that combination, put 0 instead of NA</li>
    <li data-correct="false">Multiply everything by zero</li>
    <li data-correct="false">Round to zero decimals</li>
  </ul>
  <div class="explanation">When pivoting count data wider, cells with no matching observations default to NA. <code>values_fill = 0</code> replaces those NAs with 0, which is usually what you want for counts.</div>
</div>

<div class="quiz" id="q11">
  <div class="q">11. <code>across(c(PM25, PM10, O3), \(x) mean(x, na.rm = TRUE))</code> inside <code>summarise()</code> —</div>
  <ul class="qopts">
    <li data-correct="false">Drops three columns</li>
    <li data-correct="true">Applies the mean function to PM25, PM10, and O3 — one column per pollutant in the output</li>
    <li data-correct="false">Pivots the table</li>
    <li data-correct="false">Counts the three columns</li>
  </ul>
  <div class="explanation"><code>across()</code> applies the same function to multiple columns at once. Saves writing three separate <code>mean(...)</code> calls.</div>
</div>

<div class="quiz" id="q12">
  <div class="q">12. <code>filter(region == "Burrard Peninsula") |> summarise(mean = mean(PM25, na.rm = TRUE))</code> is the dplyr equivalent of which Excel formula?</div>
  <ul class="qopts">
    <li data-correct="false">=AVERAGE(PM25)</li>
    <li data-correct="true">=AVERAGEIF(region, "Burrard Peninsula", PM25)</li>
    <li data-correct="false">=SUMIFS(...)</li>
    <li data-correct="false">=VLOOKUP(...)</li>
  </ul>
  <div class="explanation"><code>AVERAGEIF</code> averages one column subject to a criterion. In dplyr: filter to the criterion, then summarise the mean.</div>
</div>

<div class="quiz" id="q13">
  <div class="q">13. A by-station summary returns 6 rows in the result. The cleaned tibble has 742 rows. That means —</div>
  <ul class="qopts">
    <li data-correct="false">736 rows were lost</li>
    <li data-correct="false">There is a bug</li>
    <li data-correct="true">There are 6 unique stations with measured PM2.5; the 742 rows were collapsed by group</li>
    <li data-correct="false">The summary is wrong</li>
  </ul>
  <div class="explanation">Summarise collapses the rows of each group into one row. 742 rows distributed across 6 stations → 6 summary rows. Nothing was "lost".</div>
</div>

<div class="quiz" id="q14">
  <div class="q">14. If you forget <code>na.rm = TRUE</code> in <code>summarise(mean = mean(PM25))</code> and the PM25 column has any NA values, what happens?</div>
  <ul class="qopts">
    <li data-correct="false">R uses 0 for missing values</li>
    <li data-correct="true">The mean returns NA</li>
    <li data-correct="false">R skips the NAs silently</li>
    <li data-correct="false">An error is thrown</li>
  </ul>
  <div class="explanation">R refuses to silently ignore NAs by default. Always add <code>na.rm = TRUE</code> to aggregations on columns that might contain NAs.</div>
</div>

<div class="quiz" id="q15">
  <div class="q">15. The result of <code>summarise()</code> is —</div>
  <ul class="qopts">
    <li data-correct="false">A vector</li>
    <li data-correct="true">A new tibble (smaller than the input)</li>
    <li data-correct="false">A list</li>
    <li data-correct="false">A chart</li>
  </ul>
  <div class="explanation"><code>summarise()</code> returns a tibble. This is why you can pipe its output into another verb, like <code>arrange()</code> or <code>pivot_wider()</code>.</div>
</div>

<div class="quiz" id="q16">
  <div class="q">16. To get the standard deviation in dplyr, you use —</div>
  <ul class="qopts">
    <li data-correct="false">stdev()</li>
    <li data-correct="true">sd()</li>
    <li data-correct="false">stddev()</li>
    <li data-correct="false">STDEV()</li>
  </ul>
  <div class="explanation"><code>sd()</code> is the base R standard-deviation function. Like all aggregations, add <code>na.rm = TRUE</code> if needed.</div>
</div>

<div class="quiz" id="q17">
  <div class="q">17. Which is the closer R analogue of an Excel PivotTable refresh after editing the underlying Data sheet?</div>
  <ul class="qopts">
    <li data-correct="false">Manually clicking each summarise() call</li>
    <li data-correct="true">Re-rendering the Quarto document, which re-runs every chunk in order</li>
    <li data-correct="false">There is no analogue</li>
    <li data-correct="false">install.packages()</li>
  </ul>
  <div class="explanation">Re-rendering re-runs the cleaning and the summarising. Numbers in the report are always in sync with the current data. No PivotTable refresh bug.</div>
</div>

<div class="quiz" id="q18">
  <div class="q">18. The dplyr pipeline <code>data |> count(category) |> pivot_wider(names_from = category, values_from = n, values_fill = 0)</code> produces —</div>
  <ul class="qopts">
    <li data-correct="false">A bar chart</li>
    <li data-correct="false">A long-format tibble</li>
    <li data-correct="true">A one-row wide tibble with one column per category, holding the count of each</li>
    <li data-correct="false">An error</li>
  </ul>
  <div class="explanation">Count produces a long-format table (one row per category). pivot_wider turns it into a single row with one column per category — the simplest cross-tab.</div>
</div>

<div class="quiz" id="q19">
  <div class="q">19. <code>group_by(region) |> summarise(across(where(is.numeric), \(x) mean(x, na.rm = TRUE)))</code> does what?</div>
  <ul class="qopts">
    <li data-correct="false">Drops every numeric column</li>
    <li data-correct="true">Returns one row per region with the mean of every numeric column</li>
    <li data-correct="false">Computes the global mean</li>
    <li data-correct="false">Renames the columns</li>
  </ul>
  <div class="explanation"><code>where(is.numeric)</code> is a column-selection predicate that picks every numeric column. <code>across()</code> applies the mean to each. Powerful but use sparingly.</div>
</div>

<div class="quiz" id="q20">
  <div class="q">20. <strong>Reflection.</strong> Now that you can produce summary tables in dplyr, when (if ever) would you still reach for an Excel PivotTable?</div>
  <textarea placeholder="Type your answer here..."></textarea>
  <button class="reveal-btn" onclick="this.nextElementSibling.classList.toggle('show')">Show common answers</button>
  <div class="sample">Common reflections: <em>for quick one-off exploration where I do not need to share the result</em> · <em>for sharing with non-R colleagues</em> · <em>when the deliverable is itself an Excel workbook</em> · <em>for the slicer interactivity — clicking buttons is sometimes the right answer</em> · <em>almost never anymore — once I am fluent in dplyr, the same logic is faster to type than to click</em>.</div>
</div>

<script>
(function() {
  document.querySelectorAll('.quiz').forEach(function(q) {
    q.querySelectorAll('ul.qopts li').forEach(function(opt) {
      opt.addEventListener('click', function() {
        if (q.dataset.answered) return;
        q.dataset.answered = '1';
        var isCorrect = opt.getAttribute('data-correct') === 'true';
        if (isCorrect) opt.classList.add('correct');
        else {
          opt.classList.add('wrong');
          q.querySelectorAll('ul.qopts li').forEach(function(o) {
            if (o.getAttribute('data-correct') === 'true') o.classList.add('correct');
          });
        }
        var exp = q.querySelector('.explanation');
        if (exp) exp.classList.add('show');
      });
    });
  });
})();
</script>
```

# AI as a debugging companion

::: {.callout-tip}
## Useful prompts for this chapter

- *"In R, my `group_by(region) |> summarise(mean(PM25))` returns
  NA for some groups, even though the column has values. Why?"*

- *"Show me how to convert a long-format summary tibble into a
  wide-format cross-tab with `pivot_wider()`. Use these column
  names: ..."*

- *"What is the difference between `group_by() |> summarise()`
  and `group_by() |> mutate()`?"* (Answer: summarise collapses
  to one row per group; mutate keeps every row and broadcasts
  the result across the group.)

## When *not* to use AI

- Do not let AI choose what to summarise. The choice of mean
  vs median vs 90th percentile is an analytical decision —
  yours to make.
- Do not let AI invent column names that are not in your tibble.
  Always cross-check with `names(your_tibble)`.
:::

# Reading

- *R for Data Science* — chapter 5 *Data tidying* (covers
  pivot_wider and pivot_longer), chapter 13 *Quantitative
  exploratory data analysis*. Free at <https://r4ds.hadley.nz/>.
- The dplyr cheatsheet (PDF): <https://rstudio.github.io/cheatsheets/data-transformation.pdf>
- The tidyr cheatsheet (PDF): <https://rstudio.github.io/cheatsheets/tidyr.pdf>

# Instructor notes {.unnumbered}

::: {.callout-note appearance="minimal" collapse="true"}
## Pacing and emphasis

- This chapter is shorter on new verbs than Chapter 6 — only
  `summarise()` (deeper), `pivot_wider()`, `count()`,
  `n_distinct()`, and `across()`. Spend the saved time on the
  **PivotTable parity exercise** (Exercise 7), which makes the
  Excel-to-R bridge concrete by computing the same number both
  ways.
- The **`.groups = "drop"`** habit is worth repeating. Without
  it, downstream pipelines silently inherit grouping and produce
  surprising results.
- **`across()`** is powerful but overwhelming for first-time R
  users. Mention it once, demo it once, then let learners use
  whichever pattern (plain summarise vs across) feels clearer.

## Expected output checklist (matches the top-of-chapter callout)

1. `frst232_ch07_learner.html` (rendered).
2. `frst232_ch07_learner.qmd` (source).
3. Group-lab worksheet or screenshot.

## Materials provided alongside this chapter

| File | Purpose |
|---|---|
| `airdata3.csv` | Same raw file from Ch 6. Already in `data/`. |
| `frst232_ch07_learner.qmd` | Quarto starter with TODO chunks. |
| `frst232_ch07_solutions.qmd` | Completed solutions. **Do not distribute.** |
| `quiz_ch07.html` | Standalone interactive quiz. |

## Sample numbers (reference)

| Quantity | Approximate value |
|---|---|
| Overall mean PM2.5 (cleaned tibble) | ~6 μg/m³ |
| Overall median PM2.5 | ~4 μg/m³ |
| Number of unique stations with PM2.5 | 6 |
| Number of unique regions in cleaned tibble | 4 |
| Highest single-hour PM2.5 reading | ~42 μg/m³ |

## Common stumbling points

- **Confusing `summarise()` with `summary()`.** Walk through
  the distinction explicitly in the lecture: one is a verb that
  returns a new tibble; the other is a base R function that
  prints a text summary.
- **Forgetting `na.rm = TRUE`.** Causes group-mean columns full
  of NA. The fix is one argument; the habit takes a few weeks
  to form.
- **`pivot_wider()` syntax confusion.** `names_from` vs
  `values_from` is easy to swap. The mnemonic: *names_from* is
  the column whose **names** become headers; *values_from* is
  the column whose **values** fill the cells.
- *Excel-to-R bridge:* Exercise 7 (PivotTable parity) is the
  most important exercise. Insist that learners actually open
  Excel and verify the numbers match. The visceral
  "they're identical" moment is the chapter's pedagogical
  payoff.
:::
