---
title: "Visualize Forestry Data with ggplot2"
---

```{r setup-ch09}
#| include: false
library(tidyverse)
library(readxl)
library(lubridate)
library(here)
library(knitr)
library(scales)
```

::: {.callout-note appearance="simple"}
### Data for this chapter
This chapter uses the **BC silviculture**
([`bc_disturbance_reforestation.xlsx`](../data/bc_disturbance_reforestation.xlsx){download="bc_disturbance_reforestation.xlsx"})
and **BC air-quality**
([`airdata3.csv`](../data/airdata3.csv){download="airdata3.csv"}) files.
Click each file name to download it directly, then save it in your `data/`
folder. For sources and details, see the
[Datasets Used in This Book](../datasets.qmd) page.
:::

## Chapter goals {.unnumbered}

By the end of this chapter you will be able to:

- Explain the **grammar of graphics**: data →
  [aesthetic mappings](../glossary.qmd#aesthetic-mapping)
  → geometric objects → coordinate system → theme.
- Build a **basic plot in three lines** using
  `ggplot(data, aes(...)) + geom_*()`.
- Use the **six most common geoms**: `geom_point()`,
  `geom_line()`, `geom_col()` / `geom_bar()`,
  `geom_histogram()`, `geom_boxplot()`, and `geom_smooth()`.
- Add **multiple layers** to a single plot and understand the
  order in which they draw.
- Use **`facet_wrap()`** for small multiples (one panel per
  category) and **`facet_grid()`** for two-dimensional grids.
- Polish a plot with **labels** (`labs()`), **scales**
  (`scale_*_continuous()`, `scale_*_date()`), **colours**
  (`scale_color_*()` / `scale_fill_*()`), and **themes**
  (`theme_minimal()`, `theme()` overrides).
- Save a plot to disk with **`ggsave()`** at a sensible size and
  resolution.
- Recognize when long-format data is needed (and when to call
  `pivot_longer()` before plotting).

## Expected output for this chapter {.unnumbered}

::: {.callout-tip appearance="default"}
### What you will hand in

1. A **rendered Quarto HTML report**
   (`frst232_ch09_learner.html`) containing at least:
   - One **chart** of BC reforestation by fiscal year — a bar or line
     chart; pick the type that best shows change over time.
   - One **line chart** with two or more series overlaid (e.g.,
     harvested vs natural disturbance vs reforestation across
     years).
   - One **histogram** of hourly O₃ readings.
   - One **box plot** of O₃ by region.
   - One **faceted** plot (small multiples) showing O₃ over
     the days of January 2000, faceted by station.
   - At least one chart with **custom labels** (`labs(title,
     subtitle, x, y, caption)`).
2. One chart saved to `outputs/` as a `.png` using `ggsave()`.
3. A short **group-lab worksheet** (or screenshot) submitted to
   the course site.
:::

## Where we are

Through Chapters 4–8 you imported, inspected, cleaned, summarised,
reshaped, and combined forestry data. Chapter 9 turns those
tibbles into **charts** — the most efficient way for a reader to
understand a pattern that would be invisible in a table.

A picture is also the most efficient way to **lie** with data,
which is why ggplot2's grammar makes every visual choice
explicit. Every axis, scale, colour, and label is in your code.
Choices that Excel hides in menu dialogs are written down where
your reader (and future you) can see them.

## Excel-chart-to-ggplot bridge

| Excel chart | ggplot2 geom + extras |
|---|---|
| Column / bar chart | `geom_col()` (values you supply) / `geom_bar()` (counts of categories) |
| Line chart | `geom_line()` |
| Scatter plot | `geom_point()` |
| Histogram | `geom_histogram()` |
| Box / whisker plot | `geom_boxplot()` |
| Area chart | `geom_area()` |
| Pie chart | `geom_col() + coord_polar()` — but **avoid** (pie charts are hard to read; use a bar chart) |
| Trend line / regression line | `geom_smooth(method = "lm")` |
| Stacked bar | `geom_col(position = "stack")` |
| Side-by-side (grouped) bar | `geom_col(position = "dodge")` |
| Multiple charts side by side (small multiples) | `facet_wrap(~ category)` |
| Manual axis-range setting | `scale_y_continuous(limits = c(0, 100))` |
| Chart title / subtitle / data label | `labs(title = "...", subtitle = "...", caption = "...")` |
| Saving as PNG | `ggsave("outputs/myplot.png", width = 7, height = 5)` |

## The grammar of graphics — three sentences

A ggplot is built from three pieces:

1. **Data** — a tibble.
2. **Aesthetic mappings** — which column maps to which visual
   property (x, y, color, fill, size, shape).
3. **Geometric objects** — the shapes that get drawn (points,
   lines, bars).

A ggplot call says, *"Take this tibble, map these columns to
these visual properties, and draw these shapes."*

```{r minimal-ggplot}
#| fig-cap: "BC reforested area (ha) by fiscal year, 1987–2023."
#| fig-alt: "Line chart of BC reforested area in hectares by fiscal year, 1987 to 2023, fluctuating from year to year."
bc <- read_excel(here("data", "bc_disturbance_reforestation.xlsx"),
                 sheet = "Data")

ggplot(data = bc, mapping = aes(x = Fiscal_Year, y = Reforestation_ha)) +
  geom_line()
```

Three things to notice:

- `ggplot()` opens the plot. It does not draw anything yet.
- `aes()` says *"map `Fiscal_Year` to the x-axis and
  `Reforestation_ha` to the y-axis"*.
- `geom_line()` draws lines connecting the (x, y) points.
- The `+` sign **adds layers** — not the pipe `|>`. This is
  ggplot2's one bit of unusual syntax. Read `+` as *"and also draw"*.

## `geom_point()` — scatter plot

```{r geom-point}
#| fig-cap: "BC annual harvested area versus natural-disturbance area (ha), one point per year."
#| fig-alt: "Scatter plot of BC annual harvested area against natural-disturbance area; most years cluster at low values with a few high natural-disturbance outliers."
ggplot(bc, aes(x = Harvested_ha, y = Natural_Disturbance_ha)) +
  geom_point()
```

Each year is one dot. No relationship is obvious at this scale,
but the chart immediately reveals a few **outlier years** (large
natural-disturbance values) — useful for spotting the 2017
wildfire spike.

### Mapping a third variable to colour

```{r point-colour}
#| fig-cap: "Harvested versus natural-disturbance area (ha), points coloured by fiscal year."
#| fig-alt: "Scatter plot of harvested area versus natural-disturbance area, points coloured by fiscal year along a continuous colour gradient."
ggplot(bc, aes(x = Harvested_ha,
               y = Natural_Disturbance_ha,
               colour = Fiscal_Year)) +
  geom_point(size = 3)
```

`colour = Fiscal_Year` adds a third dimension. A continuous
variable like year gets a gradient legend.

## `geom_line()` — time series

```{r line-multiple-series}
#| fig-cap: "Harvested, natural-disturbance, and reforested area (ha) by fiscal year."
#| fig-alt: "Line chart with three series — harvested, natural-disturbance, and reforested area in hectares — plotted against fiscal year, each drawn with its own colour and line type."
# First reshape to long format so each series is a row
bc_long <- bc |>
  select(Fiscal_Year, Harvested_ha,
         Natural_Disturbance_ha, Reforestation_ha) |>
  pivot_longer(cols = -Fiscal_Year,
               names_to = "series",
               values_to = "area_ha")

ggplot(bc_long, aes(x = Fiscal_Year, y = area_ha, colour = series,
                    linetype = series)) +
  geom_line(linewidth = 1)
```

**Why long format**: with one row per (year, series) combination,
ggplot can draw three lines in one call by mapping `colour =
series`. With wide format (three columns) you would need three
separate `geom_line()` calls — much less elegant.

**This is why Chapter 8 spent time on `pivot_longer()`.** For
plotting, long is almost always the right shape.

## `geom_col()` vs `geom_bar()`

::: {.callout-note collapse="true"}
### AI prompt to check your work
Use this **after** your own attempt. It should check your reasoning, not hand you the answer:

> *"I want a chart that shows `[change over time / a distribution / a comparison across groups]` from `[columns]`, and I picked `[geom_*]`. Check whether that geom is the clearest choice for my message, suggest a better alternative if there is one, and list the labels and units the figure needs. Do not produce the final chart code for me."*
:::

The two bar geoms are confused often. The rule:

- `geom_col()` — you give it `y` values to plot.
- `geom_bar()` — you give it `x` values and it **counts** them.

```{r col-vs-bar}
#| fig-cap: "BC reforested area (ha) by fiscal year (geom_col)."
#| fig-alt: "Bar chart of BC reforested area in hectares for each fiscal year, 1987 to 2023."
# geom_col — you supply the y value
ggplot(bc, aes(x = Fiscal_Year, y = Reforestation_ha)) +
  geom_col(fill = "#2D6A4F")
```

```{r geom-bar-count}
#| fig-cap: "Count of disturbance events by type in a small sample (geom_bar)."
#| fig-alt: "Bar chart counting disturbance events by type in a small sample, with fire the most frequent category."
# geom_bar — it counts for you
# Build a small tibble with one row per disturbance event
events <- tibble(
  year = c(2017, 2017, 2017, 2018, 2019, 2019, 2020),
  type = c("fire", "fire", "pest", "fire", "fire", "wind", "fire")
)
ggplot(events, aes(x = type)) +
  geom_bar(fill = "#2D6A4F")
```

`geom_bar` produced the count of each `type`. `geom_col` would
have required a numeric column to use as height.

::: {.callout-note}
### A fresh import for this chapter
The `airdata_clean` object below is a **new import made here** for
plotting — it is *not* the cleaned, region-joined tibble of the same name
from Chapter 6. The `region` column is added later, in the box-plot
example (via the station → region lookup). If you copy code that uses
`region`, run that `left_join()` first, or you will see
`object 'region' not found`.
:::

## `geom_histogram()` — distribution of one variable

```{r histogram}
#| fig-cap: "Distribution of hourly ozone (O₃, ppb)."
#| fig-alt: "Histogram of hourly ozone readings in parts per billion; the distribution is strongly right-skewed, with most hours low and a long tail of higher values."
airdata_clean <- read_csv(here("data", "airdata3.csv"),
                          show_col_types = FALSE) |>
  filter(!is.na(O3))

ggplot(airdata_clean, aes(x = O3)) +
  geom_histogram(binwidth = 2, fill = "#2D6A4F", colour = "white")
```

`binwidth = 2` puts each bar in a 2 ppb bucket. The histogram
reveals the **heavily right-skewed** O₃ distribution: most
hours are clean, a few are sharply elevated. This is the kind of
pattern that a summary table can hint at but only a chart makes
obvious.

## `geom_boxplot()` — distributions by group

```{r boxplot}
#| fig-cap: "Hourly ozone (O₃, ppb) by Metro Vancouver region."
#| fig-alt: "Box plots of hourly ozone in parts per billion for each Metro Vancouver region, comparing medians, interquartile ranges, and outliers across regions."
# First add a region column (from the Ch 6 lookup)
station_regions <- tribble(
  ~location,                          ~region,
  "Burnaby South",                    "Burrard Peninsula",
  "Coquitlam Douglas College",        "North-East Sector",
  "Langley Central",                  "Fraser Valley West",
  "Maple Ridge Golden Ears School",   "North-East Sector",
  "New Westminster Sapperton Park",   "Burrard Peninsula",
  "North Delta",                      "South of Fraser",
  "North Vancouver Second Narrows",   "North Shore",
  "Pitt Meadows Airport",             "North-East Sector",
  "Port Coquitlam North",             "North-East Sector",
  "Port Moody Rocky Point Park",      "North-East Sector",
  "Richmond South",                   "South of Fraser",
  "Tsawwassen",                       "South of Fraser",
  "Vancouver Clark Drive",            "Vancouver",
  "Vancouver International Airport #2","South of Fraser",
  "Vancouver Kitsilano",              "Vancouver",
  "West Vancouver Lions Gate",        "North Shore"
)

airdata_region <- airdata_clean |>
  left_join(station_regions, by = "location") |>
  filter(!is.na(region))

ggplot(airdata_region, aes(x = region, y = O3)) +
  geom_boxplot(fill = "#52B788")
```

A boxplot shows the median (the line inside the box), the
interquartile range (the box itself), and outlier observations
(the dots). Reading regional O₃ differences from a single
chart is much faster than reading a summary table.

## `geom_smooth()` — a fitted trend

```{r smooth}
#| fig-cap: "Reforested area (ha) by fiscal year with a loess trend line."
#| fig-alt: "Scatter plot of reforested area by fiscal year with a loess trend curve and a shaded 95 percent confidence band."
ggplot(bc, aes(x = Fiscal_Year, y = Reforestation_ha)) +
  geom_point() +
  geom_smooth(method = "loess", se = TRUE)
```

`geom_smooth()` fits a smooth curve through the points. `method
= "loess"` is a local-regression smoother — flexible, makes no
parametric assumptions. `method = "lm"` would give you a straight
linear-regression line.

The shaded ribbon is the 95% confidence band. Toggle with
`se = FALSE` if you do not want it.

::: {.callout-tip}
### Layers draw in order

Every `+ geom_*()` adds a layer on top of the previous one. In
the example above, `geom_point()` draws the dots first, then
`geom_smooth()` draws the line on top. Reversing the order would
hide the dots behind the ribbon.

Read `+` as *"and also draw"*, in order.
:::

## Faceting — small multiples

When you have one chart you want repeated for each group, use
`facet_wrap()`:

```{r facet-wrap}
#| fig-cap: "Hourly ozone (O₃, ppb) distribution, one panel per region."
#| fig-alt: "Small-multiple histograms of hourly ozone in parts per billion, one panel per Metro Vancouver region."
ggplot(airdata_region, aes(x = O3)) +
  geom_histogram(binwidth = 2, fill = "#2D6A4F", colour = "white") +
  facet_wrap(~ region)
```

One panel per region. Same axes by default — change with
`scales = "free_y"` if a single y-axis is too compressed.

`facet_grid(rows ~ cols)` produces a 2D grid; less common but
useful when you have two grouping variables.

## Polish — labels, scales, theme

A scientific or professional chart needs labels, formatted
axes, and a clean theme:

```{r polish}
#| fig-cap: "BC reforestation area (ha), 1987–2023 (polished chart)."
#| fig-alt: "Polished bar chart titled BC Reforestation Area 1987 to 2023, showing reforested hectares per fiscal year with a comma-formatted y-axis, title, subtitle, and source caption."
ggplot(bc, aes(x = Fiscal_Year, y = Reforestation_ha)) +
  geom_col(fill = "#2D6A4F") +
  scale_y_continuous(labels = label_comma()) +
  scale_x_continuous(breaks = seq(1990, 2025, by = 5)) +
  labs(
    title    = "BC Reforestation Area, 1987–2023",
    subtitle = "Annual area reforested (planted or naturally regenerated)",
    x        = "Fiscal year",
    y        = "Reforested area (hectares)",
    caption  = "Source: Environmental Reporting BC (Open Government Licence — BC)"
  ) +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold"),
        plot.subtitle = element_text(colour = "grey40"))
```

Key polishing pieces:

- **`scale_y_continuous(labels = label_comma())`** — adds
  thousands separators to the y-axis (8,264,063 not 8264063).
- **`scale_x_continuous(breaks = ...)`** — controls which years
  appear on the x-axis.
- **`labs(...)`** — title, subtitle, axis labels, caption.
- **`theme_minimal()`** — a clean theme. Other built-ins:
  `theme_classic()`, `theme_bw()`, `theme_void()`.
- **`theme(...)`** — fine-grained overrides like fonts and
  colours.

## Saving a plot

```{r save-plot}
#| eval: false
my_plot <- ggplot(bc, aes(x = Fiscal_Year, y = Reforestation_ha)) +
  geom_col(fill = "#2D6A4F") +
  labs(title = "BC Reforestation Area")

ggsave(here("outputs", "bc_reforestation.png"),
       plot   = my_plot,
       width  = 7,
       height = 5,
       dpi    = 300)
```

Always set explicit `width` and `height` (in inches by default)
and a high `dpi` for print-quality output. `ggsave()` infers
the file format from the extension (`.png`, `.pdf`, `.svg`).

## Seven geoms in one place

```{r summary-table}
#| echo: false
tibble(
  Geom               = c("`geom_point()`", "`geom_line()`",
                         "`geom_col()`", "`geom_bar()`",
                         "`geom_histogram()`", "`geom_boxplot()`",
                         "`geom_smooth()`"),
  `What it draws`    = c("Dots at (x, y)", "Lines connecting (x, y)",
                         "Bars with height = y", "Bars with height = count(x)",
                         "Bars binning a single variable",
                         "Box + whiskers per group",
                         "Smooth trend curve"),
  `Typical use`      = c("Two continuous variables",
                         "One time series (or one per colour)",
                         "Pre-computed values",
                         "Counting category occurrences",
                         "Distribution of one continuous variable",
                         "Distributions across groups",
                         "Trend overlay on a scatter")
) |> kable()
```

## A worked composite — O₃ by station, faceted by region

A single chart that uses several of this chapter's tools:

```{r composite}
#| fig-cap: "Hourly ozone (O₃, ppb) over January 2000, one panel per region."
#| fig-alt: "Faceted line chart of hourly ozone over January 2000, one panel per region, with a separate coloured line for each monitoring station."
airdata_region <- airdata_clean |>
  mutate(date_parsed = mdy(Date)) |>
  left_join(station_regions, by = "location") |>
  filter(!is.na(region), !is.na(date_parsed))

ggplot(airdata_region,
       aes(x = date_parsed, y = O3, colour = location)) +
  geom_line(linewidth = 0.7, alpha = 0.85) +
  facet_wrap(~ region, ncol = 1, scales = "free_y") +
  scale_x_date(date_breaks = "1 week", date_labels = "%b %d") +
  labs(
    title    = "Hourly O₃, January 2000",
    subtitle = "Metro Vancouver air-quality monitoring network",
    x        = NULL,
    y        = "O₃ (ppb)",
    colour   = "Station",
    caption  = "Source: Environmental Reporting BC"
  ) +
  theme_minimal(base_size = 10) +
  theme(legend.position = "right",
        axis.text = element_text(size = 7),
        plot.title = element_text(face = "bold"))
```

Six tools applied: `geom_line`, `aes(colour = ...)`,
`facet_wrap`, `scale_x_date`, `labs`, `theme_minimal` plus
`theme()` overrides. **One chart, six choices, every choice in
the code.**

## Active learning

### Activity 1 — Bar chart

Build a bar chart of `Reforestation_ha` by `Fiscal_Year` for the
BC silviculture file. Make the bars dark green.

### Activity 2 — Three-line chart

Reshape `bc` to long format with three series (Harvested,
Natural_Disturbance, Reforestation). Plot all three as lines on
one chart, coloured by series.

### Activity 3 — Histogram with sensible bins

Plot a histogram of O₃ with `binwidth = 1`. Then try
`binwidth = 5`. Then `binwidth = 0.5`. Which bin width shows
the distribution shape most clearly?

### Activity 4 — Boxplot by region

Plot O₃ by region using `geom_boxplot()`. Add a `coord_flip()`
to make the regions read horizontally — easier with long region
names.

### Activity 5 — Facet a time series

Plot O₃ over time, faceted by station. Try `facet_wrap(~
location)` with the default `scales = "fixed"`. Then try
`scales = "free_y"`. Which makes patterns easier to compare?

### Activity 6 — Polish and save

Take any chart from this chapter. Add a title, subtitle, axis
labels, and a caption. Save it as a 7"x5" PNG with
`ggsave(here("outputs", "myplot.png"))`. Confirm the file
exists.

## Practice Demo Lab

::: {.callout-important}
### Practice demo only
This is a **guided practice lab in the OER book**, designed to help you
learn the workflow before completing the official lab assignment on
**Canvas**. The Canvas lab is the graded assignment. Use this activity to
practice, compare your work with the reference solution, and learn how to
write an AI verification prompt.
:::

This demo lab has **six parts** — work them in order:

1. **Your attempt.** Work through the tasks below.
2. **Reference solution.** Compare against the **worked examples earlier in
   this chapter** (your public reference). The graded Canvas lab has its own
   private answer key — do not copy demo solutions into a Canvas submission.
3. **Compare your work with the reference.** Where did it match? Where did
   it differ? What caused the difference, and how did you fix it?
4. **Write your own AI verification prompt.** Ask an AI to *check* your
   reasoning, code, and outputs — not to produce the answer for you.
5. **Model AI verification prompt.**

   > *"I am doing a FRST 232 practice demo lab. My dataset is `[file name]`
   > (columns `[columns]`). I attempted `[paste your steps or code]` and got
   > `[paste the output]`. Explain what my work does line by line, tell me
   > whether it answers the task, and list the checks I can run to verify it
   > against the chapter's worked example. Do not give me the final answer."*

6. **Reflection: what changed after checking?** In two or three sentences —
   did your attempt hold up? what did you fix? what will you check first
   next time?


::: {.callout-note appearance="default"}
### Lab: Build a chart that tells a story

Work in groups of 3 or 4.

**Task 1 — Pick a question.** Each group picks one forestry
question their chart will answer, e.g.:

- *Did BC harvest area decline after 2010?*
- *Which region had the most extreme O₃ hours in January 2000?*
- *How does the relationship between harvest and reforestation
  change over decades?*

**Task 2 — Choose the right geom.** Discuss: scatter, line,
bar, histogram, or boxplot? Pick the one that makes the answer
visible.

**Task 3 — Build the plot.** One member shares their screen and
the group types the code together. Iterate on bin widths, scales,
colours.

**Task 4 — Polish.** Add `labs()`, pick a theme, format the axes
with `scale_*_continuous(labels = label_comma())` or
`scale_x_date()`.

**Task 5 — Save.** Before saving, run this quick **plot checklist**:

- Is the **geom** right for the question (bar / line / point / box)?
- Are **x** and **y** the variables you intended?
- Do the **axis labels include units**?
- Are the **title / subtitle** informative, and is a **source caption**
  present?
- Will it save into **`outputs/`** (not your Desktop)?

Then save the chart to `outputs/` as PNG with `ggsave()`.

**Task 6 — Caption it.** Write a one-sentence caption that states the
**headline finding** in plain language. A useful template:

> *"This chart shows [variable] by [group / time]. The main pattern is
> [finding], measured in [unit]."*

**Task 7 — Submit.** Each group submits the PNG, the code, and
the caption.
:::

::: {.callout-tip}
## Reference solution — download
[**Open the Chapter 9 reference solution (rendered HTML)**](../lab-solutions/ch09-r-reference-solution.html) — the completed, rendered solution for this demo lab. Open it **after** your own attempt and use it to
check your work.

*This is the **public practice** solution. The **graded lab on Canvas** has
its own private answer key — do not copy this into a Canvas submission.*
:::

## Exercises

1. **Bar chart.** Recreate the polished BC reforestation bar
   chart from the *Polish* section above. Confirm it renders
   identically.

2. **Three-series line chart.** Pivot the BC file to long
   format and plot Harvested, Natural_Disturbance, and
   Reforestation as three coloured lines on one chart.

3. **Histogram comparison.** Plot histograms of O₃ with
   three different binwidths in three separate chunks. In one
   paragraph, justify your final binwidth choice.

4. **Box plot.** Plot O₃ by region. Add `coord_flip()`.
   Which region has the highest median? The widest spread?

5. **Faceted plot.** Plot O₃ over time, faceted by station,
   with one panel per station. Use `scales = "free_y"`.

6. **Polished composite.** Take any chart from exercises 1–5
   and add a title, subtitle, x label, y label, and caption.
   Save it as a 7×5 inch PNG in `outputs/`.

7. **Smooth trend.** Plot Reforestation vs Year as points, with
   `geom_smooth(method = "lm")` overlaid. Is the trend
   increasing or decreasing? By how much per year (look at the
   slope of the fitted line)?

8. **Optional, harder.** Plot mean O₃ by day-of-month for
   each region as lines on the same chart (`colour = region`).
   Which region has the widest range across the month?

## Optional, advanced

### Plot themes

Built-in themes you can swap in: `theme_minimal()`,
`theme_classic()`, `theme_bw()`, `theme_void()`, `theme_grey()`
(the default). For publication-quality plots in journals,
`theme_classic()` is usually safest.

### Colour palettes

ggplot2's default colour palette is fine for exploration but
weak for publication. Better:

- **`scale_colour_brewer(palette = "Set2")`** — categorical,
  printer-friendly.
- **`scale_colour_viridis_d()`** — categorical, colour-blind safe.
- **`scale_colour_viridis_c()`** — continuous, colour-blind safe.

For BC and Canadian forestry palettes, the `colorspace` package
includes many options.

### `patchwork` for multi-panel layouts

If you want two completely different plots side by side (not just
facets of one plot), use the `patchwork` package:

```r
library(patchwork)
p1 / p2          # stack p1 on top of p2
p1 | p2          # p1 and p2 side by side
(p1 | p2) / p3   # two-column top row, single-row bottom
```

### Interactive plots

`plotly::ggplotly(p)` converts most ggplot objects into
interactive HTML charts (hover for values, zoom, pan). Useful
for web reports.

## A glimpse ahead: the Integrated Case Study

The next required chapter, **Integrated Case Study: Choosing the Right
Tools**, is where the separate skills come together. Instead of being told
which function to practice, you are handed a **question and a dataset** and
decide the workflow yourself:

- Read a realistic forestry question and choose the steps it needs.
- Combine **import, inspect, clean, `filter`, `mutate`, `group_by`,
  `summarise`, `pivot`, `left_join`**, and **`ggplot2`** in the right
  order.
- Justify *why* each tool is the efficient choice — and when *not* to use
  one.
- Verify every result with row counts, missing-value checks, and a quick
  plot, then interpret it for a non-technical reader.

The O₃-by-station chart you just built is one piece; the case study is
where *you* decide which pieces a real question calls for.

*Spatial mapping is available as an **optional** extension — see the
[Optional: Spatial Data Analysis in R](../optional-spatial.qmd) page — but
it is not required for the case study.*

## Self-assessment quiz {#sec-quiz-ch09}

::: {.content-visible when-format="pdf"}
*The self-assessment quiz is interactive: it appears in the online (HTML)
book, and you can also take it on Canvas. This PDF does not display the
interactive quiz questions.*
:::

```{=html}
<style>
.quiz { background: #F4F7F4; border-left: 4px solid #2D6A4F; padding: 1em 1.5em; margin: 1.5em 0; border-radius: 6px; }
.quiz .q { font-weight: 600; color: #1B4332; margin-bottom: 0.6em; }
.quiz ul.qopts { list-style: none; padding: 0; margin: 0; }
.quiz ul.qopts li { padding: 0.55em 0.9em; margin: 0.3em 0; background: white; border: 1px solid #d8e4d8; border-radius: 4px; cursor: pointer; transition: background 0.12s; }
.quiz ul.qopts li:hover { background: #E8F1E9; }
.quiz ul.qopts li.correct { background: #C8E6C9; border-color: #2D6A4F; font-weight: 600; }
.quiz ul.qopts li.wrong { background: #FFCDD2; border-color: #B05C50; }
.quiz .explanation { display: none; margin-top: 0.7em; padding: 0.7em 1em; background: #FFF8E8; border-left: 3px solid #B7791F; border-radius: 4px; font-size: 0.95em; }
.quiz .explanation.show { display: block; }
.quiz textarea { width: 100%; min-height: 4.5em; padding: 0.6em; border: 1px solid #d8e4d8; border-radius: 4px; font-family: inherit; font-size: 0.95em; resize: vertical; margin-top: 0.3em; }
.quiz .reveal-btn { background: #2D6A4F; color: white; border: none; padding: 0.45em 0.9em; margin-top: 0.5em; border-radius: 4px; cursor: pointer; font-size: 0.9em; }
.quiz .reveal-btn:hover { background: #1B4332; }
.quiz .sample { display: none; margin-top: 0.6em; padding: 0.6em 0.9em; background: #FFF8E8; border-left: 3px solid #B7791F; border-radius: 4px; font-size: 0.95em; font-style: italic; }
.quiz .sample.show { display: block; }
</style>

<div class="quiz" id="q1" data-hints="Strip a ggplot to its minimum — what cannot be left out?||The table, the column-to-visual mapping, and the shape to draw.||data, aesthetic mappings, geometric objects."><div class="q">1. The three core pieces of a ggplot are —</div><ul class="qopts"><li data-correct="false">data, file, output</li><li data-correct="true">data, aesthetic mappings, geometric objects</li><li data-correct="false">x, y, title</li><li data-correct="false">colour, size, theme</li></ul><div class="explanation">Data + aes() + geom_*() are the minimum. Scales, themes, facets, and labels are polish.</div></div>

<div class="quiz" id="q2" data-hints="aes connects DATA to visual properties.||It maps columns to x, y, colour, size, shape.||Aesthetic mapping — maps columns to visual properties."><div class="q">2. <code>aes()</code> in ggplot2 stands for and does what?</div><ul class="qopts"><li data-correct="false">"aesthetic" — sets the chart's colours</li><li data-correct="true">"aesthetic mapping" — maps columns in your tibble to visual properties of the chart</li><li data-correct="false">"area + estimate + scale" — a statistical function</li><li data-correct="false">"axis encoding system" — sets the axes</li></ul><div class="explanation">aes() connects data columns to visual properties (x, y, colour, fill, size, shape).</div></div>

<div class="quiz" id="q3" data-hints="ggplot predates the pipe and uses a different connector.||Read it as 'and also draw'.||The + sign."><div class="q">3. ggplot layers are added with —</div><ul class="qopts"><li data-correct="false">the pipe |></li><li data-correct="true">the + sign</li><li data-correct="false">commas inside ggplot()</li><li data-correct="false">a separate function for each layer</li></ul><div class="explanation">ggplot uses + (not |>) because it was designed before the pipe was popular. Read + as "and also draw".</div></div>

<div class="quiz" id="q4" data-hints="geom_col uses your supplied y values as bar heights.||So you must give it both coordinates.||Both x and y."><div class="q">4. <code>geom_col()</code> requires you to supply —</div><ul class="qopts"><li data-correct="false">Only x</li><li data-correct="true">Both x and y (the y value is the bar height)</li><li data-correct="false">A colour palette</li><li data-correct="false">Nothing — it counts for you</li></ul><div class="explanation"><code>geom_col</code> uses your supplied y values as bar heights. <code>geom_bar</code> counts x values for you instead.</div></div>

<div class="quiz" id="q5" data-hints="geom_bar counts the x values for you.||The bar heights are those counts.||Bars whose heights are the counts of each x value."><div class="q">5. <code>geom_bar()</code> (with no y aesthetic) draws —</div><ul class="qopts"><li data-correct="false">Lines</li><li data-correct="true">Bars whose heights are the counts of each x value</li><li data-correct="false">Empty bars</li><li data-correct="false">A histogram</li></ul><div class="explanation">If you have raw events (one row per occurrence), use <code>geom_bar</code> to count. If you have pre-computed counts, use <code>geom_col</code>.</div></div>

<div class="quiz" id="q6" data-hints="A histogram buckets a numeric column into bins.||binwidth sets the bucket width.||It bins values into 2-unit buckets and plots the count in each."><div class="q">6. <code>geom_histogram(binwidth = 2)</code> on a numeric column —</div><ul class="qopts"><li data-correct="false">Plots two bars</li><li data-correct="true">Buckets the values into 2-unit-wide bins and plots the count in each bin</li><li data-correct="false">Plots the first 2 values</li><li data-correct="false">Returns an error</li></ul><div class="explanation">binwidth controls the bucket size. Smaller bins = more detail; larger bins = smoother shape.</div></div>

<div class="quiz" id="q7" data-hints="A boxplot is a five-number summary.||The line in the box is the median, not the mean.||Median, interquartile range, and outliers."><div class="q">7. <code>geom_boxplot()</code> shows —</div><ul class="qopts"><li data-correct="false">The mean</li><li data-correct="true">The median (line in box), interquartile range (box), and outliers (dots)</li><li data-correct="false">The sum</li><li data-correct="false">A linear regression</li></ul><div class="explanation">The boxplot is a robust five-number summary. The line in the box is the median, not the mean.</div></div>

<div class="quiz" id="q8" data-hints="Faceting makes small multiples.||One panel per distinct value of the variable.||One panel per region."><div class="q">8. <code>facet_wrap(~ region)</code> produces —</div><ul class="qopts"><li data-correct="false">A single chart</li><li data-correct="true">One panel per unique value of region (small multiples)</li><li data-correct="false">An error</li><li data-correct="false">A chart with region on the x-axis</li></ul><div class="explanation">Faceting creates small-multiple panels. Same chart, repeated per group.</div></div>

<div class="quiz" id="q9" data-hints="To draw three lines by colour you need a single series column.||That requires reshaping to long.||Long format, with a series column for the colour mapping."><div class="q">9. For plotting three time series on one chart, the data should be in —</div><ul class="qopts"><li data-correct="false">Wide format (one column per series)</li><li data-correct="true">Long format (one row per series × time, with a 'series' column for the colour mapping)</li><li data-correct="false">Either, no difference</li><li data-correct="false">A list, not a tibble</li></ul><div class="explanation">Long format lets you map <code>colour = series</code> and draw all three lines in one call. Wide format requires three separate <code>geom_line</code> calls.</div></div>

<div class="quiz" id="q10" data-hints="Which ggplot function writes a plot to a file?||It infers the format from the file extension.||ggsave('outputs/myplot.png', width, height)."><div class="q">10. To save a ggplot as a PNG file, use —</div><ul class="qopts"><li data-correct="false">save_plot()</li><li data-correct="true">ggsave("outputs/myplot.png", width = 7, height = 5)</li><li data-correct="false">writePNG()</li><li data-correct="false">export()</li></ul><div class="explanation"><code>ggsave</code> infers the format from the extension. Always specify width and height explicitly.</div></div>

<div class="quiz" id="q11" data-hints="labs sets the text around the plot.||Title, subtitle, axis labels, caption.||It adds or replaces chart labels."><div class="q">11. <code>labs(title = "...", x = "...")</code> does what?</div><ul class="qopts"><li data-correct="false">Adds new data</li><li data-correct="true">Adds or replaces chart labels (title, axis labels, captions)</li><li data-correct="false">Sets the colour palette</li><li data-correct="false">Saves the file</li></ul><div class="explanation"><code>labs()</code> is the consolidated labels function — title, subtitle, x, y, colour, caption.</div></div>

<div class="quiz" id="q12" data-hints="method='lm' fits a straight line, not a flexible curve.||It also shows a confidence band.||A linear regression line plus its confidence band."><div class="q">12. <code>geom_smooth(method = "lm")</code> overlays —</div><ul class="qopts"><li data-correct="false">A loess smoother</li><li data-correct="true">A linear regression line plus its confidence band</li><li data-correct="false">A boxplot</li><li data-correct="false">A histogram</li></ul><div class="explanation"><code>method = "lm"</code> fits a straight line. <code>method = "loess"</code> fits a flexible local-regression curve. Default for small samples is loess.</div></div>

<div class="quiz" id="q13" data-hints="It is one of several built-in themes.||An alternative to the default grey-background look.||A built-in clean theme (one of several)."><div class="q">13. <code>theme_minimal()</code> is —</div><ul class="qopts"><li data-correct="false">A required base layer for every chart</li><li data-correct="true">One of several built-in clean themes; an alternative to the default grey-background theme</li><li data-correct="false">A colour palette</li><li data-correct="false">A way to hide the chart</li></ul><div class="explanation">Other built-ins: theme_classic, theme_bw, theme_void. Pick whatever matches the venue (publication, web, slides).</div></div>

<div class="quiz" id="q14" data-hints="label_comma formats the axis numbers.||It inserts thousands separators.||Adds thousands separators to the y-axis labels."><div class="q">14. <code>scale_y_continuous(labels = label_comma())</code> does what?</div><ul class="qopts"><li data-correct="false">Logs the y-axis</li><li data-correct="true">Adds thousands separators to y-axis labels (8,264,063 instead of 8264063)</li><li data-correct="false">Reverses the axis</li><li data-correct="false">Adds a legend</li></ul><div class="explanation">From the scales package. Other useful formatters: label_percent, label_dollar, label_date.</div></div>

<div class="quiz" id="q15" data-hints="Mapping a column to colour must go where data is mapped.||That is inside aes(); outside aes() just sets one fixed colour.||aes(colour = third_variable)."><div class="q">15. Mapping a third variable to colour is done inside —</div><ul class="qopts"><li data-correct="false">labs()</li><li data-correct="true">aes(colour = third_variable)</li><li data-correct="false">theme(colour = ...)</li><li data-correct="false">geom_*(colour = ...)</li></ul><div class="explanation">Inside aes(), colour MAPS a column to the colour aesthetic (legend appears). Outside aes(), colour = "blue" SETS the colour for all elements (no legend).</div></div>

<div class="quiz" id="q16" data-hints="A line connects points in a specific order.||It follows the x aesthetic; unsorted x zig-zags.||In the order of the x values."><div class="q">16. <code>geom_line()</code> connects points —</div><ul class="qopts"><li data-correct="false">In a random order</li><li data-correct="true">In the order of the x aesthetic</li><li data-correct="false">By their colour</li><li data-correct="false">Counter-clockwise</li></ul><div class="explanation">geom_line draws in x-order. If your data is not sorted by x, the line zig-zags. Sort first or use geom_path for arbitrary order.</div></div>

<div class="quiz" id="q17" data-hints="Dodge changes how grouped bars are positioned.||The default is to stack them.||It places bars side by side instead of stacked."><div class="q">17. <code>position = "dodge"</code> on a bar chart —</div><ul class="qopts"><li data-correct="false">Stacks bars</li><li data-correct="true">Places bars side by side instead of on top of each other</li><li data-correct="false">Hides the bars</li><li data-correct="false">Reorders the bars</li></ul><div class="explanation">Default for geom_col with a fill aesthetic is "stack". <code>position = "dodge"</code> produces grouped bars (e.g., comparing harvested and reforested area side-by-side per year).</div></div>

<div class="quiz" id="q18" data-hints="facet_grid forces a fixed rows-by-columns layout.||Use it with two grouping variables.||A two-dimensional grid of small multiples."><div class="q">18. <code>facet_grid(rows ~ cols)</code> produces —</div><ul class="qopts"><li data-correct="false">A single chart</li><li data-correct="true">A two-dimensional grid of small multiples</li><li data-correct="false">A pie chart</li><li data-correct="false">A scatter plot</li></ul><div class="explanation">facet_wrap wraps panels into a flexible layout; facet_grid forces a rows × cols layout — useful when you have two grouping variables.</div></div>

<div class="quiz" id="q19" data-hints="Pie charts make you compare angles, which the eye reads poorly.||A bar chart of the same data is clearer.||Usually replaced by a bar chart."><div class="q">19. Pie charts in ggplot2 require <code>coord_polar()</code> and are —</div><ul class="qopts"><li data-correct="false">Strongly recommended for forestry data</li><li data-correct="true">Usually replaced by a bar chart, because angles are harder to compare than lengths</li><li data-correct="false">The default chart type</li><li data-correct="false">Impossible to make</li></ul><div class="explanation">Pie charts are harder to read accurately. A bar chart of the same data is almost always clearer. Use pie only if the venue requires it.</div></div>

<div class="quiz" id="q20"><div class="q">20. <strong>Reflection.</strong> Which Excel chart type did you use most often before this chapter? Which ggplot2 geom replaces it? What changes about how you make charts?</div><textarea placeholder="Type your answer here..."></textarea><button class="reveal-btn" onclick="this.nextElementSibling.classList.toggle('show')">Show common answers</button><div class="sample">Common reflections: <em>I used Excel bar charts most — geom_col replaces them, and now every label, colour, and axis is written down in code I can re-run</em> · <em>Excel scatter plots were a few clicks but I never knew which buttons set what — ggplot makes the choices explicit</em> · <em>I never used boxplots in Excel because they were hard; in ggplot they are one line</em> · <em>The biggest change is faceting — small multiples are trivial in ggplot but a manual nightmare in Excel</em>.</div></div>

<script>
(function() {
  document.querySelectorAll('.quiz').forEach(function(q) {
    q.querySelectorAll('ul.qopts li').forEach(function(opt) {
      opt.addEventListener('click', function() {
        if (q.dataset.answered) return;
        q.dataset.answered = '1';
        var isCorrect = opt.getAttribute('data-correct') === 'true';
        if (isCorrect) opt.classList.add('correct');
        else {
          opt.classList.add('wrong');
          q.querySelectorAll('ul.qopts li').forEach(function(o) {
            if (o.getAttribute('data-correct') === 'true') o.classList.add('correct');
          });
        }
        var exp = q.querySelector('.explanation');
        if (exp) exp.classList.add('show');
      });
    });
  });
})();
</script>
```

## AI as a debugging companion

::: {.callout-tip}
### Useful prompts

- *"My ggplot only draws one line, but my long-format tibble has
  three series. What did I miss in aes()?"* (Almost always:
  forgot to add `group = series` or `colour = series`.)

- *"How do I make the x-axis show dates like 'Jan 15' instead of
  '2000-01-15' in ggplot?"* (Answer: `scale_x_date(date_labels = "%b %d")`.)

- *"My chart's y-axis says 8.26e+06 instead of 8,264,063. How
  do I fix it?"* (Answer: `scale_y_continuous(labels = label_comma())`.)
:::

## Reading

- *R for Data Science* — chapters 1 *Data visualisation* and 12
  *Communication*. Free at <https://r4ds.hadley.nz/>.
- *ggplot2: Elegant Graphics for Data Analysis* (Wickham,
  Navarro, Pedersen) — the canonical reference. Free at
  <https://ggplot2-book.org/>.
- The ggplot2 cheatsheet (PDF):
  <https://rstudio.github.io/cheatsheets/data-visualization.pdf>

## Instructor notes {.unnumbered}

::: {.callout-note appearance="minimal" collapse="true"}
### Pacing and emphasis

- Spend most time on **`geom_col`, `geom_line`, `geom_histogram`,
  `geom_boxplot`, and `facet_wrap`** — these five cover 90% of
  routine forestry plotting.
- The **long-format-for-plotting** habit is the chapter's most
  important reusable insight. When a student's chart looks
  wrong, ask first: "is your data long or wide?"
- **`ggsave` matters.** Students will hand in screenshots
  instead of PNG files if you do not insist on `ggsave`.

### Materials provided

| File | Purpose |
|---|---|
| `frst232_ch09_learner.qmd` | Quarto starter with TODO chunks. |
| `frst232_ch09_solutions.qmd` | Completed solutions. **Do not distribute.** |
| `quiz_ch09.html` | Standalone interactive quiz. |

### Common stumbling points

- **`+` instead of `|>` in pipelines.** Walk through: `|>` for
  data transformation, `+` for ggplot layers.
- **`colour` outside `aes()`** instead of inside (or vice
  versa). Inside aes = data-driven; outside aes = constant.
- **Wide data plotted with `geom_line`** giving one line per
  observation instead of one per series. Fix: `pivot_longer`
  first.
- **Forgetting `group = ...`** in line charts with categorical
  x. Symptom: zig-zag lines.
:::
