---
title: "Visualize Forestry Data with ggplot2"
---

```{r setup-ch09}
#| include: false
library(tidyverse)
library(readxl)
library(lubridate)
library(here)
library(knitr)
library(scales)
```

# Chapter goals {.unnumbered}

By the end of this chapter you will be able to:

- Explain the **grammar of graphics**: data → aesthetic mappings
  → geometric objects → coordinate system → theme.
- Build a **basic plot in three lines** using
  `ggplot(data, aes(...)) + geom_*()`.
- Use the **six most common geoms**: `geom_point()`,
  `geom_line()`, `geom_col()` / `geom_bar()`,
  `geom_histogram()`, `geom_boxplot()`, and `geom_smooth()`.
- Add **multiple layers** to a single plot and understand the
  order in which they draw.
- Use **`facet_wrap()`** for small multiples (one panel per
  category) and **`facet_grid()`** for two-dimensional grids.
- Polish a plot with **labels** (`labs()`), **scales**
  (`scale_*_continuous()`, `scale_*_date()`), **colours**
  (`scale_color_*()` / `scale_fill_*()`), and **themes**
  (`theme_minimal()`, `theme()` overrides).
- Save a plot to disk with **`ggsave()`** at a sensible size and
  resolution.
- Recognize when long-format data is needed (and when to call
  `pivot_longer()` before plotting).

# Expected output for this chapter {.unnumbered}

::: {.callout-tip appearance="default"}
## What you will hand in

1. A **rendered Quarto HTML report**
   (`frst232_ch09_learner.html`) containing at least:
   - One **bar chart** of BC reforestation by fiscal year.
   - One **line chart** with two or more series overlaid (e.g.,
     harvested vs natural disturbance vs reforestation across
     years).
   - One **histogram** of hourly O₃ readings.
   - One **box plot** of O₃ by region.
   - One **faceted** plot (small multiples) showing O₃ over
     the days of January 2000, faceted by station.
   - At least one chart with **custom labels** (`labs(title,
     subtitle, x, y, caption)`).
2. One chart saved to `outputs/` as a `.png` using `ggsave()`.
3. A short **group-lab worksheet** (or screenshot) submitted to
   the course site.
:::

# Where we are

Through Chapters 4–8 you imported, inspected, cleaned, summarised,
reshaped, and combined forestry data. Chapter 9 turns those
tibbles into **charts** — the most efficient way for a reader to
understand a pattern that would be invisible in a table.

A picture is also the most efficient way to **lie** with data,
which is why ggplot2's grammar makes every visual choice
explicit. Every axis, scale, colour, and label is in your code.
Choices that Excel hides in menu dialogs are written down where
your reader (and future you) can see them.

# Excel-chart-to-ggplot bridge

| Excel chart | ggplot2 geom + extras |
|---|---|
| Column / bar chart | `geom_col()` (values you supply) / `geom_bar()` (counts of categories) |
| Line chart | `geom_line()` |
| Scatter plot | `geom_point()` |
| Histogram | `geom_histogram()` |
| Box / whisker plot | `geom_boxplot()` |
| Area chart | `geom_area()` |
| Pie chart | `geom_col() + coord_polar()` — but **avoid** (pie charts are hard to read; use a bar chart) |
| Trend line / regression line | `geom_smooth(method = "lm")` |
| Stacked bar | `geom_col(position = "stack")` |
| Side-by-side (grouped) bar | `geom_col(position = "dodge")` |
| Multiple charts side by side (small multiples) | `facet_wrap(~ category)` |
| Manual axis-range setting | `scale_y_continuous(limits = c(0, 100))` |
| Chart title / subtitle / data label | `labs(title = "...", subtitle = "...", caption = "...")` |
| Saving as PNG | `ggsave("outputs/myplot.png", width = 7, height = 5)` |

# The grammar of graphics — three sentences

A ggplot is built from three pieces:

1. **Data** — a tibble.
2. **Aesthetic mappings** — which column maps to which visual
   property (x, y, color, fill, size, shape).
3. **Geometric objects** — the shapes that get drawn (points,
   lines, bars).

A ggplot call says, *"Take this tibble, map these columns to
these visual properties, and draw these shapes."*

```{r minimal-ggplot}
bc <- read_excel(here("data", "bc_disturbance_reforestation.xlsx"),
                 sheet = "Data")

ggplot(data = bc, mapping = aes(x = Fiscal_Year, y = Reforestation_ha)) +
  geom_line()
```

Three things to notice:

- `ggplot()` opens the plot. It does not draw anything yet.
- `aes()` says *"map `Fiscal_Year` to the x-axis and
  `Reforestation_ha` to the y-axis"*.
- `geom_line()` draws lines connecting the (x, y) points.
- The `+` sign **adds layers** — not the pipe `|>`. This is
  ggplot2's one syntactic quirk. Read `+` as *"and also draw"*.

# `geom_point()` — scatter plot

```{r geom-point}
ggplot(bc, aes(x = Harvested_ha, y = Natural_Disturbance_ha)) +
  geom_point()
```

Each year is one dot. No relationship is obvious at this scale,
but the chart immediately reveals a few **outlier years** (large
natural-disturbance values) — useful for spotting the 2017
wildfire spike.

## Mapping a third variable to colour

```{r point-colour}
ggplot(bc, aes(x = Harvested_ha,
               y = Natural_Disturbance_ha,
               colour = Fiscal_Year)) +
  geom_point(size = 3)
```

`colour = Fiscal_Year` adds a third dimension. A continuous
variable like year gets a gradient legend.

# `geom_line()` — time series

```{r line-multiple-series}
# First reshape to long format so each series is a row
bc_long <- bc |>
  select(Fiscal_Year, Harvested_ha,
         Natural_Disturbance_ha, Reforestation_ha) |>
  pivot_longer(cols = -Fiscal_Year,
               names_to = "series",
               values_to = "area_ha")

ggplot(bc_long, aes(x = Fiscal_Year, y = area_ha, colour = series)) +
  geom_line(linewidth = 1)
```

**Why long format**: with one row per (year, series) combination,
ggplot can draw three lines in one call by mapping `colour =
series`. With wide format (three columns) you would need three
separate `geom_line()` calls — much less elegant.

**This is why Chapter 8 spent time on `pivot_longer()`.** For
plotting, long is almost always the right shape.

# `geom_col()` vs `geom_bar()`

The two bar geoms are confused often. The rule:

- `geom_col()` — you give it `y` values to plot.
- `geom_bar()` — you give it `x` values and it **counts** them.

```{r col-vs-bar}
# geom_col — you supply the y value
ggplot(bc, aes(x = Fiscal_Year, y = Reforestation_ha)) +
  geom_col(fill = "#2D6A4F")
```

```{r geom-bar-count}
# geom_bar — it counts for you
# Build a small tibble with one row per disturbance event
events <- tibble(
  year = c(2017, 2017, 2017, 2018, 2019, 2019, 2020),
  type = c("fire", "fire", "pest", "fire", "fire", "wind", "fire")
)
ggplot(events, aes(x = type)) +
  geom_bar(fill = "#2D6A4F")
```

`geom_bar` produced the count of each `type`. `geom_col` would
have required a numeric column to use as height.

# `geom_histogram()` — distribution of one variable

```{r histogram}
airdata_clean <- read_csv(here("data", "airdata3.csv"),
                          show_col_types = FALSE) |>
  filter(!is.na(O3))

ggplot(airdata_clean, aes(x = O3)) +
  geom_histogram(binwidth = 2, fill = "#2D6A4F", colour = "white")
```

`binwidth = 2` puts each bar in a 2 ppb bucket. The histogram
reveals the **heavily right-skewed** O₃ distribution: most
hours are clean, a few are sharply elevated. This is the kind of
pattern that a summary table can hint at but only a chart makes
obvious.

# `geom_boxplot()` — distributions by group

```{r boxplot}
# First add a region column (from the Ch 6 lookup)
station_regions <- tribble(
  ~location,                          ~region,
  "Burnaby South",                    "Burrard Peninsula",
  "Coquitlam Douglas College",        "North-East Sector",
  "Langley Central",                  "Fraser Valley West",
  "Maple Ridge Golden Ears School",   "North-East Sector",
  "New Westminster Sapperton Park",   "Burrard Peninsula",
  "North Delta",                      "South of Fraser",
  "North Vancouver Second Narrows",   "North Shore",
  "Pitt Meadows Airport",             "North-East Sector",
  "Port Coquitlam North",             "North-East Sector",
  "Port Moody Rocky Point Park",      "North-East Sector",
  "Richmond South",                   "South of Fraser",
  "Tsawwassen",                       "South of Fraser",
  "Vancouver Clark Drive",            "Vancouver",
  "Vancouver International Airport #2","South of Fraser",
  "Vancouver Kitsilano",              "Vancouver",
  "West Vancouver Lions Gate",        "North Shore"
)

airdata_region <- airdata_clean |>
  left_join(station_regions, by = "location") |>
  filter(!is.na(region))

ggplot(airdata_region, aes(x = region, y = O3)) +
  geom_boxplot(fill = "#52B788")
```

A boxplot shows the median (the line inside the box), the
interquartile range (the box itself), and outlier observations
(the dots). Reading regional O₃ differences from a single
chart is much faster than reading a summary table.

# `geom_smooth()` — a fitted trend

```{r smooth}
ggplot(bc, aes(x = Fiscal_Year, y = Reforestation_ha)) +
  geom_point() +
  geom_smooth(method = "loess", se = TRUE)
```

`geom_smooth()` fits a smooth curve through the points. `method
= "loess"` is a local-regression smoother — flexible, makes no
parametric assumptions. `method = "lm"` would give you a straight
linear-regression line.

The shaded ribbon is the 95% confidence band. Toggle with
`se = FALSE` if you do not want it.

::: {.callout-tip}
## Layers draw in order

Every `+ geom_*()` adds a layer on top of the previous one. In
the example above, `geom_point()` draws the dots first, then
`geom_smooth()` draws the line on top. Reversing the order would
hide the dots behind the ribbon.

Read `+` as *"and also draw"*, in order.
:::

# Faceting — small multiples

When you have one chart you want repeated for each group, use
`facet_wrap()`:

```{r facet-wrap}
ggplot(airdata_region, aes(x = O3)) +
  geom_histogram(binwidth = 2, fill = "#2D6A4F", colour = "white") +
  facet_wrap(~ region)
```

One panel per region. Same axes by default — change with
`scales = "free_y"` if a single y-axis is too compressed.

`facet_grid(rows ~ cols)` produces a 2D grid; less common but
useful when you have two grouping variables.

# Polish — labels, scales, theme

A scientific or professional chart needs labels, formatted
axes, and a clean theme:

```{r polish}
ggplot(bc, aes(x = Fiscal_Year, y = Reforestation_ha)) +
  geom_col(fill = "#2D6A4F") +
  scale_y_continuous(labels = label_comma()) +
  scale_x_continuous(breaks = seq(1990, 2025, by = 5)) +
  labs(
    title    = "BC Reforestation Area, 1987–2023",
    subtitle = "Annual area reforested (planted or naturally regenerated)",
    x        = "Fiscal year",
    y        = "Reforested area (hectares)",
    caption  = "Source: Environmental Reporting BC (Open Government Licence — BC)"
  ) +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold"),
        plot.subtitle = element_text(colour = "grey40"))
```

Key polishing pieces:

- **`scale_y_continuous(labels = label_comma())`** — adds
  thousands separators to the y-axis (8,264,063 not 8264063).
- **`scale_x_continuous(breaks = ...)`** — controls which years
  appear on the x-axis.
- **`labs(...)`** — title, subtitle, axis labels, caption.
- **`theme_minimal()`** — a clean theme. Other built-ins:
  `theme_classic()`, `theme_bw()`, `theme_void()`.
- **`theme(...)`** — fine-grained overrides like fonts and
  colours.

# Saving a plot

```{r save-plot}
#| eval: false
my_plot <- ggplot(bc, aes(x = Fiscal_Year, y = Reforestation_ha)) +
  geom_col(fill = "#2D6A4F") +
  labs(title = "BC Reforestation Area")

ggsave(here("outputs", "bc_reforestation.png"),
       plot   = my_plot,
       width  = 7,
       height = 5,
       dpi    = 300)
```

Always set explicit `width` and `height` (in inches by default)
and a high `dpi` for print-quality output. `ggsave()` infers
the file format from the extension (`.png`, `.pdf`, `.svg`).

# Six geoms in one place

```{r summary-table}
#| echo: false
tibble(
  Geom               = c("`geom_point()`", "`geom_line()`",
                         "`geom_col()`", "`geom_bar()`",
                         "`geom_histogram()`", "`geom_boxplot()`",
                         "`geom_smooth()`"),
  `What it draws`    = c("Dots at (x, y)", "Lines connecting (x, y)",
                         "Bars with height = y", "Bars with height = count(x)",
                         "Bars binning a single variable",
                         "Box + whiskers per group",
                         "Smooth trend curve"),
  `Typical use`      = c("Two continuous variables",
                         "One time series (or one per colour)",
                         "Pre-computed values",
                         "Counting category occurrences",
                         "Distribution of one continuous variable",
                         "Distributions across groups",
                         "Trend overlay on a scatter")
) |> kable()
```

# A worked composite — O₃ by station, faceted by region

A single chart that uses several of this chapter's tools:

```{r composite}
airdata_region <- airdata_clean |>
  mutate(date_parsed = mdy(Date)) |>
  left_join(station_regions, by = "location") |>
  filter(!is.na(region), !is.na(date_parsed))

ggplot(airdata_region,
       aes(x = date_parsed, y = O3, colour = location)) +
  geom_line(linewidth = 0.7, alpha = 0.85) +
  facet_wrap(~ region, ncol = 1, scales = "free_y") +
  scale_x_date(date_breaks = "1 week", date_labels = "%b %d") +
  labs(
    title    = "Hourly O₃, January 2000",
    subtitle = "Metro Vancouver air-quality monitoring network",
    x        = NULL,
    y        = "O₃ (ppb)",
    colour   = "Station",
    caption  = "Source: Environmental Reporting BC"
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom",
        plot.title = element_text(face = "bold"))
```

Six tools applied: `geom_line`, `aes(colour = ...)`,
`facet_wrap`, `scale_x_date`, `labs`, `theme_minimal` plus
`theme()` overrides. **One chart, six choices, every choice in
the code.**

# Active learning

## Activity 1 — Bar chart

Build a bar chart of `Reforestation_ha` by `Fiscal_Year` for the
BC silviculture file. Make the bars dark green.

## Activity 2 — Three-line chart

Reshape `bc` to long format with three series (Harvested,
Natural_Disturbance, Reforestation). Plot all three as lines on
one chart, coloured by series.

## Activity 3 — Histogram with sensible bins

Plot a histogram of O₃ with `binwidth = 1`. Then try
`binwidth = 5`. Then `binwidth = 0.5`. Which bin width shows
the distribution shape most clearly?

## Activity 4 — Boxplot by region

Plot O₃ by region using `geom_boxplot()`. Add a `coord_flip()`
to make the regions read horizontally — easier with long region
names.

## Activity 5 — Facet a time series

Plot O₃ over time, faceted by station. Try `facet_wrap(~
location)` with the default `scales = "fixed"`. Then try
`scales = "free_y"`. Which makes patterns easier to compare?

## Activity 6 — Polish and save

Take any chart from this chapter. Add a title, subtitle, axis
labels, and a caption. Save it as a 7"x5" PNG with
`ggsave(here("outputs", "myplot.png"))`. Confirm the file
exists.

# Group lab activity

::: {.callout-note appearance="default"}
## Lab: Build a chart that tells a story

Work in groups of 3 or 4.

**Task 1 — Pick a question.** Each group picks one forestry
question their chart will answer, e.g.:

- *Did BC harvest area decline after 2010?*
- *Which region had the most extreme O₃ hours in January 2000?*
- *How does the relationship between harvest and reforestation
  change over decades?*

**Task 2 — Choose the right geom.** Discuss: scatter, line,
bar, histogram, or boxplot? Pick the one that makes the answer
visible.

**Task 3 — Build the plot.** One member shares their screen and
the group types the code together. Iterate on bin widths, scales,
colours.

**Task 4 — Polish.** Add `labs()`, pick a theme, format the axes
with `scale_*_continuous(labels = label_comma())` or
`scale_x_date()`.

**Task 5 — Save.** Save the chart to `outputs/` as PNG with
`ggsave()`.

**Task 6 — Caption it.** Write a one-sentence caption that states
the **headline finding** of the chart in plain language.

**Task 7 — Submit.** Each group submits the PNG, the code, and
the caption.
:::

# Exercises

1. **Bar chart.** Recreate the polished BC reforestation bar
   chart from the *Polish* section above. Confirm it renders
   identically.

2. **Three-series line chart.** Pivot the BC file to long
   format and plot Harvested, Natural_Disturbance, and
   Reforestation as three coloured lines on one chart.

3. **Histogram comparison.** Plot histograms of O₃ with
   three different binwidths in three separate chunks. In one
   paragraph, justify your final binwidth choice.

4. **Box plot.** Plot O₃ by region. Add `coord_flip()`.
   Which region has the highest median? The widest spread?

5. **Faceted plot.** Plot O₃ over time, faceted by station,
   with one panel per station. Use `scales = "free_y"`.

6. **Polished composite.** Take any chart from exercises 1–5
   and add a title, subtitle, x label, y label, and caption.
   Save it as a 7×5 inch PNG in `outputs/`.

7. **Smooth trend.** Plot Reforestation vs Year as points, with
   `geom_smooth(method = "lm")` overlaid. Is the trend
   increasing or decreasing? By how much per year (look at the
   slope of the fitted line)?

8. **Optional, harder.** Plot mean O₃ by day-of-month for
   each region as lines on the same chart (`colour = region`).
   Which region has the widest range across the month?

# Optional, advanced

## Plot themes

Built-in themes you can swap in: `theme_minimal()`,
`theme_classic()`, `theme_bw()`, `theme_void()`, `theme_grey()`
(the default). For publication-quality plots in journals,
`theme_classic()` is usually safest.

## Colour palettes

ggplot2's default colour palette is fine for exploration but
weak for publication. Better:

- **`scale_colour_brewer(palette = "Set2")`** — categorical,
  printer-friendly.
- **`scale_colour_viridis_d()`** — categorical, colour-blind safe.
- **`scale_colour_viridis_c()`** — continuous, colour-blind safe.

For BC and Canadian forestry palettes, the `colorspace` package
includes many options.

## `patchwork` for multi-panel layouts

If you want two completely different plots side by side (not just
facets of one plot), use the `patchwork` package:

```r
library(patchwork)
p1 / p2          # stack p1 on top of p2
p1 | p2          # p1 and p2 side by side
(p1 | p2) / p3   # two-column top row, single-row bottom
```

## Interactive plots

`plotly::ggplotly(p)` converts most ggplot objects into
interactive HTML charts (hover for values, zoom, pan). Useful
for web reports.

# A glimpse of Chapter 10: From plots to maps

Chapter 10 moves from charts to **spatial data**:

- The `sf` package — *simple features* for vector spatial data
  (points, lines, polygons).
- Coordinate reference systems (CRS) and why they matter.
- Reading and writing shapefiles and GeoJSON.
- Basic spatial operations: `st_buffer`, `st_intersection`,
  `st_transform`.
- A first map of the BC air-quality stations as points.

The chart you just built showing O₃ by station becomes much
more powerful when each station is a **dot on a map of BC**.
Chapter 10 walks through that move.

# Self-assessment quiz {#sec-quiz-ch09}

```{=html}
<style>
.quiz { background: #F4F7F4; border-left: 4px solid #2D6A4F; padding: 1em 1.5em; margin: 1.5em 0; border-radius: 6px; }
.quiz .q { font-weight: 600; color: #1B4332; margin-bottom: 0.6em; }
.quiz ul.qopts { list-style: none; padding: 0; margin: 0; }
.quiz ul.qopts li { padding: 0.55em 0.9em; margin: 0.3em 0; background: white; border: 1px solid #d8e4d8; border-radius: 4px; cursor: pointer; transition: background 0.12s; }
.quiz ul.qopts li:hover { background: #E8F1E9; }
.quiz ul.qopts li.correct { background: #C8E6C9; border-color: #2D6A4F; font-weight: 600; }
.quiz ul.qopts li.wrong { background: #FFCDD2; border-color: #B05C50; }
.quiz .explanation { display: none; margin-top: 0.7em; padding: 0.7em 1em; background: #FFF8E8; border-left: 3px solid #B7791F; border-radius: 4px; font-size: 0.95em; }
.quiz .explanation.show { display: block; }
.quiz textarea { width: 100%; min-height: 4.5em; padding: 0.6em; border: 1px solid #d8e4d8; border-radius: 4px; font-family: inherit; font-size: 0.95em; resize: vertical; margin-top: 0.3em; }
.quiz .reveal-btn { background: #2D6A4F; color: white; border: none; padding: 0.45em 0.9em; margin-top: 0.5em; border-radius: 4px; cursor: pointer; font-size: 0.9em; }
.quiz .reveal-btn:hover { background: #1B4332; }
.quiz .sample { display: none; margin-top: 0.6em; padding: 0.6em 0.9em; background: #FFF8E8; border-left: 3px solid #B7791F; border-radius: 4px; font-size: 0.95em; font-style: italic; }
.quiz .sample.show { display: block; }
</style>

<div class="quiz" id="q1"><div class="q">1. The three core pieces of a ggplot are —</div><ul class="qopts"><li data-correct="false">data, file, output</li><li data-correct="true">data, aesthetic mappings, geometric objects</li><li data-correct="false">x, y, title</li><li data-correct="false">colour, size, theme</li></ul><div class="explanation">Data + aes() + geom_*() are the minimum. Scales, themes, facets, and labels are polish.</div></div>

<div class="quiz" id="q2"><div class="q">2. <code>aes()</code> in ggplot2 stands for and does what?</div><ul class="qopts"><li data-correct="false">"aesthetic" — sets the chart's colours</li><li data-correct="true">"aesthetic mapping" — maps columns in your tibble to visual properties of the chart</li><li data-correct="false">"area + estimate + scale" — a statistical function</li><li data-correct="false">"axis encoding system" — sets the axes</li></ul><div class="explanation">aes() connects data columns to visual properties (x, y, colour, fill, size, shape).</div></div>

<div class="quiz" id="q3"><div class="q">3. ggplot layers are added with —</div><ul class="qopts"><li data-correct="false">the pipe |></li><li data-correct="true">the + sign</li><li data-correct="false">commas inside ggplot()</li><li data-correct="false">a separate function for each layer</li></ul><div class="explanation">ggplot uses + (not |>) because it was designed before the pipe was popular. Read + as "and also draw".</div></div>

<div class="quiz" id="q4"><div class="q">4. <code>geom_col()</code> requires you to supply —</div><ul class="qopts"><li data-correct="false">Only x</li><li data-correct="true">Both x and y (the y value is the bar height)</li><li data-correct="false">A colour palette</li><li data-correct="false">Nothing — it counts for you</li></ul><div class="explanation"><code>geom_col</code> uses your supplied y values as bar heights. <code>geom_bar</code> counts x values for you instead.</div></div>

<div class="quiz" id="q5"><div class="q">5. <code>geom_bar()</code> (with no y aesthetic) draws —</div><ul class="qopts"><li data-correct="false">Lines</li><li data-correct="true">Bars whose heights are the counts of each x value</li><li data-correct="false">Empty bars</li><li data-correct="false">A histogram</li></ul><div class="explanation">If you have raw events (one row per occurrence), use <code>geom_bar</code> to count. If you have pre-computed counts, use <code>geom_col</code>.</div></div>

<div class="quiz" id="q6"><div class="q">6. <code>geom_histogram(binwidth = 2)</code> on a numeric column —</div><ul class="qopts"><li data-correct="false">Plots two bars</li><li data-correct="true">Buckets the values into 2-unit-wide bins and plots the count in each bin</li><li data-correct="false">Plots the first 2 values</li><li data-correct="false">Returns an error</li></ul><div class="explanation">binwidth controls the bucket size. Smaller bins = more detail; larger bins = smoother shape.</div></div>

<div class="quiz" id="q7"><div class="q">7. <code>geom_boxplot()</code> shows —</div><ul class="qopts"><li data-correct="false">The mean</li><li data-correct="true">The median (line in box), interquartile range (box), and outliers (dots)</li><li data-correct="false">The sum</li><li data-correct="false">A linear regression</li></ul><div class="explanation">The boxplot is a robust five-number summary. The line in the box is the median, not the mean.</div></div>

<div class="quiz" id="q8"><div class="q">8. <code>facet_wrap(~ region)</code> produces —</div><ul class="qopts"><li data-correct="false">A single chart</li><li data-correct="true">One panel per unique value of region (small multiples)</li><li data-correct="false">An error</li><li data-correct="false">A chart with region on the x-axis</li></ul><div class="explanation">Faceting creates small-multiple panels. Same chart, repeated per group.</div></div>

<div class="quiz" id="q9"><div class="q">9. For plotting three time series on one chart, the data should be in —</div><ul class="qopts"><li data-correct="false">Wide format (one column per series)</li><li data-correct="true">Long format (one row per series × time, with a 'series' column for the colour mapping)</li><li data-correct="false">Either, no difference</li><li data-correct="false">A list, not a tibble</li></ul><div class="explanation">Long format lets you map <code>colour = series</code> and draw all three lines in one call. Wide format requires three separate <code>geom_line</code> calls.</div></div>

<div class="quiz" id="q10"><div class="q">10. To save a ggplot as a PNG file, use —</div><ul class="qopts"><li data-correct="false">save_plot()</li><li data-correct="true">ggsave("outputs/myplot.png", width = 7, height = 5)</li><li data-correct="false">writePNG()</li><li data-correct="false">export()</li></ul><div class="explanation"><code>ggsave</code> infers the format from the extension. Always specify width and height explicitly.</div></div>

<div class="quiz" id="q11"><div class="q">11. <code>labs(title = "...", x = "...")</code> does what?</div><ul class="qopts"><li data-correct="false">Adds new data</li><li data-correct="true">Adds or replaces chart labels (title, axis labels, captions)</li><li data-correct="false">Sets the colour palette</li><li data-correct="false">Saves the file</li></ul><div class="explanation"><code>labs()</code> is the consolidated labels function — title, subtitle, x, y, colour, caption.</div></div>

<div class="quiz" id="q12"><div class="q">12. <code>geom_smooth(method = "lm")</code> overlays —</div><ul class="qopts"><li data-correct="false">A loess smoother</li><li data-correct="true">A linear regression line plus its confidence band</li><li data-correct="false">A boxplot</li><li data-correct="false">A histogram</li></ul><div class="explanation"><code>method = "lm"</code> fits a straight line. <code>method = "loess"</code> fits a flexible local-regression curve. Default for small samples is loess.</div></div>

<div class="quiz" id="q13"><div class="q">13. <code>theme_minimal()</code> is —</div><ul class="qopts"><li data-correct="false">A required base layer for every chart</li><li data-correct="true">One of several built-in clean themes; an alternative to the default grey-background theme</li><li data-correct="false">A colour palette</li><li data-correct="false">A way to hide the chart</li></ul><div class="explanation">Other built-ins: theme_classic, theme_bw, theme_void. Pick whatever matches the venue (publication, web, slides).</div></div>

<div class="quiz" id="q14"><div class="q">14. <code>scale_y_continuous(labels = label_comma())</code> does what?</div><ul class="qopts"><li data-correct="false">Logs the y-axis</li><li data-correct="true">Adds thousands separators to y-axis labels (8,264,063 instead of 8264063)</li><li data-correct="false">Reverses the axis</li><li data-correct="false">Adds a legend</li></ul><div class="explanation">From the scales package. Other useful formatters: label_percent, label_dollar, label_date.</div></div>

<div class="quiz" id="q15"><div class="q">15. Mapping a third variable to colour is done inside —</div><ul class="qopts"><li data-correct="false">labs()</li><li data-correct="true">aes(colour = third_variable)</li><li data-correct="false">theme(colour = ...)</li><li data-correct="false">geom_*(colour = ...)</li></ul><div class="explanation">Inside aes(), colour MAPS a column to the colour aesthetic (legend appears). Outside aes(), colour = "blue" SETS the colour for all elements (no legend).</div></div>

<div class="quiz" id="q16"><div class="q">16. <code>geom_line()</code> connects points —</div><ul class="qopts"><li data-correct="false">In a random order</li><li data-correct="true">In the order of the x aesthetic</li><li data-correct="false">By their colour</li><li data-correct="false">Counter-clockwise</li></ul><div class="explanation">geom_line draws in x-order. If your data is not sorted by x, the line zig-zags. Sort first or use geom_path for arbitrary order.</div></div>

<div class="quiz" id="q17"><div class="q">17. <code>position = "dodge"</code> on a bar chart —</div><ul class="qopts"><li data-correct="false">Stacks bars</li><li data-correct="true">Places bars side by side instead of on top of each other</li><li data-correct="false">Hides the bars</li><li data-correct="false">Reorders the bars</li></ul><div class="explanation">Default for geom_col with a fill aesthetic is "stack". <code>position = "dodge"</code> produces grouped bars (e.g., comparing harvested and reforested area side-by-side per year).</div></div>

<div class="quiz" id="q18"><div class="q">18. <code>facet_grid(rows ~ cols)</code> produces —</div><ul class="qopts"><li data-correct="false">A single chart</li><li data-correct="true">A two-dimensional grid of small multiples</li><li data-correct="false">A pie chart</li><li data-correct="false">A scatter plot</li></ul><div class="explanation">facet_wrap wraps panels into a flexible layout; facet_grid forces a rows × cols layout — useful when you have two grouping variables.</div></div>

<div class="quiz" id="q19"><div class="q">19. Pie charts in ggplot2 require <code>coord_polar()</code> and are —</div><ul class="qopts"><li data-correct="false">Strongly recommended for forestry data</li><li data-correct="true">Usually replaced by a bar chart, because angles are harder to compare than lengths</li><li data-correct="false">The default chart type</li><li data-correct="false">Impossible to make</li></ul><div class="explanation">Pie charts are harder to read accurately. A bar chart of the same data is almost always clearer. Use pie only if the venue requires it.</div></div>

<div class="quiz" id="q20"><div class="q">20. <strong>Reflection.</strong> Which Excel chart type did you use most often before this chapter? Which ggplot2 geom replaces it? What changes about how you make charts?</div><textarea placeholder="Type your answer here..."></textarea><button class="reveal-btn" onclick="this.nextElementSibling.classList.toggle('show')">Show common answers</button><div class="sample">Common reflections: <em>I used Excel bar charts most — geom_col replaces them, and now every label, colour, and axis is written down in code I can re-run</em> · <em>Excel scatter plots were a few clicks but I never knew which buttons set what — ggplot makes the choices explicit</em> · <em>I never used boxplots in Excel because they were hard; in ggplot they are one line</em> · <em>The biggest change is faceting — small multiples are trivial in ggplot but a manual nightmare in Excel</em>.</div></div>

<script>
(function() {
  document.querySelectorAll('.quiz').forEach(function(q) {
    q.querySelectorAll('ul.qopts li').forEach(function(opt) {
      opt.addEventListener('click', function() {
        if (q.dataset.answered) return;
        q.dataset.answered = '1';
        var isCorrect = opt.getAttribute('data-correct') === 'true';
        if (isCorrect) opt.classList.add('correct');
        else {
          opt.classList.add('wrong');
          q.querySelectorAll('ul.qopts li').forEach(function(o) {
            if (o.getAttribute('data-correct') === 'true') o.classList.add('correct');
          });
        }
        var exp = q.querySelector('.explanation');
        if (exp) exp.classList.add('show');
      });
    });
  });
})();
</script>
```

# AI as a debugging companion

::: {.callout-tip}
## Useful prompts

- *"My ggplot only draws one line, but my long-format tibble has
  three series. What did I miss in aes()?"* (Almost always:
  forgot to add `group = series` or `colour = series`.)

- *"How do I make the x-axis show dates like 'Jan 15' instead of
  '2000-01-15' in ggplot?"* (Answer: `scale_x_date(date_labels = "%b %d")`.)

- *"My chart's y-axis says 8.26e+06 instead of 8,264,063. How
  do I fix it?"* (Answer: `scale_y_continuous(labels = label_comma())`.)
:::

# Reading

- *R for Data Science* — chapters 1 *Data visualisation* and 12
  *Communication*. Free at <https://r4ds.hadley.nz/>.
- *ggplot2: Elegant Graphics for Data Analysis* (Wickham,
  Navarro, Pedersen) — the canonical reference. Free at
  <https://ggplot2-book.org/>.
- The ggplot2 cheatsheet (PDF):
  <https://rstudio.github.io/cheatsheets/data-visualization.pdf>

# Instructor notes {.unnumbered}

::: {.callout-note appearance="minimal" collapse="true"}
## Pacing and emphasis

- Spend most time on **`geom_col`, `geom_line`, `geom_histogram`,
  `geom_boxplot`, and `facet_wrap`** — these five cover 90% of
  routine forestry plotting.
- The **long-format-for-plotting** habit is the chapter's most
  important reusable insight. When a student's chart looks
  wrong, ask first: "is your data long or wide?"
- **`ggsave` matters.** Students will hand in screenshots
  instead of PNG files if you do not insist on `ggsave`.

## Materials provided

| File | Purpose |
|---|---|
| `frst232_ch09_learner.qmd` | Quarto starter with TODO chunks. |
| `frst232_ch09_solutions.qmd` | Completed solutions. **Do not distribute.** |
| `quiz_ch09.html` | Standalone interactive quiz. |

## Common stumbling points

- **`+` instead of `|>` in pipelines.** Walk through: `|>` for
  data transformation, `+` for ggplot layers.
- **`colour` outside `aes()`** instead of inside (or vice
  versa). Inside aes = data-driven; outside aes = constant.
- **Wide data plotted with `geom_line`** giving one line per
  observation instead of one per series. Fix: `pivot_longer`
  first.
- **Forgetting `group = ...`** in line charts with categorical
  x. Symptom: zig-zag lines.
:::
