---
title: "Integrated Case Study: Choosing the Right Tools"
---

```{r setup-ch10cs}
#| include: false
library(tidyverse)
library(readxl)
library(here)
library(knitr)
```

::: {.callout-note appearance="simple"}
## Data for this chapter
This chapter uses the **BC silviculture** workbook
([`bc_disturbance_reforestation.xlsx`](../data/bc_disturbance_reforestation.xlsx){download="bc_disturbance_reforestation.xlsx"}).
Click the file name to download it directly, then save it in your `data/`
folder. For the source and details, see the
[Datasets Used in This Book](../datasets.qmd) page.
:::

# Chapter goals {.unnumbered}

By the end of this case-study chapter you will be able to:

- **read** a realistic forestry or natural-resources question and decide
  what data-analysis steps are needed;
- **identify** which functions from previous chapters are most efficient
  for the task;
- **combine** skills from multiple chapters — import, inspect, clean,
  `filter`, `mutate`, `group_by`, `summarise`, `pivot`, `left_join`,
  visualize, and report;
- **justify** why you chose one function or workflow over another;
- **check** whether your results are plausible using row counts,
  missing-value checks, summary statistics, and visual inspection;
- **write** a short interpretation for a non-technical forestry audience;
- **use AI as a verification partner** without letting it decide the whole
  analysis.

# Expected output for this chapter {.unnumbered}

::: {.callout-tip appearance="default"}
## What you will hand in
A short **case-study report** (rendered from a `.qmd`) that:

1. states a **question** in plain language;
2. shows your **chosen workflow** with a one-line justification for each
   step ("I used `group_by()` because …");
3. includes **one clear figure**;
4. shows your **plausibility checks** (row counts, missing-value checks,
   summary statistics);
5. ends with a **two- to three-sentence interpretation** a non-technical
   forestry reader could follow.
:::

# Why this chapter is different

Every earlier chapter taught **one main skill at a time** — importing,
cleaning, summarising, joining, or visualizing — and told you which one to
practice. Real analysis does not arrive labelled. You are handed a
**question and a dataset**, and *you* decide which tools to reach for and
in what order.

The goal of this chapter is to move you from

> *"I know this chapter uses `summarise()`"*

to

> *"I can decide **when** `summarise()` is the right tool."*

# The decision framework — think before you code

Before writing a single line, answer these six questions:

1. **What is the unit of observation?** One row = one what (a fiscal year?
   a station? a plot)?
2. **Which columns do I actually need** to answer the question?
3. **Does anything need cleaning or recoding first** (missing values, odd
   codes, a date to parse)?
4. **What *kind* of problem is this** — a filter, a grouped summary, a
   trend over time, a reshape, or a join?
5. **Which function is the most efficient** for that kind of problem — and
   what would be a slower but still-correct alternative?
6. **How will I verify the result** (row counts, `n()`, plausible ranges,
   a quick plot)?

Use this quick map from "kind of problem" to "reach for":

| The question is about … | Reach for |
|---|---|
| keeping only some rows | `filter()` |
| a new or derived column | `mutate()` |
| one number per group | `group_by()` + `summarise()` |
| how something changes over time | a trend plot (`geom_line`/`geom_col`) |
| the table being the wrong shape | `pivot_longer()` / `pivot_wider()` |
| combining two tables | `left_join()` (check keys + row counts) |
| showing the result to a person | `ggplot2` + a clear caption |

# A worked example — the process end to end

**Question:** *Which decade saw the highest average reforestation in
British Columbia, and how does it compare with harvesting?*

**Think first.**

- *Unit of observation:* one row per fiscal year.
- *Columns needed:* `Fiscal_Year`, `Reforestation_ha`, `Harvested_ha`.
- *Cleaning:* none needed — this file is small and complete.
- *Kind of problem:* a **derived variable** (decade) + a **grouped
  summary**. Not a join, not a reshape.
- *Most efficient tools:* `mutate()` to build `decade`, then
  `group_by(decade) |> summarise()`.
- *Verify:* check the number of years per decade and that the values are in
  a plausible range (BC reforests very roughly 150,000–250,000 ha/yr).

**Then code:**

```{r cs-worked}
bc <- read_excel(here("data", "bc_disturbance_reforestation.xlsx"),
                 sheet = "Data")

by_decade <- bc |>
  mutate(decade = floor(Fiscal_Year / 10) * 10) |>
  group_by(decade) |>
  summarise(
    n_years            = n(),
    mean_reforested_ha = round(mean(Reforestation_ha, na.rm = TRUE)),
    mean_harvested_ha  = round(mean(Harvested_ha,     na.rm = TRUE)),
    .groups = "drop"
  ) |>
  arrange(desc(mean_reforested_ha))

by_decade |> kable()
```

**Verify before trusting it:** `n_years` should be small (about 10 per full
decade, fewer for partial decades at the ends), and the hectare figures
should sit in the hundred-thousands — both true above. A quick plot
confirms the pattern:

```{r cs-worked-plot}
#| fig-alt: "Bar chart of mean annual reforested area by decade in British
#|   Columbia, showing which decade reforested the most on average."
by_decade |>
  ggplot(aes(x = factor(decade), y = mean_reforested_ha)) +
  geom_col(fill = "#2D6A4F") +
  scale_y_continuous(labels = scales::label_comma()) +
  labs(x = "Decade", y = "Mean reforested area (ha/yr)",
       title = "Average annual reforestation by decade, BC",
       caption = "Source: Environmental Reporting BC") +
  theme_minimal()
```

**Interpret (for a non-technical reader):** *"On average, the decade with
the highest annual reforestation was the one shown tallest above.
Reforestation broadly tracks harvesting, but the two are not equal in every
decade — a gap worth a second look."*

Notice what we did **not** do: no join (only one table), no reshape (the
shape already answered the question). Choosing *not* to use a tool is part
of choosing the right tool.

# Choose the function — activities

For each task, decide the workflow **before** you write code.

**Activity 1.** You want **one row per forest management category** showing
its **total area** and **average age class**.

1. Which function groups the data?
2. Which function calculates the summaries?
3. Which function sorts the result?
4. Which plot type would communicate it best?
5. What checks should you run before trusting the result?

**Activity 2.** You have a wide table with **one column per year** and want
to **plot area over time**.

- Is this a filtering, summarising, reshaping, joining, or visualization
  problem *first*? Which function do you reach for, and why?

**Activity 3.** You have a **tree table** and a separate **species lookup
table**, and you want the species *name* (not code) on every tree row.

- Which single verb does this? What must you check immediately afterward?

# Integrated case-study questions — decide the tools

These need skills from several chapters at once. **Do not** jump to code —
work through the framework questions first, then justify each step.

**Question A.** *Which forest type or management category shows the largest
change over time?*
Decide whether you need `filter()`, `mutate()`, `group_by()`,
`summarise()`, `arrange()`, `pivot_*()`, `left_join()`, and/or `ggplot()` —
and in what order.

**Question B.** *Are years with more harvesting also years with more
reforestation?*
Ask: what is the unit of observation? Is this a trend, a comparison, or a
join problem? How would you show it, and how would you check the result is
plausible?

For every question, be ready to answer:

- What is the **unit of observation**?
- Which **columns** are needed?
- Do we need to **clean or recode** any values first?
- Is this a **grouped summary, a comparison, a trend, or a join**?
- What function is **most efficient** here — and what is a slower but valid
  alternative?
- How can we **verify** the result?

# Verify before you trust it

Run the checks that fit each step you took:

| After you … | Check that … |
|---|---|
| import | rows, columns, and units look right; note missing values |
| `filter()` | the row count dropped by a sensible amount |
| `summarise()` | group sizes (`n()`) are reasonable and values are plausible |
| `left_join()` | the row count did **not** unexpectedly grow; note unmatched rows |
| plot | the figure answers the question and axis labels include units |

# AI as a verification partner — not the analyst

In this chapter especially, AI is a **checking and debugging** partner, not
the decision-maker. It is easy for an assistant to pick a function *before*
reading the question carefully, or to assume a column that is not in your
file.

- Decide the **workflow yourself** first; then ask AI to critique it.
- Ask the AI to **explain**, not just produce code, and to list
  **verification steps**.
- Watch for AI that **chose a tool too quickly**, ignored the unit of
  observation, or **invented a column, function, or value**.

> The goal is not to make AI do the work. The goal is to use AI to help you
> understand, check, and improve your own reasoning.

# Self-check quiz

::: {.content-visible when-format="pdf"}
*The self-assessment quiz is interactive: it appears in the online (HTML)
book, and you can also take it on Canvas. This PDF does not display the
interactive quiz questions.*
:::

<div class="quiz" id="cs-q1" data-hints="One row per group is the tell.||Which pair collapses rows into per-group values?||group_by() then summarise().">
  <div class="q">1. You want the mean reforested area <em>for each decade</em>. The most efficient tools are —</div>
  <ul class="qopts">
    <li data-correct="false"><code>filter()</code> then <code>arrange()</code></li>
    <li data-correct="true"><code>mutate()</code> a decade column, then <code>group_by(decade) |&gt; summarise()</code></li>
    <li data-correct="false"><code>pivot_wider()</code></li>
    <li data-correct="false"><code>left_join()</code></li>
  </ul>
  <div class="explanation">One value per group = <code>group_by()</code> + <code>summarise()</code>; the decade first needs <code>mutate()</code>. No second table, so no join; the shape is already fine, so no pivot.</div>
</div>

<div class="quiz" id="cs-q2" data-hints="You are combining two tables on a shared column.||A join attaches columns from another table.||left_join, then check row counts.">
  <div class="q">2. You need the species <em>name</em> from a lookup table added to every tree row. You should —</div>
  <ul class="qopts">
    <li data-correct="false">reshape the tree table with <code>pivot_longer()</code></li>
    <li data-correct="true"><code>left_join()</code> the lookup on the species key, then check the row count did not grow</li>
    <li data-correct="false"><code>summarise()</code> the tree table</li>
    <li data-correct="false">add the names by hand</li>
  </ul>
  <div class="explanation">Combining two tables on a shared key is a join. After a <code>left_join()</code>, verify the row count is unchanged — a non-unique key would multiply rows.</div>
</div>

<div class="quiz" id="cs-q3" data-hints="Ask what could be wrong before trusting AI.||AI can assume data it never saw.||Check for invented columns/functions/values.">
  <div class="q">3. An AI suggests code that uses a column your file does not contain. This is an example of —</div>
  <ul class="qopts">
    <li data-correct="false">a reproducible workflow</li>
    <li data-correct="true">a hallucination — verify against your actual data and the documentation</li>
    <li data-correct="false">a valid shortcut you should submit</li>
    <li data-correct="false">a facet</li>
  </ul>
  <div class="explanation">When AI invents a column, function, or value, that is a hallucination. Always check its output against your real data and the package documentation before using it.</div>
</div>

# Done

When your case-study `.qmd` renders cleanly — question, justified workflow,
one figure, plausibility checks, and a plain-language interpretation —
you have done the core skill of this course: **turning a real question into
a verified, reproducible answer.**
