---
title: "Spatial Data in R with sf"
---

```{r setup-ch10}
#| include: false
library(tidyverse)
library(sf)
library(here)
library(knitr)
```

# Chapter goals {.unnumbered}

By the end of this chapter you will be able to:

- Distinguish **vector** spatial data (points, lines, polygons)
  from **raster** spatial data (grids), and explain where each
  fits in a forestry workflow.
- Explain the role of a **CRS** (coordinate reference system)
  and recognize three common ones: WGS84 (EPSG:4326), BC Albers
  (EPSG:3005), and Web Mercator (EPSG:3857).
- Build an **sf** point object from a tibble of latitude and
  longitude using `st_as_sf()`.
- Read and write spatial files with `st_read()` and `st_write()`
  for shapefiles, GeoJSON, and GeoPackage formats.
- Reproject between CRSs with `st_transform()`.
- Compute basic spatial operations: **`st_distance()`**,
  **`st_buffer()`**, **`st_intersection()`**, and **`st_area()`**.
- Plot spatial data inline in a Quarto report with
  **`geom_sf()`** (a ggplot2 geom).
- Recognize that **`sf` objects are tibbles** with a special
  geometry column — every dplyr verb still works.

# Expected output for this chapter {.unnumbered}

::: {.callout-tip appearance="default"}
## What you will hand in

1. A **rendered Quarto HTML report**
   (`frst232_ch10_learner.html`) containing:
   - The Metro Vancouver air-quality stations as an **sf point
     object** built from the lat/lon tibble provided in this
     chapter.
   - A simple **`geom_sf()` plot** of the stations.
   - The **pairwise distance matrix** (in metres) between every
     pair of stations.
   - A **5 km buffer** drawn around any one station, with both
     the original point and the buffer polygon plotted on the
     same chart.
   - One example of a **CRS transformation** (e.g., from WGS84
     to BC Albers) with a one-sentence note on why you would
     reproject.
2. The corresponding **`.qmd` source file**.
3. A short **group-lab worksheet** (or screenshot) submitted to
   the course site.
:::

# Where we are

Through Chapter 9 you turned tibbles into charts. Chapter 10
turns tibbles into **maps**. The same air-quality stations you
have been working with all term — Burnaby South, Coquitlam,
Langley Central, etc. — are not just rows in a tibble. They are
**points in space** with latitudes and longitudes. Once R knows
that, you can:

- Compute distances between them.
- Buffer them by a radius.
- Intersect them with watersheds, forest-district polygons, or
  fire perimeters.
- Plot them on a map alongside any other spatial layer.

This is what *"spatial data analysis"* means in practice.

# Installing sf

The `sf` package depends on three system libraries: **GDAL**,
**GEOS**, and **PROJ**. On Posit Cloud, Windows, and Mac these
are bundled — `install.packages("sf")` just works. On a fresh
Linux install you may need to install the system libraries
separately (`sudo apt install libgdal-dev libgeos-dev libproj-dev`).

```r
install.packages("sf")
library(sf)
```

If you cannot get `sf` to install locally, **switch to Posit
Cloud for this chapter**. The cloud environment has everything
pre-installed.

# What is an `sf` object?

An `sf` object is **a tibble with one extra column called
`geometry`**, plus a CRS attached. Every regular dplyr verb
(`filter`, `select`, `mutate`, `summarise`, joins) still works
on it. The geometry column is *sticky* — most operations preserve
it automatically.

That single design decision — *"spatial data is just a tibble"* —
is why `sf` integrates so cleanly with the rest of the tidyverse.

# The two spatial paradigms — vector and raster

| | Vector | Raster |
|---|---|---|
| What it is | Discrete shapes: points, lines, polygons | A grid of cells, each with a value |
| Examples | Monitoring stations (points), rivers (lines), forest-district boundaries (polygons) | Elevation, canopy cover, satellite imagery |
| R package | `sf` (this chapter) | `terra` (a separate, more advanced topic) |
| File formats | Shapefile (`.shp`), GeoJSON (`.geojson`), GeoPackage (`.gpkg`) | GeoTIFF (`.tif`), NetCDF (`.nc`) |
| Forestry uses | Cut blocks, stream networks, ownership boundaries, sample plots | DEM (elevation), forest cover, fire scars from satellite |

This chapter covers **vector** spatial data only. Raster in R
(`terra` package) is a sequel topic, not required for FRST 232.

# Coordinate Reference Systems (CRS) — the most-skipped step

Every spatial dataset has a **CRS** — the coordinate system its
geometries are expressed in. Common ones for BC forestry work:

| CRS | EPSG code | Units | Where used |
|---|---|---|---|
| WGS84 (latitude / longitude) | **4326** | degrees | GPS, web data, GeoJSON from internet |
| BC Albers Equal Area | **3005** | metres | BC Government data, all distance/area calculations in BC |
| Web Mercator | **3857** | metres | Online maps (Google Maps, Leaflet basemaps) |
| NAD83 / UTM Zone 10N | **26910** | metres | Field GPS in southern BC |

Two rules:

1. **A CRS is meaningless without units.** Lat/lon is in degrees;
   BC Albers is in metres. Asking *"what is the distance between
   these two points?"* requires a metric CRS.
2. **Two layers must share a CRS to be combined.** If your
   stations are in WGS84 and a forest-district polygon is in BC
   Albers, you must reproject one of them first.

`st_transform(data, crs = 3005)` reprojects to a target CRS.
Always check `st_crs(data)` after loading any new file.

# Build an `sf` object from a tibble

The most common starting point: you have a tibble with `latitude`
and `longitude` columns (e.g., GPS-collected field plots, station
locations). Convert it to `sf` with `st_as_sf()`:

```{r make-sf}
# Approximate Metro Vancouver air-quality station coordinates
stations <- tribble(
  ~location,                          ~lon,        ~lat,
  "Burnaby South",                    -122.985,    49.215,
  "Coquitlam Douglas College",        -122.798,    49.252,
  "Langley Central",                  -122.660,    49.105,
  "Maple Ridge Golden Ears School",   -122.620,    49.215,
  "New Westminster Sapperton Park",   -122.890,    49.235,
  "North Delta",                      -122.910,    49.135,
  "North Vancouver Second Narrows",   -123.015,    49.295,
  "Pitt Meadows Airport",             -122.700,    49.215,
  "Port Coquitlam North",             -122.770,    49.265,
  "Port Moody Rocky Point Park",      -122.860,    49.290
)

stations_sf <- stations |>
  st_as_sf(coords = c("lon", "lat"), crs = 4326)

stations_sf
```

What changed:

- The `lon` and `lat` columns disappeared.
- A new column **`geometry`** appeared, holding the point for
  each row.
- A `Geometry type:`, `Dimension:`, `Bounding box:`, and
  `Geodetic CRS:` header now appears on print.

The tibble still has 10 rows; you can still call `filter()`,
`select()`, `mutate()`, `summarise()`, and the joins on it.

::: {.callout-tip}
## `coords = c("lon", "lat")` — order matters

The first element is the **x** axis (longitude), the second is
**y** (latitude). Many beginners write `c("lat", "lon")` —
common bug, easy to spot (your points end up off the coast of
Somalia, where lat 49 swapped with lon −122 plots).
:::

# Reproject — from WGS84 to BC Albers

For any distance or area calculation in BC, reproject to **BC
Albers (EPSG:3005)**, which has units in metres:

```{r reproject}
stations_bcalbers <- stations_sf |>
  st_transform(crs = 3005)

st_crs(stations_bcalbers)$Name      # human-readable name
st_bbox(stations_bcalbers)           # bounding box in metres
```

Note the bounding-box numbers: they are now ~1 million metres
across (BC Albers' origin is far to the west), and the units are
metres, not degrees.

# `st_distance()` — pairwise distances

Once your data is in a metric CRS, distance computations are
trivial:

```{r distances}
distance_matrix <- st_distance(stations_bcalbers)
# Result is a 10 x 10 matrix of distances in metres
distance_matrix[1:4, 1:4]
```

To turn the matrix into a tidy distance tibble for plotting:

```{r distance-tidy}
dist_df <- as_tibble(distance_matrix,
                     .name_repair = ~ stations$location) |>
  mutate(from = stations$location) |>
  pivot_longer(-from, names_to = "to", values_to = "distance_m") |>
  mutate(distance_km = as.numeric(distance_m) / 1000) |>
  filter(from != to)

dist_df |>
  arrange(desc(distance_km)) |>
  head(5)
```

The pair with the longest distance is typically *North Vancouver
Second Narrows* to *Langley Central* — opposite corners of Metro
Vancouver.

# `st_buffer()` — circles around points

`st_buffer(data, dist)` returns a polygon for each input
geometry, expanded by `dist` units. In a metric CRS, `dist` is
in metres.

```{r buffer}
burnaby_buffer <- stations_bcalbers |>
  filter(location == "Burnaby South") |>
  st_buffer(dist = 5000)   # 5 km

burnaby_buffer
```

Practical use: *"which forest stands are within 5 km of a
proposed road?"* Buffer the road by 5 km, then intersect with
the stand polygons. That is a one-line spatial-overlay analysis.

# `st_intersection()` — overlay two layers

`st_intersection(a, b)` returns the geometric overlap of two
spatial datasets. Useful for *"which stations are inside this
buffer?"* and similar overlay questions:

```{r intersection}
inside_buffer <- st_intersection(stations_bcalbers, burnaby_buffer |> select(geometry))
inside_buffer$location
```

Returns the stations that fall inside the 5 km buffer around
Burnaby South — typically a small handful (New Westminster
Sapperton Park, North Delta).

# `st_area()` — area of a polygon

For polygon geometries, `st_area()` returns the area in the units
of the current CRS (square metres for BC Albers):

```{r area}
buffer_area_m2 <- st_area(burnaby_buffer)
buffer_area_m2

# In hectares (1 ha = 10,000 m²)
as.numeric(buffer_area_m2) / 10000
```

A 5 km buffer is a circle of radius 5,000 m, so area = π × 5000²
≈ 78.5 million m² ≈ 7,854 ha. The numbers should match (within
rounding for the BC Albers projection).

# Reading and writing files

## Reading

```r
# Shapefile (the .shp is the main file; .shx, .dbf, .prj must be alongside)
districts <- st_read(here("data", "forest_districts.shp"))

# GeoJSON
parks <- st_read(here("data", "parks.geojson"))

# GeoPackage (modern, single-file format — preferred)
fire <- st_read(here("data", "fire_perimeters.gpkg"))
```

`st_read()` automatically detects the format from the extension
and applies the CRS recorded in the file's metadata. After every
read, **always check `st_crs(data)`** to confirm what you got.

## Writing

```r
st_write(stations_sf, here("outputs", "stations.geojson"))
st_write(stations_sf, here("outputs", "stations.gpkg"))
```

**Recommendation:** prefer **GeoPackage** (`.gpkg`) for new
files. Shapefiles have a 10-character column-name limit, no
field-name spaces, and split a single dataset across four files —
all annoyances that GeoPackage fixes.

# Plotting with `geom_sf()`

`geom_sf()` is a ggplot2 geom that draws sf geometries. It works
the same as any other geom — no new grammar to learn:

```{r geom-sf-basic}
ggplot(stations_sf) +
  geom_sf(size = 3, colour = "#2D6A4F") +
  labs(title = "Metro Vancouver Air-Quality Monitoring Stations") +
  theme_minimal()
```

`geom_sf()` is CRS-aware: even though the data is in WGS84
(degrees), ggplot draws the axes correctly.

## Adding the buffer as a layer

```{r geom-sf-with-buffer}
ggplot() +
  geom_sf(data = burnaby_buffer |> st_transform(4326),
          fill = "#52B788", alpha = 0.3, colour = "#1B4332") +
  geom_sf(data = stations_sf, size = 3, colour = "#1B4332") +
  geom_sf_text(data = stations_sf, aes(label = location),
               size = 3, nudge_y = 0.012, check_overlap = TRUE) +
  labs(title = "5 km buffer around Burnaby South",
       caption = "Source: Environmental Reporting BC; CRS: WGS84") +
  theme_minimal()
```

Notice: the buffer was computed in BC Albers (metric), then
reprojected back to WGS84 for plotting alongside the lat/lon
stations. **Always reproject before combining layers.**

# A real spatial file: the CSRD parks layer

Everything so far used a small synthetic tibble. Now we read a
**real** open-data file: the **park point locations** published
by the Columbia Shuswap Regional District (CSRD). It is a single
GeoJSON file — no shapefile bundle, no repository clone. (The
download is scripted in `scripts/download_csrd_parks.R`.)

```{r csrd-read}
parks <- st_read(here("data", "csrd_parks.geojson"), quiet = TRUE)

# Always inspect after reading
nrow(parks)                  # how many features?
st_geometry_type(parks)[1]   # points, lines, or polygons?
st_crs(parks)$input          # which CRS did the file declare?
st_bbox(parks)               # bounding box (here: degrees, WGS84)
```

The file has `r nrow(parks)` park points in WGS84. Before any
distance or area work we reproject to **BC Albers (EPSG:3005)**,
exactly as we did for the stations:

```{r csrd-reproject}
parks_bcalbers <- parks |> st_transform(3005)
st_crs(parks_bcalbers)$Name   # confirm the reprojection
st_bbox(parks_bcalbers)       # bounding box now in metres
```

Now a **static thematic map**, coloured by park type, using the
`geom_sf()` grammar from above:

```{r csrd-map}
#| fig-alt: "Map of Columbia Shuswap Regional District park point
#|   locations coloured by park type, spread across the region."
ggplot(parks) +
  geom_sf(aes(colour = ParkType), size = 2, alpha = 0.85) +
  scale_colour_viridis_d() +
  labs(title = "CSRD parks by type",
       colour = "Park type",
       caption = "Source: Columbia Shuswap Regional District Open Data") +
  theme_minimal()
```

This is the static counterpart to the interactive `leaflet` map
you will build in Chapter 11 — the same data, the same question,
two presentation choices.

# Downloading BC data — the BC Data Catalogue

Most BC forestry spatial data is freely available from the **BC
Data Catalogue**: <https://catalogue.data.gov.bc.ca/>.

Useful starting layers:

| Layer | Catalogue search | Format |
|---|---|---|
| BC provincial boundary | "BC boundary" | Shapefile, GeoJSON |
| Forest districts | "natural resource districts" | Shapefile |
| Forest tenure cut blocks | "consolidated cutblocks" | Shapefile |
| Fire perimeters | "fire perimeters historical" | Shapefile, GeoJSON |
| Watersheds | "FWA watershed groups" | Shapefile |

For a working forestry analyst, this catalogue is the most
important single bookmark in BC. Every dataset is published
under the **Open Government Licence — British Columbia** and
distributes as a shapefile or GeoJSON.

To use one of these in your project:

1. Download from the catalogue.
2. Unzip into `data/` inside your `frst232/` project.
3. Read with `st_read(here("data", "filename.shp"))`.
4. Check `st_crs(data)` and reproject to BC Albers if needed.
5. Plot with `geom_sf()` or combine with other layers using
   `st_intersection()`.

# Spatial joins — `st_join()`

When you have **two layers** and want to attach attributes of
one to the other based on spatial position, use `st_join()`:

```r
# Suppose `districts_sf` is a polygon layer of forest districts.
# Add the district name to each station based on which district it falls in:
stations_with_district <- st_join(stations_sf, districts_sf, join = st_within)
```

`st_within` means *"join station to district where the station
is inside the district"*. Other join predicates: `st_intersects`,
`st_touches`, `st_overlaps`, `st_nearest_feature`.

A spatial join is the GIS equivalent of `left_join()` from
Chapter 6 — except the key is **geometric position**, not a
shared column.

# `sf` objects are tibbles — all dplyr verbs work

```{r dplyr-on-sf}
# filter — works
stations_sf |> filter(str_starts(location, "P"))

# select — works (geometry stays sticky)
stations_sf |> select(location)

# summarise — also works (returns a multi-point or aggregated geometry)
stations_sf |>
  mutate(group = if_else(str_detect(location, "Vancouver"), "Vancouver", "Other")) |>
  group_by(group) |>
  summarise(n = n(), .groups = "drop")
```

The geometry column comes along for the ride. This is the single
biggest reason `sf` works so well: nothing new to learn beyond
the spatial operations themselves.

# Active learning

## Activity 1 — Build the sf object

Build `stations_sf` from the `stations` tibble. Confirm
`nrow(stations_sf)` is 10 and `st_crs(stations_sf)$Name` says
`"WGS 84"`.

## Activity 2 — Reproject

Reproject `stations_sf` to BC Albers (EPSG:3005). What units
does `st_bbox()` now report?

## Activity 3 — Distances

Compute the pairwise distance matrix in metres. What is the
distance (in km) from `Burnaby South` to `Langley Central`?

## Activity 4 — Buffer

Build a 10 km buffer around `North Vancouver Second Narrows`.
Which stations fall inside it? Use `st_intersection()` to find
out.

## Activity 5 — Plot

Plot all the stations with `geom_sf()` and add a title. Save
the plot as a 7×5 inch PNG.

# Group lab activity

::: {.callout-note appearance="default"}
## Lab: Build a station map together

Work in groups of 3 or 4. Each member starts from the same
`stations` tibble.

**Task 1 — Build the sf object.** Each member runs `st_as_sf()`
to build `stations_sf`. Confirm everyone gets the same result.

**Task 2 — Pick a buffer radius.** Discuss: for a smoke-exposure
analysis, what radius makes sense around each station?
(Often 5 km for community exposure, larger for regional
estimates.) Pick a number.

**Task 3 — Buffer all stations.** Apply `st_buffer()` to all 10
stations at the chosen radius.

**Task 4 — Find overlapping pairs.** Use `st_intersects()` to
find pairs of stations whose buffers overlap. Hint:
`st_intersects(buffers_sf, buffers_sf)` returns a list of
indices.

**Task 5 — Plot it.** Build a single `geom_sf()` plot showing
all the buffer polygons (with `fill` mapped to a colour) and
all the station points on top.

**Task 6 — Caption.** Write a one-sentence caption explaining
what the chart shows.

**Task 7 — Submit.** Submit the rendered chart, the buffer
radius chosen, the code, and the caption.
:::

# Exercises

1. **Build sf.** Build `stations_sf` from the `stations` tibble.
   Confirm with `class(stations_sf)` that the result is an `sf`
   object.

2. **Reproject.** Reproject `stations_sf` to BC Albers
   (EPSG:3005). Compare `st_bbox()` of both — what are the units
   of each?

3. **Pairwise distances.** Compute and print the pairwise
   distance matrix in **kilometres**. What is the longest
   distance? Between which two stations?

4. **Buffer + intersection.** Build a 7.5 km buffer around
   `Burnaby South`. Use `st_intersection()` to find which other
   stations fall inside it. Report the list of station names.

5. **`geom_sf` plot.** Plot all stations with `geom_sf()`, then
   add labels with `geom_sf_text(aes(label = location), size =
   3)`. Save as a PNG.

6. **Write a GeoJSON.** Save `stations_sf` to
   `outputs/stations.geojson` with `st_write()`. Open the
   resulting file in a text editor — what does the JSON
   structure look like?

7. **Area of a buffer.** Compute the area in hectares of the
   5 km buffer around any station. Confirm it is approximately
   π × 5000² / 10,000 ≈ 7,854 ha.

8. **Optional, harder.** Download the BC provincial boundary
   from the BC Data Catalogue. Load it with `st_read()`. Plot
   it as a polygon with `geom_sf()`, then overlay your stations
   on top.

# Optional, advanced

## Spatial joins

A spatial join attaches attributes of one layer to another
based on geometric predicates. Most common predicates:

- `st_within()` — point inside polygon
- `st_intersects()` — any geometric overlap
- `st_touches()` — share a boundary but do not overlap
- `st_nearest_feature()` — nearest neighbour

```r
st_join(points, polygons, join = st_within)
```

## Caching downloaded files

After you `st_read()` a downloaded shapefile, save it locally as
a GeoPackage:

```r
districts <- st_read(here("data", "districts.shp"))
st_write(districts, here("data", "districts.gpkg"), delete_dsn = TRUE)
```

Subsequent reads from `.gpkg` are faster and avoid the
shapefile's quirks.

## Validating geometries

Sometimes a downloaded polygon layer has invalid geometries
(self-intersections, slivers). Use:

```r
districts_valid <- districts |> st_make_valid()
```

before any overlay operations.

# A glimpse of Chapter 11: From static maps to interactive maps

In Chapter 11 you take the same `stations_sf` object and put it
on an **interactive web map** using the **leaflet** package.
Pan, zoom, click for popups — the difference between a static
PNG and a slippy map is the difference between a report and a
tool.

| Static (Ch 10, geom_sf) | Interactive (Ch 11, leaflet) |
|---|---|
| Fixed extent | Pan + zoom |
| No popups | Click a point for info |
| Print-friendly | Web-deliverable |
| One basemap | Many basemap choices (street, satellite, terrain) |

The two are complementary. Static maps for reports; interactive
maps for dashboards and field handoff.

# Self-assessment quiz {#sec-quiz-ch10}

```{=html}
<style>
.quiz { background: #F4F7F4; border-left: 4px solid #2D6A4F; padding: 1em 1.5em; margin: 1.5em 0; border-radius: 6px; }
.quiz .q { font-weight: 600; color: #1B4332; margin-bottom: 0.6em; }
.quiz ul.qopts { list-style: none; padding: 0; margin: 0; }
.quiz ul.qopts li { padding: 0.55em 0.9em; margin: 0.3em 0; background: white; border: 1px solid #d8e4d8; border-radius: 4px; cursor: pointer; transition: background 0.12s; }
.quiz ul.qopts li:hover { background: #E8F1E9; }
.quiz ul.qopts li.correct { background: #C8E6C9; border-color: #2D6A4F; font-weight: 600; }
.quiz ul.qopts li.wrong { background: #FFCDD2; border-color: #B05C50; }
.quiz .explanation { display: none; margin-top: 0.7em; padding: 0.7em 1em; background: #FFF8E8; border-left: 3px solid #B7791F; border-radius: 4px; font-size: 0.95em; }
.quiz .explanation.show { display: block; }
.quiz textarea { width: 100%; min-height: 4.5em; padding: 0.6em; border: 1px solid #d8e4d8; border-radius: 4px; font-family: inherit; font-size: 0.95em; resize: vertical; margin-top: 0.3em; }
.quiz .reveal-btn { background: #2D6A4F; color: white; border: none; padding: 0.45em 0.9em; margin-top: 0.5em; border-radius: 4px; cursor: pointer; font-size: 0.9em; }
.quiz .reveal-btn:hover { background: #1B4332; }
.quiz .sample { display: none; margin-top: 0.6em; padding: 0.6em 0.9em; background: #FFF8E8; border-left: 3px solid #B7791F; border-radius: 4px; font-size: 0.95em; font-style: italic; }
.quiz .sample.show { display: block; }
</style>

<div class="quiz" id="q1"><div class="q">1. The two main paradigms for representing spatial data are —</div><ul class="qopts"><li data-correct="false">2D and 3D</li><li data-correct="true">vector (points, lines, polygons) and raster (grid of cells)</li><li data-correct="false">SVG and PDF</li><li data-correct="false">XML and JSON</li></ul><div class="explanation">Vector for discrete shapes; raster for continuous fields like elevation or satellite imagery.</div></div>

<div class="quiz" id="q2"><div class="q">2. An <code>sf</code> object in R is —</div><ul class="qopts"><li data-correct="false">A completely new data structure unrelated to tibbles</li><li data-correct="true">A tibble with one extra column called geometry, plus a CRS</li><li data-correct="false">A shapefile</li><li data-correct="false">A list of coordinates</li></ul><div class="explanation">That tibble-like design is why every dplyr verb still works on an sf object.</div></div>

<div class="quiz" id="q3"><div class="q">3. The CRS code for WGS84 (latitude / longitude) is —</div><ul class="qopts"><li data-correct="false">3005</li><li data-correct="true">4326</li><li data-correct="false">3857</li><li data-correct="false">26910</li></ul><div class="explanation">EPSG:4326 = WGS84. EPSG:3005 = BC Albers. EPSG:3857 = Web Mercator.</div></div>

<div class="quiz" id="q4"><div class="q">4. The CRS code for BC Albers (the standard for BC Government data) is —</div><ul class="qopts"><li data-correct="false">4326</li><li data-correct="true">3005</li><li data-correct="false">3857</li><li data-correct="false">26910</li></ul><div class="explanation">EPSG:3005 — equal-area projection used for nearly all BC forestry spatial data. Units are metres.</div></div>

<div class="quiz" id="q5"><div class="q">5. You have point data in lat/lon (degrees). To compute the distance between two points in metres, you must —</div><ul class="qopts"><li data-correct="false">Use Pythagoras on the degrees</li><li data-correct="true">Reproject to a metric CRS (e.g., BC Albers, EPSG:3005), then use st_distance()</li><li data-correct="false">Convert degrees to metres manually with 111000</li><li data-correct="false">Use raster algebra</li></ul><div class="explanation">Distance in degrees is meaningless because degrees of longitude vary with latitude. Reproject to a metric CRS first.</div></div>

<div class="quiz" id="q6"><div class="q">6. <code>st_as_sf(coords = c("lon", "lat"), crs = 4326)</code> tells R that —</div><ul class="qopts"><li data-correct="false">lon is y and lat is x</li><li data-correct="true">lon (x) and lat (y) columns in the tibble define point geometries in WGS84</li><li data-correct="false">The data should be plotted but not stored</li><li data-correct="false">The data is in BC Albers</li></ul><div class="explanation">Order matters: first is x (longitude), second is y (latitude). Swapping them is a common bug.</div></div>

<div class="quiz" id="q7"><div class="q">7. <code>st_transform(data, crs = 3005)</code> does what?</div><ul class="qopts"><li data-correct="false">Adds an attribute column</li><li data-correct="true">Reprojects the data to BC Albers, transforming the coordinates so they are still in the right place but in metres</li><li data-correct="false">Deletes the geometry</li><li data-correct="false">Plots the data</li></ul><div class="explanation">Reprojection is mathematical — the geometries move from one coordinate system to another while keeping the real-world position.</div></div>

<div class="quiz" id="q8"><div class="q">8. <code>st_buffer(point, dist = 5000)</code> in a CRS with metres returns —</div><ul class="qopts"><li data-correct="false">A point 5,000 metres away</li><li data-correct="true">A polygon representing a circle of radius 5,000 metres around the point</li><li data-correct="false">A line of length 5,000 m</li><li data-correct="false">An error</li></ul><div class="explanation">In a metric CRS, dist is in metres. So 5000 = 5 km buffer.</div></div>

<div class="quiz" id="q9"><div class="q">9. <code>st_intersection(layer_a, layer_b)</code> returns —</div><ul class="qopts"><li data-correct="false">The union of both layers</li><li data-correct="true">The geometric overlap of the two layers</li><li data-correct="false">Only the attributes of layer_b</li><li data-correct="false">An error</li></ul><div class="explanation">Intersection is the geometric AND — the area, line, or points where both layers overlap.</div></div>

<div class="quiz" id="q10"><div class="q">10. <code>geom_sf()</code> is —</div><ul class="qopts"><li data-correct="false">A completely separate plotting system from ggplot2</li><li data-correct="true">A ggplot2 geom that draws sf geometries (points, lines, polygons) using ggplot's grammar</li><li data-correct="false">A function in the leaflet package</li><li data-correct="false">A way to save sf objects</li></ul><div class="explanation"><code>geom_sf()</code> uses the same ggplot grammar as every other geom — same +, same labs(), same themes.</div></div>

<div class="quiz" id="q11"><div class="q">11. To read a shapefile in R, you use —</div><ul class="qopts"><li data-correct="false">read_csv()</li><li data-correct="true">st_read()</li><li data-correct="false">read_excel()</li><li data-correct="false">readRDS()</li></ul><div class="explanation">st_read() reads any vector spatial format and auto-detects from the extension.</div></div>

<div class="quiz" id="q12"><div class="q">12. The recommended new-data file format (over shapefile) is —</div><ul class="qopts"><li data-correct="false">CSV with WKT</li><li data-correct="true">GeoPackage (.gpkg) — single file, no column-name limits</li><li data-correct="false">Excel</li><li data-correct="false">Plain text</li></ul><div class="explanation">GeoPackage is a modern single-file format that fixes shapefile's limitations (10-char column names, no spaces, 4 separate files per layer).</div></div>

<div class="quiz" id="q13"><div class="q">13. A spatial join (st_join) is the GIS equivalent of —</div><ul class="qopts"><li data-correct="false">filter()</li><li data-correct="true">left_join() — but matched on geometric position instead of a shared column</li><li data-correct="false">summarise()</li><li data-correct="false">pivot_wider()</li></ul><div class="explanation">Where left_join matches on a shared key column, st_join matches on geometric predicates like "within" or "intersects".</div></div>

<div class="quiz" id="q14"><div class="q">14. You loaded a new spatial file. What is the first command you should run?</div><ul class="qopts"><li data-correct="false">plot(data)</li><li data-correct="false">summary(data)</li><li data-correct="true">st_crs(data) to confirm the coordinate reference system</li><li data-correct="false">nrow(data)</li></ul><div class="explanation">Without knowing the CRS, you cannot safely combine, transform, or compute distances. Always check the CRS first.</div></div>

<div class="quiz" id="q15"><div class="q">15. <code>st_area()</code> on a polygon in BC Albers returns area in —</div><ul class="qopts"><li data-correct="false">degrees squared</li><li data-correct="true">square metres</li><li data-correct="false">hectares</li><li data-correct="false">acres</li></ul><div class="explanation">BC Albers' units are metres, so areas come out in m². To convert to hectares, divide by 10,000.</div></div>

<div class="quiz" id="q16"><div class="q">16. dplyr verbs on an sf object —</div><ul class="qopts"><li data-correct="false">Strip the geometry column</li><li data-correct="true">Keep the geometry column automatically (it is "sticky")</li><li data-correct="false">Do not work — you need separate spatial verbs</li><li data-correct="false">Convert the data to a base R data.frame</li></ul><div class="explanation">The geometry column survives filter, select, mutate, summarise, and joins. To drop it explicitly, use st_drop_geometry().</div></div>

<div class="quiz" id="q17"><div class="q">17. When you have two sf layers and want to combine them on a plot, you must first —</div><ul class="qopts"><li data-correct="false">Convert both to data.frames</li><li data-correct="true">Make sure they share a CRS (or reproject one to match the other)</li><li data-correct="false">Save both to disk</li><li data-correct="false">Pivot one of them longer</li></ul><div class="explanation">Plotting or any geometric operation on two mismatched-CRS layers either errors or gives nonsense. Reproject first.</div></div>

<div class="quiz" id="q18"><div class="q">18. For BC forestry spatial data, the canonical source is —</div><ul class="qopts"><li data-correct="false">A textbook</li><li data-correct="true">The BC Data Catalogue (https://catalogue.data.gov.bc.ca/) — Open Government Licence BC</li><li data-correct="false">Wikipedia</li><li data-correct="false">A private consulting firm</li></ul><div class="explanation">Forest districts, cut blocks, fire perimeters, watersheds, ownership — all openly published.</div></div>

<div class="quiz" id="q19"><div class="q">19. The single most common bug when building an sf object from a lat/lon tibble is —</div><ul class="qopts"><li data-correct="false">Forgetting to load tidyverse</li><li data-correct="true">Passing coords = c("lat", "lon") instead of c("lon", "lat") — first is x, second is y</li><li data-correct="false">Using too many decimal places</li><li data-correct="false">The wrong package version</li></ul><div class="explanation">Swapping lat and lon plots your BC points off the coast of Somalia. Look at the bounding box to spot it.</div></div>

<div class="quiz" id="q20"><div class="q">20. <strong>Reflection.</strong> Until now your forestry analysis treated rows as anonymous records. How does adding spatial location change what questions you can answer?</div><textarea placeholder="Type your answer here..."></textarea><button class="reveal-btn" onclick="this.nextElementSibling.classList.toggle('show')">Show common answers</button><div class="sample">Common reflections: <em>I can now ask "how close" instead of just "how many"</em> · <em>buffers let me reason about exposure zones, fire risk, road influence</em> · <em>intersection lets me cross datasets that share nothing but location</em> · <em>plotting on a map immediately surfaces patterns that a table hides</em> · <em>the biggest unlock is that BC Data Catalogue is now usable from R — every shapefile becomes a line of code instead of a manual import</em>.</div></div>

<script>
(function() {
  document.querySelectorAll('.quiz').forEach(function(q) {
    q.querySelectorAll('ul.qopts li').forEach(function(opt) {
      opt.addEventListener('click', function() {
        if (q.dataset.answered) return;
        q.dataset.answered = '1';
        var isCorrect = opt.getAttribute('data-correct') === 'true';
        if (isCorrect) opt.classList.add('correct');
        else {
          opt.classList.add('wrong');
          q.querySelectorAll('ul.qopts li').forEach(function(o) {
            if (o.getAttribute('data-correct') === 'true') o.classList.add('correct');
          });
        }
        var exp = q.querySelector('.explanation');
        if (exp) exp.classList.add('show');
      });
    });
  });
})();
</script>
```

# AI as a debugging companion

::: {.callout-tip}
## Useful prompts

- *"My ggplot of an sf object plots the points off the coast of
  Africa. What did I do wrong?"* (Answer: swapped lat and lon
  in `coords = c(...)`.)

- *"How do I find which of my stations are inside this
  polygon?"* (Answer: `st_join(stations, polygon, join =
  st_within)` or `st_intersection(stations, polygon)`.)

- *"My `st_distance()` returns numbers in millions. What units
  are those in?"* (Answer: the units of the current CRS — for
  BC Albers, metres.)
:::

# Reading

- *Geocomputation with R* (Lovelace, Nowosad, Muenchow) — the
  canonical free book on R spatial. Chapters 1–3 are required.
  <https://r.geocompx.org/>
- The `sf` package vignettes:
  <https://r-spatial.github.io/sf/articles/>
- BC Data Catalogue:
  <https://catalogue.data.gov.bc.ca/>
- EPSG code lookup:
  <https://epsg.io/>

# Instructor notes {.unnumbered}

::: {.callout-note appearance="minimal" collapse="true"}
## Pacing and emphasis

- The hardest concept this chapter is **CRS**. Spend a full
  lecture on it. Demo the "off the coast of Africa" bug live —
  show how degrees vs metres change everything.
- The **"sf objects are tibbles"** message is the single biggest
  unlock. Once learners realise every dplyr verb still works,
  spatial work stops feeling like a different world.
- **Posit Cloud** is strongly recommended for this chapter. Local
  sf installs sometimes fail on Linux and Mac without the right
  system libraries; cloud sidesteps all of it.

## Materials provided

| File | Purpose |
|---|---|
| `frst232_ch10_learner.qmd` | Quarto starter with TODO chunks. |
| `frst232_ch10_solutions.qmd` | Completed solutions. **Do not distribute.** |
| `quiz_ch10.html` | Standalone interactive quiz. |

## Reference numbers

| Quantity | Approximate value |
|---|---|
| `nrow(stations_sf)` | 10 |
| `st_crs(stations_sf)$Name` | "WGS 84" |
| Burnaby South → Langley Central distance | ~30 km |
| North Vancouver → Langley Central distance (longest pair) | ~37 km |
| Area of 5 km buffer | ~7,854 ha |

## Common stumbling points

- **Coords order.** `coords = c("lat", "lon")` is the
  single most common bug. Insist on always checking with
  `st_bbox()` after building.
- **CRS confusion.** Symptom: distances of "0.0003" (degrees)
  or "millions" (metres). Always check `st_crs()` and convert
  to a metric CRS before measuring.
- **Plotting two layers with mismatched CRSs.** ggplot will
  often just plot the second layer in the first layer's CRS,
  giving a wrong-looking map. Reproject explicitly.
- **Shapefile column-name surprises.** A shapefile column
  named `forest_district_name_long` will arrive truncated to
  `forst_dst_`. Use GeoPackage going forward.
:::
