---
title: "Map Forestry Data with leaflet"
---

```{r setup-ch11}
#| include: false
library(tidyverse)
library(sf)
library(leaflet)
library(here)
library(knitr)
```

::: {.callout-note appearance="simple"}
## Data for this chapter
This chapter uses the **CSRD parks** layer (`csrd_parks.geojson`); the
optional air-quality examples reuse `airdata3.csv`. Download from the
[Datasets Used in This Book](../datasets.qmd) page into your `data/`
folder.
:::

# Chapter goals {.unnumbered}

By the end of this chapter you will be able to:

- Explain the difference between a **static** map (Ch 10,
  `geom_sf`) and an **interactive** web map (this chapter,
  `leaflet`), and choose the right tool for each deliverable.
- Build a basic **leaflet** map in three lines:
  `leaflet() |> addTiles() |> addMarkers()`.
- Add **base tiles** from default OpenStreetMap and from
  alternative providers (terrain, satellite, light/dark).
- Place **markers** and **circle markers** on the map, with
  size and colour driven by data columns.
- Attach **popups** and **labels** that show information when a
  reader clicks or hovers.
- Map quantitative data with **`colorNumeric()`** and categorical
  data with **`colorFactor()`** palettes, and add a matching
  **`addLegend()`**.
- Combine multiple layers with a **layers-control** widget that
  lets the reader toggle each layer on or off.
- Embed a leaflet map inside a rendered Quarto HTML file —
  the map remains interactive in the final report.
- Recognize when **`leaflet` data must be in WGS84 (EPSG:4326)**
  and reproject if needed.

# Expected output for this chapter {.unnumbered}

::: {.callout-tip appearance="default"}
## What you will hand in

1. A **rendered Quarto HTML report**
   (`frst232_ch11_learner.html`) containing at least:
   - A **basic leaflet map** of the 10 Metro Vancouver
     monitoring stations with default markers and click-popups
     showing station names.
   - A **circle-marker map** where circle colour is driven by
     each station's **mean O₃** (computed from the cleaned
     air-quality tibble in Chapter 6), with a matching legend.
   - A **multi-layer map** combining the stations and 5 km
     buffer polygons (from Chapter 10), with a layers-control
     widget to toggle each layer.
   - At least one map using a non-default tile provider
     (e.g., Esri WorldImagery for satellite or
     CartoDB.Positron for clean light).
2. The corresponding **`.qmd` source file**.
3. A short **group-lab worksheet** (or screenshot) submitted to
   the course site.
:::

# Where we are

In Chapter 10 you built `stations_sf` — a tibble of 10 BC
monitoring stations with point geometries. You plotted it with
`geom_sf()` and got a clean static map suitable for a printed
report.

This chapter takes the same `stations_sf` and turns it into an
**interactive web map**. Click any station to see its name. Pan
and zoom freely. Switch between street, satellite, and terrain
basemaps. The reader does not need R installed — the map lives
in the rendered HTML.

# Static map vs interactive map — when to use which

| | Static (Ch 10) | Interactive (this chapter) |
|---|---|---|
| Library | `ggplot2::geom_sf()` | `leaflet` |
| Output | PNG, PDF, or SVG | HTML widget |
| Reader experience | Fixed view | Pan, zoom, click, toggle |
| Where it works | Print, slides, papers | Web, dashboards, Quarto HTML |
| File size | Small (KB) | Larger (MB — includes JS) |
| Build time | Fast | Slower |
| Customization | Full ggplot grammar | Many leaflet options, but not everything |

Use **`geom_sf`** for any chart that goes into a paper or a
PowerPoint slide. Use **`leaflet`** for any chart that goes into
an HTML report, a Quarto deliverable, or a dashboard your reader
will explore.

The two are complementary. A working analyst publishes both — a
static map in the printed memo, an interactive map in the linked
HTML.

# A note about Excel

Excel does not have a direct equivalent of leaflet. You could
embed a Bing Maps chart in Excel 365 (limited), or paste a
screenshot of a web map. Neither is interactive in the leaflet
sense. **This is one of the places where R does something
Excel cannot.**

# Installing leaflet

```r
install.packages("leaflet")
library(leaflet)
```

`leaflet` is much lighter than `sf` — no system dependencies,
installs in seconds on every platform. If you already have
Posit Cloud from Chapter 10, `leaflet` is pre-installed there
too.

# The leaflet pipeline pattern

Every leaflet map is built with the pipe `|>` (or `%>%`) and the
`add*()` family of functions:

```{r minimal-leaflet}
#| eval: false
leaflet() |>
  addTiles() |>
  addMarkers(lng = -123.117, lat = 49.247, popup = "Vancouver, BC")
```

Three things to notice:

1. **`leaflet()`** starts an empty map.
2. **`addTiles()`** adds the default basemap (OpenStreetMap).
3. **`addMarkers()`** drops a pin.

Read each `|>` as *"and also add"*. This is the same layering
pattern as ggplot's `+`, just with the pipe.

::: {.callout-tip}
## leaflet uses `|>` (not `+`)

Unlike `ggplot2`, which connects layers with `+`, leaflet uses
the pipe `|>`. This makes sense — each `add*()` call takes a
leaflet map object and returns a modified one, so it composes
like every other dplyr-style operation.
:::

# Build the stations sf object (carrying forward from Ch 10)

```{r build-stations}
stations <- tribble(
  ~location,                          ~lon,        ~lat,
  "Burnaby South",                    -122.985,    49.215,
  "Coquitlam Douglas College",        -122.798,    49.252,
  "Langley Central",                  -122.660,    49.105,
  "Maple Ridge Golden Ears School",   -122.620,    49.215,
  "New Westminster Sapperton Park",   -122.890,    49.235,
  "North Delta",                      -122.910,    49.135,
  "North Vancouver Second Narrows",   -123.015,    49.295,
  "Pitt Meadows Airport",             -122.700,    49.215,
  "Port Coquitlam North",             -122.770,    49.265,
  "Port Moody Rocky Point Park",      -122.860,    49.290
)

stations_sf <- stations |>
  st_as_sf(coords = c("lon", "lat"), crs = 4326)
```

::: {.callout-warning}
## leaflet requires WGS84 (EPSG:4326)

The leaflet JavaScript library expects all coordinates in
**latitude/longitude (WGS84)**. If your sf object is in BC
Albers or any other projected CRS, **reproject before passing
to leaflet**:

```r
stations_for_leaflet <- stations_sf |> st_transform(4326)
```

leaflet will sometimes warn if you forget; sometimes it just
plots your data in the wrong place. Make the reprojection
explicit.
:::

# Your first interactive map

```{r first-map}
#| eval: false
leaflet(data = stations_sf) |>
  addTiles() |>
  addMarkers(popup = ~location)
```

When this renders, you get:

- A draggable, zoomable map centered on Metro Vancouver.
- 10 default blue teardrop pins, one per station.
- A popup with the station name on each click.

Three new ideas in that small example:

- **`data = stations_sf`** — leaflet auto-detects the geometry
  column and finds the lat/lon.
- **`popup = ~location`** — the `~` (tilde) is leaflet's *formula
  notation*: *"the value of the column called `location`"*. You
  will see this pattern everywhere in leaflet.
- The map auto-centres and auto-zooms to fit all the points.

# Markers, circles, and popups

## `addMarkers()` — default pins

```r
leaflet(stations_sf) |>
  addTiles() |>
  addMarkers(popup = ~location, label = ~location)
```

- **`popup`**: shows on click.
- **`label`**: shows on hover.

## `addCircleMarkers()` — circles whose size and colour can vary

When the data column matters (not just the location), use circle
markers:

```r
leaflet(stations_sf) |>
  addTiles() |>
  addCircleMarkers(
    radius      = 8,
    color       = "#1B4332",
    fillColor   = "#52B788",
    fillOpacity = 0.8,
    weight      = 2,
    popup       = ~location
  )
```

## Driving size and colour from a data column

Attach a mean-O₃ value to each station (carrying forward from
Ch 7) and use it to size/colour the circles:

```{r o3-circles}
#| eval: false
# Suppose airdata_clean is loaded from Ch 6
station_means <- airdata_clean |>
  group_by(location) |>
  summarise(mean_o3 = mean(O3, na.rm = TRUE), .groups = "drop")

stations_o3 <- stations_sf |>
  left_join(station_means, by = "location")

pal <- colorNumeric(
  palette = c("#52B788", "#B7791F", "#B05C50"),
  domain  = stations_o3$mean_o3,
  na.color = "grey50"
)

leaflet(stations_o3) |>
  addTiles() |>
  addCircleMarkers(
    radius      = ~ifelse(is.na(mean_o3), 5, mean_o3 + 4),
    color       = "#1B4332",
    fillColor   = ~pal(mean_o3),
    fillOpacity = 0.85,
    weight      = 1.5,
    popup       = ~paste0("<b>", location, "</b><br>",
                          "Mean O₃: ",
                          ifelse(is.na(mean_o3), "no data",
                                 paste0(round(mean_o3, 1), " ppb")))
  ) |>
  addLegend(
    "bottomright",
    pal     = pal,
    values  = ~mean_o3,
    title   = "Mean O₃ (ppb)",
    opacity = 1
  )
```

Three new ideas:

- **`colorNumeric(palette, domain)`** builds a palette function
  that maps numeric values to colours. Use it inside
  `fillColor = ~pal(column)`.
- **`addLegend()`** draws a matching legend. The palette
  function and the data column must match what the markers used.
- **HTML in popups** — you can use `<b>`, `<br>`, links —
  anything HTML.

# Tile providers — change the basemap

`addTiles()` adds the default OpenStreetMap. For other basemaps,
use `addProviderTiles()`:

```r
leaflet(stations_sf) |>
  addProviderTiles("CartoDB.Positron") |>     # clean light theme
  addMarkers(popup = ~location)
```

Useful providers:

| Provider | When to use |
|---|---|
| `OpenStreetMap.Mapnik` (default) | General-purpose, free |
| `CartoDB.Positron` | Clean light theme — popular for data viz |
| `CartoDB.DarkMatter` | Dark theme |
| `Esri.WorldImagery` | Satellite imagery |
| `Esri.WorldTopoMap` | Topographic with relief |
| `OpenTopoMap` | Open-source topo, good for forestry |
| `Stadia.AlidadeSmooth` | Subtle smooth basemap |

Full provider list: <https://leaflet-extras.github.io/leaflet-providers/preview/>.

# Polygons — adding the buffers from Ch 10

Polygon layers (like the 5 km buffers from Ch 10) go in with
`addPolygons()`:

```r
buffers_sf <- stations_sf |>
  st_transform(3005) |>
  st_buffer(dist = 5000) |>
  st_transform(4326)

leaflet() |>
  addProviderTiles("CartoDB.Positron") |>
  addPolygons(data = buffers_sf,
              fillColor = "#52B788",
              fillOpacity = 0.25,
              color = "#1B4332",
              weight = 1) |>
  addCircleMarkers(data = stations_sf,
                   radius = 5,
                   color = "#1B4332",
                   fillColor = "#1B4332",
                   fillOpacity = 1,
                   popup = ~location)
```

Note the **CRS dance**: buffers are computed in BC Albers (3005,
metric — required for the buffer distance), then reprojected to
WGS84 (4326) for leaflet. **Always end in WGS84 before passing to
leaflet.**

# Multiple layers + layers control

When a map has multiple groups, `addLayersControl()` lets the
reader toggle them:

```r
leaflet() |>
  addProviderTiles("OpenStreetMap.Mapnik", group = "Street") |>
  addProviderTiles("Esri.WorldImagery", group = "Satellite") |>
  addProviderTiles("OpenTopoMap", group = "Topo") |>
  addCircleMarkers(data = stations_sf,
                   group = "Stations",
                   popup = ~location) |>
  addPolygons(data = buffers_sf,
              group = "5 km buffers",
              fillColor = "#52B788", fillOpacity = 0.3) |>
  addLayersControl(
    baseGroups    = c("Street", "Satellite", "Topo"),
    overlayGroups = c("Stations", "5 km buffers"),
    options       = layersControlOptions(collapsed = FALSE)
  )
```

- **`baseGroups`** are mutually exclusive (radio buttons — one
  basemap at a time).
- **`overlayGroups`** are independently toggleable (checkboxes).

The reader sees a control box in the top-right of the map.

# `setView()` and `fitBounds()` — control the initial view

By default, leaflet zooms to fit the data. Override with:

```r
# Centre on a specific lat/lon and zoom level
leaflet() |>
  addTiles() |>
  setView(lng = -122.9, lat = 49.22, zoom = 11)

# Or fit a specific bounding box
leaflet() |>
  addTiles() |>
  fitBounds(lng1 = -123.1, lat1 = 49.10,
            lng2 = -122.6, lat2 = 49.30)
```

Zoom level: 0 (whole world) to ~18 (street level). For a
Metro Vancouver overview, zoom 10 or 11 is right.

# Saving and embedding

When you render a Quarto document, leaflet maps appear **inline**
in the rendered HTML. The map is fully interactive — your reader
can pan, zoom, click — without opening R.

To save a leaflet map as a standalone HTML file:

```r
library(htmlwidgets)
saveWidget(my_leaflet_map,
           file = here("outputs", "stations_map.html"),
           selfcontained = TRUE)
```

`selfcontained = TRUE` bundles all the JavaScript into a single
file — easy to email, share, or open offline.

# A complete worked example

```{r complete-example}
#| eval: false
# All together: stations + buffers + O₃ colour, layered + legend

# 1. Build the buffer layer (compute in metric CRS, reproject for leaflet)
buffers_sf <- stations_sf |>
  st_transform(3005) |>
  st_buffer(dist = 5000) |>
  st_transform(4326)

# 2. Attach O₃ (assume station_means is from Ch 6/7)
stations_o3 <- stations_sf |>
  left_join(station_means, by = "location")

# 3. Palette for O₃
pal <- colorNumeric(palette = c("#52B788", "#B7791F", "#B05C50"),
                    domain = stations_o3$mean_o3,
                    na.color = "grey60")

# 4. Build the map
leaflet() |>
  addProviderTiles("CartoDB.Positron", group = "Light") |>
  addProviderTiles("Esri.WorldImagery", group = "Satellite") |>
  addPolygons(data = buffers_sf,
              group = "5 km buffers",
              fillColor = "#52B788", fillOpacity = 0.2,
              color = "#1B4332", weight = 1) |>
  addCircleMarkers(data = stations_o3,
                   group = "Stations",
                   radius      = ~ifelse(is.na(mean_o3), 5, mean_o3 + 4),
                   fillColor   = ~pal(mean_o3),
                   fillOpacity = 0.9,
                   color       = "#1B4332",
                   weight      = 1.5,
                   popup       = ~paste0("<b>", location, "</b><br>",
                                          "Mean O₃: ",
                                          ifelse(is.na(mean_o3), "no data",
                                                 paste0(round(mean_o3, 1),
                                                        " ppb")))) |>
  addLegend("bottomright",
            pal = pal, values = stations_o3$mean_o3,
            title = "Mean O₃ (ppb)",
            na.label = "No data",
            opacity = 1) |>
  addLayersControl(
    baseGroups    = c("Light", "Satellite"),
    overlayGroups = c("Stations", "5 km buffers"),
    options       = layersControlOptions(collapsed = FALSE)
  ) |>
  setView(lng = -122.85, lat = 49.22, zoom = 11)
```

This map renders as a single interactive widget. A reader can
zoom out to BC, zoom in to a station, click for the popup, toggle
the buffers off, switch to satellite. **One R block, full
dashboard.**

# A real interactive map: CSRD parks

::: {.content-visible when-format="html"}

The maps above used the synthetic station tibble so you could see
each piece in isolation. Here is the real thing — the **Columbia
Shuswap Regional District parks** layer from Chapter 10, as a live
interactive map you can pan, zoom, and click:

```{r csrd-leaflet}
#| eval: !expr knitr::is_html_output()
parks <- st_read(here("data", "csrd_parks.geojson"), quiet = TRUE)

pal <- colorFactor("viridis", domain = parks$ParkType)

leaflet(parks) |>
  addProviderTiles("CartoDB.Positron") |>
  addCircleMarkers(
    radius      = 5,
    color       = ~pal(ParkType),
    fillOpacity = 0.85,
    stroke      = FALSE,
    label       = ~ParkName,
    popup       = ~paste0("<b>", ParkName, "</b><br>",
                          "Type: ",   ParkType, "<br>",
                          "Status: ", ParkStatus)
  ) |>
  addLegend("bottomright", pal = pal, values = ~ParkType,
            title = "Park type", opacity = 1)
```

*Map: Columbia Shuswap Regional District park locations, coloured
by park type. Source: CSRD Open Data. Non-interactive description:
the parks are distributed across the CSRD — clustering near Salmon
Arm and following the Shuswap Lake and Columbia River corridors. A
static `geom_sf` version carrying the same information appears in
Chapter 10 for readers who cannot use the interactive widget.*

This is the interactive counterpart to the static map in Chapter
10 — **same data, same question, two presentation choices.** Use
the interactive version when exploration helps (a reader wants to
find a specific park); use the static version when the map must
work in print, in a PDF, or for a screen-reader user.

:::

::: {.content-visible when-format="pdf"}

The interactive CSRD parks map appears in the HTML edition of this book.
In print, see the static `geom_sf` map of the same CSRD parks layer in
Chapter 10 — *same data, same question, two presentation choices.*

:::

# Active learning

## Activity 1 — Minimal map

Build a leaflet map showing only one location: your university
campus. Use `addTiles()` and `addMarkers()` with a popup.

## Activity 2 — All ten stations

Build a leaflet map of `stations_sf` with default markers and
hover labels.

## Activity 3 — Circle markers

Replace the default markers with `addCircleMarkers()`. Set
`radius = 7`, `fillColor = "#2D6A4F"`, `fillOpacity = 0.85`.

## Activity 4 — Tile provider experiment

Build the same map three times with three different tile
providers (default, `CartoDB.Positron`, `Esri.WorldImagery`).
Which feels best for a forestry deliverable?

## Activity 5 — Add a polygon layer

Take any one station, build a 5 km buffer, and add it to the
map as `addPolygons()`. Make sure both the buffer and the
station markers are in WGS84 before plotting.

## Activity 6 — Save as standalone HTML

Use `htmlwidgets::saveWidget()` to save your map as
`outputs/my_map.html`. Open it in a browser — confirm it
works offline.

# Group lab activity

::: {.callout-note appearance="default"}
## Lab: Build the chapter capstone interactive map

Work in groups of 3 or 4. The lab builds the multi-layer
interactive map specified in the chapter's *Expected output*.

**Task 1 — Build `stations_sf`** (everyone confirms 10 rows in
WGS84, same as Ch 10).

**Task 2 — Compute mean O₃ per station** using the Chapter 6
cleaning pipeline + `group_by(location) |> summarise(mean_o3 =
mean(O3, na.rm = TRUE))`. Join the mean back onto
`stations_sf` with `left_join(by = "location")`.

**Task 3 — Build a O₃ palette** with `colorNumeric()`.
Discuss as a group: which colour scheme makes sense for an
air-quality variable? (Many groups choose green→yellow→red.)

**Task 4 — Build the map.** Add tiles, add circle markers
with size/colour driven by mean_o3, add a legend, add
informative popups.

**Task 5 — Add the buffer overlay.** Build 5 km buffers around
every station and add them as a separate group with
`addPolygons()`. Add a layers-control widget so the reader can
toggle buffers on and off.

**Task 6 — Pick the basemap.** Discuss which provider tile
makes the air-quality colours pop. Add at least two basemap
options to the layers control.

**Task 7 — Submit.** Each group submits the rendered HTML
file and a one-sentence caption explaining the headline
finding.

This lab fits comfortably inside a single hands-on session.
:::

# Exercises

1. **Minimal map.** Build a leaflet map of `stations_sf` with
   default tiles and markers. Confirm the popup shows the
   station name.

2. **Circle markers.** Replace `addMarkers()` with
   `addCircleMarkers()`. Set radius, fill colour, and
   fill opacity manually.

3. **Tile providers.** Build the same map three times with
   three different `addProviderTiles()` calls. Embed all three
   in your `.qmd`.

4. **Colour by O₃.** Compute mean O₃ per station (from
   Ch 6 cleaning). Use `colorNumeric()` to colour the circles.
   Add `addLegend()`.

5. **Popup with HTML.** Build popups that include the station
   name in bold, a line break, and the mean O₃ rounded to
   one decimal.

6. **Add a polygon layer.** Build 5 km buffers around all
   stations and add them with `addPolygons()`. Confirm both
   layers are in WGS84 before plotting.

7. **Layers control.** Add an `addLayersControl()` with two
   base groups (street, satellite) and two overlay groups
   (stations, buffers).

8. **Save standalone.** Use `htmlwidgets::saveWidget()` to save
   your final map as `outputs/stations_map.html`. Open it
   directly in a browser.

# Optional, advanced

## Markers with custom icons

```r
fire_icon <- makeIcon(
  iconUrl = "https://leafletjs.com/examples/custom-icons/leaf-red.png",
  iconWidth = 24, iconHeight = 38
)
leaflet(stations_sf) |>
  addTiles() |>
  addMarkers(icon = fire_icon, popup = ~location)
```

For forestry use cases, custom icons can distinguish point
types (fire ignitions, plot centres, sampling locations).

## Adding a search box

The `leaflet.extras` add-on package provides extras including
search, drawing tools, and measurement widgets:

```r
library(leaflet.extras)
leaflet(stations_sf) |>
  addTiles() |>
  addMarkers(popup = ~location, group = "stations") |>
  addSearchFeatures(targetGroups = "stations",
                    options = searchFeaturesOptions(zoom = 12))
```

## Heatmaps

For dense point data (e.g., wildfire ignitions over years),
`leaflet.extras::addHeatmap()` produces a smooth density
overlay:

```r
library(leaflet.extras)
leaflet(ignition_points) |>
  addTiles() |>
  addHeatmap(radius = 12)
```

# A glimpse of Chapter 12: From maps to presentations

In Chapter 12 you take everything from Chapters 4–11 — your
cleaned data, summary tables, charts, and maps — and assemble
them into a **portfolio**:

- A Quarto **revealjs presentation** for an in-class talk.
- A Quarto **HTML report** showcasing your best analyses.
- A printable **PDF** for a deliverable submission.

The leaflet maps you built in this chapter become **slides** in
the presentation — fully interactive when displayed live.

# Self-assessment quiz {#sec-quiz-ch11}

```{=html}
<style>
.quiz { background: #F4F7F4; border-left: 4px solid #2D6A4F; padding: 1em 1.5em; margin: 1.5em 0; border-radius: 6px; }
.quiz .q { font-weight: 600; color: #1B4332; margin-bottom: 0.6em; }
.quiz ul.qopts { list-style: none; padding: 0; margin: 0; }
.quiz ul.qopts li { padding: 0.55em 0.9em; margin: 0.3em 0; background: white; border: 1px solid #d8e4d8; border-radius: 4px; cursor: pointer; transition: background 0.12s; }
.quiz ul.qopts li:hover { background: #E8F1E9; }
.quiz ul.qopts li.correct { background: #C8E6C9; border-color: #2D6A4F; font-weight: 600; }
.quiz ul.qopts li.wrong { background: #FFCDD2; border-color: #B05C50; }
.quiz .explanation { display: none; margin-top: 0.7em; padding: 0.7em 1em; background: #FFF8E8; border-left: 3px solid #B7791F; border-radius: 4px; font-size: 0.95em; }
.quiz .explanation.show { display: block; }
.quiz textarea { width: 100%; min-height: 4.5em; padding: 0.6em; border: 1px solid #d8e4d8; border-radius: 4px; font-family: inherit; font-size: 0.95em; resize: vertical; margin-top: 0.3em; }
.quiz .reveal-btn { background: #2D6A4F; color: white; border: none; padding: 0.45em 0.9em; margin-top: 0.5em; border-radius: 4px; cursor: pointer; font-size: 0.9em; }
.quiz .reveal-btn:hover { background: #1B4332; }
.quiz .sample { display: none; margin-top: 0.6em; padding: 0.6em 0.9em; background: #FFF8E8; border-left: 3px solid #B7791F; border-radius: 4px; font-size: 0.95em; font-style: italic; }
.quiz .sample.show { display: block; }
</style>

<div class="quiz" id="q1" data-hints="leaflet is an R wrapper around a JavaScript library.||Your R code emits the HTML and JS.||It generates HTML/JS wrapping leaflet.js."><div class="q">1. leaflet builds a web map by —</div><ul class="qopts"><li data-correct="false">Asking Google Maps</li><li data-correct="true">Generating HTML + JavaScript that wraps the leaflet.js library</li><li data-correct="false">Drawing into a static PNG</li><li data-correct="false">Calling ggplot</li></ul><div class="explanation">leaflet is an R wrapper around leaflet.js — your R code produces the HTML/JS that runs in a browser.</div></div>

<div class="quiz" id="q2" data-hints="Each add* takes a map and returns a modified map.||So they chain — but not with ggplot's +.||The pipe |> (or %>%)."><div class="q">2. The leaflet pipeline pattern uses —</div><ul class="qopts"><li data-correct="false">+ between calls (like ggplot2)</li><li data-correct="true">|> (or %>%) between leaflet() and the add*() functions</li><li data-correct="false">commas inside a single call</li><li data-correct="false">no chaining</li></ul><div class="explanation">Each add* function takes a leaflet map and returns a modified one, so they compose with the pipe.</div></div>

<div class="quiz" id="q3" data-hints="leaflet.js speaks latitude/longitude.||Not a projected or metric CRS.||EPSG:4326 (WGS84) — reproject with st_transform(4326)."><div class="q">3. leaflet requires its input data to be in CRS —</div><ul class="qopts"><li data-correct="false">EPSG:3005 (BC Albers)</li><li data-correct="true">EPSG:4326 (WGS84, lat/lon)</li><li data-correct="false">EPSG:3857 (Web Mercator)</li><li data-correct="false">Any CRS</li></ul><div class="explanation">leaflet.js expects lat/lon. Reproject with st_transform(4326) before plotting.</div></div>

<div class="quiz" id="q4" data-hints="Which add* call, with no arguments, adds OpenStreetMap tiles?||It is the simplest tile call.||addTiles()."><div class="q">4. To add the default OpenStreetMap basemap, you use —</div><ul class="qopts"><li data-correct="false">addBackground()</li><li data-correct="true">addTiles()</li><li data-correct="false">addBasemap()</li><li data-correct="false">addLayer()</li></ul><div class="explanation"><code>addTiles()</code> with no arguments adds the default OpenStreetMap Mapnik tiles.</div></div>

<div class="quiz" id="q5" data-hints="A separate function pulls named third-party tile sets.||Pass a provider name like Esri.WorldImagery.||addProviderTiles()."><div class="q">5. To use a different tile provider (e.g., satellite imagery), you use —</div><ul class="qopts"><li data-correct="false">addTiles("satellite")</li><li data-correct="true">addProviderTiles("Esri.WorldImagery")</li><li data-correct="false">addTiles(satellite = TRUE)</li><li data-correct="false">setBasemap()</li></ul><div class="explanation">leaflet bundles many third-party tile providers. Browse the full list at https://leaflet-extras.github.io/leaflet-providers/preview/.</div></div>

<div class="quiz" id="q6" data-hints="The tilde means 'look up this column in the data'.||Similar to how aes() works in ggplot.||The value of the location column in the tibble."><div class="q">6. In leaflet, the formula notation <code>popup = ~location</code> means —</div><ul class="qopts"><li data-correct="false">Literally the string "location"</li><li data-correct="true">The value of the column called <code>location</code> in the data tibble</li><li data-correct="false">The opposite of location</li><li data-correct="false">A function called location</li></ul><div class="explanation">The tilde is leaflet's way of saying "look up this column in the data" — similar to how aes() works in ggplot.</div></div>

<div class="quiz" id="q7" data-hints="Circle markers can vary with the data; default pins are fixed.||Use circles when the values matter visually.||Circles whose size and colour you drive from data columns."><div class="q">7. <code>addCircleMarkers()</code> differs from <code>addMarkers()</code> in that —</div><ul class="qopts"><li data-correct="false">It only works with raster data</li><li data-correct="true">It draws circles whose size and colour you can drive from data columns; default markers are fixed pins</li><li data-correct="false">It is faster but less accurate</li><li data-correct="false">It is for polygons</li></ul><div class="explanation">Circle markers are the leaflet equivalent of a sized/coloured scatter. Use them whenever the data values matter visually.</div></div>

<div class="quiz" id="q8" data-hints="leaflet builds a palette FUNCTION from your values.||You then pass fillColor = ~pal(column).||colorNumeric(palette, domain)."><div class="q">8. To map a numeric column to colour, you use —</div><ul class="qopts"><li data-correct="false">scale_colour_continuous()</li><li data-correct="true">colorNumeric(palette, domain) — and pass the resulting function as fillColor = ~pal(column)</li><li data-correct="false">addColour()</li><li data-correct="false">paint()</li></ul><div class="explanation">leaflet's colour-palette functions (colorNumeric, colorBin, colorFactor) build a function that maps values to colours.</div></div>

<div class="quiz" id="q9" data-hints="It is optional, not required for every map.||It reuses your palette and column to draw a key.||An optional layer that draws a legend."><div class="q">9. <code>addLegend()</code> is —</div><ul class="qopts"><li data-correct="false">A required call for every map</li><li data-correct="true">An optional layer that draws a legend matching a palette and data values</li><li data-correct="false">A way to add a title</li><li data-correct="false">A debugging tool</li></ul><div class="explanation"><code>addLegend</code> takes the same palette function and column you used for the markers, and draws a legend box anchored to a corner.</div></div>

<div class="quiz" id="q10" data-hints="Filled shapes, not points or single lines.||Think buffers, districts, watersheds.||Polygon geometries."><div class="q">10. <code>addPolygons()</code> is for —</div><ul class="qopts"><li data-correct="false">Points</li><li data-correct="false">Lines only</li><li data-correct="true">Polygon geometries (e.g., buffers, forest districts, watersheds)</li><li data-correct="false">Markers</li></ul><div class="explanation"><code>addPolygons</code> draws filled polygon shapes. <code>addPolylines</code> draws line geometries.</div></div>

<div class="quiz" id="q11" data-hints="Base groups are mutually exclusive; overlays are independent.||Radio buttons versus checkboxes.||Base groups = radio (one basemap), overlay groups = checkboxes (any combination)."><div class="q">11. <code>addLayersControl()</code> with <code>baseGroups</code> and <code>overlayGroups</code> —</div><ul class="qopts"><li data-correct="false">Hides the map</li><li data-correct="true">Adds a toggle widget where base groups are radio buttons (one at a time) and overlay groups are checkboxes (any combination)</li><li data-correct="false">Removes all layers</li><li data-correct="false">Saves the map</li></ul><div class="explanation">baseGroups are mutually exclusive (basemaps); overlayGroups can each be toggled independently.</div></div>

<div class="quiz" id="q12" data-hints="It controls where the map opens.||Centre and zoom on first render.||Sets the initial centre and zoom level."><div class="q">12. <code>setView(lng, lat, zoom)</code> —</div><ul class="qopts"><li data-correct="false">Adds a marker at the centre</li><li data-correct="true">Sets the initial centre and zoom level when the map first renders</li><li data-correct="false">Saves the map</li><li data-correct="false">Adds an overlay</li></ul><div class="explanation"><code>setView</code> controls the starting view. Without it, leaflet auto-zooms to fit the data.</div></div>

<div class="quiz" id="q13" data-hints="leaflet outputs an HTML widget, not a picture.||In rendered HTML it keeps working.||Remains fully interactive — pan, zoom, click, toggle."><div class="q">13. A leaflet map embedded in a Quarto HTML report —</div><ul class="qopts"><li data-correct="false">Renders as a static PNG</li><li data-correct="true">Remains fully interactive in the rendered HTML — the reader can pan, zoom, click, and toggle layers</li><li data-correct="false">Requires R to be installed on the reader's computer</li><li data-correct="false">Cannot be embedded</li></ul><div class="explanation">leaflet outputs an HTML widget. When rendered into a self-contained HTML file, the widget remains fully interactive.</div></div>

<div class="quiz" id="q14" data-hints="Which function writes a widget to a self-contained HTML file?||Not ggsave — that is for ggplots.||htmlwidgets::saveWidget()."><div class="q">14. To save a leaflet map as a standalone HTML file, you use —</div><ul class="qopts"><li data-correct="false">ggsave()</li><li data-correct="true">htmlwidgets::saveWidget()</li><li data-correct="false">write_csv()</li><li data-correct="false">save()</li></ul><div class="explanation"><code>saveWidget(my_map, "outputs/map.html", selfcontained = TRUE)</code> bundles all the JS into one shareable file.</div></div>

<div class="quiz" id="q15" data-hints="leaflet treats coordinates as lat/lon.||If they are metric (BC Albers), they land far out to sea.||The data is not in WGS84 — you forgot st_transform(4326)."><div class="q">15. The most likely cause of leaflet plotting your points in the middle of the ocean is —</div><ul class="qopts"><li data-correct="false">A bug in leaflet</li><li data-correct="true">Your sf data is not in WGS84 (EPSG:4326) — you forgot to st_transform(4326) before passing it to leaflet</li><li data-correct="false">Your internet is down</li><li data-correct="false">Your zoom level is wrong</li></ul><div class="explanation">leaflet expects lat/lon. If you pass BC Albers (metres) data, leaflet treats the metric coordinates as if they were lat/lon, and your points land somewhere in the Atlantic.</div></div>

<div class="quiz" id="q16" data-hints="Buffering needs metric units, but leaflet needs lat/lon.||Do the buffer in BC Albers, then convert it back.||Buffer in EPSG:3005, then st_transform back to 4326."><div class="q">16. To compute a 5 km buffer for use in leaflet —</div><ul class="qopts"><li data-correct="false">Use st_buffer in WGS84 with dist = 5000</li><li data-correct="true">Reproject to BC Albers (metric), st_buffer with dist = 5000, then reproject the buffer back to WGS84 for leaflet</li><li data-correct="false">leaflet handles buffers automatically</li><li data-correct="false">It is not possible</li></ul><div class="explanation">Buffering requires metric units. Compute in EPSG:3005, then reproject back to EPSG:4326 for leaflet.</div></div>

<div class="quiz" id="q17" data-hints="What can a web map do that a printed one cannot?||Pan, zoom, click, toggle.||Interactivity."><div class="q">17. The biggest advantage of leaflet over a static map is —</div><ul class="qopts"><li data-correct="false">Higher resolution</li><li data-correct="true">Interactivity — pan, zoom, click for popups, toggle layers</li><li data-correct="false">Lower file size</li><li data-correct="false">No CRS required</li></ul><div class="explanation">leaflet trades file size and print-friendliness for interactivity. Pick it for web/HTML deliverables.</div></div>

<div class="quiz" id="q18" data-hints="Does Excel produce real interactive web maps?||This is a place R does something Excel cannot.||No direct equivalent."><div class="q">18. Excel has —</div><ul class="qopts"><li data-correct="false">A built-in leaflet equivalent</li><li data-correct="true">No direct equivalent — interactive web mapping is one of the places R does something Excel cannot</li><li data-correct="false">Better web mapping</li><li data-correct="false">Identical functionality</li></ul><div class="explanation">Excel's mapping is limited to embedded Bing charts. For real interactive web mapping, R (leaflet) or a dedicated GIS tool is the choice.</div></div>

<div class="quiz" id="q19" data-hints="popup is on click; label is the lighter, always-handy one.||It appears when you move the mouse over a marker.||Hover."><div class="q">19. <code>popup</code> shows on click; <code>label</code> shows on —</div><ul class="qopts"><li data-correct="false">Page load</li><li data-correct="true">Hover</li><li data-correct="false">Double-click</li><li data-correct="false">Right-click</li></ul><div class="explanation">Use label for short always-visible info on hover; popup for richer content on click.</div></div>

<div class="quiz" id="q20"><div class="q">20. <strong>Reflection.</strong> Now that you can publish interactive maps, which forestry analyses you have done in Excel feel like they would have been more useful as interactive maps? Why?</div><textarea placeholder="Type your answer here..."></textarea><button class="reveal-btn" onclick="this.nextElementSibling.classList.toggle('show')">Show common answers</button><div class="sample">Common reflections: <em>my 2017 wildfire analysis would have come alive as an interactive map of fire perimeters with clickable details</em> · <em>monitoring station tables are abstract; the same data on a clickable map immediately raises questions</em> · <em>the cleaned air-quality file becomes a dashboard when each station is a clickable popup</em> · <em>any analysis with a spatial component benefits from being explorable — Excel's tables hide the geographic story</em>.</div></div>

<script>
(function() {
  document.querySelectorAll('.quiz').forEach(function(q) {
    q.querySelectorAll('ul.qopts li').forEach(function(opt) {
      opt.addEventListener('click', function() {
        if (q.dataset.answered) return;
        q.dataset.answered = '1';
        var isCorrect = opt.getAttribute('data-correct') === 'true';
        if (isCorrect) opt.classList.add('correct');
        else {
          opt.classList.add('wrong');
          q.querySelectorAll('ul.qopts li').forEach(function(o) {
            if (o.getAttribute('data-correct') === 'true') o.classList.add('correct');
          });
        }
        var exp = q.querySelector('.explanation');
        if (exp) exp.classList.add('show');
      });
    });
  });
})();
</script>
```

# AI as a debugging companion

::: {.callout-tip}
## Useful prompts

- *"My leaflet map shows my points in the middle of the
  Atlantic Ocean. My sf data is in BC Albers. What did I miss?"*
  (Answer: reproject to WGS84 before passing to leaflet.)

- *"How do I add a legend that matches the colour palette I
  used for my circles?"*

- *"My popups show the literal word 'location' instead of the
  station name. What's wrong?"* (Answer: missing the `~` in
  `popup = ~location`.)
:::

# Reading

- *Leaflet for R* — the canonical reference, free:
  <https://rstudio.github.io/leaflet/>
- *Geocomputation with R* — chapter 8 *Making maps with R*:
  <https://r.geocompx.org/adv-map.html>
- Tile providers preview:
  <https://leaflet-extras.github.io/leaflet-providers/preview/>
- HTML widgets gallery: <https://www.htmlwidgets.org/>

# Instructor notes {.unnumbered}

::: {.callout-note appearance="minimal" collapse="true"}
## Pacing and emphasis

- The chapter has **two big concepts**: the pipeline pattern
  with `|>` (different from `+` in ggplot), and the
  WGS84-required-for-leaflet rule. Demo both in the first
  lecture.
- The **circle marker + colorNumeric + addLegend trio** is
  the chapter's most powerful pattern. Spend time getting
  learners fluent with it.
- **Layers control** is a small UI feature but turns a single
  map into a mini-dashboard. Worth emphasising.

## Materials provided

| File | Purpose |
|---|---|
| `frst232_ch11_learner.qmd` | Quarto starter with TODO chunks. |
| `frst232_ch11_solutions.qmd` | Completed solutions. **Do not distribute.** |
| `quiz_ch11.html` | Standalone interactive quiz. |

## Reference numbers

| Quantity | Approximate value |
|---|---|
| Number of stations on the map | 10 |
| Map centre (lon, lat) | (-122.85, 49.22) |
| Useful initial zoom | 10 or 11 |
| Number of stations with mean O₃ | ~6 (the rest are NA in the cleaned file) |
| Mean O₃ range (used for the colour palette) | ~2 to ~10 ppb |

## Common stumbling points

- **Points in the wrong place.** Almost always: sf data in BC
  Albers (3005) was passed to leaflet without
  `st_transform(4326)`. Symptom: pins land in the Atlantic
  or Indian Ocean.
- **Popup shows literal column name.** Symptom: every popup
  says "location". Fix: add the `~` — `popup = ~location`.
- **Buffer computed in WGS84.** Symptom: buffer is a
  micro-shape (5000 degrees instead of 5000 metres) or
  errors. Fix: buffer in 3005, reproject back.
- **Legend palette mismatch.** Symptom: the legend's colours
  do not match the markers'. Fix: use the same `pal` object
  and the same column for both `addCircleMarkers(fillColor =
  ~pal(col))` and `addLegend(pal = pal, values = ~col)`.
- **Forgetting `library(leaflet)`.** Symptom: *"could not
  find function 'leaflet'"*. Confirm setup chunk loads it.
:::
