---
title: "Group Presentations and Portfolio Outputs"
---

```{r setup-ch12}
#| include: false
library(tidyverse)
library(here)
library(knitr)
```

::: {.callout-note appearance="simple"}
## Data for this chapter
This chapter uses the **BC silviculture** workbook
(`bc_disturbance_reforestation.xlsx`). Download it from the
[Datasets Used in This Book](../datasets.qmd) page and put it in your
`data/` folder.
:::

# Chapter goals {.unnumbered}

By the end of this chapter you will be able to:

- Build a **Quarto revealjs presentation** by changing one line
  of YAML in a `.qmd` file.
- Apply the **one-idea-per-slide** rule and use slide separators
  (`---` or `## Heading`) effectively.
- Embed R code chunks (charts, tables, leaflet maps) directly in
  slides so the slide refreshes whenever the data changes.
- Add **incremental reveals**, **speaker notes**, and **footnote
  citations** to slides.
- Assemble a **portfolio** — a single Quarto site or HTML
  collection — that showcases your best analyses from
  Chapters 4–11.
- Cite **data sources** (BC Data Catalogue, Environmental
  Reporting BC) and **R packages** properly.
- Apply an **open licence** (CC BY 4.0) to your deliverables so
  they can be shared and built upon.
- Publish a rendered Quarto site to a free hosting service
  (Quarto Pub, GitHub Pages, Netlify).

# Expected output for this chapter {.unnumbered}

::: {.callout-tip appearance="default"}
## What you will hand in

1. A **rendered group presentation** in revealjs HTML format,
   covering one analysis from the term (10–15 slides for a
   ~12-minute talk).
2. A **personal portfolio** — at least three of your previous
   chapter deliverables (rendered HTML reports) collected into
   a single linked landing page.
3. A **citation file** (`CITATION.md`) listing the data sources,
   R packages, and any external code used.
4. A **licence statement** (CC BY 4.0 or equivalent) at the
   bottom of your portfolio landing page.
5. The corresponding **`.qmd` source files** for both the
   presentation and the portfolio.
:::

# Where we are

Through Chapters 1–11 you produced a body of work: cleaned
datasets, summary reports, charts, static maps, interactive
maps. Each rendered HTML stands alone. **Chapter 12 is about
packaging.**

Two questions this chapter answers:

1. **How do I present what I built** to a non-R audience in a
   12-minute talk?
2. **How do I publish what I built** so a future employer, TA,
   or classmate can find it and build on it?

Both answers use the same tool you have been using since
Chapter 5: Quarto.

# Quarto outputs — three formats from one file

The same `.qmd` source can produce different outputs by changing
the YAML `format` field. Through this course you have already
used:

| `format:` | What it produces |
|---|---|
| `html` | A single HTML report — your chapter deliverables. |

This chapter adds two more:

| `format:` | What it produces |
|---|---|
| `revealjs` | A revealjs HTML presentation (slides). |
| `pdf` | A printable PDF (requires a TeX install). |

A single Quarto **project** (a folder of `.qmd` files plus a
`_quarto.yml` config) can produce **all three formats from the
same content**. The same chart that appears on slide 7 of your
revealjs talk can appear on page 12 of your PDF report and at
section 3 of your HTML portfolio — without ever copy-pasting.

That single-source-many-outputs property is the difference
between PowerPoint+Word+PDF as separate deliverables and Quarto
as a publishing system.

# PowerPoint-to-Quarto bridge

| What you would do in PowerPoint | What you do in Quarto revealjs |
|---|---|
| New slide | `## Slide title` (h2 heading) or `---` separator |
| Two-column layout | `::: {.columns}` / `::: {.column width="50%"}` |
| Title slide | YAML header (`title:`, `subtitle:`, `author:`) |
| Speaker notes | `::: {.notes}` block |
| Animated reveal | `::: {.incremental}` block or `{.fragment}` class |
| Paste a chart | An R code chunk that re-renders the chart |
| Paste a screenshot of a map | A leaflet chunk that produces a live, interactive map |
| Embed a video | `<video>` tag in HTML cell |
| Insert a footnote | `^[footnote text]` or proper Markdown citation |
| Apply a template | `theme:` in YAML (`default`, `dark`, `simple`, `sky`, etc.) |
| Save as PDF | Render once to `revealjs` (HTML); press **E** key during the talk to export PDF |
| Edit the company logo every slide | Set it once in `_quarto.yml`; every slide inherits |

**Every PowerPoint move has a Quarto equivalent, and most are one
line of code instead of a menu hunt.**

# Your first revealjs file

Change one line of YAML at the top of any `.qmd` file:

```yaml
---
title: "BC Reforestation Trends, 1987–2023"
author: "Your name"
date: today
format: revealjs
---
```

Render. Open the resulting `.html`. You are presenting.

## Slide separators — two patterns

Inside the document, two patterns create a new slide:

**Pattern 1 — heading-based:**

```markdown
## Slide 1 title

content

## Slide 2 title

more content
```

Each `## ` heading starts a new slide. **Recommended for most
talks** — the title is built into the slide automatically.

**Pattern 2 — horizontal-rule-based:**

```markdown
A slide with no title.

---

Another slide with no title.
```

The `---` (three dashes on a blank line) creates an
untitled slide. Useful for image-only slides or section
transitions.

::: {.callout-tip}
## Heading hierarchy in revealjs

- **`# Section Title`** (h1) — creates a *section divider* slide.
- **`## Slide title`** (h2) — creates a *content slide*.
- **`### Subheading`** (h3) — used within a slide.

A talk with three sections, each containing three slides, has
`#` × 3 and `##` × 9.
:::

# The one-idea-per-slide rule

The strongest discipline in slide design: **one slide, one
idea**. If your slide has two unrelated points, split it.

Symptoms of slides that need splitting:

- More than ~50 words of body text.
- More than one chart per slide.
- A title with the word *"and"* in it (*"Methods and Results"*).
- Bullet points that wrap to a third line.

A revealjs deck of 20 short slides is almost always clearer than
the same content jammed into 8 dense slides.

# Embedding R output in slides

Code chunks work the same way in revealjs as in HTML reports —
the output appears on the slide where the chunk lives:

```` markdown
## BC reforestation trend

```{r}
#| echo: false
#| fig.width: 8
#| fig.height: 4.5
#| fig-alt: "Bar chart of BC reforested area in hectares by fiscal year, used as a slide figure."
library(tidyverse)
library(readxl)
library(here)

bc <- read_excel(here("data", "bc_disturbance_reforestation.xlsx"),
                 sheet = "Data")

ggplot(bc, aes(x = Fiscal_Year, y = Reforestation_ha)) +
  geom_col(fill = "#2D6A4F") +
  labs(title = NULL,
       x = "Fiscal year", y = "Reforested area (ha)") +
  theme_minimal(base_size = 18)
```
````

Notice the **`base_size = 18`** — slide text needs to be larger
than report text. A 12-point chart that reads fine in a printed
report is illegible on a projector. **Always increase font sizes
in slide charts.**

# Speaker notes

A `::: {.notes}` block adds notes that you see in presenter mode
but the audience never sees:

```markdown
## BC reforestation trend

[chart]

::: {.notes}
Talking points: emphasise the post-2010 increase, mention the
2018 dip caused by silviculture-budget cuts, prompt audience
to predict 2024.
:::
```

Press **`S`** during the live presentation to open presenter mode
in a separate window.

# Incremental reveals

Sometimes you want bullets to appear one at a time:

```markdown
## Three findings

::: {.incremental}
- Reforestation has risen ~50% since 2010.
- Natural disturbance peaked in 2017 wildfires.
- The ratio reforestation : disturbance crossed 1 in 2019.
:::
```

The reader sees the first bullet, presses → to reveal the
second, again for the third. Use sparingly — incremental
reveals can frustrate audiences if overused.

# Two-column layout

For comparing two charts or putting text next to an image:

```markdown
## Static vs interactive maps

::: {.columns}
::: {.column width="50%"}
**Static** — `geom_sf`

- One PNG
- Fixed view
- Goes in printed reports
:::

::: {.column width="50%"}
**Interactive** — `leaflet`

- HTML widget
- Pan, zoom, click
- Goes in HTML deliverables
:::
:::
```

# A complete revealjs `.qmd` skeleton

Copy this as a starting point for a 10-slide talk:

```yaml
---
title: "BC Air Quality, January 2000"
subtitle: "An FRST 232 group presentation"
author: "Members A, B, C, D"
date: 2027-04-10
format:
  revealjs:
    theme: simple
    incremental: false
    transition: slide
    slide-number: true
    chalkboard: false
    embed-resources: true
    code-overflow: wrap
---

# Background

## The question

What did January 2000 PM2.5 look like across Metro Vancouver?

::: {.notes}
Set up the question — the file we cleaned in Ch 6 is hourly PM2.5
across 16 stations.
:::

## The data

::: {.incremental}
- 11,904 rows × 27 columns raw
- 16 stations across Metro Vancouver
- Source: Environmental Reporting BC
- Licence: Open Government Licence — BC
:::

# Cleaning

## The pipeline

[paste the Ch 6 dplyr pipeline as a code chunk with `echo: true`]

# Findings

## Per-station means

[paste the bar chart of mean PM2.5 by station]

## Regional differences

[paste the box plot from Ch 9]

## On a map

[paste the leaflet capstone map from Ch 11]

# Conclusion

## What we found

One sentence. Big and centered.

## Thank you

Questions?
```

Render. Adjust. Iterate.

# Building a portfolio

A **portfolio** is a single landing page that links to your
previously rendered analyses. Your future employer, TA, or
classmate visits one URL and can browse everything you have
built.

## Simple folder structure

```
frst232/
├── _quarto.yml                  ← project config
├── index.qmd                    ← portfolio landing page
├── docs/                        ← rendered output (auto-generated)
└── chapters/
    ├── ch05.qmd
    ├── ch06.qmd
    ├── ...
```

## Minimal `_quarto.yml`

```yaml
project:
  type: website
  output-dir: docs

website:
  title: "FRST 232 — Forestry Data Analysis"
  navbar:
    left:
      - href: index.qmd
        text: Home
      - href: chapters/ch05.qmd
        text: Ch 5 — Inspection
      - href: chapters/ch06.qmd
        text: Ch 6 — Cleaning
      - href: chapters/ch07.qmd
        text: Ch 7 — Summarising
      - href: chapters/ch09.qmd
        text: Ch 9 — Charts
      - href: chapters/ch11.qmd
        text: Ch 11 — Maps

format:
  html:
    theme: cosmo
    toc: true
```

`quarto render` produces a complete navigable site in `docs/`.
Drop that folder anywhere — a USB stick, a web server, GitHub
Pages — and it renders as a real website.

## A minimal `index.qmd`

```markdown
---
title: "FRST 232 Portfolio"
---

# About this site

A collection of forestry data analyses produced for FRST 232
(UBC, Fall 2026). Each linked report covers a different
methodological topic, using openly licensed BC and Ontario
forestry data.

## Highlights

- [Chapter 6 — Cleaning the BC air-quality file](chapters/ch06.html)
- [Chapter 9 — Visualising reforestation trends](chapters/ch09.html)
- [Chapter 11 — Interactive map of monitoring stations](chapters/ch11.html)

## Author

Your name · [LinkedIn / email]

## Licence

All content © 2026 Your Name, released under
[CC BY 4.0](https://creativecommons.org/licenses/by/4.0/).
Code is released under the MIT licence.

## Data sources

- BC silviculture: Environmental Reporting BC
- BC air quality: Environmental Reporting BC
- Ontario forest stats: Government of Ontario
```

# Citation — citing data, code, and packages

Every analysis you publish should make three things easy to
find:

1. **The data source.** Who published it? Where? Under what
   licence?
2. **The code dependencies.** Which R packages did you use?
3. **The methods.** Which previously published work did you
   build on?

## A `CITATION.md` in your project root

```markdown
# Citations

## Data

- BC Government, Environmental Reporting BC. *Trends in
  Silviculture in British Columbia*. Open Government Licence —
  BC. <https://catalogue.data.gov.bc.ca/dataset/indicator-summary-data-trends-in-silviculture-in-b-c->.
  Accessed 2026-09-15.

- Government of Ontario. *Forest Resources of Ontario 2021*.
  Open Government Licence — Ontario.
  Accessed 2026-09-22.

## R packages

- Wickham H et al. (2019). Welcome to the tidyverse. *Journal of
  Open Source Software*, 4(43), 1686.
- Pebesma E (2018). Simple Features for R: Standardized Support
  for Spatial Vector Data. *The R Journal* 10(1), 439–446.
- Cheng J, Karambelkar B, Xie Y (2024). leaflet: Create
  Interactive Web Maps with the JavaScript "Leaflet" Library.
```

In R, `citation("package_name")` returns the BibTeX entry for any
package. Copy the entries into your `CITATION.md`.

# Licensing — choose one openly

Three commonly used licences for student work:

| Licence | What it allows | When to use |
|---|---|---|
| **CC BY 4.0** | Anyone can reuse and modify, as long as they credit you | Reports, charts, written analyses — the default for OER work. |
| **CC BY-NC 4.0** | Same as above but no commercial use | If you want to restrict business reuse |
| **MIT** | Free reuse of code, must include copyright notice | Code (R scripts, Quarto sources) |

Many real-world projects combine: **CC BY for prose and figures,
MIT for code**. Both can co-exist in one repository.

Add a one-line statement to the bottom of your portfolio:

```markdown
© 2026 Your Name · Content licensed under CC BY 4.0 · Code under MIT
```

# Publishing — three free hosts

Once `quarto render` produces a `docs/` folder, you can drop it
on:

| Host | What it gives you | Effort |
|---|---|---|
| **Quarto Pub** | A unique URL like `yourname.quarto.pub/portfolio/`. Free. | One command: `quarto publish quarto-pub` |
| **GitHub Pages** | Free hosting on `yourname.github.io/repo-name`. Needs a GitHub account. | Push to GitHub, enable Pages in repo settings |
| **Netlify** | Same idea, more enterprise features. Free tier generous. | Drag-and-drop the `docs/` folder onto the Netlify dashboard |

All three are zero-cost for student-scale traffic. The first one
is the easiest by far — `quarto publish quarto-pub` from your
project root, sign in, done.

# Active learning

## Activity 1 — Convert a chapter to slides

Take your `frst232_ch09_learner.qmd` (charts chapter). Change
the YAML `format:` from `html` to `revealjs`. Re-render. Inspect
what changes — which charts make readable slides, which need
bigger fonts.

## Activity 2 — One idea per slide

Take the rendered HTML from Activity 1. Identify any slide with
more than one chart or more than ~50 words of text. Split each
into two slides.

## Activity 3 — Two-column layout

Build a two-column slide: left column has a chart, right column
has three bullet takeaways.

## Activity 4 — Citation

Run `citation("tidyverse")` and `citation("sf")` in the R
console. Copy the output into a `CITATION.md` file in your
`frst232/` project root.

## Activity 5 — Apply a licence

Add a one-line licence statement to the bottom of your
`index.qmd` (CC BY 4.0 for content, MIT for code).

## Activity 6 — Publish

Run `quarto publish quarto-pub` from your project root. Sign
in to Quarto Pub when prompted. Confirm the resulting URL
loads in a browser.

# Group lab activity — In-class presentations

::: {.callout-note appearance="default"}
## Lab: The synthesis presentations

The lab session for this chapter is given over to **in-class
group presentations**. Each group presents one analysis from
the term, using a revealjs deck they have built in advance.

**Format**

- 4 groups × 12 minutes each (10 min talk + 2 min Q&A).
- Each group covers one chapter's headline analysis (Ch 6
  cleaning, Ch 7 summarising, Ch 9 charts, Ch 11 maps).
- Each group member speaks for at least 2 minutes.

**Required slide structure**

1. **Title slide** — title, group members, date.
2. **The question** — what forestry question are we answering?
3. **The data** — source, licence, row/column counts.
4. **The method** — show the actual R code (≤ 10 lines).
5. **The result** — one chart, table, or map.
6. **Interpretation** — one sentence headline + one paragraph
   nuance.
7. **Limitations** — what does this analysis NOT show?
8. **Acknowledgements** — citations of data sources and
   packages.

**Peer evaluation**

Every learner submits a peer-eval form for every other group,
scoring:

- Clarity of the question (0–3)
- Quality of the slides (0–3)
- Depth of interpretation (0–3)
- Acknowledgement of limitations (0–3)

**Submission**

- The rendered revealjs HTML.
- The source `.qmd`.
- The peer-eval forms.
:::

# Exercises

These are the take-home exercises for this chapter.

1. **Change the format.** Take any Chapter 5–11 learner `.qmd`
   you have rendered. Change the YAML `format:` line from `html`
   to `revealjs`. Re-render. What changes?

2. **Slide separators.** In your converted file, identify three
   places where you should add `## ` headings or `---`
   separators to break a long page into multiple slides.

3. **One slide, one chart.** Refactor one chapter so each
   slide contains exactly one chart (or table, or map). Compare
   the new deck length to the original page count.

4. **Speaker notes.** Add `::: {.notes}` blocks to three of
   your slides with talking points only you will see.

5. **Two-column layout.** Build one slide with a two-column
   layout — a chart on the left, three bullet takeaways on the
   right.

6. **Build a portfolio.** Create `_quarto.yml`, `index.qmd`,
   and a navbar that links to at least three of your previous
   chapter deliverables. Run `quarto render` and inspect the
   resulting `docs/` folder.

7. **CITATION.md.** Write a `CITATION.md` file listing your
   three data sources and at least five R packages you used
   through the course.

8. **Publish.** Run `quarto publish quarto-pub` and share the
   resulting URL with one classmate. Confirm it loads on their
   computer without any R install.

# Optional, advanced

## Slide themes and templates

revealjs ships with multiple built-in themes:

```yaml
format:
  revealjs:
    theme: [default, dark, sky, beige, simple, serif, blood, night, moon, solarized]
```

Pick one that matches the venue. For most academic/professional
talks, `simple` or `serif` is safest. `dark` works well in
dim conference rooms.

## Custom CSS

If a built-in theme is not quite right:

```yaml
format:
  revealjs:
    theme: simple
    css: my-custom.css
```

Then in `my-custom.css`:

```css
.reveal h1, .reveal h2, .reveal h3 {
  color: #1B4332;
}
.reveal blockquote {
  border-left: 4px solid #2D6A4F;
  padding-left: 1em;
}
```

## Print to PDF during the talk

While presenting in revealjs, press **`E`** to enter print mode,
then **Ctrl+P** / **Cmd+P** to save as PDF. Useful for handouts.

## Multi-author Quarto projects

For a group presentation, share the `.qmd` source via Git
(Chapter 12 references for more). Each member edits one section.
Render only when merging.

# A glimpse of Chapter 13: Wrap-up and final exam preparation

Chapter 13 is the term-end consolidation:

- A complete review of all 12 chapters.
- The Excel-to-R bridge in one summary table.
- The final-exam structure and topics.
- Where to go next: advanced R, dedicated GIS courses,
  reproducibility frameworks.

The chapter lab is the **final exam itself** — closed-AI,
open-notes, covering everything from Chapter 1 to Chapter 12.

# Self-assessment quiz {#sec-quiz-ch12}

```{=html}
<style>
.quiz { background: #F4F7F4; border-left: 4px solid #2D6A4F; padding: 1em 1.5em; margin: 1.5em 0; border-radius: 6px; }
.quiz .q { font-weight: 600; color: #1B4332; margin-bottom: 0.6em; }
.quiz ul.qopts { list-style: none; padding: 0; margin: 0; }
.quiz ul.qopts li { padding: 0.55em 0.9em; margin: 0.3em 0; background: white; border: 1px solid #d8e4d8; border-radius: 4px; cursor: pointer; transition: background 0.12s; }
.quiz ul.qopts li:hover { background: #E8F1E9; }
.quiz ul.qopts li.correct { background: #C8E6C9; border-color: #2D6A4F; font-weight: 600; }
.quiz ul.qopts li.wrong { background: #FFCDD2; border-color: #B05C50; }
.quiz .explanation { display: none; margin-top: 0.7em; padding: 0.7em 1em; background: #FFF8E8; border-left: 3px solid #B7791F; border-radius: 4px; font-size: 0.95em; }
.quiz .explanation.show { display: block; }
.quiz textarea { width: 100%; min-height: 4.5em; padding: 0.6em; border: 1px solid #d8e4d8; border-radius: 4px; font-family: inherit; font-size: 0.95em; resize: vertical; margin-top: 0.3em; }
.quiz .reveal-btn { background: #2D6A4F; color: white; border: none; padding: 0.45em 0.9em; margin-top: 0.5em; border-radius: 4px; cursor: pointer; font-size: 0.9em; }
.quiz .reveal-btn:hover { background: #1B4332; }
.quiz .sample { display: none; margin-top: 0.6em; padding: 0.6em 0.9em; background: #FFF8E8; border-left: 3px solid #B7791F; border-radius: 4px; font-size: 0.95em; font-style: italic; }
.quiz .sample.show { display: block; }
</style>

<div class="quiz" id="q1" data-hints="Quarto is single-source, many-outputs.||Change just the output format, not the content.||Set format from html to revealjs in the YAML."><div class="q">1. To turn an HTML Quarto report into a presentation, you change —</div><ul class="qopts"><li data-correct="false">Every code chunk</li><li data-correct="true">The YAML <code>format:</code> line from html to revealjs</li><li data-correct="false">The entire R session</li><li data-correct="false">Nothing — they are the same</li></ul><div class="explanation">Quarto's single-source-many-outputs design means one YAML change reformats the same content as slides.</div></div>

<div class="quiz" id="q2" data-hints="It is about clarity, not a literal word count.||Split when you have two unrelated ideas.||Each slide should make one clear point."><div class="q">2. The "one idea per slide" rule means —</div><ul class="qopts"><li data-correct="false">Use exactly one word per slide</li><li data-correct="true">Each slide should make one clear point — split if you have two unrelated ideas</li><li data-correct="false">Use one slide per chapter</li><li data-correct="false">It is just a suggestion</li></ul><div class="explanation">Dense slides confuse audiences. Split when you have two ideas, two charts, or 50+ words.</div></div>

<div class="quiz" id="q3" data-hints="Slides are separated by headings or a rule.||A heading, or a horizontal rule on a blank line.||A ## heading or a --- rule."><div class="q">3. In revealjs, a new slide is created by —</div><ul class="qopts"><li data-correct="false">Pressing Enter twice</li><li data-correct="true">A `## ` heading or a `---` horizontal rule on a blank line</li><li data-correct="false">A {slide} block</li><li data-correct="false">A separate file per slide</li></ul><div class="explanation">Heading-based separators (## Title) are most common; --- creates an untitled slide.</div></div>

<div class="quiz" id="q4" data-hints="Heading level sets the slide's role.||h1 is the highest level — it divides sections.||A section-divider slide."><div class="q">4. A `# Section Title` (h1) in revealjs creates —</div><ul class="qopts"><li data-correct="false">A regular content slide</li><li data-correct="true">A section-divider slide between groups of content slides</li><li data-correct="false">An error</li><li data-correct="false">A speaker note</li></ul><div class="explanation">h1 = section divider; h2 = content slide; h3 = subheading within a slide.</div></div>

<div class="quiz" id="q5" data-hints="Content you see but the audience does not.||Visible in presenter mode.||Speaker notes."><div class="q">5. `::: {.notes}` blocks contain —</div><ul class="qopts"><li data-correct="false">Code that does not run</li><li data-correct="true">Speaker notes — visible to you in presenter mode, hidden from the audience</li><li data-correct="false">Footnotes</li><li data-correct="false">References</li></ul><div class="explanation">Press S during the live presentation to open presenter mode in a separate window.</div></div>

<div class="quiz" id="q6" data-hints="Slides run code just like reports do.||The output lands on the slide with the chunk.||They run and embed their output on the slide."><div class="q">6. R code chunks in revealjs slides —</div><ul class="qopts"><li data-correct="false">Do not run</li><li data-correct="true">Run and embed their output (charts, tables, maps) on the slide where they live</li><li data-correct="false">Run only on the title slide</li><li data-correct="false">Require a separate library</li></ul><div class="explanation">Same as in HTML reports — the chunk runs, the output appears.</div></div>

<div class="quiz" id="q7" data-hints="Projector audiences sit far from the screen.||Bigger text reads better at a distance.||A larger base_size (about 16-20)."><div class="q">7. When embedding a chart in a slide, you should set —</div><ul class="qopts"><li data-correct="false">A smaller font than for a report</li><li data-correct="true">A larger base_size (e.g., 16–20) so the chart reads on a projector</li><li data-correct="false">No font change</li><li data-correct="false">Black text only</li></ul><div class="explanation">Projector audiences sit further away than report readers. Increase chart font sizes accordingly.</div></div>

<div class="quiz" id="q8" data-hints="The bullets are revealed progressively.||One at a time as the presenter advances.||Bullets that appear one at a time."><div class="q">8. `::: {.incremental}` blocks produce —</div><ul class="qopts"><li data-correct="false">A loop in the code</li><li data-correct="true">Bullets that appear one at a time as the presenter advances</li><li data-correct="false">A counter</li><li data-correct="false">A footnote</li></ul><div class="explanation">Use sparingly — overusing incremental reveals frustrates audiences.</div></div>

<div class="quiz" id="q9" data-hints="Columns nest inside a parent block.||A .columns parent with child .column blocks.||A .columns block containing two .column blocks."><div class="q">9. A two-column slide layout uses —</div><ul class="qopts"><li data-correct="false">Two separate files</li><li data-correct="true">A `::: {.columns}` parent block with two `::: {.column width="50%"}` child blocks</li><li data-correct="false">A table</li><li data-correct="false">Tabs</li></ul><div class="explanation">Adjust widths as needed (e.g., 40/60). The columns blocks must be inside a columns block.</div></div>

<div class="quiz" id="q10" data-hints="It is more than one document — it is navigable.||A landing page plus linked deliverables.||A Quarto website showcasing your best work."><div class="q">10. A "portfolio" in the context of this course is —</div><ul class="qopts"><li data-correct="false">A single document</li><li data-correct="true">A Quarto website (one landing page + multiple linked chapter deliverables) showcasing your best work</li><li data-correct="false">A printed binder</li><li data-correct="false">A LinkedIn profile</li></ul><div class="explanation">A landing page (index.qmd) + multiple chapter pages + a _quarto.yml = a navigable website.</div></div>

<div class="quiz" id="q11" data-hints="It is the single config file at the project root.||It defines structure, navigation, and format defaults.||The project config file."><div class="q">11. `_quarto.yml` is —</div><ul class="qopts"><li data-correct="false">A data file</li><li data-correct="true">The project config file that defines the site's structure, navigation, and format defaults</li><li data-correct="false">A backup file</li><li data-correct="false">Optional metadata</li></ul><div class="explanation">_quarto.yml is to a Quarto project what package.json is to a Node project — the single config file at the root.</div></div>

<div class="quiz" id="q12" data-hints="R has a built-in function that returns a package's citation.||Run it with the package name in the console.||citation('package_name')."><div class="q">12. To cite an R package in your report, run —</div><ul class="qopts"><li data-correct="false">help(package)</li><li data-correct="true">citation("package_name") in the R console</li><li data-correct="false">summary(package)</li><li data-correct="false">There is no standard way</li></ul><div class="explanation">citation() returns the BibTeX/plain-text reference the package authors want you to use.</div></div>

<div class="quiz" id="q13" data-hints="It is the most permissive Creative Commons licence.||Reuse and adapt freely — one condition.||Anyone can reuse/adapt, as long as they credit you."><div class="q">13. The CC BY 4.0 licence —</div><ul class="qopts"><li data-correct="false">Prohibits commercial use</li><li data-correct="true">Allows anyone to reuse and adapt your work, as long as they credit you</li><li data-correct="false">Requires payment</li><li data-correct="false">Only applies to code</li></ul><div class="explanation">CC BY 4.0 is the most permissive Creative Commons licence — credit required, otherwise free.</div></div>

<div class="quiz" id="q14" data-hints="All BC Data Catalogue datasets share one licence.||Attribution required, otherwise free to use.||Open Government Licence — British Columbia (OGL-BC)."><div class="q">14. The BC silviculture and air-quality data you used this term is licensed under —</div><ul class="qopts"><li data-correct="false">CC BY 4.0</li><li data-correct="true">Open Government Licence — British Columbia (OGL-BC)</li><li data-correct="false">MIT</li><li data-correct="false">Proprietary</li></ul><div class="explanation">All BC Government datasets in the Data Catalogue are OGL-BC — similar to CC BY 4.0 in spirit; attribution required, otherwise free to use.</div></div>

<div class="quiz" id="q15" data-hints="It pushes the rendered site to free hosting.||You get back a public URL.||Uploads the site to Quarto Pub and gives a public URL."><div class="q">15. <code>quarto publish quarto-pub</code> does what?</div><ul class="qopts"><li data-correct="false">Saves a local PDF</li><li data-correct="true">Uploads the rendered site to Quarto Pub (free hosting) and gives you a public URL</li><li data-correct="false">Compiles the document</li><li data-correct="false">Installs Quarto</li></ul><div class="explanation">Quarto Pub is RStudio/Posit's free hosting service. One command from your project root.</div></div>

<div class="quiz" id="q16" data-hints="S opens a second window for the presenter.||It shows your notes and the upcoming slide.||Presenter mode."><div class="q">16. During a live revealjs presentation, pressing <strong>S</strong> —</div><ul class="qopts"><li data-correct="false">Saves the file</li><li data-correct="true">Opens presenter mode in a separate window, showing speaker notes and the next slide</li><li data-correct="false">Stops the slideshow</li><li data-correct="false">Searches the slides</li></ul><div class="explanation">Other useful keys: F = fullscreen, E = print-PDF mode, B = blank, ESC = overview.</div></div>

<div class="quiz" id="q17" data-hints="In PowerPoint every chart is a manual paste.||In Quarto a chart is a chunk that re-runs.||Charts re-render automatically when the code changes."><div class="q">17. The biggest reproducibility advantage of revealjs over PowerPoint is —</div><ul class="qopts"><li data-correct="false">Smaller file size</li><li data-correct="true">Charts and tables re-render automatically when the underlying R code changes — no copy-pasting</li><li data-correct="false">Animations</li><li data-correct="false">More fonts</li></ul><div class="explanation">In PowerPoint, every chart is a paste. In Quarto, every chart is a chunk that re-runs.</div></div>

<div class="quiz" id="q18" data-hints="Credible analyses say what they do NOT show.||Naming caveats builds trust.||A limitations slide."><div class="q">18. A required slide in a professional analytical presentation is —</div><ul class="qopts"><li data-correct="false">A list of your hobbies</li><li data-correct="true">A "limitations" slide that names what the analysis does NOT show</li><li data-correct="false">A long bio</li><li data-correct="false">An advertisement</li></ul><div class="explanation">Naming limitations builds credibility. Audiences trust analysts who acknowledge what they do not know.</div></div>

<div class="quiz" id="q19" data-hints="Three free, student-scale hosts for Quarto sites.||Not Office apps or cloud drives.||Quarto Pub, GitHub Pages, Netlify."><div class="q">19. Three commonly used free hosts for Quarto sites are —</div><ul class="qopts"><li data-correct="false">Word, Excel, PowerPoint</li><li data-correct="true">Quarto Pub, GitHub Pages, Netlify</li><li data-correct="false">AWS, GCP, Azure (paid)</li><li data-correct="false">Dropbox, Google Drive, OneDrive</li></ul><div class="explanation">All three are free at student-scale. Quarto Pub is easiest (one command); GitHub Pages requires Git; Netlify is drag-and-drop.</div></div>

<div class="quiz" id="q20"><div class="q">20. <strong>Reflection.</strong> Compare this chapter's "single-source-many-outputs" idea (one Quarto file → HTML + slides + PDF) with what you used to do across Word + PowerPoint + PDF separately. What changes in how you would build a deliverable next time?</div><textarea placeholder="Type your answer here..."></textarea><button class="reveal-btn" onclick="this.nextElementSibling.classList.toggle('show')">Show common answers</button><div class="sample">Common reflections: <em>I'll write the analysis once and reformat for any audience</em> · <em>no more "the chart is wrong in the slides but right in the report" bug because the chart is generated from one source</em> · <em>I can re-render in 30 seconds if the data updates instead of remaking three documents</em> · <em>publishing as a URL feels more professional than emailing a .docx attachment</em>.</div></div>

<script>
(function() {
  document.querySelectorAll('.quiz').forEach(function(q) {
    q.querySelectorAll('ul.qopts li').forEach(function(opt) {
      opt.addEventListener('click', function() {
        if (q.dataset.answered) return;
        q.dataset.answered = '1';
        var isCorrect = opt.getAttribute('data-correct') === 'true';
        if (isCorrect) opt.classList.add('correct');
        else {
          opt.classList.add('wrong');
          q.querySelectorAll('ul.qopts li').forEach(function(o) {
            if (o.getAttribute('data-correct') === 'true') o.classList.add('correct');
          });
        }
        var exp = q.querySelector('.explanation');
        if (exp) exp.classList.add('show');
      });
    });
  });
})();
</script>
```

# AI as a debugging companion

::: {.callout-tip}
## Useful prompts

- *"My revealjs slide has too much text. Help me split it into
  three shorter slides."* (Paste the slide content.)

- *"My ggplot chart looks fine in HTML but the labels are tiny
  on the projector. What do I change?"* (Answer: increase
  `base_size` in `theme_*(base_size = 18)`.)

- *"Write a CITATION.md entry for the readxl R package."*
:::

# Reading

- *Quarto Presentations* guide — the canonical reference:
  <https://quarto.org/docs/presentations/revealjs/>
- *Quarto Websites* guide:
  <https://quarto.org/docs/websites/>
- Nancy Duarte, *Slide:ology* — short, opinionated, and the
  single best one-book primer on slide design.
- Creative Commons licence chooser:
  <https://creativecommons.org/choose/>
- Open Government Licence — BC:
  <https://www2.gov.bc.ca/gov/content/data/policy-standards/open-data/open-government-licence-bc>

# Instructor notes {.unnumbered}

::: {.callout-note appearance="minimal" collapse="true"}
## Lab logistics

- The lab session for this chapter is in-class group
  presentations.
- 4 groups × ~12 minutes each fits in a 60–75 minute lab.
- Bring the projector hook-up that matches the venue; have a
  PDF backup of each group's slides in case revealjs fails on
  the projector machine.
- Encourage groups to use **presenter mode (S)** during the
  talk — speaker notes are vastly more useful than memorising
  bullet points.

## Grading rubric

| Criterion | Weight | What to look for |
|---|---|---|
| Title slide + agenda | 10 % | All members named; clear topic |
| The question | 10 % | Stated as one sentence, not a paragraph |
| The data | 10 % | Source, licence, row counts |
| The method | 15 % | Real R code, not pseudocode |
| The result | 25 % | One clear chart, table, or map |
| Interpretation | 15 % | Headline + nuance |
| Limitations | 10 % | What the analysis does NOT show |
| Acknowledgements | 5 % | Data citations + R-package citations |

## Materials provided

| File | Purpose |
|---|---|
| `frst232_ch12_learner.qmd` | Quarto starter — exercises walking through revealjs and portfolio assembly. |
| `frst232_ch12_solutions.qmd` | Completed solutions. **Do not distribute.** |
| `frst232_ch12_slides_template.qmd` | A ready-to-customize revealjs presentation skeleton. |
| `quiz_ch12.html` | Standalone interactive quiz. |

## Common stumbling points

- **Chart fonts too small on projector.** Insist on
  `theme_*(base_size = 18)` in slide chunks.
- **No limitations slide.** Push back hard — limitations are
  the single best signal of professional maturity in an
  analytical presentation.
- **Wall-of-text slides.** Apply the 50-word rule live;
  refactor in front of the class.
- **Missing data citations.** Every dataset used must be
  credited on the acknowledgement slide.
- **`format: revealjs` not rendering.** Some Posit Cloud
  setups need `install.packages("rmarkdown")` even though
  Quarto is bundled. The error message is descriptive.
:::
