---
title: "Wrap-up and Final-Exam Preparation"
---

```{r setup-ch13}
#| include: false
library(tidyverse)
library(readxl)
library(here)
library(knitr)
```

::: {.callout-note appearance="simple"}
### Data for this chapter
The end-to-end case study uses the **BC silviculture** workbook
([`bc_disturbance_reforestation.xlsx`](../data/bc_disturbance_reforestation.xlsx){download="bc_disturbance_reforestation.xlsx"}).
Click the file name to download it directly, then save it in your `data/`
folder. For the source and details, see the
[Datasets Used in This Book](../datasets.qmd) page.
:::

## Chapter goals {.unnumbered}

By the end of this chapter you will be able to:

- Trace the full reproducible workflow taught in this book —
  **import → clean → summarize → reshape → join → visualize →
  render → explain** — on a single forestry dataset.
- Reproduce a complete analysis from a clean project, from a raw
  file to a rendered Quarto report.
- Apply a repeatable **debugging ladder** to any error you meet
  in the final exam (and on the job).
- Name concrete **next steps** for learning beyond this course.

## Expected output for this chapter {.unnumbered}

::: {.callout-tip appearance="default"}
### What you will produce

A complete **end-to-end case-study report**: take one of the
course datasets, start from the raw file, document your cleaning,
summarize by group, produce **one figure** (and an **optional map** only
if you used the spatial extension), and
render it as a single Quarto HTML report. This is both your final
review and the best possible exam practice.
:::

## Where we are

You started in Excel (Chapters 1–3), crossed the bridge into R
(Chapter 4), and built — one chapter at a time — every stage of a
reproducible analysis:

| Stage | You learned it in | Key functions |
|---|---|---|
| Import | Ch 4–5 | `read_excel()`, `read_csv()`, `here()` |
| Inspect | Ch 5 | `glimpse()`, `summary()`, `dim()`, `colSums(is.na())` |
| Clean | Ch 6 | `select()`, `filter()`, `rename()`, `mutate()`, `case_when()` |
| Summarize | Ch 7 | `group_by()`, `summarise()`, `count()`, `n_distinct()` |
| Reshape | Ch 8 | `pivot_longer()`, `pivot_wider()` |
| Join | Ch 8 | `left_join()`, the filtering joins, pre/post-join checks |
| Visualize | Ch 9 | `ggplot2`, geoms, facets, themes, `fig-alt` |
| Render | Ch 5–12 | Quarto: code + prose + output in one document |

This chapter threads them all together.

## End-to-end case study: BC silviculture

We use one dataset — the BC silviculture file you have known since
Chapter 1 — and run it through the whole pipeline.

### 1. Import

```{r case-import}
bc <- read_excel(here("data", "bc_disturbance_reforestation.xlsx"),
                 sheet = "Data")
dim(bc)
```

### 2. Inspect

```{r case-inspect}
glimpse(bc)
colSums(is.na(bc))   # missing values per column
```

### 3. Clean and derive

We add a **decade** grouping variable and a **reforestation
ratio** (reforested area per hectare harvested), documenting each
step as we go:

```{r case-clean}
bc_clean <- bc |>
  mutate(
    decade          = (Fiscal_Year %/% 10) * 10,
    reforest_ratio  = Reforestation_ha / Harvested_ha
  )
bc_clean |> select(Fiscal_Year, decade, Harvested_ha,
                   Reforestation_ha, reforest_ratio) |> head()
```

### 4. Summarize by group

```{r case-summary}
by_decade <- bc_clean |>
  group_by(decade) |>
  summarise(
    n_years           = n(),
    mean_harvest_ha   = round(mean(Harvested_ha, na.rm = TRUE)),
    mean_reforest_ha  = round(mean(Reforestation_ha, na.rm = TRUE)),
    mean_ratio        = round(mean(reforest_ratio, na.rm = TRUE), 2),
    .groups = "drop"
  )
by_decade |> kable()
```

Note the discipline from Chapter 7: **`n_years` is reported beside
every mean**, so a reader can see that the 1980s and 2020s
decades are partial.

### 5. Visualize — one figure

```{r case-figure}
#| fig-cap: "BC harvested and reforested area (ha) by fiscal year, 1987–2023."
#| fig-alt: "Line chart of BC harvested area (solid amber line) and
#|   reforested area (dashed green line) by fiscal year, 1987 to 2023,
#|   with the two series tracking each other and reforestation
#|   generally following harvest."
bc_long <- bc_clean |>
  select(Fiscal_Year, Harvested_ha, Reforestation_ha) |>
  pivot_longer(-Fiscal_Year, names_to = "series", values_to = "area_ha")

ggplot(bc_long, aes(Fiscal_Year, area_ha, colour = series, linetype = series)) +
  geom_line(linewidth = 1) +
  scale_colour_manual(values = c("#B7791F", "#2D6A4F"),
                      labels = c("Harvested", "Reforested")) +
  scale_linetype_manual(values = c("solid", "dashed"),
                        labels = c("Harvested", "Reforested")) +
  labs(title = "BC harvest and reforestation, 1987–2023",
       x = "Fiscal year", y = "Area (hectares)",
       colour = NULL, linetype = NULL,
       caption = "Source: Environmental Reporting BC") +
  theme_minimal()
```

### 6. Render and explain

Everything above lives in one `.qmd` file. Pressing **Render**
produces a single HTML report that another analyst can reproduce
from the raw file — no hidden clicks, no copy-paste. That
reproducibility is the whole point of this course.

## Final-exam preparation: the debugging ladder

The exam is not about memorizing function names — it is about
**recovering when something breaks**. When you hit an error,
climb this ladder one rung at a time:

1. **Read the error message.** R errors are more helpful than they
   look. Note the function named and the object named.
2. **Identify the step.** Which is wrong — the *object*, the
   *function*, or the *data*? (`object not found` → object;
   `could not find function` → package not loaded; an unexpected
   number of rows → data.)
3. **Inspect the data.** `glimpse()`, `summary()`, `dim()`,
   `names()`. Most "code" bugs are really data surprises.
4. **Reduce to a minimal example.** Strip the pipeline back to the
   smallest piece that still fails.
5. **Test one fix.** Change one thing, re-run, observe. Not three
   things at once.
6. **Document the verification.** Note what the fix was and how you
   confirmed it — a row count, a spot check, a re-render.

::: {.callout-note}
### Using AI during study (not during a closed exam)

When studying, AI is excellent for *explaining* an error or
*generating practice variants*. Paste the **exact** error, your R
and package versions, and the code you tried. Then verify the
explanation against `?function_name` and a small test — the same
discipline the rest of the book taught. In a closed-book exam,
the ladder above is what you carry in your head.
:::

## Active learning — exam practice

### Activity 1 — Full pipeline, new dataset
Run the seven-step pipeline above on a *different* course dataset
(Ontario forest stats, or the cleaned air-quality file). Produce
one summary table and one figure.

### Activity 2 — Break and fix
Introduce a deliberate bug (misspell a column, drop a
`library()`, swap `lat`/`lon`). Use the debugging ladder to find
and fix it. Write one sentence per rung.

### Activity 3 — Reproduce from clean
Move your `.qmd` and `data/` to a fresh folder and render. Does it
work with no manual steps? If not, find the hidden dependency.

## Where to go next

This course is a foundation. Natural next steps:

- **Statistical modelling** — regression, mixed models, and
  growth-and-yield models in R (FRST 231 / 430 build directly on
  this).
- **Spatial data science** — deeper `sf`, raster data with
  `terra`, and remote sensing for forest inventory.
- **Time series** — trends, seasonality, and forecasting for
  climate and disturbance data.
- **Reproducible research** — `renv`, Git/GitHub, and
  parameterized Quarto reports for production work.

Recommended reading mixes canonical and diverse voices:

- Wickham, Çetinkaya-Rundel & Grolemund (2023). *R for Data
  Science* (2e). <https://r4ds.hadley.nz>
- Lovelace, Nowosad & Muenchow (2019). *Geocomputation with R*.
  <https://r.geocompx.org>
- Kimmerer, R. W. (2013). *Braiding Sweetgrass.* Milkweed.
- BC First Nations Forestry Council. <https://www.bcfnfc.ca/>

## AI as a debugging companion {.unnumbered}

Across the book's chapters, AI has been a **verification companion**,
not an author: it explains errors, drafts starting code, and
suggests alternatives — and you check every line against the data
and the documentation. That habit, more than any single function,
is what makes you employable in a world where code is cheap and
*judgment* is scarce.

## Instructor notes {.unnumbered}

::: {.callout-note collapse="true"}
### Pacing, exam logistics, and accommodations

- **Pacing.** One session of review, one of Q&A / sample-exam
  walkthrough, one wrap-up. No graded lab this week.
- **Final exam** covers import, clean, summarize, reshape, join,
  and visualize with R, tidyverse, and ggplot2 — weighted 30% per the
  syllabus. (Spatial mapping with `sf` is an optional extension and is
  not required on the exam.)
- **Accommodations.** Confirm Centre for Accessibility
  arrangements *before* the exam (extended time, alternative
  formats, rest breaks). Where the timed format raises equity
  concerns, consider a take-home practical for the whole cohort,
  not just students with formal accommodations — see the
  *Accessibility statement* in the front matter.
- **Common stumbling points.** Forgotten `library()` calls; data
  paths that work locally but not from a clean project; reading
  `NA` as zero.
:::


---

## Working with AI on this chapter {.unnumbered}

Use AI as a **verification and debugging partner**, not an answer machine.
The goal is not to make AI do the work — it is to use AI to help you
**understand, check, and improve your own work.**

::: {.callout-note collapse="true"}
### AI prompt to check your work
Fill in the brackets and paste this into your AI assistant. It should
*explain and check*, not hand you a final assignment answer:

> *"I am working on FRST 232, this chapter. My dataset is `[file name]`
> with columns `[columns]`. My task is `[what you are trying to do]`. I
> tried `[paste your code or steps]` and got `[paste the exact output or
> error]`. Explain what this does line by line, tell me whether it answers
> the question, and list the checks I should run to verify it. Do not give
> me the final answer."*
:::

**Verify it yourself** — before trusting any result (yours *or* an AI's):

- the rows, columns, and units are what you expect;
- the row count changes sensibly after a `filter()` or a join;
- group sizes and summary values are plausible;
- the figure answers the question and axis labels include units.

**Common AI mistakes to watch for:** inventing a column, function, or value
that is not in your data; choosing a tool before reading the question; and
(in Excel) suggesting the wrong cell range or forgetting to refresh a
PivotTable.

::: {.callout-tip collapse="true"}
### Compare your solution with the reference
When a reference solution is available, do **not** just copy it — use it to
check your own work:

- **Expected result** — does your output match the target shape and values?
- **Required components** — did you include every step the task asked for?
- **If your result differs** — say *where* it matched, *where* it differed,
  *what* caused the difference, *how* you fixed it, and *what* you learned.
:::

::: {.callout-tip collapse="true"}
### AI replication challenge
Write a prompt that would guide an AI *toward* the reference solution
without handing over the final answer. Include the dataset, the columns,
the task goal, the output format you need, and the verification checks.
Then paste (or summarise) the AI's response and answer: did its approach
match the book's? did it miss a verification step? did it invent a column,
function, or value? what did your group change after checking it?
:::
