---
title: "Wrap-up and Final-Exam Preparation"
---

```{r setup-ch13}
#| include: false
library(tidyverse)
library(readxl)
library(here)
library(knitr)
library(sf)
```

::: {.callout-note appearance="simple"}
## Data for this chapter
The end-to-end case study uses the **BC silviculture**
(`bc_disturbance_reforestation.xlsx`) and **CSRD parks**
(`csrd_parks.geojson`) files. Download them from the
[Datasets Used in This Book](../datasets.qmd) page into your `data/`
folder.
:::

# Chapter goals {.unnumbered}

By the end of this chapter you will be able to:

- Trace the full reproducible workflow taught in this book —
  **import → clean → summarize → reshape → join → visualize →
  map → render → explain** — on a single forestry dataset.
- Reproduce a complete analysis from a clean project, from a raw
  file to a rendered Quarto report.
- Apply a repeatable **debugging ladder** to any error you meet
  in the final exam (and on the job).
- Name concrete **next steps** for learning beyond this course.

# Expected output for this chapter {.unnumbered}

::: {.callout-tip appearance="default"}
## What you will produce

A complete **end-to-end case-study report**: take one of the
course datasets, start from the raw file, document your cleaning,
summarize by group, produce **one figure** and **one map**, and
render it as a single Quarto HTML report. This is both your final
review and the best possible exam practice.
:::

# Where we are

You started in Excel (Chapters 1–3), crossed the bridge into R
(Chapter 4), and built — one chapter at a time — every stage of a
reproducible analysis:

| Stage | You learned it in | Key functions |
|---|---|---|
| Import | Ch 4–5 | `read_excel()`, `read_csv()`, `here()` |
| Inspect | Ch 5 | `glimpse()`, `summary()`, `dim()`, `colSums(is.na())` |
| Clean | Ch 6 | `select()`, `filter()`, `rename()`, `mutate()`, `case_when()` |
| Summarize | Ch 7 | `group_by()`, `summarise()`, `count()`, `n_distinct()` |
| Reshape | Ch 8 | `pivot_longer()`, `pivot_wider()` |
| Join | Ch 8 | `left_join()`, the filtering joins, pre/post-join checks |
| Visualize | Ch 9 | `ggplot2`, geoms, facets, themes, `fig-alt` |
| Map | Ch 10–11 | `sf`, `st_transform()`, `geom_sf()`, `leaflet` |
| Render | Ch 5–12 | Quarto: code + prose + output in one document |

This chapter threads them all together.

# End-to-end case study: BC silviculture

We use one dataset — the BC silviculture file you have known since
Chapter 1 — and run it through the whole pipeline.

## 1. Import

```{r case-import}
bc <- read_excel(here("data", "bc_disturbance_reforestation.xlsx"),
                 sheet = "Data")
dim(bc)
```

## 2. Inspect

```{r case-inspect}
glimpse(bc)
colSums(is.na(bc))   # missing values per column
```

## 3. Clean and derive

We add a **decade** grouping variable and a **reforestation
ratio** (reforested area per hectare harvested), documenting each
step as we go:

```{r case-clean}
bc_clean <- bc |>
  mutate(
    decade          = (Fiscal_Year %/% 10) * 10,
    reforest_ratio  = Reforestation_ha / Harvested_ha
  )
bc_clean |> select(Fiscal_Year, decade, Harvested_ha,
                   Reforestation_ha, reforest_ratio) |> head()
```

## 4. Summarize by group

```{r case-summary}
by_decade <- bc_clean |>
  group_by(decade) |>
  summarise(
    n_years           = n(),
    mean_harvest_ha   = round(mean(Harvested_ha, na.rm = TRUE)),
    mean_reforest_ha  = round(mean(Reforestation_ha, na.rm = TRUE)),
    mean_ratio        = round(mean(reforest_ratio, na.rm = TRUE), 2),
    .groups = "drop"
  )
by_decade |> kable()
```

Note the discipline from Chapter 7: **`n_years` is reported beside
every mean**, so a reader can see that the 1980s and 2020s
decades are partial.

## 5. Visualize — one figure

```{r case-figure}
#| fig-alt: "Line chart of BC harvested area and reforested area
#|   by fiscal year, 1987 to 2023, with the two series tracking
#|   each other and reforestation generally following harvest."
bc_long <- bc_clean |>
  select(Fiscal_Year, Harvested_ha, Reforestation_ha) |>
  pivot_longer(-Fiscal_Year, names_to = "series", values_to = "area_ha")

ggplot(bc_long, aes(Fiscal_Year, area_ha, colour = series)) +
  geom_line(linewidth = 1) +
  scale_colour_manual(values = c("#B05C50", "#2D6A4F"),
                      labels = c("Harvested", "Reforested")) +
  labs(title = "BC harvest and reforestation, 1987–2023",
       x = "Fiscal year", y = "Area (hectares)", colour = NULL,
       caption = "Source: Environmental Reporting BC") +
  theme_minimal()
```

## 6. Map — one map

A complete analysis often has a spatial dimension. Here we close
the loop with the CSRD parks layer from Chapters 10–11, reprojected
and drawn with `geom_sf()`:

```{r case-map}
#| fig-alt: "Map of Columbia Shuswap Regional District park points
#|   coloured by park type."
parks <- st_read(here("data", "csrd_parks.geojson"), quiet = TRUE)

ggplot(parks) +
  geom_sf(aes(colour = ParkType), size = 2, alpha = 0.85) +
  scale_colour_viridis_d() +
  labs(title = "CSRD parks by type", colour = "Park type",
       caption = "Source: CSRD Open Data") +
  theme_minimal()
```

## 7. Render and explain

Everything above lives in one `.qmd` file. Pressing **Render**
produces a single HTML report that another analyst can reproduce
from the raw file — no hidden clicks, no copy-paste. That
reproducibility is the whole point of this course.

# Final-exam preparation: the debugging ladder

The exam is not about memorizing function names — it is about
**recovering when something breaks**. When you hit an error,
climb this ladder one rung at a time:

1. **Read the error message.** R errors are more helpful than they
   look. Note the function named and the object named.
2. **Identify the step.** Which is wrong — the *object*, the
   *function*, or the *data*? (`object not found` → object;
   `could not find function` → package not loaded; an unexpected
   number of rows → data.)
3. **Inspect the data.** `glimpse()`, `summary()`, `dim()`,
   `names()`. Most "code" bugs are really data surprises.
4. **Reduce to a minimal example.** Strip the pipeline back to the
   smallest piece that still fails.
5. **Test one fix.** Change one thing, re-run, observe. Not three
   things at once.
6. **Document the verification.** Note what the fix was and how you
   confirmed it — a row count, a spot check, a re-render.

::: {.callout-note}
## Using AI during study (not during a closed exam)

When studying, AI is excellent for *explaining* an error or
*generating practice variants*. Paste the **exact** error, your R
and package versions, and the code you tried. Then verify the
explanation against `?function_name` and a small test — the same
discipline the rest of the book taught. In a closed-book exam,
the ladder above is what you carry in your head.
:::

# Active learning — exam practice

## Activity 1 — Full pipeline, new dataset
Run the seven-step pipeline above on a *different* course dataset
(Ontario forest stats, or the cleaned air-quality file). Produce
one summary table and one figure.

## Activity 2 — Break and fix
Introduce a deliberate bug (misspell a column, drop a
`library()`, swap `lat`/`lon`). Use the debugging ladder to find
and fix it. Write one sentence per rung.

## Activity 3 — Reproduce from clean
Move your `.qmd` and `data/` to a fresh folder and render. Does it
work with no manual steps? If not, find the hidden dependency.

# Where to go next

This course is a foundation. Natural next steps:

- **Statistical modelling** — regression, mixed models, and
  growth-and-yield models in R (FRST 231 / 430 build directly on
  this).
- **Spatial data science** — deeper `sf`, raster data with
  `terra`, and remote sensing for forest inventory.
- **Time series** — trends, seasonality, and forecasting for
  climate and disturbance data.
- **Reproducible research** — `renv`, Git/GitHub, and
  parameterized Quarto reports for production work.

Recommended reading mixes canonical and diverse voices:

- Wickham, Çetinkaya-Rundel & Grolemund (2023). *R for Data
  Science* (2e). <https://r4ds.hadley.nz>
- Lovelace, Nowosad & Muenchow (2019). *Geocomputation with R*.
  <https://r.geocompx.org>
- Kimmerer, R. W. (2013). *Braiding Sweetgrass.* Milkweed.
- BC First Nations Forestry Council. <https://www.bcfnfc.ca/>

# AI as a debugging companion {.unnumbered}

Across thirteen chapters, AI has been a **verification companion**,
not an author: it explains errors, drafts starting code, and
suggests alternatives — and you check every line against the data
and the documentation. That habit, more than any single function,
is what makes you employable in a world where code is cheap and
*judgment* is scarce.

# Instructor notes {.unnumbered}

::: {.callout-note collapse="true"}
## Pacing, exam logistics, and accommodations

- **Pacing.** One session of review, one of Q&A / sample-exam
  walkthrough, one wrap-up. No graded lab this week.
- **Final exam** covers import, clean, summarize, reshape, join,
  visualize, and map with R, tidyverse, ggplot2, and `sf` —
  weighted 30% per the syllabus.
- **Accommodations.** Confirm Centre for Accessibility
  arrangements *before* the exam (extended time, alternative
  formats, rest breaks). Where the timed format raises equity
  concerns, consider a take-home practical for the whole cohort,
  not just students with formal accommodations — see the
  *Accessibility statement* in the front matter.
- **Common stumbling points.** Forgotten `library()` calls; data
  paths that work locally but not from a clean project; reading
  `NA` as zero; `lat`/`lon` reversed in `st_as_sf()`.
:::
