---
title: "Forestry Data Analysis — <your title here>"
subtitle: "<one-line description of the question>"
author: "<your name / group>"
date: today
format:
  html:
    toc: true
    toc-depth: 2
    embed-resources: true   # one self-contained file, works offline
    fig-cap-location: bottom
---

```{r setup}
#| include: false
library(tidyverse)
library(readxl)
library(here)
library(knitr)
# library(sf)        # uncomment if your analysis is spatial
# library(leaflet)   # uncomment for an interactive map
```

# The question

State, in one or two sentences, the forestry / natural-resources
question this report answers. A good question is specific and
answerable from the data (e.g., *"How has the share of
clearcutting-with-reserves changed by decade in BC?"*).

# Data and provenance

| Property | Value |
|---|---|
| File(s) | `<filename>` |
| Source | `<agency / catalogue + URL>` |
| Licence | `<e.g., Open Government Licence — BC>` |
| Retrieved | `<YYYY-MM-DD>` |
| Rows × columns | `<n × m>` |

# Import

```{r import}
# raw <- read_excel(here("data", "<file>.xlsx"), sheet = "Data")
# raw <- read_csv(here("data", "<file>.csv"), show_col_types = FALSE)
```

# Cleaning log

Document every transformation in prose *and* code — anyone should be
able to reproduce your analysis-ready dataset from the raw file.

```{r clean}
# clean <- raw |>
#   select(...) |>
#   filter(...) |>
#   mutate(...)
```

- *Step 1 —* what you changed and why.
- *Step 2 —* …

# Summary

Report group sizes (`n()`) beside every mean.

```{r summary}
# clean |>
#   group_by(<group>) |>
#   summarise(n = n(), mean_value = mean(<var>, na.rm = TRUE), .groups = "drop") |>
#   kable()
```

# Figure

```{r figure}
#| fig-alt: "<one-sentence description of what the chart shows, for screen readers>"
# ggplot(clean, aes(x = ..., y = ...)) +
#   geom_...() +
#   labs(title = "...", x = "...", y = "...", caption = "Source: ...") +
#   theme_minimal()
```

# Map (optional)

```{r map}
#| fig-alt: "<description of the map>"
# parks <- st_read(here("data", "<layer>.geojson"), quiet = TRUE)
# ggplot(parks) + geom_sf(aes(colour = <attr>)) + theme_minimal()
```

# Discussion

What does the result mean for forestry / natural-resource management?
What are the limitations of the data?

# AI acknowledgement

If generative AI substantially contributed to code or writing here,
acknowledge it in one line (e.g., *"Draft code generated with an AI
assistant; revised and verified by the team."*).

# References

<List sources used, including the dataset and any documentation.>
