---
title: "FRST 232 — Chapter NN Learner Workbook"
subtitle: "<chapter topic> — your working copy"
author: "<your name>"
date: today
format:
  html:
    toc: true
    embed-resources: true
---

```{r setup}
#| include: false
library(tidyverse)
library(readxl)
library(here)
library(knitr)
```

> **How to use this workbook.** Each `# TODO` is a task. Replace the
> placeholder with your own code, then render (the **Render** button)
> to check it works. Keep the prose notes — they are part of the
> deliverable. Submit the rendered `.html` (and the `.qmd`) to the
> course site.

# Warm-up

```{r warmup}
# TODO: load the chapter's dataset with read_csv()/read_excel() + here()

```

# Exercise 1 — <short title>

*Goal:* <what this exercise practises>.

```{r ex1}
# TODO: your code here

```

*Your answer in words:* <one or two sentences interpreting the result>.

# Exercise 2 — <short title>

```{r ex2}
# TODO

```

# Exercise 3 — <short title>

```{r ex3}
# TODO

```

# Reflection

- What was the hardest step, and how did you debug it?
- If you used an AI assistant, what did you ask, and how did you
  verify its answer?
