---
title: "Chapter 9 — Learner Exercises"
subtitle: "Visualize Forestry Data with ggplot2"
author: "Your name here"
date: today
format:
  html:
    toc: true
    toc-location: left
    theme: cosmo
    embed-resources: true
---

```{r setup}
#| include: false
library(tidyverse)
library(readxl)
library(lubridate)
library(here)
library(knitr)
library(scales)
```

# About this file {.unnumbered}

Learner workbook for Chapter 9. Each exercise has one or more
`# TODO` chunks. Render when done; submit the `.qmd`, the
rendered `.html`, and the PNG from Exercise 6.

```{r load-data}
bc <- read_excel(here("data", "bc_disturbance_reforestation.xlsx"),
                 sheet = "Data")

station_regions <- tribble(
  ~location,                          ~region,
  "Burnaby South",                    "Burrard Peninsula",
  "Coquitlam Douglas College",        "North-East Sector",
  "Langley Central",                  "Fraser Valley West",
  "Maple Ridge Golden Ears School",   "North-East Sector",
  "New Westminster Sapperton Park",   "Burrard Peninsula",
  "North Delta",                      "South of Fraser",
  "North Vancouver Second Narrows",   "North Shore",
  "Pitt Meadows Airport",             "North-East Sector",
  "Port Coquitlam North",             "North-East Sector",
  "Port Moody Rocky Point Park",      "North-East Sector"
)

airdata_clean <- read_csv(here("data", "airdata3.csv"),
                          show_col_types = FALSE) |>
  filter(!is.na(PM25)) |>
  mutate(date_parsed = mdy(Date),
         day = day(date_parsed)) |>
  left_join(station_regions, by = "location") |>
  filter(!is.na(region))
```

---

# Exercise 1 — Bar chart

```{r ex1}
# TODO: build a bar chart of bc$Reforestation_ha by bc$Fiscal_Year.
# Use geom_col() with fill = "#2D6A4F" (dark green).

```

---

# Exercise 2 — Three-line chart

## 2a — Reshape to long format

```{r ex2a}
# TODO: select Fiscal_Year, Harvested_ha, Natural_Disturbance_ha,
# Reforestation_ha from `bc`. Pivot longer with names_to = "series"
# and values_to = "area_ha". Save as `bc_long`.

```

## 2b — Plot

```{r ex2b}
# TODO: plot bc_long with x = Fiscal_Year, y = area_ha,
# colour = series, using geom_line(linewidth = 1).

```

---

# Exercise 3 — Histogram with three binwidths

## 3a — binwidth = 1

```{r ex3a}
# TODO: histogram of airdata_clean$PM25 with binwidth = 1.

```

## 3b — binwidth = 5

```{r ex3b}
# TODO: histogram of airdata_clean$PM25 with binwidth = 5.

```

## 3c — binwidth = 0.5

```{r ex3c}
# TODO: histogram of airdata_clean$PM25 with binwidth = 0.5.

```

## 3d — Choose

> *Write your answer here.* Which binwidth shows the
> distribution shape most clearly? Why?

---

# Exercise 4 — Boxplot with coord_flip

```{r ex4}
# TODO: boxplot of PM25 by region. Add coord_flip() so the region
# names read horizontally.

```

> *Write your answer here.* Which region has the highest median?
> The widest spread?

---

# Exercise 5 — Faceted time series

```{r ex5}
# TODO: plot PM25 over date_parsed, one panel per station, using
# facet_wrap(~ location, scales = "free_y").

```

---

# Exercise 6 — Polish and save

```{r ex6-build}
# TODO: take any chart from Exercises 1-5. Add labs(title, subtitle,
# x, y, caption). Apply theme_minimal(base_size = 12). Save the
# completed plot in a variable called `my_plot`.

```

```{r ex6-save}
#| eval: false
# TODO: save my_plot to outputs/ as a 7x5 inch PNG with dpi = 300.
# Use ggsave() and here().

```

---

# Exercise 7 — Smooth trend

```{r ex7}
# TODO: scatter plot of Reforestation_ha vs Fiscal_Year with
# geom_point() and geom_smooth(method = "lm") overlaid.

```

> *Write your answer here.* Is the trend increasing or decreasing?
> By roughly how many hectares per year?

---

# Group lab — Build a chart that tells a story

## Group members

> *List your group members here.*

## The question

> *Type your group's chosen question here.*

## The chart

```{r group-chart}
# TODO: build the group's chart here.

```

## The one-sentence caption

> *Write the headline finding here.*

---

# Done

When everything renders cleanly, submit:

1. This `.qmd` source.
2. The rendered `.html`.
3. The PNG file from Exercise 6.
