---
title: "Chapter 8 — Learner Exercises"
subtitle: "Reshape and Combine Forestry Datasets Safely"
author: "Your name here"
date: today
format:
  html:
    toc: true
    toc-location: left
    theme: cosmo
    embed-resources: true
---

```{r setup}
#| include: false
library(tidyverse)
library(readxl)
library(here)
library(knitr)
```

# About this file {.unnumbered}

Learner workbook for Chapter 8. Each numbered exercise has one
or more `# TODO` chunks. Render when done; submit the `.qmd` and
the rendered `.html`.

```{r load-data}
bc <- read_excel(here("data", "bc_disturbance_reforestation.xlsx"),
                 sheet = "Data")
ontario <- read_excel(here("data", "on_forest_statistics_1.xlsx"),
                      sheet = "Data")
```

---

# Exercise 1 — Pivot longer

## 1a — Build a wide tibble

```{r ex1a}
# TODO: from `bc`, select Fiscal_Year, Harvested_ha, and
# Natural_Disturbance_ha. Save as `bc_wide`.

```

## 1b — Pivot longer

```{r ex1b}
# TODO: pivot bc_wide to long format with names_to = "disturbance_type"
# and values_to = "area_ha". Save as `bc_long`. Print head(bc_long).
# Then confirm nrow(bc_long) == nrow(bc_wide) * 2.

```

---

# Exercise 2 — All four joins on a small example

```{r ex2-setup}
plots <- tribble(
  ~plot_id, ~district,    ~area_ha,
  "P001",   "Vancouver",    12.0,
  "P002",   "Vancouver",     8.5,
  "P003",   "Squamish",     21.2,
  "P004",   "Squamish",     15.0,
  "P005",   "Nanaimo",       9.7
)

districts <- tribble(
  ~district,         ~bec_zone,
  "Vancouver",       "CWHdm",
  "Squamish",        "CWHvm",
  "Sunshine Coast",  "CWHvh"
)
```

## 2a — left_join

```{r ex2a}
# TODO: left_join plots and districts on "district". Report nrow().

```

## 2b — inner_join

```{r ex2b}
# TODO: inner_join plots and districts on "district". Report nrow().

```

## 2c — right_join

```{r ex2c}
# TODO: right_join plots and districts on "district". Report nrow().

```

## 2d — full_join

```{r ex2d}
# TODO: full_join plots and districts on "district". Report nrow().

```

## 2e — Interpret

> *Write your answer here.* Which join kept the most rows?
> Which kept the fewest? In one sentence each, what did each
> join include or exclude?

---

# Exercise 3 — Filtering joins

## 3a — anti_join

```{r ex3a}
# TODO: anti_join plots and districts on "district". What single row
# does this return, and why?

```

## 3b — semi_join

```{r ex3b}
# TODO: semi_join plots and districts on "district". How many rows?

```

## 3c — Relationship between the two

> *Write your answer here.* What is nrow(plots) compared to
> nrow(anti_join) + nrow(semi_join)? Why must they sum to that?

---

# Exercise 4 — bind_rows for multi-source data

## 4a — Tag the two files

```{r ex4a}
# TODO: build bc_tagged with Fiscal_Year, Harvested_ha, and a
# new column province = "BC".
# Build ontario_sample as a small tribble with 3 rows for 2018-2020,
# each tagged province = "ON".

```

## 4b — Combine

```{r ex4b}
# TODO: bind_rows bc_tagged and ontario_sample into `combined`.
# Print the tail(combined, 6).

```

## 4c — Summarise by province

```{r ex4c}
# TODO: group combined by province; summarise mean Harvested_ha.

```

---

# Exercise 5 — Join diagnostics

## 5a — Row count check

```{r ex5a}
# TODO: compute nrow(plots) and nrow(left_join(plots, districts, by = "district")).
# Are they equal? They should be — left_join with no duplicate keys
# in `districts` preserves the row count.

```

## 5b — Introduce a duplicate key

```{r ex5b}
# TODO: create districts_dup by adding one row that duplicates Vancouver
# but with a different bec_zone. Re-run the left_join. What is the new
# row count? What does that tell you?

```

---

# Group lab — Combine three sources

Use this section for your group-lab submission.

## Group members

> *List your group members here.*

## Common columns identified

> *List the columns that are comparable across the BC, Ontario, and
> air-quality files.*

## The combined tibble

```{r group-pipeline}
# TODO: paste your group's combined-tibble code here.

```

## Decisions made

> *In one paragraph: what choices did your group make about
> units, time aggregation, scope, or column matching that the
> code itself could not decide for you?*

---

# Done

When everything renders cleanly, submit the `.qmd` and the
rendered `.html`.
