---
title: "Chapter 5 — Learner Exercises"
subtitle: "Import and Inspect Forestry Data with R and Quarto"
author: "Your name here"
date: today
format:
  html:
    toc: true
    toc-location: left
    theme: cosmo
    embed-resources: true
---

```{r setup}
#| include: false
# Load packages quietly. If any of these fail, run:
#   install.packages(c("tidyverse", "readxl", "here", "knitr"))
library(tidyverse)
library(readxl)
library(here)
library(knitr)
```

# About this file {.unnumbered}

This is the learner workbook for Chapter 5. Each numbered
exercise is a section. Code chunks marked `# TODO` are where you
fill in your own R code.

**Folder structure expected** (same as Chapter 4):

```
frst232/
├── frst232.Rproj
├── data/
│   ├── bc_disturbance_reforestation.xlsx
│   └── on_forest_statistics_1.xlsx
├── outputs/
├── R/
└── frst232_ch05_learner.qmd   ← this file
```

When you are done, click **Render** at the top of RStudio. The
rendered `.html` is what you submit, alongside this `.qmd`
source.

---

# Exercise 1 — Vectors

## 1a — Build a vector

```{r ex1a}
# TODO: build a numeric vector called `ha` with these five values:
#   220000, 180000, 250000, 190000, 210000
# Use c().

```

## 1b — Compute summary statistics

```{r ex1b}
# TODO: compute the sum, mean, min, max, and length of `ha`.

```

## 1c — Logical comparison

```{r ex1c}
# TODO: print the result of ha > 200000.
# Expected: a logical vector of length 5 (TRUE/FALSE per element).

```

## 1d — Count and extract

```{r ex1d}
# TODO: count how many values in ha are greater than 200000.
# Hint: sum(ha > 200000)

# TODO: extract the actual values greater than 200000.
# Hint: ha[ha > 200000]

```

---

# Exercise 2 — Import both files

## 2a — Import the BC silviculture file

```{r ex2a}
# TODO: import the Data sheet of bc_disturbance_reforestation.xlsx
# Save into a variable called `bc`.

```

## 2b — Import the Ontario forest stats file

```{r ex2b}
# TODO: import the Data sheet of on_forest_statistics_1.xlsx
# Save into a variable called `ontario`.

```

---

# Exercise 3 — The six-move inspection

Apply each of `dim`, `nrow`, `ncol`, `names`, `head`, `glimpse`
to `bc`.

## 3a — Dimensions and counts

```{r ex3a}
# TODO: call dim(bc), nrow(bc), ncol(bc).

```

## 3b — Column names

```{r ex3b}
# TODO: call names(bc).

```

## 3c — First three rows

```{r ex3c}
# TODO: call head(bc, 3).

```

## 3d — Glimpse

```{r ex3d}
# TODO: call glimpse(bc).

```

## 3e — Repeat for Ontario

```{r ex3e}
# TODO: call dim(ontario) and glimpse(ontario).

```

---

# Exercise 4 — Column types

## 4a — Identify types

Look at the `glimpse()` output for both files. Answer in prose
below.

> *Write your answer here.* Which columns are `<chr>`? Which
> are `<dbl>`? Are there any `<int>` columns? Are the two files
> structured differently?

## 4b — Class check

```{r ex4b}
# TODO: use class() to confirm the type of:
#   bc$Reforestation_ha
#   ontario$PFTName

```

---

# Exercise 5 — Missing-value audit

## 5a — Per-column NA count, BC file

```{r ex5a}
# TODO: compute colSums(is.na(bc)).

```

## 5b — Per-column NA count, Ontario file

```{r ex5b}
# TODO: compute colSums(is.na(ontario)).

```

## 5c — Interpret

> *Write your answer here.* Did either file have any missing
> values? Does this match what you saw in Excel for Chapters 1
> and 2?

---

# Exercise 6 — Data dictionary

Write a small Markdown table describing every column of the BC
silviculture file. Use the format below.

| Column | Type | Unit | Description |
|---|---|---|---|
| `Fiscal_Year` | integer | year | TODO — your description |
| `Clearcutting_ha` | double | hectares | TODO |
| `Clearcutting_with_reserves_ha` | double | hectares | TODO |
| `Partial_cutting_ha` | double | hectares | TODO |
| `Harvested_ha` | double | hectares | TODO |
| `Natural_Disturbance_ha` | double | hectares | TODO |
| `Total_Disturbance_ha` | double | hectares | TODO |
| `Reforestation_ha` | double | hectares | TODO |

> Replace each "TODO" with a one-sentence description in your
> own words. Compare with the ReadMe sheet of
> `bc_disturbance_reforestation.xlsx` after you write yours.

---

# Exercise 7 — Vectorized arithmetic

## 7a — Compute the ratio

```{r ex7a}
# TODO: compute Reforestation_ha / Total_Disturbance_ha for every row.
# Save the result as a vector called `ratio`.

```

## 7b — Count years above 1

```{r ex7b}
# TODO: count how many years had ratio > 1.
# Hint: sum(ratio > 1)

```

---

# Exercise 8 — Render check

When all chunks above run without error, click **Render**
(or Ctrl+Shift+K / Cmd+Shift+K).

**Reproducibility test:** before you submit, restart R
(**Session → Restart R**, or Ctrl+Shift+F10 / Cmd+Shift+F10),
then re-render. If the document still renders cleanly, your
work is reproducible.

---

# Two-sentence summary

> *Write two sentences here* describing what the BC and Ontario
> files contain. Imagine your reader has not seen the data before.
> Mention the source (Government of...) and the time period
> covered.

---

# Peer-review section

Use this section for your group peer-review activity.

## Reviewer name

> *Type the reviewer's name here.*

## Score (out of 10)

| Criterion | Score (0–2) |
|---|---|
| Renders without error | TODO |
| dim, names, glimpse visible | TODO |
| Missing-value summary present | TODO |
| Data dictionary table | TODO |
| Two-sentence prose summary | TODO |

## One piece of feedback

> *Write your one piece of constructive feedback here.*

---

# Done

Submit:

1. This `.qmd` file (source).
2. The rendered `.html` file.
3. The peer-review section above (for the lecture peer-review
   activity).
