---
title: "Chapter 4 — Learner Exercises"
subtitle: "Start of Computing in R"
author: "Your name here"
date: today
format:
  html:
    toc: true
    toc-location: left
    theme: cosmo
    embed-resources: true
---

```{r setup}
#| include: false
# This setup chunk is included to load packages quietly.
# You may need to install them first with:
#   install.packages(c("tidyverse", "readxl", "here"))
library(tidyverse)
library(readxl)
library(here)
```

# About this file {.unnumbered}

This is the **learner workbook** for Chapter 4 of the FRST 232 OER
book. Work through each exercise in order. Code chunks marked
`# TODO` are where you fill in your own R code.

When you are done, **render** the document by clicking the
**Render** button at the top of RStudio (or pressing Ctrl+Shift+K
/ Cmd+Shift+K). The rendered HTML is what you submit.

**Folder structure expected:**

```
frst232/
├── frst232.Rproj
├── data/bc_disturbance_reforestation.xlsx
├── outputs/
├── R/
└── frst232_ch04_learner.qmd   ← this file
```

---

# Exercise 1 — Confirm your setup

## 1a — R version

Run the chunk below. The output should be a string like
`"R version 4.5.0 (2025-04-11)"`.

```{r ex1a}
# Inspect your R version. No edits needed; just run this chunk.
R.version.string
```

## 1b — Find the project root

```{r ex1b}
# TODO: print your project root path using here()
# Hint: here() with no arguments returns the root of the project.

```

**Expected:** A path string ending in `/frst232` (or whatever you
named your project folder).

## 1c — List the files in data/

```{r ex1c}
# TODO: list every file in your data/ folder using list.files() and here()
# Hint: list.files(here("data"))

```

**Expected:** A character vector that includes
`"bc_disturbance_reforestation.xlsx"`. If the result is `character(0)`
(empty), the file is not in `data/` and you need to put it there.

---

# Exercise 2 — Import the BC silviculture file

## 2a — Read the file into a tibble

```{r ex2a}
# TODO: import the Data sheet of bc_disturbance_reforestation.xlsx
# Use read_excel() with here() for the path, and sheet = "Data"
# Save the result into a variable called `bc`
# Hint:
#   bc <- read_excel(here("data", "bc_disturbance_reforestation.xlsx"),
#                    sheet = "Data")

```

## 2b — Confirm dimensions

```{r ex2b}
# TODO: call dim(bc). The result should be c(37, 8).

```

## 2c — List column names

```{r ex2c}
# TODO: call names(bc).
# The result should include "Fiscal_Year", "Harvested_ha",
# "Natural_Disturbance_ha", "Total_Disturbance_ha", "Reforestation_ha", etc.

```

## 2d — Glimpse the tibble

```{r ex2d}
# TODO: call glimpse(bc).
# This shows a one-line preview of each column with its type.

```

---

# Exercise 3 — Compute the five statistics

The five-statistic suite from the chapter: total, mean, min, max,
median of `bc$Reforestation_ha`.

## 3a — Total reforestation

```{r ex3a}
# TODO: compute sum(bc$Reforestation_ha)
# Expected: about 8,264,063 ha

```

## 3b — Mean annual reforestation

```{r ex3b}
# TODO: compute mean(bc$Reforestation_ha)
# Expected: about 223,353 ha

```

## 3c — Min, max, median (one chunk)

```{r ex3c}
# TODO: compute min, max, and median of bc$Reforestation_ha
# Three separate calls or one combined chunk — your choice.

```

## 3d — Verify against Excel

> *Write your answer here.* Open
> `bc_silviculture_ch01_workbook.xlsx` from your Chapter 1
> deliverables (or your `outputs/` folder). Compare the five
> values you computed in this exercise to the same five
> statistics on the Summary sheet of the workbook.
>
> **Do they match?** (Write "yes" or describe any differences.)

---

# Exercise 4 — The native pipe

Rewrite the following nested expression as a pipe with two or
more steps:

```r
round(mean(bc$Total_Disturbance_ha), 0)
```

## 4a — Your piped version

```{r ex4a}
# TODO: rewrite the expression above using |>
# Hint:
#   bc$Total_Disturbance_ha |> mean() |> round(0)

```

## 4b — Confirm same answer

```{r ex4b}
# TODO: confirm both versions return the same number.
# Either print both and compare visually, or use:
#   round(mean(bc$Total_Disturbance_ha), 0) ==
#     (bc$Total_Disturbance_ha |> mean() |> round(0))

```

---

# Exercise 5 — Find the year of the 2017 spike

The BC file has its largest `Natural_Disturbance_ha` value in
2017 (the record wildfire season). Use R to confirm this.

## 5a — Find the largest value

```{r ex5a}
# TODO: compute max(bc$Natural_Disturbance_ha)
# Expected: about 218,275 ha

```

## 5b — Find the year of the maximum

```{r ex5b}
# TODO: use which.max() to find the index of the maximum, then
# use that index to look up the year:
#   bc$Fiscal_Year[which.max(bc$Natural_Disturbance_ha)]
# Expected: 2017

```

---

# Exercise 6 — Optional: import the Ontario file

If you completed Chapter 2, you have
`on_forest_statistics_1.xlsx` in your `data/` folder too. Import
it the same way.

## 6a — Import

```{r ex6a}
# TODO: import the Data sheet of on_forest_statistics_1.xlsx
# Save into a variable called `ontario`

```

## 6b — Inspect

```{r ex6b}
# TODO: call dim(ontario). Expected: c(179, 8)

```

---

# Group lab — Confirmation

Use this section for your group lab submission.

## Group members

> *List your group members here, one per line.*

## Confirm sum(bc$Reforestation_ha)

```{r group-confirm}
# Already done in 3a — just re-run for the screenshot.
sum(bc$Reforestation_ha)
```

## Confirm dim(bc)

```{r group-confirm-dim}
dim(bc)
```

## Screenshot reminder

> Take a screenshot of your rendered HTML showing:
> - the project root path,
> - the column names of `bc`,
> - the total reforestation number.
>
> Submit the screenshot to the course site.

---

# Done

When the document renders cleanly with no errors and shows all
expected outputs, you have completed Chapter 4.

Re-render: **Render** button (or Ctrl+Shift+K / Cmd+Shift+K).

Submit the rendered `.html` file (and this `.qmd` source) to the
course site.
