---
title: "Chapter 11 — Learner Exercises"
subtitle: "Map Forestry Data with leaflet"
author: "Your name here"
date: today
format:
  html:
    toc: true
    toc-location: left
    theme: cosmo
    embed-resources: true
---

```{r setup}
#| include: false
# If leaflet is not installed:
#   install.packages(c("leaflet", "sf", "tidyverse", "lubridate", "here", "htmlwidgets"))
library(tidyverse)
library(sf)
library(leaflet)
library(lubridate)
library(here)
library(knitr)
library(htmlwidgets)
```

# About this file {.unnumbered}

Learner workbook for Chapter 11. Code chunks marked `# TODO` are
where you write your own R code. Render when done — the leaflet
maps stay interactive in the rendered HTML, which is the
deliverable.

```{r build-stations}
# Build stations_sf (same as Ch 10)
stations <- tribble(
  ~location,                          ~lon,        ~lat,
  "Burnaby South",                    -122.985,    49.215,
  "Coquitlam Douglas College",        -122.798,    49.252,
  "Langley Central",                  -122.660,    49.105,
  "Maple Ridge Golden Ears School",   -122.620,    49.215,
  "New Westminster Sapperton Park",   -122.890,    49.235,
  "North Delta",                      -122.910,    49.135,
  "North Vancouver Second Narrows",   -123.015,    49.295,
  "Pitt Meadows Airport",             -122.700,    49.215,
  "Port Coquitlam North",             -122.770,    49.265,
  "Port Moody Rocky Point Park",      -122.860,    49.290
)

stations_sf <- stations |>
  st_as_sf(coords = c("lon", "lat"), crs = 4326)
```

```{r compute-o3}
# Compute mean O₃ per station from the Ch 6 cleaned air-quality data
airdata_clean <- read_csv(here("data", "airdata3.csv"),
                          show_col_types = FALSE) |>
  filter(!is.na(O3))

station_means <- airdata_clean |>
  group_by(location) |>
  summarise(mean_o3 = mean(O3, na.rm = TRUE),
            n_obs = n(),
            .groups = "drop")

stations_o3 <- stations_sf |>
  left_join(station_means, by = "location")
```

---

# Exercise 1 — Your first interactive map

```{r ex1}
# TODO: build a leaflet map of stations_sf with:
#   1) leaflet(stations_sf)
#   2) addTiles()
#   3) addMarkers(popup = ~location)

```

> *Write your answer here.* When you render this document and
> click a marker, what shows up?

---

# Exercise 2 — Circle markers

```{r ex2}
# TODO: rebuild the map using addCircleMarkers() instead of
# addMarkers(). Set radius = 7, fillColor = "#2D6A4F",
# fillOpacity = 0.85, color = "#1B4332", weight = 2, popup = ~location.

```

---

# Exercise 3 — Three tile providers

## 3a — Default OpenStreetMap

```{r ex3a}
# TODO: leaflet(stations_sf) |> addTiles() |> addMarkers(popup = ~location)

```

## 3b — CartoDB.Positron (light theme)

```{r ex3b}
# TODO: same map but with addProviderTiles("CartoDB.Positron") instead

```

## 3c — Esri.WorldImagery (satellite)

```{r ex3c}
# TODO: same map but with addProviderTiles("Esri.WorldImagery")

```

> *Write your answer here.* Which tile provider feels best for a
> forestry deliverable? Why?

---

# Exercise 4 — Colour by O₃

## 4a — Build the palette

```{r ex4a}
# TODO: build a colour palette with colorNumeric().
# Use palette = c("#52B788", "#B7791F", "#B05C50") (green to red),
# domain = stations_o3$mean_o3, na.color = "grey50".
# Save as `pal`.

```

## 4b — Build the map

```{r ex4b}
# TODO: leaflet(stations_o3) |> addTiles() |>
#   addCircleMarkers(radius = ~ifelse(is.na(mean_o3), 5, mean_o3 + 4),
#                    fillColor = ~pal(mean_o3),
#                    fillOpacity = 0.85,
#                    color = "#1B4332",
#                    weight = 1.5,
#                    popup = ~location) |>
#   addLegend("bottomright", pal = pal, values = ~mean_o3,
#             title = "Mean O₃ (ppb)", opacity = 1)

```

---

# Exercise 5 — HTML popups

```{r ex5}
# TODO: rebuild the map from Ex 4 with rich HTML popups using paste0():
#   popup = ~paste0("<b>", location, "</b><br>",
#                   "Mean O₃: ",
#                   ifelse(is.na(mean_o3), "no data",
#                          paste0(round(mean_o3, 1), " ppb")), "<br>",
#                   "N observations: ", ifelse(is.na(n_obs), "0", n_obs))

```

---

# Exercise 6 — Add a buffer polygon layer

## 6a — Compute buffers in BC Albers, reproject back to WGS84

```{r ex6a}
# TODO: build buffers_sf:
#   stations_sf |> st_transform(3005) |> st_buffer(dist = 5000) |> st_transform(4326)

```

## 6b — Plot both layers

```{r ex6b}
# TODO: build a leaflet map with:
#   addProviderTiles("CartoDB.Positron")
#   addPolygons(data = buffers_sf, fillColor = "#52B788", fillOpacity = 0.25,
#               color = "#1B4332", weight = 1)
#   addCircleMarkers(data = stations_sf, radius = 5,
#                    color = "#1B4332", fillColor = "#1B4332",
#                    fillOpacity = 1, popup = ~location)

```

---

# Exercise 7 — Layers control

```{r ex7}
# TODO: build the multi-layer map from the chapter's complete
# worked example. It should include:
#   - Two basemaps (Light, Satellite) as baseGroups
#   - Stations and 5 km buffers as overlayGroups
#   - addLayersControl with options(collapsed = FALSE)
#   - setView centred at lng = -122.85, lat = 49.22, zoom = 11

```

---

# Exercise 8 — Save as standalone HTML

```{r ex8}
#| eval: false
# TODO: save the Exercise 7 map as outputs/stations_map.html using
# htmlwidgets::saveWidget(my_map, file = here("outputs", "stations_map.html"),
#                         selfcontained = TRUE)

```

---

# Group lab — Chapter capstone interactive map

## Group members

> *List your group members here.*

## Palette choice and rationale

> *What colour palette did your group choose for O₃, and why?*

## The capstone map

```{r group-capstone}
# TODO: build the chapter capstone map here.
# - Stations + buffers
# - Colour by mean O₃ with a matching legend
# - Layers control (at least two basemaps + two overlay groups)
# - Informative popups

```

## Headline finding

> *In one sentence, what does the map show?*

---

# Done

When rendered, this `.qmd` becomes an HTML report with **all
maps still interactive** — pan, zoom, click, toggle layers.
Submit:

1. This `.qmd` source.
2. The rendered `.html` (with all interactive maps embedded).
3. The standalone HTML map from Exercise 8.
