---
title: "Chapter 10 — Learner Exercises"
subtitle: "Spatial Data in R with sf"
author: "Your name here"
date: today
format:
  html:
    toc: true
    toc-location: left
    theme: cosmo
    embed-resources: true
---

```{r setup}
#| include: false
# If sf is not installed:
#   install.packages(c("sf", "tidyverse", "here"))
library(tidyverse)
library(sf)
library(here)
library(knitr)
```

# About this file {.unnumbered}

Learner workbook for Chapter 10. Code chunks marked `# TODO`
are where you write your own R code. Render the document when
done.

```{r stations-tibble}
# The Metro Vancouver air-quality station coordinates.
stations <- tribble(
  ~location,                          ~lon,        ~lat,
  "Burnaby South",                    -122.985,    49.215,
  "Coquitlam Douglas College",        -122.798,    49.252,
  "Langley Central",                  -122.660,    49.105,
  "Maple Ridge Golden Ears School",   -122.620,    49.215,
  "New Westminster Sapperton Park",   -122.890,    49.235,
  "North Delta",                      -122.910,    49.135,
  "North Vancouver Second Narrows",   -123.015,    49.295,
  "Pitt Meadows Airport",             -122.700,    49.215,
  "Port Coquitlam North",             -122.770,    49.265,
  "Port Moody Rocky Point Park",      -122.860,    49.290
)
```

---

# Exercise 1 — Build the sf object

```{r ex1}
# TODO: convert `stations` into an sf point object with CRS = 4326.
# Save as `stations_sf`. Use st_as_sf(coords = c("lon", "lat"), crs = 4326).
# Then print stations_sf and confirm:
#   - nrow(stations_sf) is 10
#   - st_crs(stations_sf)$Name says "WGS 84"

```

---

# Exercise 2 — Reproject to BC Albers

```{r ex2}
# TODO: reproject stations_sf to BC Albers (EPSG:3005).
# Save as `stations_bcalbers`. Then call st_bbox(stations_bcalbers)
# to see the bounding box. What are the units now (numbers in the
# millions, or in degrees)?

```

> *Write your answer here.* What are the units of the bounding
> box after reprojecting to BC Albers? Why is this useful?

---

# Exercise 3 — Pairwise distances

```{r ex3}
# TODO: compute the pairwise distance matrix in metres using
# st_distance() on stations_bcalbers.
# Then convert to kilometres and find the longest pair.
# Hint:
#   distance_matrix <- st_distance(stations_bcalbers)
#   dist_df <- as_tibble(distance_matrix, .name_repair = ~ stations$location) |>
#     mutate(from = stations$location) |>
#     pivot_longer(-from, names_to = "to", values_to = "distance_m") |>
#     mutate(distance_km = as.numeric(distance_m) / 1000) |>
#     filter(from != to)

```

> *Write your answer here.* Which pair of stations has the
> longest distance, and roughly how many km is it?

---

# Exercise 4 — Buffer + intersection

```{r ex4-buffer}
# TODO: build a 7.5 km buffer around Burnaby South (in BC Albers).
# Save as `burnaby_buffer`.

```

```{r ex4-intersect}
# TODO: use st_intersection() to find which OTHER stations fall
# inside burnaby_buffer.
# Print the list of station names.

```

---

# Exercise 5 — Plot with geom_sf

```{r ex5}
# TODO: build a ggplot with TWO geom_sf layers:
#   1) burnaby_buffer (reproject back to WGS84 for plotting), fill green and transparent
#   2) stations_sf, all 10 points, drawn on top
# Add labs(title, caption) and theme_minimal().

```

---

# Exercise 6 — Save as GeoJSON

```{r ex6}
#| eval: false
# TODO: write stations_sf to outputs/stations.geojson using st_write().

```

---

# Exercise 7 — Area of a buffer

```{r ex7}
# TODO: compute the area of burnaby_buffer in hectares.
# Hint: as.numeric(st_area(burnaby_buffer)) / 10000
# Confirm it is approximately pi * 5000^2 / 10000 = 7,854 ha
# (or for 7.5 km buffer: pi * 7500^2 / 10000 = ~17,671 ha)

```

---

# Group lab — Build a station map together

## Group members

> *List your group members here.*

## Buffer radius chosen

> *Type the radius (in km) your group selected, and why.*

## The map

```{r group-map}
# TODO: build the group's buffer-and-points map here, with all 10
# stations and all 10 buffers visible.

```

## Caption

> *Write a one-sentence caption explaining what the chart shows.*

---

# Done

Submit:

1. This `.qmd` source.
2. The rendered `.html`.
3. The PNG of your map (from Exercise 5 or the group map).
