---
title: "Your Presentation Title Here"
subtitle: "FRST 232 — A group presentation"
author: "Member A · Member B · Member C · Member D"
date: 2027-04-10
format:
  revealjs:
    theme: simple
    slide-number: true
    transition: slide
    incremental: false
    embed-resources: true
    code-overflow: wrap
    chalkboard: false
    logo: "" # add a path to a logo if you have one
    footer: "FRST 232 — UBC · CC BY 4.0"
---

```{r setup}
#| include: false
library(tidyverse)
library(readxl)
library(lubridate)
library(here)
library(knitr)
library(scales)
# Add: library(sf) and library(leaflet) if your talk uses spatial work
```

# Background

## The question

> Replace this slide with **one sentence** stating the question
> your analysis answers.

::: {.notes}
Talking points (only you see these in presenter mode):

- Set the scene in 15 seconds.
- Why does this question matter for forestry?
- One concrete real-world implication.
:::

## The data

::: {.incremental}
- **Source:** [name the publisher, e.g., Environmental Reporting BC]
- **Licence:** [Open Government Licence — BC]
- **Coverage:** [time period, region, units of observation]
- **Size:** [rows × columns]
- **Why we picked this dataset:** [one sentence]
:::

::: {.notes}
Acknowledge the licence here, not just at the end.
:::

# Methods and Results

## The cleaning pipeline

Show your actual R code — ideally ≤ 10 lines.

```{r method}
#| echo: true
#| eval: false
#| code-line-numbers: true

# Example placeholder — replace with your real pipeline
airdata_clean <- read_csv(here("data", "airdata3.csv")) |>
  select(-starts_with("Unnamed")) |>
  select(where(\(col) !all(is.na(col)))) |>
  mutate(date_parsed = mdy(Date)) |>
  filter(!is.na(PM25)) |>
  left_join(station_regions, by = "location")
```

::: {.notes}
Read each verb out loud. Audiences who do not know R can still
follow "filter, mutate, join".
:::

## Result — chart

```{r chart}
#| echo: false
#| fig.width: 9
#| fig.height: 4.8

# Replace with your group's headline chart.
# Use base_size 18+ so the chart reads on a projector.
bc <- read_excel(here("data", "bc_disturbance_reforestation.xlsx"),
                 sheet = "Data")

ggplot(bc, aes(x = Fiscal_Year, y = Reforestation_ha)) +
  geom_col(fill = "#2D6A4F") +
  scale_y_continuous(labels = label_comma()) +
  labs(title    = NULL,
       x        = "Fiscal year",
       y        = "Reforested area (ha)",
       caption  = "Source: Environmental Reporting BC") +
  theme_minimal(base_size = 18) +
  theme(plot.caption = element_text(colour = "grey40"))
```

## Result — table

```{r table}
#| echo: false
# Replace with your group's headline summary table
bc |>
  summarise(
    n_years     = n(),
    mean_ha     = mean(Reforestation_ha) |> round(0) |> comma(),
    median_ha   = median(Reforestation_ha) |> round(0) |> comma(),
    max_ha      = max(Reforestation_ha) |> round(0) |> comma(),
    max_year    = Fiscal_Year[which.max(Reforestation_ha)]
  ) |>
  kable(col.names = c("Years", "Mean (ha)", "Median (ha)",
                      "Max (ha)", "Max year"))
```

## Result — map

```{r map}
#| echo: false
#| eval: false

# If your talk uses a leaflet map, place it here.
# (Set eval: true and uncomment when you have the data set up.)
#
# leaflet(stations_pm25) |>
#   addProviderTiles("CartoDB.Positron") |>
#   addCircleMarkers(radius = ~mean_pm25 + 4,
#                    fillColor = ~pal(mean_pm25),
#                    fillOpacity = 0.85,
#                    color = "#1B4332", weight = 1.5,
#                    popup = ~paste0("<b>", location, "</b>")) |>
#   addLegend("bottomright", pal = pal,
#             values = stations_pm25$mean_pm25,
#             title = "Mean PM2.5 (μg/m³)")
```

# Interpretation

## What we found

::: {.incremental}
- **Headline:** [one sentence stating the main finding]
- **Magnitude:** [the number that proves it]
- **Context:** [why this matters — for forestry, policy, or
  research]
:::

::: {.notes}
The headline must be defensible. If a sceptic asks "really?"
have one number ready.
:::

## Comparison

Build a two-column slide if you have a useful comparison.

::: {.columns}
::: {.column width="50%"}
**Our result**

- key number 1
- key number 2
- key number 3
:::
::: {.column width="50%"}
**Baseline or comparison**

- baseline number 1
- baseline number 2
- baseline number 3
:::
:::

# Conclusion

## What we know

::: {.incremental}
- **One:** [first conclusion]
- **Two:** [second conclusion]
- **Three:** [third conclusion]
:::

## Limitations

::: {.incremental}
- **Data scope:** [e.g., one month of measurements; one
  province; one decade]
- **Methodology:** [e.g., did not adjust for X; treated Y as
  comparable to Z]
- **Generalisability:** [where these results would NOT hold]
:::

::: {.notes}
A frank limitations slide makes you more credible, not less.
Audiences who hear acknowledged limitations trust the rest of
the talk more.
:::

## What we could do next

::: {.incremental}
- More years of data
- Stratify by [variable not yet considered]
- Combine with [companion dataset]
:::

# Acknowledgements

## Data

- BC Government, Environmental Reporting BC. Open Government
  Licence — BC.
- [Any other data source here]

## R packages

tidyverse · readxl · dplyr · ggplot2 · lubridate · sf · leaflet ·
Quarto

## Course

FRST 232 — Computer Applications in Forestry
Dr. Suborna Ahmed · UBC Faculty of Forestry · Fall 2026

::: {.notes}
Always credit the data. Quietly mention which packages you used.
End on the course name + instructor.
:::

## Questions

Thank you.

::: {.notes}
End on a clean slide with one word ("Questions" or "Thank you").
NEVER end on the acknowledgements slide — it gives the audience
nowhere to look during Q&A.
:::
