---
title: "Chapter 7 — Learner Exercises"
subtitle: "Summarize Forestry Data Overall and by Group"
author: "Your name here"
date: today
format:
  html:
    toc: true
    toc-location: left
    theme: cosmo
    embed-resources: true
---

```{r setup}
#| include: false
# If any of these fail, run:
#   install.packages(c("tidyverse", "lubridate", "here", "knitr"))
library(tidyverse)
library(lubridate)
library(here)
library(knitr)
```

# About this file {.unnumbered}

This is the learner workbook for Chapter 7. Code chunks marked
`# TODO` are where you fill in your own R code.

**Folder structure expected** (same as Chapters 4–6):

```
frst232/
├── frst232.Rproj
├── data/
│   └── airdata3.csv                                     ← from Ch 6
├── outputs/
│   └── airdata3_clean.csv                               ← from Ch 6 (optional)
├── R/
└── frst232_ch07_learner.qmd                             ← this file
```

You will start from `airdata_clean` — the same tibble you built
in Chapter 6. The first chunk below rebuilds the Chapter 6
pipeline so this file can render on its own.

---

# Setup — rebuild the cleaned tibble

```{r ch6-pipeline}
# This chunk reproduces the Chapter 6 cleaning so the rest of
# the document can run. If you completed Chapter 6, you can
# alternatively load outputs/airdata3_clean.csv directly.

station_regions <- tribble(
  ~location,                          ~region,
  "Burnaby South",                    "Burrard Peninsula",
  "Coquitlam Douglas College",        "North-East Sector",
  "Langley Central",                  "Fraser Valley West",
  "Maple Ridge Golden Ears School",   "North-East Sector",
  "New Westminster Sapperton Park",   "Burrard Peninsula",
  "North Delta",                      "South of Fraser",
  "North Vancouver Second Narrows",   "North Shore",
  "Pitt Meadows Airport",             "North-East Sector",
  "Port Coquitlam North",             "North-East Sector",
  "Port Moody Rocky Point Park",      "North-East Sector"
)

airdata_clean <- read_csv(here("data", "airdata3.csv"),
                          show_col_types = FALSE) |>
  select(-starts_with("Unnamed"), -any_of("...1")) |>
  select(where(\(col) !all(is.na(col)))) |>
  mutate(date_parsed = mdy(Date),
         year = year(date_parsed),
         month = month(date_parsed),
         day = day(date_parsed)) |>
  filter(!is.na(PM25)) |>
  mutate(pm25_band = case_when(
    PM25 < 10  ~ "low",
    PM25 < 25  ~ "moderate",
    PM25 < 50  ~ "high",
    TRUE       ~ "very high"
  )) |>
  left_join(station_regions, by = "location")

dim(airdata_clean)
```

---

# Exercise 1 — Overall summary

Write a single `summarise()` call on `airdata_clean` that returns:

- `n_obs` (row count via `n()`),
- `n_stations` (via `n_distinct(location)`),
- `mean_pm25`, `median_pm25`, `sd_pm25`,
- `min_pm25`, `max_pm25`.

Remember `na.rm = TRUE` for the numeric aggregations.

```{r ex1}
# TODO: write the overall summarise() call here.

```

> *Write a one-sentence interpretation here.* What is the overall
> mean PM2.5 across the cleaned tibble, and how does it compare
> to the maximum?

---

# Exercise 2 — By-station summary

Group by `location`, summarise `n_obs`, `mean_pm25`, `median_pm25`,
and `max_pm25`. Sort descending by `mean_pm25`. Use
`.groups = "drop"`.

```{r ex2}
# TODO: write the by-station summarise pipeline.

```

> *Write your answer here.* Which station has the **highest**
> mean PM2.5 in this month? Which has the lowest?

---

# Exercise 3 — By-day summary

Group by `day` (the day-of-month integer). Summarise `n_obs` and
`mean_pm25` per day. Sort descending by `mean_pm25`.

```{r ex3}
# TODO: write the by-day summarise pipeline.

```

> *Write your answer here.* Which day in January 2000 had the
> **highest provincial mean PM2.5** across all stations? How
> many observations support that mean?

---

# Exercise 4 — Cross-tabulation with `pivot_wider()`

Build a cross-tab: rows are `region`, columns are `pm25_band`,
values are counts. Use `values_fill = 0` so empty cells show 0.

```{r ex4}
# TODO: count by region and pm25_band, then pivot_wider.
# Hint:
#   airdata_clean |>
#     count(region, pm25_band) |>
#     pivot_wider(names_from = pm25_band, values_from = n,
#                 values_fill = 0)

```

> *Write your answer here.* Which region has the most "high"
> or "very high" PM2.5 hours? Which region had the fewest?

---

# Exercise 5 — `across()` for multiple pollutants

Group by `region`. Use `across()` to compute the mean of every
listed pollutant in one summarise call: `PM25`, `PM10`, `O3`,
`NO2`.

```{r ex5}
# TODO: write the across() summarise pipeline.
# Hint:
#   airdata_clean |>
#     group_by(region) |>
#     summarise(across(c(PM25, PM10, O3, NO2),
#                      \(x) mean(x, na.rm = TRUE)),
#               .groups = "drop")

```

> *Write your answer here.* Which pollutant has the highest
> regional variability (largest difference between regions)?

---

# Exercise 6 — 90th-percentile PM2.5 per station

Group by `location`. Use `quantile(PM25, 0.9, na.rm = TRUE)` to
compute the 90th-percentile PM2.5 per station. Arrange descending.

```{r ex6}
# TODO: write the per-station 90th-percentile pipeline.

```

> *Write your answer here.* The 90th percentile tells you the
> level that PM2.5 stays *below* 90% of the time. Which station
> has the highest 90th-percentile reading? How does this ranking
> compare to the mean ranking from Exercise 2?

---

# Exercise 7 — PivotTable parity

This exercise asks you to **verify the same numbers two ways** —
once in R, once in Excel.

## 7a — Save the cleaned tibble to CSV

```{r ex7a}
# TODO: save airdata_clean to outputs/airdata3_clean.csv using
# write_csv() and here().

```

## 7b — Open in Excel

Open `outputs/airdata3_clean.csv` in Excel. Build a PivotTable:

- **Rows**: `region`
- **Columns**: `pm25_band`
- **Values**: `Count of PM25`

Take a screenshot of the PivotTable.

## 7c — Confirm parity

> *Write your answer here.* Do the cell counts in your Excel
> PivotTable match the cell counts in your Exercise 4
> `pivot_wider()` result? They should match to the unit. If they
> do not, which row × column cell differs, and what might be the
> cause?

---

# Group lab — Summary report

Use this section for your group-lab submission.

## Group members

> *List your group members here.*

## The question your group chose

> *Type the one forestry question your group is answering.*

## The pipeline

```{r group-pipeline}
# TODO: paste the dplyr pipeline that answers your group's question.

```

## The one-sentence caption

> *Write the headline finding of your summary table in one
> sentence.*

---

# Cleaning log addendum

If your Chapter 6 cleaning log was for the cleaning steps, this
chapter's log entries are for the **summarising steps**:

| Date | Step | What I did | Why |
|---|---|---|---|
| TODO | 1 | Built overall summary | Establish baseline numbers. |
| TODO | 2 | Grouped by location | Compare across stations. |
| TODO | 3 | TODO | TODO |

---

# Done

When everything above runs without error, click **Render**.
Submit:

1. This `.qmd` source file.
2. The rendered `.html`.
3. The PivotTable screenshot from Exercise 7.
4. The group-lab worksheet section.
