---
title: "Chapter 12 — Learner Exercises"
subtitle: "Group Presentations and Portfolio Outputs"
author: "Your name here"
date: today
format:
  html:
    toc: true
    toc-location: left
    theme: cosmo
    embed-resources: true
---

```{r setup}
#| include: false
library(tidyverse)
library(here)
library(knitr)
```

# About this file {.unnumbered}

Learner workbook for Chapter 12. This chapter is a **synthesis
chapter** — most exercises ask you to produce *additional* files
beyond this `.qmd`. Track what you create as you go and submit
the whole bundle at the end.

---

# Exercise 1 — Convert a chapter to slides

Pick any chapter learner `.qmd` you have already rendered as
HTML — for example `frst232_ch09_learner.qmd`.

**Steps:**

1. Copy the file to `frst232_ch09_slides.qmd` (keep the
   original `.qmd` intact).
2. Open the copy. In the YAML header, change `format: html`
   to:
   ```yaml
   format:
     revealjs:
       theme: simple
       slide-number: true
       embed-resources: true
   ```
3. Render. Open the resulting HTML — it is now a slide deck.

> *Write your answer here.* When the rendered output opened,
> which sections were the most natural slides? Which were
> uncomfortably long (a single section taking 4+ slides because
> of vertical overflow)?

---

# Exercise 2 — One idea per slide

Take the rendered HTML from Exercise 1. For each slide that has
**more than 50 words of body text or more than one chart**, edit
the source `.qmd` to split it into two slides.

> *Write your answer here.* How many original slides did you
> split, and into how many new ones? Did the deck feel clearer
> after the split?

---

# Exercise 3 — Slide separators

In your converted file, introduce a clear three-section structure
using `# Section Title` (h1) section dividers:

1. `# Background`
2. `# Methods and Results`
3. `# Conclusion`

Each h1 creates a section-divider slide. Each h2 (`## Slide
title`) underneath creates a content slide in that section.

```{r ex3-note}
# (No code needed — this is structural editing in your .qmd.)
```

> *Write your answer here.* How does the section-divider
> structure change the audience's mental map of the talk?

---

# Exercise 4 — Speaker notes and incremental reveals

Take any one slide and:

- Add a `::: {.notes}` block with 2–3 talking points only you
  will see.
- Convert a bullet list on the same slide into a
  `::: {.incremental}` block so the bullets appear one at a
  time.

Re-render and confirm:

- Pressing **S** during the live preview opens presenter mode
  with your notes visible.
- The bullets appear one at a time as you advance.

---

# Exercise 5 — Two-column layout

Build a new slide with a two-column layout. Left column: a chart
(any chart from your earlier work). Right column: three bullet
takeaways.

Use the syntax:

````markdown
## My comparison slide

::: {.columns}
::: {.column width="50%"}
[chart chunk or image here]
:::
::: {.column width="50%"}
- Takeaway 1
- Takeaway 2
- Takeaway 3
:::
:::
````

---

# Exercise 6 — Build the portfolio scaffold

In your `frst232/` project root, create three new files:

## 6a — `_quarto.yml`

```yaml
project:
  type: website
  output-dir: docs

website:
  title: "FRST 232 — Your Name"
  navbar:
    left:
      - href: index.qmd
        text: Home
      - href: chapters/ch06.qmd
        text: Cleaning
      - href: chapters/ch09.qmd
        text: Charts
      - href: chapters/ch11.qmd
        text: Maps

format:
  html:
    theme: cosmo
    toc: true
```

## 6b — `index.qmd`

```markdown
---
title: "FRST 232 Portfolio"
---

# About

A collection of forestry data analyses for FRST 232 (UBC).

## Highlights

- [Cleaning the BC air-quality file](chapters/ch06.html)
- [Visualising reforestation trends](chapters/ch09.html)
- [Interactive station map](chapters/ch11.html)

## Licence

© 2026 Your Name · CC BY 4.0 for content · MIT for code
```

## 6c — Copy chapter sources

Copy your rendered chapter `.qmd` files into `chapters/`:

```
chapters/
├── ch06.qmd
├── ch09.qmd
└── ch11.qmd
```

Then run **`quarto render`** in the terminal from the project
root. Confirm a `docs/` folder appears with `index.html` and
the chapter pages.

---

# Exercise 7 — Write a CITATION.md

Run these commands in the R console:

```{r ex7-citation}
#| eval: false
citation("tidyverse")
citation("readxl")
citation("dplyr")
citation("ggplot2")
citation("sf")
citation("leaflet")
citation("lubridate")
citation("here")
```

Create `CITATION.md` in your project root and paste the output
of each command (or just the BibTeX/plain-text reference).

Also list your **data sources** at the top of the file:

```markdown
## Data

- Environmental Reporting BC, *Trends in Silviculture in BC*.
  Open Government Licence — BC.
  https://catalogue.data.gov.bc.ca/...
- Environmental Reporting BC, BC Air Data.
- Government of Ontario, *Forest Resources of Ontario 2021*.
```

---

# Exercise 8 — Publish to Quarto Pub (optional but recommended)

From the project root in the terminal:

```bash
quarto publish quarto-pub
```

Sign in when prompted. The command produces a public URL like
`yourname.quarto.pub/frst232/`.

> *Write your answer here.* What is your published URL? Share it
> with a classmate and ask them to confirm it loads on their
> computer.

---

# Group lab — In-class presentation

The lab session IS the in-class presentations. Use this section
of your `.qmd` to track your group's preparation.

## Group members

> *List your group members here.*

## Topic chosen

> *Type your group's chosen analysis topic (e.g., "Ch 6
> cleaning of the BC air-quality file").*

## Slide structure (8 required slides)

1. Title slide — members, date
2. The question — one sentence
3. The data — source, licence, row/column counts
4. The method — show the actual R code (≤ 10 lines)
5. The result — one chart, table, or map
6. Interpretation — headline + nuance
7. Limitations — what the analysis does NOT show
8. Acknowledgements — citations of data and packages

## Peer-evaluation forms

> Keep your filled-in peer-eval forms for the other three groups
> in this section after the lab.

---

# Done

Submit at the end of the chapter:

1. This `.qmd` source.
2. The revealjs `.qmd` from Exercise 1 (renamed
   `frst232_ch12_slides.qmd`).
3. The rendered revealjs HTML.
4. Your `_quarto.yml`, `index.qmd`, and at least three chapter
   pages.
5. Your `CITATION.md`.
6. The Quarto Pub URL (optional).
