---
title: "Chapter 6 — Learner Exercises"
subtitle: "Clean and Prepare Forestry Data in R"
author: "Your name here"
date: today
format:
  html:
    toc: true
    toc-location: left
    theme: cosmo
    embed-resources: true
---

```{r setup}
#| include: false
# If any of these fail, run:
#   install.packages(c("tidyverse", "lubridate", "here", "knitr"))
library(tidyverse)
library(lubridate)
library(here)
library(knitr)
```

# About this file {.unnumbered}

This is the learner workbook for Chapter 6. Code chunks marked
`# TODO` are where you fill in your own R code.

**Folder structure expected:**

```
frst232/
├── frst232.Rproj
├── data/
│   ├── bc_disturbance_reforestation.xlsx
│   ├── on_forest_statistics_1.xlsx
│   └── airdata3.csv                      ← new this chapter
├── outputs/
├── R/
└── frst232_ch06_learner.qmd              ← this file
```

When everything renders cleanly, submit both this `.qmd` source
and the rendered `.html`.

---

# Exercise 1 — Import and inspect

## 1a — Import the raw file

```{r ex1a}
# TODO: read airdata3.csv into a tibble called `airdata_raw`.
# Use read_csv() with here() for the path.
# Add show_col_types = FALSE to silence the type-guessing message.

```

## 1b — Six-move inspection

```{r ex1b}
# TODO: call dim(), names() (or glimpse()), and
# colSums(is.na(airdata_raw)) |> sort(decreasing = TRUE) |> head(15)
# to see which columns are most missing.

```

## 1c — Note your findings

> *Write your answer here.* How many rows? How many columns?
> Which columns are 100% missing?

---

# Exercise 2 — Drop redundant columns with `select()`

## 2a — Drop the upstream-tool index columns

```{r ex2a}
# TODO: starting from airdata_raw, drop any column whose name
# starts with "Unnamed" or "...".
# Save as `airdata_step1`.
# Hint: select(-starts_with("Unnamed"), -any_of("...1"))

```

## 2b — Drop all-NA columns

```{r ex2b}
# TODO: starting from airdata_step1, drop every column where every
# value is NA. Save as `airdata_step2`.
# Hint: select(where(\(col) !all(is.na(col))))

```

## 2c — Confirm

```{r ex2c}
# TODO: call ncol(airdata_step2) and names(airdata_step2).
# You should be down to roughly half the original columns.

```

---

# Exercise 3 — Parse the date with `mutate()`

## 3a — Add a parsed-date column

```{r ex3a}
# TODO: starting from airdata_step2, use mutate() to add a column
# called `date_parsed` built from the Date column using mdy().
# Save as `airdata_step3`.

```

## 3b — Verify

```{r ex3b}
# TODO: glimpse(airdata_step3) and confirm `date_parsed` is <date>.
# Also call summary(airdata_step3$date_parsed) to see the range.

```

---

# Exercise 4 — Filter to non-missing O₃

```{r ex4}
# TODO: starting from airdata_step3, keep only rows where O3 is
# NOT missing. Save as `airdata_step4`.
# Hint: filter(!is.na(O3))
# Then report nrow(airdata_step4).

```

---

# Exercise 5 — `case_when()` for O₃ bands

```{r ex5}
# TODO: starting from airdata_step4, add a column `o3_band`:
#   O3 < 10  -> "low"
#   O3 < 20  -> "moderate"
#   O3 < 30  -> "high"
#   TRUE       -> "very high"
# Save as `airdata_step5`.
# Then call count(airdata_step5, o3_band).

```

---

# Exercise 6 — `left_join()` with a station-region lookup

## 6a — Build the lookup tibble

```{r ex6a}
station_regions <- tribble(
  ~location,                          ~region,
  "Burnaby South",                    "Burrard Peninsula",
  "Coquitlam Douglas College",        "North-East Sector",
  "Langley Central",                  "Fraser Valley West",
  "Maple Ridge Golden Ears School",   "North-East Sector",
  "New Westminster Sapperton Park",   "Burrard Peninsula",
  "North Delta",                      "South of Fraser",
  "North Vancouver Second Narrows",   "North Shore",
  "Pitt Meadows Airport",             "North-East Sector",
  "Port Coquitlam North",             "North-East Sector",
  "Port Moody Rocky Point Park",      "North-East Sector",
  "Richmond South",                   "South of Fraser",
  "Tsawwassen",                       "South of Fraser",
  "Vancouver Clark Drive",            "Vancouver",
  "Vancouver International Airport #2","South of Fraser",
  "Vancouver Kitsilano",              "Vancouver",
  "West Vancouver Lions Gate",        "North Shore"
)
station_regions
```

## 6b — Join

```{r ex6b}
# TODO: left_join airdata_step5 with station_regions, joining on
# the location column. Save as `airdata_clean`.

```

## 6c — Confirm row count is preserved

```{r ex6c}
# TODO: nrow(airdata_step5) and nrow(airdata_clean) should match.
# (left_join never drops rows from the main tibble.)

```

---

# Exercise 7 — Save the cleaned tibble

```{r ex7}
# TODO: write airdata_clean to outputs/airdata3_clean.csv using
# write_csv(). Use here() for the path.

```

---

# Exercise 8 — Per-station summary

```{r ex8}
# TODO: group airdata_clean by location, then summarise with:
#   n_obs = n()
#   mean_o3 = mean(O3, na.rm = TRUE)
#   max_o3  = max(O3, na.rm = TRUE)
# Then arrange(desc(mean_o3)).
# Print the result.

```

---

# Cleaning log

> Write your own cleaning log here. Use the table format from
> the chapter prose. One row per cleaning step you did above.

| Date | Step | What I did | Why |
|---|---|---|---|
| TODO | 1 | TODO | TODO |
| ... | ... | ... | ... |

---

# Group-lab summary

Use this section for your group-lab submission.

## Group members

> *List your group members here.*

## Highest-O₃ station (Task 5 of the group lab)

> *Type the station name and its mean O₃ here.*

## One-sentence interpretation

> *Why might that station have higher O₃? Use your knowledge
> of Metro Vancouver geography.*

---

# Done

When everything above runs without error, click **Render**.
Submit the `.qmd` and the rendered `.html`.
